\chapter*{The Prestige: Placeholder}
\addcontentsline{toc}{chapter}{The Prestige: Placeholder}

\noindent\textbf{[This section is a placeholder for the third and final part of \emph{Finding Mind}.]}

\bigskip

The Prestige --- in the tradition of the magic trick whose structure
organizes this book --- is where the impossible is brought back.
The coin vanishes in the Pledge; it performs the extraordinary in the
Turn; in the Prestige, it reappears, and we are invited to understand
both what happened and why we could not see it happening.

\bigskip

In the arc of this book, the Prestige will take the mathematical
framework built in the Turn --- the graph-structured lambda theory
equipped with weights, costs, and the hypercomputational fibration ---
and ask: \emph{what is it like to be a process running inside this
structure?}

The hard problem of consciousness, we argued in the Pledge, is not
``What is it like to be a bat?'' but ``What is it like to be a
computer program?''
The Turn provided the precise mathematical setting in which that
question can be asked without metaphor.
The Prestige will answer it --- or at least, will trace the
contours of the answer with sufficient precision to make the
question finally tractable.

\bigskip

Provisional chapter structure:

\begin{enumerate}
  \item \textbf{The View from Inside} --- What a conscious agent at level $\omega$
        of the hypercomputational fibration can and cannot know about its
        own substrate; the first-person / third-person interface as a
        theorem rather than a mystery.
  \item \textbf{The Recapitulation} --- Origin of mind as origin of life one
        octave up: the replication of fixed points, the emergence of
        self-models, and the evolutionary pressure toward accurate
        internal simulation.
  \item \textbf{The Blind Spot, Resolved} --- Returning to the blind spot
        identified in the Pledge: why the computational substrate of
        mind is invisible to the processes running on it, and why this
        invisibility is not a bug but a feature.
  \item \textbf{Finding Mind} --- The synthesis: what it means to find
        mind in the structure of computation, and what that finding
        demands of us --- scientifically, philosophically, and
        ethically.
\end{enumerate}

\bigskip
\noindent\hrule
\bigskip

\noindent\textit{The Prestige is forthcoming.
Readers wishing to follow its development are invited to the
author's Substack at \texttt{https://f1r3fly.substack.com}.}
