\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage[protrusion=true,expansion=false]{microtype}
\usepackage[colorlinks=true,linkcolor=black,citecolor=black,urlcolor=blue]{hyperref}
\usepackage{enumitem}

% ---- theorem environments ----
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}[definition]{Example}
\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{conjecture}[definition]{Conjecture}
\theoremstyle{remark}
\newtheorem{remark}[definition]{Remark}

% ---- notation ----
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\Stk}{\mathsf{Stk}}
\newcommand{\Tok}{T}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Rpos}{\mathbb{R}_{>0}}
\newcommand{\cost}{\mathsf{C}}
\newcommand{\ciGSLT}{\mathsf{ciGSLT}}
\newcommand{\iGSLT}{\mathsf{iGSLT}}
\newcommand{\Conv}{\mathsf{Conv}}
\newcommand{\Convf}{\mathsf{Conv}_{0}}
\newcommand{\grad}{\operatorname{grad}}
\newcommand{\dvg}{\operatorname{div}}
\newcommand{\im}{\operatorname{im}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\near}{\mathsf{near}}
\newcommand{\wrap}[2]{\{#1\}_{#2}}
\newcommand{\eqv}{\equiv}
\DeclareMathOperator{\Betti}{b_1}

\title{\bfseries Energy as a Virtual Token\\[0.35em]
\large No-Arbitrage, the Graph--Hodge Potential, and the Conserved Scalar of Cost-Accounted GSLTs}
\author{L.~G.~Meredith\\ \small F1R3FLY.io}
\date{July 2026 --- working note}

\begin{document}
\maketitle

\begin{abstract}
The cost-accounting construction on graph-structured lambda theories replaces a single,
undifferentiated fuel (phlogiston) by a spectrum of signature-indexed tokens: as many
``colours'' of fuel as there are economically empowered actors. Reading the coloured tokens
as the many manifest forms of energy in naive physics---chemical, mechanical, electrical---we
ask what plays the role of energy itself, the single scalar into which the engines of physics
interconvert those forms. We answer: adjoin \emph{engines} (conversion contracts) between token
types, require that no cycle of conversions manufactures value, and a \emph{virtual token}
appears---a valuation $\nu:\Sig\to\Rpos$ into which every colour is priced. We show that this
virtual token is not an additional construct layered on top of no-arbitrage but is
\emph{definitionally identical} to it: on the graph of engines, the log-rates form a $1$-cochain,
no-arbitrage is exactly its closedness, and the virtual token is the potential $0$-cochain whose
gradient it is. The graph--Hodge decomposition then splits the conversion structure canonically
into an \emph{exact} part (the virtual token, i.e.\ energy) and a \emph{cyclic} part (the
arbitrage), the dimension of the latter being the first Betti number of the engine graph.
``Shaking out arbitrage'' is the orthogonal projection onto the exact part---the unit of a
reflection, hence an idempotent monad, in pointed contrast to the non-idempotent cost monad
$\cost$. The construction compresses the $O(n^2)$ bilateral rates to the $O(n)$ potential and,
because the virtual token is a \emph{unit of account} (a monoid homomorphism out of $\Sig$) rather
than a \emph{medium of exchange} (an element of $\Sig$), it prices compound signatures without
laundering authority---so that, in the channel-based ciGSLTs that carry an object-capability
discipline at all (the mobile process calculi, whose interaction surface is nominal, as against
combinatory theories such as SKI whose rigid application constructor locates authority
structurally), that discipline survives. Finally we separate the two
faces of physical energy: the virtual token is a conserved charge (a first law, exact in the
lossless limit; a Lyapunov free energy with a second-law inequality once engines carry friction),
but it is not yet the \emph{generator} of the dynamics. We conjecture that closing that gap is an
energy--time conjugacy with the temporal token stack, and read the ontology accordingly: the
virtual token is a Noether charge, not a substance.
\end{abstract}

\section{Introduction}

The cost-accounted rho calculus \cite{CARho} and its generalisation, the cost endofunctor $\cost$
on the category $\ciGSLT$ of continued interactive graph-structured lambda theories \cite{CE},
meter computation by gating each rewrite on the consumption of a matching token. Tokens are
indexed by signatures $s\in\Sig$; the ``spectral decomposition of phlogiston'' \cite{CARho}
replaces one universal fuel by a family of signature-indexed resources, one per economically
empowered actor. The companion note \cite{CE} already flirts with a physical reading of this
family: signatures appear as an internal (flavour) index, token distributions as fields over the
interaction surfaces, and the tropical semiring recovers cost.

This note takes the physical reading one step further, along a line suggested in recent dialogue.
In secondary-school physics one meets energy in many manifest forms---chemical, mechanical,
electrical, thermal---and one meets \emph{engines} that convert one form into another. The forms
are many; energy is one. The unifying scalar is not any particular manifestation but the invariant
that all manifestations share and that the engines conserve. We reproduce this move in the
cost-accounted setting. The coloured tokens are the manifest forms. We adjoin engines---conversion
contracts between token types, of exactly the kind already present as the exchange contract
$\mathsf{Exchange}(c,v)$ of \cite[App.~B]{CARho}. We then ask for the analogue of energy: a single
scalar into which every colour converts.

Two hypotheses make the analogue exist. First, that engines are available across the token graph
(convertibility). Second---and this is the load-bearing condition---that the conversion structure
is \emph{arbitrage-free}: no cycle of conversions returns more value than it consumed. Under these
hypotheses a \emph{virtual token} emerges: a common numeraire, a valuation against which all
tokens are priced, and which agents holding positions across the token distribution may adopt in
place of picking one existing colour as reference. Our thesis is that this virtual token is the
cost-accounted transport of \emph{energy}.

The note is organised as follows. Section~\ref{sec:conv} sets up conversion systems over the
signature monoid. Section~\ref{sec:noarb} proves the pivot: no-arbitrage is definitionally the
existence of the virtual token. Section~\ref{sec:hodge} gives the canonical decomposition of a
conversion system into energy and arbitrage and identifies arbitrage-removal as an idempotent
reflection. Section~\ref{sec:compress} reads the construction as compression and resolves the
object-capability tension by distinguishing unit of account from medium of exchange.
Sections~\ref{sec:first}--\ref{sec:second} separate the first law (conservation, lossless) from
the second (dissipation, friction). Section~\ref{sec:conjugacy} states the energy--time conjugacy
as a conjecture, and Section~\ref{sec:ontology} draws the ontological conclusion.

\section{Conversion systems over a signature monoid}
\label{sec:conv}

We work over the signature structure of \cite{CE}: $(\Sig,\ast,())$ is a commutative monoid, with
$\ast$ the compound-signature operator and $()$ the unit. Token stacks $\Stk$, $S::=()\mid s::S$,
form the free graded structure whose head cells are consumed in step with reduction \cite{CE}.

\begin{definition}[Conversion system]\label{def:conv}
A \emph{conversion system} over $\Sig$ is a tuple $\mathcal{K}=(\Tok,E,r)$ where $\Tok\subseteq\Sig$
is a finite set of token types; $E\subseteq \Tok\times \Tok$ is a symmetric set of \emph{engines}
(unordered pairs, presented with both orientations); and $r:E\to\Rpos$ is a \emph{rate}, with the
\emph{reversibility} law
\[
  r(a\!\to\! b)\cdot r(b\!\to\! a)=1 .
\]
An engine $a\!\to\! b$ is read operationally: one unit of the $a$-token is consumed and
$r(a\!\to\! b)$ units of the $b$-token are released. We write $\Gamma_{\mathcal K}=(\Tok,E)$ for
the underlying (connected, without loss of generality) graph of engines.
\end{definition}

The exchange contract of \cite[App.~B.9]{CARho} is the paradigm engine: a persistent process
$\mathsf{Exchange}(c,v)$ that swaps one $c$-token for one $v$-token and back, i.e.\ an engine with
$r(c\!\to\! v)=1$. Definition~\ref{def:conv} allows arbitrary positive rates and, below, arbitrary
finite graphs of them.

\begin{definition}[Monoidal compatibility]\label{def:monoidal}
$\mathcal K$ is \emph{monoidal} when the token set is closed under $\ast$ on the fragment it names
and the rates respect composition: whenever $a\ast a'$ and $b\ast b'$ are tokens and the component
engines exist,
\[
  r(a\ast a' \to b\ast b') \;=\; r(a\!\to\! b)\cdot r(a'\!\to\! b') .
\]
\end{definition}

Monoidal compatibility is the statement that the ``value'' of a compound authority is built from
the values of its components in the same multiplicative way that $\ast$ builds the authority. It is
the hypothesis under which the potential of Section~\ref{sec:noarb} will be forced to respect
$\ast$, and hence under which the virtual token prices multi-signatures coherently.

\section{No-arbitrage is the virtual token}
\label{sec:noarb}

Fix a conversion system $\mathcal K=(\Tok,E,r)$. Pass to logarithms: the \emph{log-rate} is the
edge $1$-cochain
\[
  \ell\in C^1(\Gamma_{\mathcal K};\Real),\qquad \ell(a\!\to\! b):=\log r(a\!\to\! b),
\]
antisymmetric by reversibility, $\ell(b\!\to\! a)=-\ell(a\!\to\! b)$.

\begin{definition}[Arbitrage]\label{def:arb}
A \emph{cycle} is a closed edge-path $a_0\!\to\! a_1\!\to\!\cdots\!\to\! a_k=a_0$. Its
\emph{yield} is $\prod_{i} r(a_i\!\to\! a_{i+1})=\exp\!\big(\sum_i \ell(a_i\!\to\! a_{i+1})\big)$.
The system exhibits \emph{arbitrage} if some cycle has yield $\neq 1$; it is
\emph{arbitrage-free} if every cycle has yield $1$, equivalently if $\ell$ has vanishing
integral around every cycle.
\end{definition}

\begin{definition}[Virtual token / valuation]\label{def:val}
A \emph{valuation} (or \emph{virtual token}) for $\mathcal K$ is a function $\nu:\Tok\to\Rpos$
with
\[
  r(a\!\to\! b)=\frac{\nu(a)}{\nu(b)}\qquad\text{for every engine }a\!\to\! b .
\]
Writing $p=\log\nu:\Tok\to\Real$ for the \emph{log-price}, the condition is
$\ell(a\!\to\! b)=p(a)-p(b)$; that is, $\ell=-\,\delta p$ is (minus) the graph gradient of a
potential. When $\mathcal K$ is monoidal we additionally require $\nu$ to be a monoid
homomorphism, $\nu(a\ast a')=\nu(a)\,\nu(a')$, equivalently $p$ additive.
\end{definition}

The reading of Definition~\ref{def:val} is that one unit of $a$ and $r(a\!\to\! b)$ units of $b$
have equal value: $\nu(a)=r(a\!\to\! b)\,\nu(b)$. The virtual token is a price list; the engines
are value-preserving by construction against it.

\begin{theorem}[The fundamental equivalence]\label{thm:ftap}
For a connected conversion system $\mathcal K$ the following are equivalent.
\begin{enumerate}[label=\textup{(\roman*)}]
\item $\mathcal K$ is arbitrage-free.
\item $\ell$ is exact: $\ell=-\,\delta p$ for some potential $p:\Tok\to\Real$.
\item A valuation $\nu=\exp p$ exists.
\end{enumerate}
The potential $p$, hence the valuation $\nu$, is unique up to an additive (resp.\ multiplicative)
constant---a global choice of unit. If $\mathcal K$ is monoidal, $\nu$ may moreover be taken to
be a monoid homomorphism, and is then unique up to a multiplicative character
$\chi:\Sig\to\Rpos$.
\end{theorem}

\begin{proof}
(i)$\Rightarrow$(ii). Fix a basepoint $a_\star\in\Tok$ and set $p(a_\star)=0$. For any $a$, choose
a path $a_\star\!\to\!\cdots\!\to\! a$ and put $p(a):=-\sum \ell$ along it. Path-independence is
exactly the vanishing of $\ell$ around cycles (Definition~\ref{def:arb}), so $p$ is well defined;
by construction $\ell=-\delta p$. (ii)$\Rightarrow$(iii) is $\nu=\exp p$, and
(iii)$\Rightarrow$(i) telescopes: any cycle has yield
$\prod_i \nu(a_i)/\nu(a_{i+1})=1$. Uniqueness up to a constant is the connectedness of
$\Gamma_{\mathcal K}$ (harmonic $0$-cochains are constant). For the monoidal refinement, additivity
of $p$ on the generators propagates along $\ast$-compatible engines
(Definition~\ref{def:monoidal}); the residual freedom is a homomorphism
$\Tok\to\Real$, i.e.\ a character of $\Sig$.
\end{proof}

Theorem~\ref{thm:ftap} is the combinatorial core of the fundamental theorem of asset pricing
\cite{HK}: a price system exists iff there is no arbitrage. What matters for us is the direction of
explanation. The virtual token does not have to be \emph{built} once arbitrage is removed; its
\emph{existence is the removal}. Energy is what remains of the conversion structure after one
quotients by the arbitrage cycles.

\begin{remark}[Gauge]
The freedom in $p$ up to an additive constant is precisely the freedom to choose where potential
energy is zero; the freedom in $\nu$ up to a multiplicative constant is the freedom to choose the
unit of the numeraire. Selecting an \emph{existing} colour as reference (as one selects
``prices in dollars'') fixes this gauge at a vertex of $\Gamma_{\mathcal K}$; the virtual token is
the refusal to privilege any vertex.
\end{remark}

\section{Energy and arbitrage: the graph--Hodge decomposition}
\label{sec:hodge}

When $\mathcal K$ is \emph{not} arbitrage-free, Theorem~\ref{thm:ftap}(ii) fails: $\ell$ is not
exact. The exact obstruction is measured by discrete Hodge theory on the graph
$\Gamma_{\mathcal K}$ \cite{Hodge}. Orient the edges and equip $C^0,C^1$ with the standard inner
products. Let $\delta:C^0\to C^1$ be the coboundary (gradient) and $\delta^\ast:C^1\to C^0$ its
adjoint (negative divergence). Then
\[
  C^1 \;=\; \underbrace{\im\delta}_{\text{exact (gradient) flows}}
        \;\oplus\; \underbrace{\ker\delta^\ast}_{\text{cycle (divergence-free) flows}} ,
\]
an orthogonal direct sum.

\begin{definition}[Energy and arbitrage components]
Write the log-rate as $\ell=\ell_{\mathrm{en}}+\ell_{\mathrm{arb}}$ along this splitting:
$\ell_{\mathrm{en}}\in\im\delta$ is the \emph{energy component} (an exact, hence arbitrage-free,
rate structure) and $\ell_{\mathrm{arb}}\in\ker\delta^\ast$ is the \emph{arbitrage component} (a
divergence-free circulation with no potential).
\end{definition}

\begin{proposition}[Arbitrage is a cohomology class]\label{prop:betti}
$\mathcal K$ is arbitrage-free iff $\ell_{\mathrm{arb}}=0$. The space of arbitrage components has
dimension equal to the first Betti number of the engine graph,
\[
  \dim\ker\delta^\ast \;=\; \Betti(\Gamma_{\mathcal K}) \;=\; |E|-|\Tok|+1
\]
(for connected $\Gamma_{\mathcal K}$). In particular a tree of engines
($\Betti=0$) admits no arbitrage, and each independent cycle contributes exactly one degree of
arbitrage freedom.
\end{proposition}

\begin{proof}
The class of $\ell$ in $C^1/\im\delta\cong H^1(\Gamma_{\mathcal K};\Real)$ vanishes iff $\ell$ is
exact iff (Theorem~\ref{thm:ftap}) $\mathcal K$ is arbitrage-free; and
$\dim H^1=\Betti$ for a graph.
\end{proof}

Now the operational content of ``shaking the arbitrage out.'' The orthogonal projection
\[
  \Pi:C^1\longrightarrow \im\delta,\qquad \Pi\,\ell=\ell_{\mathrm{en}},
\]
is the least-squares replacement of the observed rates by the nearest arbitrage-free rates; it
recovers a potential $p$ (hence a virtual token) up to the harmonic constant. Two properties are
worth stating because they contrast sharply with the cost functor.

\begin{proposition}[Arbitrage-removal is an idempotent reflection]\label{prop:reflect}
Let $\Conv$ be the category of conversion systems on a fixed graph and $\Convf\hookrightarrow\Conv$
the full subcategory of arbitrage-free ones. The map $L:\ell\mapsto\Pi\ell$ is a reflector: it is
left adjoint to the inclusion, its unit $\eta_{\mathcal K}:\ell\Rightarrow \iota L\,\ell$ is the
natural transformation sending each system to its arbitrage-free projection, and the induced monad
$\iota L$ is \emph{idempotent} ($\Pi^2=\Pi$).
\end{proposition}

\begin{remark}[Two monads of opposite character]
The idempotence of Proposition~\ref{prop:reflect} is the exact foil to the cost monad
$(\cost,\eta,\mu)$ of \cite[\S9]{CE}, which is \emph{non}-idempotent: $\mu$ merges two resource
accounts rather than collapsing a redundant copy, so re-metering a metered calculus yields a finer
account. Metering \emph{accumulates}; arbitrage-removal \emph{saturates}. One squeeze suffices to
expel arbitrage, because the cyclic part of a fixed cochain is expelled once and does not return;
whereas each application of $\cost$ genuinely adds a layer of signatures and a stack-of-stacks.
The two constructions are monads on the same material with opposite idempotence, and the note
\cite{CE} makes both natural.
\end{remark}

The categorical packaging of Proposition~\ref{prop:reflect} is stated at the level of rigour a
working note allows: it presumes a choice of morphisms of conversion systems (rate-nonincreasing
graph maps are the natural candidate, matching the dissipative order of
Section~\ref{sec:second}); pinning these down and verifying the universal property in full is
deferred. The concrete content---orthogonal projection onto the exact cochains, idempotent, with
$\eta$ the assignment of the projected system---is exact.

\section{Compression, coordination, and the unit of account}
\label{sec:compress}

\subsection*{Compression}
A conversion system carries, a priori, one rate per engine: up to $O(n^2)$ real numbers for
$n=|\Tok|$ tokens with all pairwise engines. Theorem~\ref{thm:ftap} collapses this to the
potential: $O(n)$ numbers, one price per token, minus the global gauge constant.

\begin{proposition}[No-arbitrage is a rank condition]
For a complete engine graph, arbitrage-freedom is equivalent to the log-rate matrix
$L_{ab}=\log r(a\!\to\! b)$ being the \emph{difference matrix of a single potential},
$L_{ab}=p(a)-p(b)$; i.e.\ $L=p\,\mathbf{1}^{\!\top}-\mathbf{1}\,p^{\!\top}$. The bilateral data thus
compress from the $\binom{n}{2}$ independent entries of a general antisymmetric $L$ to the $n-1$
independent entries of $p$.
\end{proposition}

The compression is the coordination win read statically: agents holding positions across the token
distribution need agree only on the single broadcast field $p:\Tok\to\Real$ rather than
continuously reconcile $O(n^2)$ bilateral rates. The virtual token is the minimal sufficient
statistic of a coherent conversion structure.

\subsection*{Unit of account, not medium of exchange}
There is an immediate objection from the object-capability discipline of \cite[\S4.8.5]{CARho}:
weakening is forbidden. A compound token minted as $s_{\text{sender}}\ast s_{\text{compliance}}$
encodes ``neither party alone may act''; an engine that converted it into a single $s'$ would
launder away the compliance requirement. If the virtual token were an engine to which
\emph{everything} converts, the multi-signature guarantee would evaporate.

The resolution turns on a distinction the physics analogy makes vivid and that
Definition~\ref{def:val} already encodes: the virtual token is a \emph{unit of account}, not a
\emph{medium of exchange}.

\begin{proposition}[Pricing does not launder authority]\label{prop:noweaken}
A valuation is a monoid homomorphism $\nu:(\Sig,\ast,())\to(\Rpos,\cdot,1)$, an arrow \emph{out of}
$\Sig$; it is not an element of $\Sig$ and it induces no engine. It assigns
$\nu(s_1\ast s_2)=\nu(s_1)\,\nu(s_2)$ without providing any reduction that consumes an
$s_1\ast s_2$-token in exchange for an $s'$-token. Hence pricing a compound authority is compatible
with the prohibition on weakening: the number $\nu(s_1\ast s_2)$ commits no holder to accepting
$s_1$ alone.
\end{proposition}

Nobody ever \emph{holds} virtual tokens; one \emph{prices} in them. A unit of account requires no
custody, no issuer, and no depletable pool, and it privileges no agent's colour---which is exactly
why it lowers coordination cost below that of anointing an existing token.

\begin{remark}[Scope: object capabilities are specific to nominal-surface theories]
\label{rmk:scope}
The objection this subsection answers, and hence the guarantee it secures, is meaningful only for
those ciGSLTs that carry an object-capability discipline in the first place. By \cite[\S10]{CE},
whether proximity to a resource stack confers authority is dictated by the equational theory of the
interaction constructor $K$. When $K$ is non-free---the associative--commutative $\mid$ of the
channel-based mobile process calculi (rho, $\pi$)---the surface is carried \emph{nominally} by a
channel name, authority is possession of a matching name, and the weakening prohibition of
\cite[\S4.8.5]{CARho} is a genuine capability constraint; there Proposition~\ref{prop:noweaken} is
the load-bearing statement that pricing does not forge that possession. When $K$ is free---the
rigid application spine of combinatory theories such as SKI, or of the $\lambda$-calculus, where
authority is carried \emph{structurally} by unforgeable position---there is no nominal capability
to launder, and the weakening question does not arise in the same form: the valuation is then
simply a grading, with no authority discipline to protect. So the claim ``the object-capability
discipline survives'' should be read with its antecedent: \emph{where} such a discipline is
present---the nominal-surface calculi---pricing is compatible with it; for the structural-surface
theories the virtual token is a pure grading and the point is vacuous rather than false.
\end{remark}

\begin{remark}[Dual to the $s_0$-collapse]
Setting every signature to a single distinguished $s_0$ recovers undifferentiated phlogiston
\cite[\S6, Rmk.~5.1]{CARho}---the monochromatic limit, which \emph{discards} the spectrum. The
virtual token is the dual move: keep the spectrum---chemical, mechanical, electrical stay distinct
as colours---and adjoin a conserved scalar \emph{above} them that makes them commensurable. Energy
does not abolish the difference between chemical and electrical potential; it grades them onto one
axis. The $s_0$-collapse erases colour; the valuation $\nu$ measures it.
\end{remark}

\section{Conservation: the first law}
\label{sec:first}

Extend $\nu$ additively from tokens to stacks and multisets by
$\nu(s::S)=\nu(s)+\nu(S)$, $\nu(())=0$ (the temporal monoid $::$ read through the price). This is
the \emph{total value} of a token supply.

\begin{proposition}[Value conservation in the lossless regime]\label{prop:first}
In an arbitrage-free conversion system, the total value $\nu$ of a token supply is invariant under
every engine operation: an engine $a\!\to\! b$ replaces one $a$-cell of value $\nu(a)$ by
$r(a\!\to\! b)$ $b$-cells of total value $r(a\!\to\! b)\,\nu(b)=\nu(a)$. Consequently $\nu$ is a
conserved quantity of the splitter/joiner token economy of \cite[App.~A.6]{CARho}.
\end{proposition}

Proposition~\ref{prop:first} upgrades the \emph{conservation of authority} of
\cite[Prop.~4.7]{CARho}---the invariance of the consumed signature \emph{multiset} across every
partition of a join---to a scalar law. Where \cite{CARho} conserves a multiset, the virtual token
conserves its total price. This is the first law: in the reversible idealisation, value is neither
created nor destroyed by conversion, only redistributed across colours.

\section{Friction: the second law}
\label{sec:second}

The platform's design philosophy is \emph{valuable friction} \cite[\S5.3]{CARho}: every
resource-consuming operation carries an explicit cost. Applied to the engines themselves, this says
real conversion is \emph{lossy}. Model a lossy engine by a rate strictly below the value-preserving
one: converting $a\!\to\! b$ yields
$\rho(a\!\to\! b)=r(a\!\to\! b)\,e^{-\theta(a\to b)}$ with dissipation
$\theta(a\!\to\! b)\ge 0$.

\begin{proposition}[Dissipation breaks exactness into an inequality]\label{prop:second}
With lossy engines the log-rate cochain $\hat\ell=\ell-\theta$ satisfies, around every oriented
cycle,
\[
  \oint \hat\ell \;=\; -\oint \theta \;\le\; 0 ,
\]
with equality iff the cycle is frictionless. Hence no exact potential exists in general; instead
the total value $\nu$ (computed against any fixed reference potential) is \emph{non-increasing}
along engine operations, and strictly decreasing whenever a lossy engine fires. $\nu$ is thus a
Lyapunov function for the token economy.
\end{proposition}

\begin{remark}[Free energy, exergy, and mortality]
Proposition~\ref{prop:second} is the second law arriving on schedule. The frictionless virtual
token of Section~\ref{sec:first} is a conserved \emph{energy}; the realistic, friction-bearing
virtual token is a monotone \emph{free energy} (exergy)---the part of the value still able to do
computational work, dissipated by $\oint\theta$ as an entropy analogue. This is already the ambient
picture of \cite{MC}, where metabolism is motion on a Ruelle--Bowen landscape and death is the
absorbing set: a process whose free energy is exhausted deadlocks by starvation
\cite[Def.~1]{MC}. The Lyapunov $\nu$ of Proposition~\ref{prop:second} is a certificate of that
descent---value degrades monotonically toward the starvation boundary. The clean conserved energy
is the reversible limit $\theta\equiv 0$; the cost-accounted calculus, being frictionful by
design, lives in the dissipative regime, and its honest scalar is free energy.
\end{remark}

\section{The generative half: an energy--time conjecture}
\label{sec:conjugacy}

Physical energy wears two hats. It is the conserved charge of Sections~\ref{sec:first}--\ref{sec:second}, and it is the
\emph{generator of time evolution}---the Hamiltonian. The virtual token, so far, wears only the
first hat: it is a valuation, not a generator of reduction. Yet the missing conjugate variable is
already present in the construction. By \cite[Prop.~4.1]{CE} the temporal token stack $::$ is the
reified clock: its consumed prefix is an exact operational modulus, the number of forced redexes,
the run length itself. Time, in a cost-accounted GSLT, is $::$.

\begin{conjecture}[Energy--time conjugacy]\label{conj:conjugacy}
In an arbitrage-free, frictionless cost-accounted GSLT, the total virtual-token value $\nu$ is the
conserved quantity \emph{conjugate to translation along the temporal monoid} $::$, in the sense of
a Noether \cite{Noether} correspondence: the path-independence symmetry of conversion
(Theorem~\ref{thm:ftap}) is the internal symmetry whose conserved charge is $\nu$, and its coupling
to $::$-translation makes $\nu$ the generator of a step of reduction. Under friction the
correspondence degrades to the second-law inequality of Proposition~\ref{prop:second}: $\nu$
generates a \emph{descent} rather than a symmetry.
\end{conjecture}

Conjecture~\ref{conj:conjugacy} is offered in the spirit of the physical correspondences of
\cite[\S13]{CE}, not as a theorem. What it would buy is the second hat: an identification of the
conserved charge with the generator of dynamics, which in physics is a non-trivial coincidence
(the Hamiltonian is both). The cost-accounted setting is unusual in letting the two be prised apart
and studied separately---the charge is manifest, the generator conjectural---and in offering a
concrete candidate ($::$) for the conjugate time.

\section{Ontology: a charge, not a substance}
\label{sec:ontology}

We can now answer the ontological question that motivated the note. \emph{What is the virtual
token?} It is not a substance---not a stuff one holds, transfers, or points at.
Proposition~\ref{prop:noweaken} makes this precise on the syntactic side (it is an arrow out of
$\Sig$, never an inhabitant), and Theorem~\ref{thm:ftap} makes it precise on the semantic side (it
is the potential forced into existence by a symmetry of the conversion dynamics---the
path-independence of engines). Its ontological status is that of a \emph{Noether charge}: the
conserved scalar associated with an invariance, in the deflationary reading of energy under which
energy \emph{is} ``the conserved quantity associated with time-translation symmetry'' rather than a
substance the world contains. Both are invariants of an equivalence, not occupants of it.

Honesty requires the counter-reading. Physical energy earns a thicker ontology than mere
bookkeeping because it also \emph{generates} the dynamics: the same object that is conserved is the
Hamiltonian that pushes the state forward in time, and their coincidence is substantive. The
virtual token, as established here, discharges only the bookkeeping role; the generative role is
Conjecture~\ref{conj:conjugacy}. So the sharp form of the ontological question is not settled but
sharpened: \emph{is energy a conserved charge that happens to generate time, or two coincident
things?} In the cost-accounted GSLT one may hold the two apart---the charge proven, the generator
open---and watch whether they want to merge. That is more than the physics affords, and it is the
reason the transported notion is worth taking seriously rather than as mere analogy.

\section*{Further work}
\addcontentsline{toc}{section}{Further work}

Three threads deserve promotion beyond working-note rigour.
First, the reflection of Proposition~\ref{prop:reflect}: fix the morphisms of $\Conv$ (the
dissipative order of Section~\ref{sec:second} suggests rate-nonincreasing maps), and verify the
adjunction and idempotence against them, situating the arbitrage-removal monad beside the cost
monad $\cost$ of \cite{CE} in one $2$-categorical picture.
Second, the free-energy reading of Section~\ref{sec:second}: identify the state function whose
descent Proposition~\ref{prop:second} certifies with the exergy of a Ruelle--Bowen potential on the
transition graph \cite{MC}, and make the ``entropy production'' $\oint\theta$ a genuine thermodynamic
quantity rather than an analogy.
Third, Conjecture~\ref{conj:conjugacy}: specify the symmetry action of $::$-translation and the
Noether pairing precisely enough to decide whether the conserved $\nu$ generates a reduction step,
closing---or refuting---the energy--time conjugacy.

\bigskip
\noindent\emph{E Pluribus, Unum Pretium.}

\begin{thebibliography}{99}

\bibitem{CARho} L.~G.~Meredith, \emph{Cost-Accounted Rho Calculus: A Spectral Decomposition of
Phlogiston}, F1R3FLY.io, 2026.

\bibitem{CE} L.~G.~Meredith, \emph{Continued Interactive GSLTs and the Cost Endofunctor: A
construction one level up from cost-accounted Rholang}, F1R3FLY.io, 2026.

\bibitem{MC} L.~G.~Meredith, \emph{Mortal Computation: Mortality, the Genome, and Life-History in
Cost-Accounted GSLTs}, F1R3FLY.io working note, 2026.

\bibitem{RHO} L.~G.~Meredith and M.~Radestock, \emph{A reflective higher-order calculus},
Electronic Notes in Theoretical Computer Science, vol.~141, no.~5, pp.~49--67, 2005.

\bibitem{HK} J.~M.~Harrison and D.~M.~Kreps, \emph{Martingales and arbitrage in multiperiod
securities markets}, Journal of Economic Theory, vol.~20, no.~3, pp.~381--408, 1979.

\bibitem{Hodge} X.~Jiang, L.-H.~Lim, Y.~Yao, and Y.~Ye, \emph{Statistical ranking and combinatorial
Hodge theory}, Mathematical Programming, vol.~127, no.~1, pp.~203--244, 2011.

\bibitem{Girard} J.-Y.~Girard, \emph{Linear logic}, Theoretical Computer Science, vol.~50, no.~1,
pp.~1--101, 1987.

\bibitem{Noether} E.~Noether, \emph{Invariante Variationsprobleme}, Nachrichten von der
Gesellschaft der Wissenschaften zu G\"ottingen, Mathematisch-Physikalische Klasse, pp.~235--257,
1918.

\bibitem{MacLane} S.~Mac~Lane, \emph{Categories for the Working Mathematician}, 2nd ed., Graduate
Texts in Mathematics, vol.~5, Springer, 1998.

\end{thebibliography}

\end{document}
