\documentclass[11pt,a4paper]{article}

%% ── packages ──────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{stmaryrd}            % \llbracket, \rrbracket
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{booktabs}
\usepackage{listings}

%% ── listings style ────────────────────────────────────────────────
\lstdefinestyle{rholang}{
  basicstyle=\small\ttfamily,
  keywordstyle=\bfseries,
  commentstyle=\itshape\color{gray},
  morekeywords={for,new,let,in,match},
  literate={<-}{{$\leftarrow$}}2
           {!}{{!}}1
           {|}{{$\mid$}}1,
  breaklines=true,
  frame=single,
  framesep=6pt,
  xleftmargin=1em,
  xrightmargin=1em,
  aboveskip=\medskipamount,
  belowskip=\medskipamount,
}

%% ── theorem environments ──────────────────────────────────────────
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

%% ── macros ────────────────────────────────────────────────────────
\newcommand{\lsem}{\llbracket}
\newcommand{\rsem}{\rrbracket}

% calculus syntax
\newcommand{\Par}{\mathbin{|}}
\newcommand{\forc}{\mathbf{for}}
\newcommand{\send}[2]{#1 !\!\left( #2 \right)}
\newcommand{\recv}[3]{\forc\!\left( #1 \leftarrow #2 \right) #3}
\newcommand{\nil}{\mathbf{0}}
\newcommand{\drop}[1]{{\ast} #1}
\newcommand{\quot}[1]{@#1}
\newcommand{\subst}[3]{#1\{#2 / #3\}}
\newcommand{\signed}[2]{\{#1\}_{#2}}

% signatures and tokens
\newcommand{\comp}{\mathbin{*}}
\newcommand{\hashf}{\texttt{\#}}
\newcommand{\bigParop}{\mathop{\big|}\displaylimits}
\newcommand{\bigCompop}{\mathop{\text{\Large$\ast$}}\displaylimits}

% structural equivalence and reduction
\newcommand{\Seq}{\equiv_{\!S}}
\newcommand{\Neq}{\equiv_{\!N}}
\newcommand{\red}{\rightsquigarrow}

% free names
\newcommand{\FN}{\mathrm{FN}}

%% ── headers ───────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textsc{F1R3FLY.io}}
\fancyhead[R]{\small\textsc{A Spectral Decomposition of Phlogiston}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

%% ── title ─────────────────────────────────────────────────────────
\title{%
  \textbf{Cost-Accounted Rho Calculus}\\[6pt]
  \large A Spectral Decomposition of Phlogiston}
\author{%
  L.\ Gregory Meredith\thanks{F1R3FLY.io; F1R3FLY Industries.
  \texttt{lgreg.meredith@f1r3fly.io}}%
}
\date{May 2026}

%% ══════════════════════════════════════════════════════════════════
\begin{document}
\maketitle

\begin{abstract}
We construct a cost-accounted version of the rho calculus by
exhibiting the rho calculus as an object of the category
\textbf{ciGSLT} of continued interactive graph-structured
lambda theories and then applying the cost
endofunctor~$\mathcal{C}$ to it.  The instantiation identifies
the rho interaction cut ($K = |$, $K_p = \mathbf{for}$,
$K_e = {!}$, contraction = Milner's pseudo-application);
applying~$\mathcal{C}$ produces a calculus in which signatures
are a species of name, token stacks are first-class terms, and
every continuation is a metered thunk by construction; and
Turing completeness lets us translate the result back to pure
rho via the internalisation adjunction.  The signature grammar
is parameterized by a type variable~$G$ representing the
cryptographic backend, making the calculus generic over
signature schemes.  The object-capability analysis of the
general construction ensures that the AC nature of parallel
composition does not introduce ambient authority: located
resource stacks and the nearness predicate (channel-name
equality) confine authority to holders of matching channel
names.

We identify the for-comprehension as the transaction primitive
of the calculus: the binding specification is the transaction,
the substitution is the witness, and application to the
continuation is the state update.  Individual transactions are
atomic by construction (the subject-reduction lemma for
wrapping).  Multi-step transactions are made atomic by a
decidable static analysis at deployment-acceptance time: a
linear resource proof verifying that every for-comprehension in
the deployment's call graph is funded before any step executes.

Because the rho calculus is the formal core of Rholang, this
instantiation serves as the design specification for
rearchitecting phlogiston accounting in Rholang~1.4 (MeTTaIL)
and the F1R3FLY platform.
\end{abstract}

\tableofcontents
\bigskip

%% ══════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

The rho calculus~\cite{MeredithRadestock2005} is a reflective
higher-order process calculus: channels are the quoted codes of
processes, and processes may be recovered from channels by
dequotation.  As the formal core of
Rholang~\cite{RholangSpec}, the rho calculus underpins the
computation and storage model of the F1R3FLY platform.

Rholang's cost-accounting mechanism --- the \emph{phlogiston}
(phlo) system --- gates every deploy behind a token balance
associated with a digital signature.  In a companion
paper~\cite{CostEndofunctor} we showed that this construction
is an instance of a generic endofunctor~$\mathcal{C}$ on the
category \textbf{ciGSLT} of \emph{continued interactive
graph-structured lambda theories}: interactive theories
presented as an \emph{interaction cut} (contact constructor~$K$,
two introductions $K_p(I, C_p)$ and $K_e(J, C_e)$ meeting
along matching interaction surfaces, contraction
$\mathit{compute}$), equipped with a computable section of
their structural-congruence quotient and with continuation
slots sorted as wrapped terms by construction.

The present paper constructs a cost-accounted rho calculus by
exhibiting the rho calculus as an object of \textbf{ciGSLT} and
applying~$\mathcal{C}$ to it, developed in sufficient detail to
serve as the design specification for rearchitecting phlogiston
accounting in Rholang~1.4 (MeTTaIL) and the F1R3FLY platform.
The development proceeds in three phases:

\begin{enumerate}[label=\textbf{Phase \arabic*.}]
  \item \textbf{Instantiation.}  We exhibit the rho calculus
    as an object of \textbf{ciGSLT} by identifying the
    interaction-cut components: $K = \Par$ (associative--commutative), $K_p = \forc$ with surface
    $I = x$ and program continuation $C_p = \lambda y.\,T$,
    $K_e = {!}$ with surface $J = x$ and degenerate
    environment continuation $C_e = -$, and $\mathit{compute}$
    is Milner's pseudo-application (semantic substitution with
    empty release).  The section is $\hashf\!{} = \mathit{digest}
    \circ \mathit{cf}$ where $\mathit{cf}$ is a canonical-form
    function on processes.

  \item \textbf{Cost construction.}  Applying the endofunctor
    $\mathcal{C}$ to this instantiation produces the
    cost-accounted rho calculus: signatures become a species
    of name, token stacks become first-class terms, the
    $\forc$/$!$ operators range over signed terms, and $\Par$
    becomes polymorphic over signed-term/stack pairs.  The five
    rewrite rules are the gated family (R1)--(R3) of the
    general construction, specialised to the rho interaction cut.

  \item \textbf{Internalisation.}  Because the rho calculus is
    Turing-complete, the internalisation adjunction
    (Adjunction~II of~\cite{CostEndofunctor}) applies: the
    entire cost-accounting apparatus can be encoded back into
    the pure rho calculus via a source-to-source translation.
    This translation is a compositional compiler pass that
    emits pure Rholang~1.1, giving the platform immediate
    access to cost-accounted semantics while the native
    implementation is developed.
\end{enumerate}

\noindent
The key design moves, now understood as consequences of the
general construction, are:

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Signatures are names.}  The name sort is
    extended so that $x, y \;::=\; \quot{T} \mid s$, where~$s$
    ranges over signatures.  Signatures and quoted processes
    coexist as first-class channel names.
  \item \textbf{Section and quotient disentangled.}  In the
    general construction~\cite{CostEndofunctor}, the section
    $\hashf = \mathit{digest} \circ \mathit{cf}$ (authentication)
    and the structural-congruence quotient (communication) are
    independent.  In rho, the operator~$\quot{\cdot}$ serves
    both roles, which is why the cost story previously appeared
    to need ``two axes of reflection.''  They are not two
    signature schemes; one axis ($\quot{\cdot}$ up to~$\Seq$)
    is the communication quotient the LTS already carries, the
    other ($\hashf = \mathit{digest} \circ \mathit{cf}$) is the
    authentication section the cost functor adds.
  \item \textbf{Signed terms pervade the syntax.}  Every
    continuation slot is re-sorted to the wrapped-term sort~$T$,
    so every redex is born inside a wrapper $\signed{\cdot}{s}$
    --- the ``wrapping by construction'' discipline of the
    general theory.
  \item \textbf{Token stacks are first-class terms.}  Token
    stacks $S \;::=\; () \mid s : S$ are the temporal
    monoid of the general construction (Section~4
    of~\cite{CostEndofunctor}), reifying the computation's time
    axis as a term.
  \item \textbf{$\Par$ is polymorphic.}  Parallel composition
    is the interaction constructor~$K$, made polymorphic by
    the cost functor to range over signed-term/stack and
    stack/stack pairs (Remark~2.2 of~\cite{CostEndofunctor}:
    $K$ is the only constructor that needs overloading).
  \item \textbf{Generic cryptography.}  The signature grammar
    is parameterized by a type variable~$G$ representing the
    ground signature scheme (Ed25519, secp256k1, OQS, etc.),
    making the calculus independent of any particular
    cryptographic backend.
\end{enumerate}

\noindent
Because the rho calculus \emph{is} the core of Rholang, this
instantiation is not a theoretical exercise --- it is the
design specification for rearchitecting phlogiston accounting in
Rholang~1.4 (MeTTaIL) and the F1R3FLY platform.

%% ══════════════════════════════════════════════════════════════════
\section{Motivation: From Run-to-Completion to Concurrent Acceptance}
\label{sec:motivation}

The redesign presented in this paper is motivated by a
fundamental architectural limitation in the current F1R3FLY
implementation, inherited from RChain.

\subsection{The Run-to-Completion Model}
\label{sec:run-to-completion}

In the current architecture, each deployment submitted to a
F1R3Node is executed under \emph{run-to-completion} semantics:
the RSpace tuplespace engine accepts a deployment, executes it
to termination (or until phlogiston is exhausted), commits the
resulting state changes, and only then accepts the next
deployment.  This model introduces two related problems:

\begin{enumerate}[label=(\roman*)]
  \item \textbf{Implicit transaction boundaries.}  The entire
    deployment is treated as a single, implicit transaction.
    There is no formal specification of what constitutes a
    transaction --- the deployment boundary \emph{is} the
    transaction boundary, by convention rather than by
    design.  This conflates the unit of submission (the
    deployment) with the unit of atomicity (the transaction),
    making it impossible to express deployments that contain
    multiple independent transactions or to share
    transactional structure across deployments.

  \item \textbf{Limited availability.}  Because RSpace
    executes deployments sequentially, it is \emph{locked}
    for the duration of each deployment's execution.  No
    other deployment can be accepted, pattern-matched, or
    reduced while the current deployment is running.  The
    throughput of the F1R3Node is bounded by the execution
    time of the slowest deployment in the queue.  This is a
    severe bottleneck: a single expensive deployment blocks
    all other activity on the node.
\end{enumerate}

\subsection{The Merge Analysis Burden}
\label{sec:merge-burden}

To recover concurrency in the current architecture, F1R3FLY
(following RChain) employs \emph{merge analysis}: after
multiple deployments have been executed independently (e.g.,
by different validators or in speculative parallel
execution), their resulting states must be \emph{merged}
into a single consistent state.  The merge analysis
determines whether the independently computed states are
compatible (no conflicting writes to the same tuplespace
entries) and, if so, how to combine them.

Merge analysis is expensive, complex, and a persistent
source of bugs.  It requires tracking read and write sets
for every deployment, computing conflict graphs, and
resolving or rejecting conflicting state transitions.
Moreover, the analysis is \emph{post-hoc}: deployments are
executed speculatively, and conflicts are detected only
after the work has been done.  Wasted computation is the
norm, not the exception.

\subsection{The New Model: Acceptance by Linear Proof}
\label{sec:new-model}

The cost-accounted rho calculus eliminates both problems by
replacing run-to-completion with \emph{concurrent acceptance
gated by linear resource proofs}.

Under the new model:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Explicit transaction boundaries.}  The
    for-comprehension is the transaction primitive
    (Section~\ref{sec:for-as-txn}).  Transaction boundaries
    are determined by the syntactic structure of signed terms,
    not by deployment boundaries.  A deployment may contain
    multiple transactions; multiple deployments may
    participate in a single financial transaction (via shared
    signature channels).

  \item \textbf{Concurrent acceptance.}  RSpace can accept
    deployments \emph{at any time}, provided each deployment
    carries a linear resource proof
    (Section~\ref{sec:linear-proof}) demonstrating that
    sufficient tokens exist to complete all of its
    transactions.  Multiple deployments can be active
    simultaneously, each drawing from its own committed token
    supply.  RSpace is never locked by a single deployment.

  \item \textbf{Throughput.}  Because acceptance is gated by a
    static check (linear in the AST size) rather than by
    sequential execution, the F1R3Node can accept and begin
    executing deployments in parallel.  The throughput
    bottleneck shifts from ``one deployment at a time'' to
    ``as many deployments as there are non-conflicting token
    supplies'' --- a dramatic improvement.

  \item \textbf{Merge via signatures and cost accounting.}
    The need for post-hoc merge analysis is substantially
    reduced.  In the cost-accounted calculus, conflicts are
    detected \emph{at acceptance time} through the linear
    resource proof: two deployments that compete for the same
    tokens are two proof obligations competing for the same
    linear hypotheses, and at most one can succeed
    (Section~\ref{sec:deploy-boundaries}).  Merging is no
    longer a separate analysis pass over execution traces ---
    it is a consequence of the signature and cost-accounting
    structure of the calculus itself.

    Specifically, deployments signed by \emph{different}
    signatures draw from \emph{disjoint} token pools and
    cannot conflict.  They may be executed in parallel and
    their results composed without merge analysis.
    Deployments signed by the \emph{same} signature (or
    overlapping compound signatures) compete for the same
    token pool and are serialized by the linear proof ---
    again without merge analysis.  The only case requiring
    attention is deployments that interact via shared
    data channels, and even here the cost-accounting
    structure provides a natural serialization order.
\end{enumerate}

\begin{remark}[From blocking to budgeting]
  The architectural shift can be summarized in a single
  sentence: the current model asks ``is the tuplespace
  free?''\ (blocking); the new model asks ``is this
  deployment funded?''\ (budgeting).  The first question
  serializes all activity.  The second question can be
  answered independently for each deployment, enabling
  concurrency proportional to the diversity of the
  token supply.
\end{remark}

\subsection{New Capability: Multi-Signature Execution}
\label{sec:multisig-motivation}

Beyond streamlining execution and increasing throughput, the
cost-accounted calculus introduces a feature absent from the
current architecture: \emph{$n$-party multi-signature
execution}.

The compound signature $s_1 \comp s_2 \comp \cdots \comp s_n$
in the signed term
$\signed{P}{s_1 \comp \cdots \comp s_n}$
requires that \emph{all} $n$ parties explicitly sign off on the
execution of~$P$.  The rewrite rules enforce this structurally:
the process cannot fire unless tokens from every component
signature are present.  There is no separate multi-sig
verification layer, no threshold-signature scheme, no
off-chain coordination protocol.  Multi-sig is a consequence
of the grammar.

\paragraph{Programmable tokenized capabilities.}
The interaction between compound signatures and token stacks
yields a rich \emph{object-capability} model with
programmable amplification and attenuation:

\begin{description}[style=nextline, leftmargin=2em, nosep]
  \item[Attenuation.]
    A token stack limits what a capability can do.  A process
    signed by~$s$ with a two-token stack $s : s : ()$ can
    perform at most two funded actions.  By controlling the
    depth of the token stack granted to a subprocess, a parent
    process \emph{attenuates} the subprocess's authority.
    This is token-metered capability attenuation: the child
    can do no more than the parent has budgeted.

  \item[Amplification.]
    Composing signatures via $\comp$ \emph{amplifies}
    authority.  A process signed by $s_1 \comp s_2$ can
    perform actions that neither~$s_1$ nor~$s_2$ could
    authorize alone.  For example, a funds-transfer contract
    might require $\signed{\mathit{transfer}}
    {s_{\mathit{sender}} \comp s_{\mathit{compliance}}}$,
    meaning the transfer executes only if both the sender and
    the compliance officer have signed.  Neither party alone
    can move funds; their composed authority can.

  \item[Programmable delegation.]
    Because signatures are names and token stacks are
    first-class terms, a process can \emph{delegate}
    authority by sending tokens on a signature channel.  A
    manager process holding $s_{\mathit{mgr}} : S$ can send a
    portion of its stack to a subordinate process, granting
    the subordinate a metered budget of the manager's
    authority.  When the subordinate exhausts its budget, it
    halts --- no revocation protocol needed.

  \item[Just-in-time funding.]
    Because token stacks are first-class terms, they can be
    \emph{constructed and supplied dynamically} --- not merely
    pre-allocated at deployment time.  A contract can mint a
    token stack $s : s : ()$ and send it on a signature
    channel at the precise moment it is needed, funding a
    capability on an as-needed basis.

    This enables natural \emph{on-boarding} patterns: a
    platform contract grants a new user a small slush fund
    (a shallow token stack) to explore the system.  The
    user's initial deployments are funded by this grant,
    letting them experiment without acquiring tokens of
    their own.  Once the slush fund is exhausted, further
    execution requires the user to acquire tokens through
    the token economy --- purchase, earn, or receive via
    delegation.  The transition from subsidized exploration
    to self-funded participation is not a policy check
    external to the calculus; it is a \emph{resource
    exhaustion event} within it.
\end{description}

\noindent
This capability model is \emph{intrinsic} to the calculus:
amplification, attenuation, delegation, and just-in-time
funding are not library features or design patterns but
direct consequences of the signature algebra and token-stack
semantics.  The F1R3FLY platform's token-attenuated
object-capability model for agentic AI
governance~\cite{VACRADA} is a direct application of this
structure.

\paragraph{Simplified minting.}
Because token stacks are first-class terms that can be
constructed, communicated, and composed within the calculus,
\emph{minting} --- the creation of new tokens --- is a
straightforward operation: a minting contract sends a
freshly constructed token stack on the appropriate signature
channel.  No privileged runtime hook or external oracle is
required.  The protocol's phlogiston minting at epoch
boundaries, the slush-fund grants for new users, and the
fee-to-stake conversion all reduce to the same primitive:
constructing a token stack and sending it on a channel.
This uniformity dramatically simplifies the implementation
of token economies, including ERC-20-like fungible token
contracts, which become expressible as standard
cost-accounted Rholang programs --- a direction we develop
in a companion paper.

%% ══════════════════════════════════════════════════════════════════
\section{The Rho Calculus as a Continued Interactive GSLT}
\label{sec:instantiation}

This section presents the rho calculus as an object of
\textbf{ciGSLT} by identifying each component of the
interaction-cut presentation~\cite{CostEndofunctor}.

\subsection{The Interaction-Cut Components}
\label{sec:cut-components}

\begin{definition}[Rho interaction cut]
  \label{def:rho-cut}
  The rho calculus is a continued interactive GSLT with:
  \begin{description}[style=nextline, leftmargin=2em, nosep]
    \item[Interaction constructor $K = \Par$.]
      Associative--commutative (an unordered bag of processes).
      This is the spatial monoid of the construction: it records
      what is in contact with what.
    \item[Program introduction $K_p = \forc$.]
      The for-comprehension $\recv{y}{x}{T}$, with interaction
      surface $I = x$ (the channel) and program continuation
      $C_p = \lambda y.\,T$ (the abstraction on~$y$ over the
      continuation~$T$).
    \item[Environment introduction $K_e = {!}$.]
      The send $\send{x}{U}$, with interaction surface
      $J = x$ (the same channel) and environment continuation
      $C_e = -$ (degenerate: asynchronous output carries no
      sender continuation).
    \item[Surface match: $\mathit{near}(I, J)$.]
      Name equality of channels: $I = x = J$.  Because $K = \Par$
      is AC, the surface must be carried nominally (by the
      channel name) rather than structurally (by position) ---
      see Section~\ref{sec:ocap-authority}.
    \item[Contraction: $\mathit{compute}$.]
      Milner's pseudo-application: semantic substitution with
      empty release.
      $\mathit{compute}(x, x, \lambda y.\,T, U) = (T\{\quot{U}/y\}, -)$.
      The datum~$U$ migrates from the environment continuation
      into the program continuation through the binder~$y$.
      There is no environment residual (asynchronous output).
    \item[Section: $\hashf = \mathit{digest} \circ \mathit{cf}$.]
      A collision-resistant digest composed with a canonical-form
      function~$\mathit{cf}$ on the $\Seq$-quotient of
      processes.  This is the authentication section; the
      communication quotient $\quot{\cdot}$ up to~$\Seq$ is a
      separate, pre-existing structure.
    \item[Wrappability.]
      The contraction $\mathit{compute}$ is well-sorted on
      wrapped continuations: substituting a wrapped term
      $\signed{Q}{s}$ into a position yields a wrapped term,
      so $\mathit{compute}$ preserves the sort~$T$ for free.
  \end{description}
\end{definition}

\begin{remark}[Asynchronous output as degenerate $K_e$]
  \label{rem:async}
  In the synchronous $\pi$-calculus, the environment continuation
  is the pair $C_e = [Q_1, Q_2]$ of sent process and sender
  continuation.  Rholang's asynchronous output has $C_e = -$:
  the sender does not wait, so there is no continuation.  This
  is the temporal counterpart of $\lambda$'s degenerate
  environment surface: where $\lambda$ has a degenerate
  (structurally carried) surface, rho has a degenerate (absent)
  environment residual.
\end{remark}

\subsection{Object Capabilities and Ambient Authority}
\label{sec:ocap-authority}

Because $K = \Par$ is associative--commutative, the interaction
surfaces must be carried \emph{nominally} rather than
\emph{structurally}~\cite{CostEndofunctor}.  In a rigid $K$
(such as $\mathit{App}$ in $\lambda$), position is
unforgeable and the context $K([\,], S)$ is itself the
capability to draw on the stack~$S$ --- authority is
structural.  But in the AC bag of rho, position is
non-exclusive: any process in the bag floats into adjacency
with any stack.  Taking mere adjacency as authority would be
\emph{ambient authority}~\cite{MillerYeeShapiro2003,
MillerPhD2006} --- the antithesis of an object-capability
discipline.

The recovery is the \emph{nearness predicate}:
$\mathit{near}(I, J)$ requires that the receiver's surface
(channel name~$x$) match the token's surface before
interaction fires.  Authority is conferred not by position in
the bag but by possession of a matching channel name --- which,
for unforgeable channels created by $\mathbf{new}$, is a
genuine capability.

\begin{definition}[Located resource stack]
  \label{def:located-stack}
  A resource stack \emph{located at a surface}
  $S(I, s : S')$ is a stack indexed by the interaction
  surface~$I$ (a channel name) that may draw on it.  In rho,
  this is a token stack sitting on a specific channel: only
  processes holding a matching channel name can consume tokens
  from it.
\end{definition}

\noindent
The located stack turns ``reachable by whatever is adjacent''
into ``reachable by a matching surface.''  This is the form on
which the linear resource proofs of
Section~\ref{sec:linear-proof} operate: the static analysis
checks token availability \emph{per located surface}, not
globally over all processes in the bag.

\begin{remark}[Funding slots as located stacks]
  The funding-slot pattern of Section~\ref{sec:funding-slots} is
  precisely a located resource stack: the unforgeable channel
  created by $\mathbf{new}$ is the surface~$I$, and the token
  stack deposited on it is $S(I, \cdot)$.  The
  object-capability security of the funding slot --- only
  processes explicitly given the channel name can deposit or
  withdraw tokens --- is a consequence of the general
  ocap analysis of the AC case.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{The Cost-Accounted Rho Calculus}
\label{sec:calculus}

\subsection{Syntax}
\label{sec:syntax}

The calculus has four mutually defined syntactic categories:
\emph{processes}, \emph{names}, \emph{signed terms}, and
\emph{signatures}.

\begin{definition}[Processes and names]
  \label{def:proc-name}
  \begin{align}
    P, Q &\;::=\; \nil
          \;\mid\; \recv{y}{x}{T}
          \;\mid\; \send{x}{U}
          \;\mid\; P \Par Q
          \;\mid\; \drop{x}
    \label{eq:proc-syntax}
    \\[4pt]
    x, y &\;::=\; \quot{T} \;\mid\; s
    \label{eq:name-syntax}
  \end{align}
\end{definition}

\noindent
The intended readings carry over from the pure rho calculus,
with one critical change: the for-comprehension
$\recv{y}{x}{T}$ has a \emph{signed term}~$T$ as its
continuation, and the send $\send{x}{U}$ carries a signed
term~$U$ as its payload.

\begin{definition}[Signed terms and token stacks]
  \label{def:signed-term}
  \begin{align}
    T, U &\;::=\; \signed{P}{s}
          \;\mid\; T \Par U
          \;\mid\; S
    \label{eq:signed-term-syntax}
    \\[4pt]
    S &\;::=\; ()
       \;\mid\; s : S
    \label{eq:token-syntax}
  \end{align}
\end{definition}

\noindent
A signed term is one of:
\begin{description}[style=nextline, leftmargin=2em]
  \item[$\signed{P}{s}$]
    A process~$P$ signed by signature~$s$.  The braces
    delimit the scope of the signature.
  \item[$T \Par U$]
    Parallel composition of signed terms --- the same
    $\Par$ operator as at the process level, lifted to
    signed terms.
  \item[$S$]
    A token stack, representing a phlogiston balance.
\end{description}

\noindent
A token stack $s_1 : s_2 : \cdots : s_n : ()$ is read as a
sequence of signature-indexed balance layers.  The empty
stack~$()$ represents a depleted balance.

\begin{definition}[Signatures, parameterized by~$G$]
  \label{def:sig}
  \begin{equation}
    s(G) \;::=\; g
          \;\mid\; \hashf P
          \;\mid\; s \comp s
    \label{eq:sig-syntax}
  \end{equation}
  where $g \in G$ is a \emph{ground signature} drawn from the
  cryptographic backend~$G$.
\end{definition}

\noindent
The three forms of signature are:
\begin{description}[style=nextline, leftmargin=2em]
  \item[$g$]
    A \emph{ground signature}: an opaque value from the
    cryptographic backend~$G$ (e.g., an Ed25519 public key, a
    secp256k1 key hash).  The calculus is parametrically
    polymorphic in~$G$.
  \item[$\hashf P$]
    \emph{Cryptographic quoting}: the hash of a process~$P$,
    yielding a signature.  This is the authentication-axis
    analog of structural quoting~$\quot{P}$.
  \item[$s_1 \comp s_2$]
    \emph{Signature composition}: the compound authority of
    $s_1$ and~$s_2$ acting jointly.
\end{description}

\begin{remark}[Two axes of reflection]
  \label{rem:two-axes}
  The calculus now has two quoting operators:
  \begin{center}
    \begin{tabular}{lll}
      \textbf{Operator} & \textbf{Axis} & \textbf{Purpose} \\
      \midrule
      $\quot{T}$ & Structural  & Signed term $\to$ channel name \\
      $\hashf P$ & Cryptographic & Process $\to$ signature \\
    \end{tabular}
  \end{center}
  Both are forms of \emph{reflection}: they take a computational
  entity and produce a \emph{name} (a datum that can be
  communicated, compared, and used as an address).  The
  difference is that $\quot{T}$ preserves full structural
  information --- including the signature and cost provenance
  of~$T$, and the signed term can be recovered
  via~$\drop{x}$ --- while $\hashf P$ is a one-way,
  collision-resistant digest, suitable for authentication but
  not for recovery.
\end{remark}

\subsection{Parametric Polymorphism of Parallel Composition}
\label{sec:par-poly}

The parallel composition operator~$\Par$ appears at two levels:

\begin{enumerate}[nosep]
  \item $P \Par Q$ --- parallel composition of processes.
  \item $T \Par U$ --- parallel composition of signed terms.
\end{enumerate}

\noindent
This is a parametrically polymorphic operator: the same symbol,
the same algebraic structure (commutative monoid with
appropriate identity), but ranging over different sorts.  At the
process level the identity is~$\nil$; at the signed-term level
the identity is the empty stack~$()$.

\subsection{Free Names}
\label{sec:fn}

The binding discipline carries over from the pure rho calculus:
$\recv{y}{x}{T}$ binds~$y$.  Free names are computed
recursively over both processes and signed terms:

\begin{align*}
  \FN(\nil) &= \emptyset \\
  \FN(\recv{y}{x}{T})
    &= \bigl(\FN(T) \setminus \{y\}\bigr) \cup \{x\} \\
  \FN(\send{x}{U})
    &= \FN(U) \cup \{x\} \\
  \FN(P \Par Q) &= \FN(P) \cup \FN(Q) \\
  \FN(\drop{x}) &= \{x\}
\end{align*}

\noindent
Extended to signed terms:
\begin{align*}
  \FN(\signed{P}{s})
    &= \FN(P) \cup \FN_s(s) \\
  \FN(T \Par U)
    &= \FN(T) \cup \FN(U) \\
  \FN(())   &= \emptyset \\
  \FN(s : S)
    &= \FN_s(s) \cup \FN(S)
\end{align*}

\noindent
where $\FN_s$ extracts names from signatures:
\begin{align*}
  \FN_s(g) &= \emptyset \\
  \FN_s(\hashf P) &= \FN(P) \\
  \FN_s(s_1 \comp s_2)
    &= \FN_s(s_1) \cup \FN_s(s_2)
\end{align*}

\subsection{Structural Equivalence}
\label{sec:struct-equiv}

Structural equivalence~$\Seq$ is the smallest equivalence
relation on processes including $\alpha$-equivalence and making
$(P, \Par, \nil)$ a commutative monoid:

\begin{align*}
  P \Par \nil    &\Seq P           \\
  P \Par Q       &\Seq Q \Par P    \\
  (P \Par Q) \Par R &\Seq P \Par (Q \Par R)
\end{align*}

\noindent
Likewise, at the signed-term level, $(T, \Par, ())$ forms a
commutative monoid under~$\Seq$:

\begin{align*}
  T \Par ()      &\Seq T           \\
  T \Par U       &\Seq U \Par T    \\
  (T \Par U) \Par V &\Seq T \Par (U \Par V)
\end{align*}

\subsection{Substitution}
\label{sec:subst}

Substitution $T\{\quot{U}/y\}$ is the standard capture-avoiding
substitution of~$\quot{U}$ for~$y$ in~$T$, extended to signed
terms.  Following the pure rho calculus, the only non-trivial
case is dequotation:
\[
  (\drop{x})\{\quot{U}/y\}
  \;=\;
  \begin{cases}
    U & \text{if } y \Neq x, \\
    \drop{x} & \text{otherwise,}
  \end{cases}
\]
where $\Neq$ denotes name equality.

\begin{remark}[Signed-term substitution]
  \label{rem:signed-subst}
  Because the name sort is $x, y \;::=\; \quot{T} \mid s$,
  the substitution $T\{\quot{U}/y\}$ naturally substitutes the
  quoted \emph{signed term}~$U$, not a bare process extracted
  from~$U$.  There is no unwrapping.  This means the signature
  and cost information associated with~$U$ is \emph{preserved
  through communication}: a signed term received on a channel
  retains its provenance.
\end{remark}

\subsection{Rewrite Rules}
\label{sec:rewrites}

The operational semantics consists of five rewrite rules, each
a variation on the \textsc{Comm} rule of the pure rho calculus
gated by token consumption.  In every rule, a communication
fires only when matching tokens are present, and the
consumed tokens are released as residual terms.

\bigskip
\noindent
\textbf{Rule 1} --- Single signature, whole redex:
\begin{equation}
  \boxed{
    \signed{\recv{y}{x}{T} \Par \send{x}{U}}{s}
    \;\Par\; s : S
    \;\;\red\;\;
    \subst{T}{\quot{U}}{y} \Par S
  }
  \label{eq:rule1}
\end{equation}

\noindent
The entire redex (input and output) is signed by~$s$.  A single
layer of token is consumed; the stack remainder~$S$ is released.

\bigskip
\noindent
\textbf{Rule 2} --- Compound signature, whole redex, split
tokens:
\begin{equation}
  \boxed{
    \signed{\recv{y}{x}{T} \Par \send{x}{U}}{s_1 \comp s_2}
    \;\Par\; s_1 : S \Par s_2 : S'
    \;\;\red\;\;
    \subst{T}{\quot{U}}{y} \Par S \Par S'
  }
  \label{eq:rule2}
\end{equation}

\noindent
The redex is signed by the compound authority~$s_1 \comp s_2$.
Both component signatures must supply tokens; both remainders
are released.

\bigskip
\noindent
\textbf{Rule 3} --- Compound signature, whole redex, combined
token:
\begin{equation}
  \boxed{
    \signed{\recv{y}{x}{T} \Par \send{x}{U}}{s_1 \comp s_2}
    \;\Par\; (s_1 \comp s_2) : S
    \;\;\red\;\;
    \subst{T}{\quot{U}}{y} \Par S
  }
  \label{eq:rule3}
\end{equation}

\noindent
The compound authority~$s_1 \comp s_2$ holds a single combined
token.  One layer is consumed; the remainder~$S$ is released.

\bigskip
\noindent
\textbf{Rule 4} --- Split processes, split tokens:
\begin{equation}
  \boxed{
    \signed{\recv{y}{x}{T}}{s_1}
    \;\Par\; \signed{\send{x}{U}}{s_2}
    \;\Par\; s_1 : S \Par s_2 : S'
    \;\;\red\;\;
    \subst{T}{\quot{U}}{y} \Par S \Par S'
  }
  \label{eq:rule4}
\end{equation}

\noindent
Input and output are signed separately.  Each signer supplies
their own token.

\bigskip
\noindent
\textbf{Rule 5} --- Split processes, combined token:
\begin{equation}
  \boxed{
    \signed{\recv{y}{x}{T}}{s_1}
    \;\Par\; \signed{\send{x}{U}}{s_2}
    \;\Par\; (s_1 \comp s_2) : S
    \;\;\red\;\;
    \subst{T}{\quot{U}}{y} \Par S
  }
  \label{eq:rule5}
\end{equation}

\noindent
Input and output are signed separately, but the token is held
jointly under the compound authority.

\bigskip
\noindent
As in the pure rho calculus, these rules are closed under
parallel context and structural equivalence:
\begin{gather}
  \frac{P \red P'}{P \Par Q \red P' \Par Q}
  \qquad\text{(\textsc{Par})}
  \label{eq:par-ctx}
  \\[6pt]
  \frac{P \Seq P' \qquad P' \red Q' \qquad Q' \Seq Q}
       {P \red Q}
  \qquad\text{(\textsc{Struct})}
  \label{eq:struct-ctx}
\end{gather}

\subsection{Summary of the Five Rules}
\label{sec:rule-summary}

The five rules can be organized along two axes:

\begin{center}
  \begin{tabular}{l|cc}
    & \textbf{Split tokens} & \textbf{Combined token} \\
    \midrule
    \textbf{Whole redex, single sig}
      & Rule~1 & Rule~1 \\
    \textbf{Whole redex, compound sig}
      & Rule~2 & Rule~3 \\
    \textbf{Split processes}
      & Rule~4 & Rule~5 \\
  \end{tabular}
\end{center}

\noindent
Every rule follows the same pattern: \emph{communication is
gated by token consumption}.  The variations arise from the
granularity of signing (whole redex vs.\ split processes) and
the granularity of token holding (split vs.\ combined).

\subsection{Cost Accounting for Joins}
\label{sec:joins}

Rholang is based not on the binary rho calculus but on a rho
calculus with \emph{joins}: a single for-comprehension may
await messages on several channels at once, firing only when
all are present:
\begin{equation}
  \recv{y_1}{x_1}{} \,\&\, \cdots \,\&\, \recv{y_N}{x_N}{}\,P
  \;\Par\; \send{x_1}{Q_1} \;\Par\; \cdots \;\Par\; \send{x_N}{Q_N}
  \;\red\;
  P\{\quot{Q_1}/y_1, \ldots, \quot{Q_N}/y_N\}
  \label{eq:join-rule}
\end{equation}
In interaction-cut terms (Section~\ref{sec:instantiation}), a
join is an \emph{$N$-ary cut}: one receiver introduction with
$N$ interaction surfaces $x_1, \ldots, x_N$, and $N$ separate
sender introductions, each matching one surface.  The
contraction fires only when all $N$ rendezvous succeed
simultaneously --- the join is an all-or-nothing atomic
synchronization across $N$ channels.  This built-in atomicity
is what makes cost accounting for joins both richer and, in a
precise sense, safer than the binary case.

\subsubsection{The Base Cases}

\paragraph{Case J1 (whole-join signing, single funding signature).}
The entire join redex --- the receiver together with all $N$
sends --- is wrapped under a single signature~$s_1$, and a
single token funds it:
\begin{multline}
  \signed{
    (\recv{y_1}{x_1}{} \,\&\, \cdots \,\&\, \recv{y_N}{x_N}{})
    \signed{P}{s_2}
    \Par \send{x_1}{U_1} \Par \cdots \Par \send{x_N}{U_N}
  }{s_1}
  \Par s_1 : S
  \\
  \red\;
  \signed{P\{\quot{U_1}/y_1, \ldots, \quot{U_N}/y_N\}}{s_2}
  \Par S
  \label{eq:join-J1}
\end{multline}
The whole $N$-way rendezvous-and-match is a \emph{single}
funded event under~$s_1$: one token consumed, regardless
of~$N$.  The continuation~$P$ carries its own
signature~$s_2$ (the linear-transfer pattern of
Section~\ref{sec:sugar}: $s_1$ funds the join, $s_2$ funds the
continuation), and the substitutions for all $N$ bindings are
applied simultaneously.  This is the join analogue of Rule~1.

\paragraph{Case J2 (separately-signed participants, single combined token).}
Each participant is signed independently --- the receiver
under~$s_1$, each sender~$\send{x_i}{U_i}$ under its own~$t_i$
--- and a single combined token keyed to the compound signature
funds the join:
\begin{multline}
  \signed{
    (\recv{y_1}{x_1}{} \,\&\, \cdots \,\&\, \recv{y_N}{x_N}{})
    \signed{P}{s_2}
  }{s_1}
  \Par \signed{\send{x_1}{U_1}}{t_1} \Par \cdots
  \Par \signed{\send{x_N}{U_N}}{t_N}
  \\
  \Par\; (s_1 \comp t_1 \comp \cdots \comp t_N) : S
  \;\red\;
  \signed{P\{\quot{U_1}/y_1, \ldots, \quot{U_N}/y_N\}}{s_2}
  \Par S
  \label{eq:join-J2}
\end{multline}
This is a genuine multi-party join: $N+1$ distinct economic
actors must all be present for the synchronization.  The
crucial design choice is that the token's key is the
\emph{fusion} of all participants' signatures
$s_1 \comp t_1 \comp \cdots \comp t_N$, so the whole join
fires on one combined payment or not at all.  There is no way
to debit $t_1$ and $t_2$ and then block waiting for~$x_3$ ---
the single combined cell is consumed atomically the moment all
surfaces match.  The partial-funding hazard that arises
\emph{across} multi-step transactions
(Section~\ref{sec:partial-funding}) is structurally impossible
\emph{within} a join.  This is the join analogue of Rule~3,
generalized so that the compound signature fuses the receiver's
authority with all $N$ senders' authorities.

\subsubsection{Per-Clause Signing and the Two-Sided Surface}

The receiver signature need not be monolithic.  Each binding
clause may carry its own signature, pushing the signing down
onto the individual rendezvous surfaces:
\begin{equation}
  \forc\!\left(
    \signed{y_1 \leftarrow x_1}{s_1} \,\&\, \cdots \,\&\,
    \signed{y_N \leftarrow x_N}{s_N}
  \right) P
  \label{eq:join-per-clause}
\end{equation}
Each clause $\signed{y_i \leftarrow x_i}{s_i}$ is one
receiver-side introduction with surface~$x_i$, funded by~$s_i$
independently.  This gives the receiver side the same
combinatorial freedom as the sender side, and sharpens the
interaction-cut reading: per channel~$i$, the rendezvous
on~$x_i$ requires \emph{both}~$s_i$ (the right to receive
on~$x_i$) \emph{and}~$t_i$ (the right to send~$U_i$).  The
join is thus a two-sided capability check: one needs authority
to offer each message and authority to consume it, and the
join gathers all $2N$ authorities at once.

\subsubsection{The General Schema}

For $N$-ary joins, enumerating cases is hopeless; we present a
single schema quantified over partitions.  Three independent
axes of granularity, all over the index set
$\{1, \ldots, N\}$, govern how a join may be signed and funded:

\begin{description}[style=nextline, leftmargin=2em, nosep]
  \item[Axis C (receiver clauses).]
    How the binding clauses $\{y_i \leftarrow x_i\}$ are
    grouped under receiver signatures, ranging from a single
    monolithic $s_1$ over the whole $\forc$ (as in J1, J2) to
    a distinct $s_i$ per clause (as
    in~\eqref{eq:join-per-clause}).
  \item[Axis B (sends).]
    How the sends $\send{x_i}{U_i}$ are grouped under sender
    signatures, from all sends under one signature to each
    individually signed.
  \item[Axis A (token presentation).]
    How the resulting signatures' tokens are partitioned across
    stacks, from a single fully combined cell
    $s_1 \comp \cdots \comp t_N : S$ to fully split
    $s_1 : S_1 \Par \cdots \Par t_N : S_N$, and every
    grouping in between.
\end{description}

\begin{definition}[Join cost schema]
  \label{def:join-schema}
  Let a join carry receiver clauses grouped by a partition~$\pi_C$
  of $\{1,\ldots,N\}$ (clauses in the same block share a receiver
  signature) and sends grouped by a partition~$\pi_B$.  Let the
  required authorities form the multiset
  \[
    \mathcal{A}
    \;=\;
    \{\sigma_b : b \in \pi_C\}
    \;\uplus\;
    \{\tau_b : b \in \pi_B\},
  \]
  where $\sigma_b$ (resp.\ $\tau_b$) is the signature on
  block~$b$.  For \emph{any} partition~$\pi_A$ of~$\mathcal{A}$
  into composite tokens --- one stack-headed cell per block of
  $\pi_A$, keyed by the $\comp$-product of the signatures in
  that block --- the join fires when every surface matches and
  every cell of~$\pi_A$ is present, consuming all of them and
  releasing their stack remainders:
  \[
    (\text{join under } \pi_C, \pi_B)
    \;\Par\;
    \bigParop_{b \in \pi_A}
      \Bigl( \bigCompop_{a \in b} a \Bigr) : S_b
    \;\red\;
    \signed{P\{\quot{U_1}/y_1, \ldots, \quot{U_N}/y_N\}}{s_2}
    \;\Par\;
    \bigParop_{b \in \pi_A} S_b .
  \]
\end{definition}

\begin{proposition}[Conservation of authority]
  \label{prop:join-conservation}
  Across every instance of the schema --- for all choices of
  $\pi_C$, $\pi_B$, $\pi_A$ --- the multiset of signatures
  consumed when a join fires is exactly
  \[
    \{s_1, \ldots, s_N\} \;\uplus\; \{t_1, \ldots, t_N\},
  \]
  one receiver-authority and one sender-authority per channel.
  Grouping (along any axis) changes the bookkeeping; it never
  changes the total.  The join either gathers one token's worth
  of every participant's authority, or it does not fire.
\end{proposition}

\noindent
The binary Rules~1--5 are the $N = 1$ special cases; Cases J1
and J2 are the fully-combined corners of the receiver axis.

\subsubsection{Regrouping Tokens: Reverse Currying}

Because $\comp$ is associative--commutative, the demand of
Definition~\ref{def:join-schema} is for a \emph{multiset} of
authorities, and any partition of that multiset into composite
tokens satisfies it.  Two composite tokens may therefore be
assembled into one, or one split into two, without changing
what they can fund:
\begin{equation}
  (s_i \comp t_i) : S \;\Par\; (s_j \comp t_j) : S'
  \qquad\xrightleftharpoons[\;\mathrm{Split}\;]{\;\mathrm{Join}\;}\qquad
  (s_i \comp t_i \comp s_j \comp t_j) : (S \cdot S')
  \label{eq:reverse-curry}
\end{equation}
We call the right-to-left direction \emph{reverse Currying}:
it assembles a large composite authority from smaller ones,
the iso between a curried and an uncurried presentation of the
same compound signature.  The splitter and joiner
infrastructure processes
(Appendix~\ref{app:infrastructure}) are the operational
witnesses of this iso: $\mathrm{Join}$ assembles,
$\mathrm{Split}$ disassembles.  Reverse Currying preserves the
conserved multiset of Proposition~\ref{prop:join-conservation};
it is the sanctioned way to regroup tokens to match a join's
demand.

\subsubsection{Weakening Is Forbidden}

The dangerous dual operation is \emph{weakening}: consuming a
composite token $s_i \comp s_j$ to discharge a demand for
$s_i$ alone, discarding the $s_j$ component.  We forbid it.

The reason is the multi-signature semantics of
Section~\ref{sec:multisig-motivation}.  A token minted as
$s_{\mathit{sender}} \comp s_{\mathit{compliance}}$ encodes
``neither party alone may act.''  If such a token could be
partially consumed as just $s_{\mathit{sender}}$, the
compliance requirement would evaporate --- the multi-sig
guarantee would be defeated by weakening.  A composite
signature must therefore be discharged \emph{in full}, or
\emph{explicitly} $\mathrm{Split}$ first (an observable
reduction step that leaves the $s_j$ component live on its own
stack, not discarded).

\begin{remark}[Two monoids, two disciplines]
  The temporal monoid~$:$ and the spatial-signature
  monoid~$\comp$ behave oppositely with respect to weakening.
  The temporal stack \emph{admits} weakening in the benign
  sense that reduction consumes a head cell and keeps the tail
  --- that is simply what a step does.  The spatial-signature
  monoid \emph{does not} admit weakening: a composite must be
  discharged in full or explicitly split, never silently
  partially consumed.  Reverse Currying
  (\eqref{eq:reverse-curry}) --- assembling and disassembling
  composites via $\mathrm{Join}$ and $\mathrm{Split}$ --- is
  the only sanctioned way to regroup~$\comp$, and it conserves
  the multiset.
\end{remark}

\begin{remark}[The tensor--hom adjunction]
  \label{rem:tensor-hom}
  Reverse Currying and the linear-transfer sugar~$\multimap$
  (Section~\ref{sec:sugar}) are two halves of a single
  structure.  The compound operator~$\comp$ is a tensor on
  signatures; the transfer~$\multimap$ is its right adjoint
  (internal hom), so that
  \[
    (s_1 \comp s_2) \multimap s_3
    \;\cong\;
    s_1 \multimap (s_2 \multimap s_3).
  \]
  Currying relates joint funding ($s_1 \comp s_2$ funds in one
  combined event) to staged funding ($s_1$ funds, then $s_2$
  funds the continuation).  The full development of this
  adjunction --- and whether the token discipline is a model of
  intuitionistic linear logic with $\comp = \otimes$ and
  $\multimap$ the linear function space --- we leave to the
  sequel.
\end{remark}

\begin{remark}[Joins and the static analysis]
  \label{rem:join-static}
  A join blocks until all $N$ senders are present, which may be
  data-dependent.  The overcharge-and-refund discipline of
  Section~\ref{sec:data-dependent} extends cleanly: the demand
  of a join is the fixed multiset of
  Proposition~\ref{prop:join-conservation}, independent of the
  partition axes, so the conservative bound counts $2N$
  authorities per join.  Reverse Currying does not change the
  count, and the prohibition on weakening keeps the demand a
  clean multiset equality.  Whether all $N$ senders \emph{will}
  arrive is the data-dependent question, handled exactly as
  before --- by funding the worst case and refunding the
  unforced remainder.
\end{remark}

\subsection{Syntactic Sugar}
\label{sec:sugar}

Two abbreviations streamline the common case of signing a
for-comprehension together with its continuation.

\begin{definition}[Uniform signing]
  \label{def:sugar-uniform}
  \begin{equation}
    \signed{\recv{y}{x}{P}}{s}
    \;\;\triangleq\;\;
    \signed{
      \recv{y}{x}{\signed{P}{s}}
    }{s}
    \label{eq:sugar-uniform}
  \end{equation}
\end{definition}

\noindent
When a for-comprehension and its continuation are signed by
the \emph{same} signature~$s$, the abbreviation on the left
expands to two nested signed layers on the right.  The outer
$\signed{\cdot}{s}$ pays for the transaction (rendezvous and
matching); the inner $\signed{\cdot}{s}$ pays for executing
the continuation.  This is the common case for single-party
execution: both the communication and its effect are
authorized and funded by the same signer.

\begin{definition}[Linear transfer of authority ($\multimap$)]
  \label{def:sugar-lollipop}
  \begin{equation}
    \signed{\recv{y}{x}{P}}{s_1 \multimap s_2}
    \;\;\triangleq\;\;
    \signed{
      \recv{y}{x}{\signed{P}{s_2}}
    }{s_1}
    \label{eq:sugar-lollipop}
  \end{equation}
\end{definition}

\noindent
The \emph{lollipop} notation
$\signed{\cdot}{s_1 \multimap s_2}$ denotes a
\emph{transfer of authority}: signature~$s_1$ authorizes and
funds the communication, but signature~$s_2$ authorizes and
funds the continuation.  The $\multimap$ is deliberately
chosen to evoke the linear implication of linear logic:
$s_1$'s resource is consumed to establish the communication,
and $s_2$'s resource funds the resulting computation.

\begin{example}[Cross-party service invocation]
  \label{ex:cross-party}
  A client~$c$ invokes a service operated by provider~$p$.
  The client pays for the call; the provider pays for the
  execution:
  \[
    \signed{
      \recv{\mathit{result}}{\mathit{service}}{\mathit{handleResult}}
    }{c \multimap p}
    \;\;=\;\;
    \signed{
      \recv{\mathit{result}}{\mathit{service}}{
        \signed{\mathit{handleResult}}{p}
      }
    }{c}
  \]
  The client's token funds the rendezvous with the service
  channel; the provider's token funds the execution of
  $\mathit{handleResult}$.  Neither party can impose costs
  on the other beyond what the sugar makes explicit.
\end{example}

\begin{remark}[Composability]
  Both sugar forms compose with compound signatures.  For
  example,
  $\signed{\recv{y}{x}{P}}{(s_1 \comp s_2) \multimap s_3}$
  requires both~$s_1$ and~$s_2$ to authorize the
  communication, while $s_3$ alone funds the continuation.
  Chains of transfers are also expressible:
  nested for-comprehensions can each carry their own
  $\multimap$, threading authority through a pipeline of
  parties.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Design Analysis}
\label{sec:analysis}

\subsection{Signatures as Names: Eliminating the Translation Layer}
\label{sec:sig-as-name}

In a previous formulation, the cost-accounted calculus was
specified as an extension of the pure rho calculus, and a
compositional translation $\mathcal{N}\lsem\cdot\rsem$ mapped
signatures to names, tokens to messages, and signed processes
to fuel-gated wrappers.  While that translation demonstrated
expressibility, it introduced an indirection: the
implementation had to maintain two representations and a mapping
between them.

The present formulation eliminates this indirection.  By
declaring $x, y \;::=\; \quot{T} \mid s$, signatures \emph{are}
names.  There is nothing to translate.  The implementation
maintains a single namespace in which quoted signed terms and
signatures coexist.  This has direct consequences for the
Rholang runtime:

\begin{itemize}[nosep]
  \item The tuplespace (RSpace) stores both data channels
    ($\quot{T}$) and signature channels ($s$) in the same
    structure.
  \item Pattern matching in for-comprehensions can match on
    both kinds of name, enabling contracts that dispatch on
    the kind of authority presented.
  \item Namespace isolation is enforced by the existing
    namespace logic~\cite{MeredithRadestock2005}, now extended
    to cover the signature sort.
\end{itemize}

\subsection{Section and Quotient Disentangled}
\label{sec:crypto-quote}

The general construction~\cite{CostEndofunctor} identifies two
maps relating a theory's free syntax~$T$ to its
structural-congruence quotient~$T/{\Seq}$: the
\emph{quotient}~$T \twoheadrightarrow T/{\Seq}$, which respects~$\Seq$
and is the structure a channel constructor exposes (in rho,
$\quot{P}$ with $\quot{P} = \quot{Q} \iff P \Seq Q$); and the
\emph{section}~$\mathit{cf} : T \to T/{\Seq}$, a choice of
canonical representative per class, which composed with a
collision-resistant digest gives the signature constructor
$\hashf = \mathit{digest} \circ \mathit{cf}$.

The quotient is not new --- it is the LTS the rho calculus
already carries.  The section is the authentication mechanism
the cost functor adds.  In rho, the operator~$\quot{\cdot}$
happens to serve both roles (viewed two ways), which is why the
earlier formulation appeared to need ``two axes of
reflection.''  The general construction pulls them apart: one
axis is the communication quotient, the other is the
authentication section, and the signature is $\hashf$ alone.

In the F1R3FLY platform, $\hashf P$ will be implemented as a
configurable hash function (SHA-256, Blake2b, etc.) applied to
the serialized process tree.  The parameterization of~$s$ by~$G$
ensures that the choice of hash function is a deployment
decision, not a calculus-level commitment.

\subsection{Valuable Friction at the Syntactic Level}
\label{sec:valuable-friction}

The F1R3FLY platform's design philosophy of \emph{valuable
friction} holds that every resource-consuming operation should
carry an explicit cost, and that this cost is not merely a
runtime check but a \emph{structural feature} of the language.
The revised calculus embodies this principle:

\begin{itemize}[nosep]
  \item A for-comprehension $\recv{y}{x}{T}$ cannot fire unless
    its continuation~$T$ includes (possibly nested) signed
    terms with matching tokens.
  \item A send $\send{x}{U}$ carries a signed term~$U$ as
    payload, so the receiver inherits the sender's cost
    commitments.
  \item There is no way to write a communicating process that
    ``escapes'' the cost layer, because the grammar does not
    allow bare processes in the positions where communication
    occurs.
\end{itemize}

\subsection{Token Stacks as First-Class Terms}
\label{sec:token-stacks}

Token stacks $S \;::=\; () \mid s : S$ are lists of signatures,
representing layered balances.  Their inclusion in the
signed-term grammar ($T \;::=\; \cdots \mid S$) means that
tokens are not inert metadata --- they are terms that appear in
parallel compositions, flow through channels as parts of signed
messages, and participate in reduction.

This design enables several patterns:

\begin{itemize}[nosep]
  \item \textbf{Token delegation:}  A contract can receive a
    token stack on a channel and use it to fund subsequent
    communications.
  \item \textbf{Token splitting:}  A stack $s_1 : s_2 : S$ can
    be destructured by separate rules consuming $s_1$ and $s_2$
    independently.
  \item \textbf{Token markets:}  Because tokens are terms, they
    can be exchanged, auctioned, and composed using standard
    Rholang contract patterns.
\end{itemize}

\subsection{Genericity over the Cryptographic Backend}
\label{sec:genericity}

The signature grammar is parameterized: $s(G) \;::=\; g \mid
\hashf P \mid s \comp s$, where $g \in G$ and $G$ is a type
parameter.  This makes the entire calculus --- and therefore
the Rholang implementation --- generic over the cryptographic
backend.

Instantiating~$G$ yields a concrete calculus:
\begin{itemize}[nosep]
  \item $G = \mathrm{Ed25519}$:  ground signatures are Ed25519
    public keys.
  \item $G = \mathrm{secp256k1}$:  ground signatures are
    Ethereum-compatible key hashes.
  \item $G = \mathrm{Ed25519} + \mathrm{secp256k1}$:  a
    coproduct backend supporting both schemes.
  \item $G = \mathrm{OQS}$:  ground signatures are
    post-quantum signature schemes from the Open Quantum Safe
    (liboqs) library, such as CRYSTALS-Dilithium, FALCON, or
    SPHINCS$^+$.  In ongoing work, F1R3FLY has identified a
    migration plan for replacing classical cryptographic
    signatures with OQS backends, making the platform
    quantum-resistant.  The parametricity of the calculus in~$G$
    ensures that this migration requires no changes to the
    cost-accounting semantics --- only the instantiation of~$G$.
\end{itemize}

\noindent
The $\comp$ operator composes signatures regardless of their
ground type, enabling cross-scheme multi-sig patterns ---
including hybrid classical/post-quantum multi-sig, where one
component is a classical key and the other is a post-quantum
key, providing defense in depth during the transition period.
The implementation must provide a ground-signature equality
predicate and a hash function for each instantiation of~$G$;
the calculus itself is agnostic.

\subsection{From Phlogiston to Spectrum: The Decomposition of
  Cost}
\label{sec:spectrum}

In the original Rholang architecture, \emph{phlogiston} was a
singular, undifferentiated resource: every deploy consumed units
of the same universal fuel, regardless of who signed it or what
it computed.  The cost-accounted calculus decomposes this
singular resource into a \emph{spectrum} of signature-indexed
resources --- as many distinct fuels as there are economically
empowered actors.

Two metaphors illuminate this decomposition:

\paragraph{The prism.}
Phlogiston is white light.  The signature algebra acts as a
prism, splitting the undifferentiated beam into its many
constituent frequencies.  Each signature~$s$ indexes a distinct
spectral line --- a resource pool visible only to processes
signed by~$s$.  Compound signatures $s_1 \comp s_2$ produce
interference patterns: new frequencies that emerge only when
two spectral components are superposed.  The linear resource
proof at deployment-acceptance time is a \emph{spectral
analysis}: it verifies that every frequency the deployment
requires is present with sufficient intensity.

\paragraph{Algorithmic chemistry.}
Alternatively, one may view the calculus as an \emph{algorithmic
chemistry} in which economic agency is the catalyst for
computation.  Processes are reagents; tokens are catalytic
substrates; and the rewrite rules are reaction rules that fire
only when the appropriate catalyst is present.  A signed
process $\signed{P}{s}$ is a reaction waiting for the
catalyst~$s$.  A token stack $s : S$ is a quantity of catalytic
substrate.  Communication is a catalyzed reaction: the substrate
is consumed, the reagents are transformed, and the products
(the continuation plus the remaining substrate) are released.

In this reading, the diversity of signatures corresponds to the
diversity of catalysts in a chemical system.  Different
reactions require different catalysts.  A compound
signature~$s_1 \comp s_2$ is a \emph{co-catalyst}: both
substances must be present for the reaction to proceed.  The
economy of the F1R3FLY platform is an algorithmic chemistry in
which the diversity and abundance of catalysts determine the
rate and variety of computation.

\begin{remark}[Phlogiston as a limit case]
  The original phlogiston model is recovered by assigning all
  actors the same signature~$s_0$.  In the prism metaphor, this
  collapses the spectrum back to white light.  In the chemistry
  metaphor, this reduces the system to a single universal
  catalyst.  The cost-accounted calculus generalizes from this
  monochromatic / mono-catalytic limit to a full spectrum of
  economic agency.
\end{remark}

\subsection{Funding Slots via Unforgeable Channels}
\label{sec:funding-slots}

The linear-transfer sugar
$\signed{\recv{y}{x}{P}}{s_1 \multimap s_2}$ requires the
continuation's funder~$s_2$ to be named at compile time.  In
practice, one often wants to write a contract whose
continuation can be funded by \emph{any} party willing to pay
--- without knowing their identity in advance.  Adding
existential quantification over signatures would achieve this
but at the cost of complicating the linear resource proofs
(the demand function~$\Delta_s$ would need to range over
unknown signatures).

A simpler and more natural solution exploits a feature that
Rholang already provides: \emph{unforgeable channels} via
$\mathbf{new}$.  Rather than existentially quantifying over
the funder's signature, a contract creates a fresh funding
channel --- a \emph{funding slot} --- and listens for tokens
on it.  Anyone who knows the channel can deposit tokens there.
The linear proof at acceptance time checks whether tokens are
present on the funding slot, not who deposited them.

\begin{definition}[Funding slot]
  \label{def:funding-slot}
  A funding slot is a pattern in which a contract creates an
  unforgeable channel and uses it as the signature for a
  continuation:
  \[
    \mathbf{new}\;\mathit{slot}\;\mathbf{in}\;
    \bigl(
      \signed{\recv{y}{x}{P}}
        {s_1 \multimap \mathit{slot}}
      \;\Par\;
      \send{\quot{\mathit{slot}}}{(\cdots)}
    \bigr)
  \]
  where the send on~$\quot{\mathit{slot}}$ publishes the
  funding-slot address to potential funders.
\end{definition}

\noindent
The funding slot decouples three concerns:

\begin{description}[style=nextline, leftmargin=2em, nosep]
  \item[Authorization (who may initiate).]
    Controlled by $s_1$: only the signer of~$s_1$ can
    trigger the communication.
  \item[Funding (who pays for the continuation).]
    Open: anyone who deposits tokens on~$\mathit{slot}$.
  \item[Execution (what happens).]
    Determined by~$P$: the continuation runs once the
    slot is funded.
\end{description}

\begin{example}[Sponsored execution]
  \label{ex:sponsored}
  A new user~$u$ wants to run a deployment~$D$ but holds no
  tokens.  A sponsor~$\sigma$ is willing to fund the
  execution.  The on-boarding contract creates a funding slot:
  \begin{align*}
    \mathbf{new}\;\mathit{slot}\;\mathbf{in}\;
    \bigl(\,
      &\signed{\recv{\mathit{dep}}{u}{D}}
        {u \multimap \mathit{slot}}
      \\[-2pt]
      &\;\Par\;
      \send{\quot{\mathit{slot}}}
        {\mathit{slot} : \mathit{slot} : ()}
    \,\bigr)
  \end{align*}

  \noindent
  The sponsor funds the slot by depositing tokens:
  \[
    \send{\Sigma\lsem\mathit{slot}\rsem}{
      \mathit{slot} : \mathit{slot} : ()
    }
  \]
  The user~$u$ authorizes and triggers the communication
  (consuming $u$-tokens); the sponsor's tokens on the
  funding slot pay for the continuation.  The linear proof
  sees concrete tokens on a concrete channel --- no
  existential quantification is needed.

  When the slush fund is exhausted, the user must either
  attract another sponsor or acquire tokens of their own.
  The transition from subsidized to self-funded participation
  is a resource-exhaustion event, not a policy decision.
\end{example}

\begin{remark}[Unforgeability and security]
  The $\mathbf{new}$ operator guarantees that the funding
  slot is \emph{unforgeable}: only processes that have been
  explicitly given the channel name can deposit tokens on it.
  This prevents unauthorized parties from draining the slot
  or injecting spurious tokens.  The capability-security
  properties of unforgeable channels, already well-understood
  in the rho calculus, carry over directly to funding slots.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Relationship to the Pure Rho Calculus}
\label{sec:relationship}

The cost-accounted calculus is a \emph{conservative extension}
of the pure rho calculus.  Setting $T = \signed{P}{s_0}$ for a
distinguished ``free'' signature~$s_0$ with unbounded token
supply, and collapsing the signed-term and process levels,
recovers the pure rho calculus exactly:

\[
  \signed{\recv{y}{x}{\signed{P'}{s_0}} \Par
    \send{x}{\signed{Q}{s_0}}}{s_0}
  \;\Par\; s_0 : S
  \;\;\red\;\;
  \signed{P'}{s_0}\{\quot{\signed{Q}{s_0}}/y\} \Par S
\]

\noindent
With $s_0 : S$ always available (supplied by a persistent
infrastructure process), this reduces to the standard
\textsc{Comm} rule.  The free signature~$s_0$ plays the role of
an ``always-funded'' system account, modeling the behavior of
the pure calculus where cost is not a concern.

This embedding also clarifies the implementation path: the
Rholang~1.4 runtime need only implement the cost-accounted
calculus, and the ``free'' mode (for testing, development, and
system-level contracts) is obtained by fixing~$s_0$ and
providing an inexhaustible token supply.

%% ══════════════════════════════════════════════════════════════════
\section{Implications for the F1R3FLY Platform}
\label{sec:implications}

\subsection{Rholang 1.4 (MeTTaIL) Integration}

MeTTaIL~\cite{MeTTaIL}, the Rholang~1.4 language, provides
means to define languages as triples of types (syntactic
categories), equations on types, and rewrites on types.  The
cost-accounted rho calculus is itself such a triple:

\begin{itemize}[nosep]
  \item \textbf{Types:}  $P$, $T$, $S$, $s(G)$, and the name
    sort.
  \item \textbf{Equations:}  Structural equivalence ($\Seq$)
    at both the process and signed-term levels.
  \item \textbf{Rewrites:}  Rules~1--5 and the contextual
    closure rules.
\end{itemize}

\noindent
This means the cost-accounted calculus can be \emph{defined
within MeTTaIL as a GSLT} (graph-structured lambda theory),
inheriting MeTTaIL's tooling for type checking, bisimulation,
and OSLF-generated logics.

\subsection{RSpace and the Tuplespace}

In the current F1R3FLY architecture, RSpace is a tuple-space
storage engine keyed by channel names.  The revised calculus
requires RSpace to store entries keyed by both~$\quot{T}$
and~$s$.  Since both are names, this is a matter of extending
the serialization format to accommodate the signature sort,
not a structural change to RSpace.

Token stacks~$S$ are stored as values in signature-keyed
entries.  The five rewrite rules correspond to tuplespace
match-and-consume operations that the reducer already performs,
extended with a signature-matching predicate.

\subsection{Casper CBC Consensus}

The cost-accounting rules interact with Casper CBC consensus at
the validator level:  a proposed block is valid only if every
communication it contains is backed by matching tokens.  The
revised calculus makes this validity check a syntactic
predicate: the block's contents must be well-typed in the
cost-accounted grammar, with token stacks present for every
signed communication.

\subsection{Implementation Path}

\begin{enumerate}[nosep]
  \item \textbf{Phase 1:}  Extend the Rholang parser and AST
    to include the signed-term, token-stack, and parameterized
    signature sorts.  Implement the five rewrite rules in the
    reducer.
  \item \textbf{Phase 2:}  Extend RSpace serialization to
    handle signature-keyed entries and token-stack values.
  \item \textbf{Phase 3:}  Define the cost-accounted calculus as
    a GSLT in MeTTaIL, enabling formal verification via OSLF.
  \item \textbf{Phase 4:}  Expose the signature and token APIs
    to user contracts, enabling composable metering,
    cross-chain fee markets, and delegated funding patterns.
\end{enumerate}

%% ══════════════════════════════════════════════════════════════════
\section{Transactions, Atomicity, and Static Analysis}
\label{sec:transactions}

The cost-accounted calculus reveals a precise correspondence
between the for-comprehension and the concept of a
\emph{transaction}.  This section develops that correspondence,
identifies a partial-funding vulnerability in multi-step
transactions, and shows that the vulnerability is eliminated by
a decidable static analysis at deployment-acceptance time.

\subsection{The For-Comprehension as Transaction Primitive}
\label{sec:for-as-txn}

Every for-comprehension $\recv{y}{x}{T}$ in the cost-accounted
calculus is a \emph{transaction} in the precise sense that it
decomposes into two funded computational actions:

\begin{enumerate}[nosep]
  \item \textbf{Rendezvous at the channel.}  The runtime must
    locate a matching send $\send{x}{U}$ on channel~$x$.  This
    is a search operation over the tuplespace, and its cost is
    paid by one token from the signer's stack.
  \item \textbf{Matching.}  The runtime must verify that the
    pattern~$y$ matches the offered payload~$U$ and compute
    the substitution $\{\quot{U}/y\}$.  This is a
    pattern-matching and substitution operation, and its cost
    is paid by a second token.
\end{enumerate}

\noindent
The \emph{transaction} is exactly what appears between the
round braces of the for-comprehension: the binding
specification $(y \leftarrow x)$.  Everything else --- the
continuation~$T$, the surrounding signed-term structure, the
token stack --- is the \emph{context} in which the transaction
executes.

\begin{definition}[Transaction, witness, state update]
  \label{def:txn}
  Given a signed for-comprehension
  $\signed{\recv{y}{x}{T}}{s}$ and a matching send
  $\send{x}{U}$:
  \begin{description}[style=nextline, leftmargin=2em, nosep]
    \item[Transaction]
      The binding specification $(y \leftarrow x)$: the
      declaration that we seek a value on channel~$x$ and
      will bind it to~$y$.
    \item[Witness]
      The substitution $\{\quot{U}/y\}$: the proof that the
      transaction succeeded.  A message was present on~$x$, it
      matched the pattern, and here is the concrete value that
      was bound.
    \item[State update]
      The application of the substitution to the
      continuation, $\subst{T}{\quot{U}}{y}$: the
      computational effect of the transaction on the system
      state.
    \item[Atomicity]
      The substitution is either applied or not.  There is no
      intermediate state.  The rewrite rule
      \[
        \signed{\recv{y}{x}{T} \Par \send{x}{U}}{s}
        \Par s : S
        \;\red\;
        \subst{T}{\quot{U}}{y} \Par S
      \]
      fires as a single, indivisible step.  If the token is
      absent or the send is missing, nothing happens ---
      the process blocks.
  \end{description}
\end{definition}

\begin{remark}[Database atomicity by construction]
  \label{rem:db-atomicity}
  It is important to distinguish two levels of atomicity that
  the community frequently conflates:
  \emph{database-transaction atomicity} and
  \emph{financial-transaction atomicity}.

  \textbf{Database-transaction atomicity} is the property of a
  single communication step: the substitution is either applied
  or not, the token is either consumed or not, the state is
  either updated or not.  In the cost-accounted rho calculus,
  this is a \emph{theorem} of the operational semantics ---
  the rewrite rules are defined so that a communication either
  fires completely or does not fire at all.  There is no rule
  that performs a ``partial'' communication.  No locks,
  two-phase commit, or rollback logs are needed.

  \textbf{Financial-transaction atomicity} is a different
  property: a multi-step operation (e.g., a payment decomposed
  into debit and credit) should either complete entirely or not
  begin at all.  This is \emph{not} guaranteed by the rewrite
  rules alone --- it requires the static analysis developed in
  Section~\ref{sec:linear-proof} and the deployment-boundary
  semantics of Section~\ref{sec:deploy-boundaries}.
\end{remark}

\subsection{The Partial Funding Vulnerability}
\label{sec:partial-funding}

Single-step transactions are atomic by construction.  The
vulnerability arises in \emph{multi-step} transactions: a
deployment that chains two or more for-comprehensions
sequentially may have enough tokens to fund the first step
but not the second.

Consider a payment from Alice to Bob, decomposed into two
operations:

\begin{enumerate}[nosep]
  \item \textbf{Debit:}  Extract a purse of amount~$n$ from
    Alice's wallet.
  \item \textbf{Credit:}  Deposit the purse into Bob's wallet.
\end{enumerate}

\noindent
Each operation is a for-comprehension, each wrapped in a
signed term $\signed{\cdots}{\mathit{Alice}}$, each consuming
tokens from Alice's stack.  If Alice's stack has enough tokens
for the debit but not the credit, the system reaches an
inconsistent intermediate state: funds extracted from Alice's
wallet, sitting in a purse, but the deposit into Bob's wallet
never fires because the fuel ran out.

This is a \emph{partial funding vulnerability}: the
transaction is semantically atomic (it should either complete
or not happen) but operationally non-atomic (it can stall
midway through).

\subsection{Rholang 1.2 Illustration}
\label{sec:rholang-illustration}

We illustrate the vulnerability using Rholang~1.2 syntax,
where the $\mathtt{?!}$ operator provides synchronous
(function-call-like) communication: $\mathtt{x?!(args)}$ sends
on~$x$ with a return channel and blocks until the reply.
All three code fragments below use the uniform-signing sugar
(Definition~\ref{def:sugar-uniform}):
$\signed{\recv{y}{x}{P}}{s}$ expands to
$\signed{\recv{y}{x}{\signed{P}{s}}}{s}$, so each signed
for-comprehension consumes \emph{two} tokens (one for the
communication, one for the continuation).

\medskip
\noindent
\textbf{Debit handler} --- extracts a purse from Alice's
wallet:
\begin{lstlisting}[style=rholang]
// Sugar: costs 2 Alice-tokens (comm + continuation)
{ for( amt <- aliceDebit?!( request, ... ) ){
     let ans = extractPurse( request, ... )
     in rtn!( ans )
} }_{ Alice }
\end{lstlisting}

\noindent
\textbf{Credit handler} --- deposits a purse into Bob's
account:
\begin{lstlisting}[style=rholang]
// Sugar: costs 2 Alice-tokens (comm + continuation)
{ for( amt <- bobCredit?!( ... ) ){
    new rcpt in
       addPurseAcct( amt, ... )
       in rtn!( rcpt )
} }_{ Alice }
\end{lstlisting}

\noindent
\textbf{Alice's token stack:}
\begin{lstlisting}[style=rholang]
Alice : Alice : Alice : Alice : Alice : Alice : S
\end{lstlisting}

\noindent
\textbf{Orchestrating deployment} --- chains the debit and
credit:
\begin{lstlisting}[style=rholang]
// Sugar: costs 2 Alice-tokens (comm + continuation)
{ for( amt <- aliceDebit?!( request, ... );
       rcpt <- bobCredit?!( amt, ... ) ){
     P
} }_{ Alice }
\end{lstlisting}

\noindent
The total token demand of this deployment is:
\begin{center}
  \begin{tabular}{lccc}
    \textbf{Signed layer}
      & \textbf{Comm}
      & \textbf{Cont.}
      & \textbf{Total} \\
    \midrule
    Orchestrator's $\signed{\cdots}{\mathit{Alice}}$
      & 1 & 1 & 2 \\
    Debit handler's $\signed{\cdots}{\mathit{Alice}}$
      & 1 & 1 & 2 \\
    Credit handler's $\signed{\cdots}{\mathit{Alice}}$
      & 1 & 1 & 2 \\
    \midrule
    \textbf{Total} & \textbf{3} & \textbf{3} & \textbf{6} \\
  \end{tabular}
\end{center}

\noindent
If Alice's stack has at least six tokens, all three
signed layers (and their continuations) are funded and the
transaction completes.  If the stack has only four tokens,
the orchestrator and the debit handler fire (consuming
four tokens), but the credit handler's signed layer finds
no token and \emph{blocks}.
Funds are stranded in the purse.

\subsection{Decomposing the Transaction Cost}
\label{sec:txn-cost-decomp}

The six-token count of Section~\ref{sec:rholang-illustration}
reflects the uniform-signing sugar expansion of three explicit
$\signed{\cdots}{\mathit{Alice}}$ layers.  However, the
$\mathtt{?!}$ operator itself desugars to a send followed by a
for-comprehension on the return channel, so each $\mathtt{?!}$
call involves a for-comprehension on \emph{each} side --- the
caller's and the handler's.  These internal for-comprehensions
are also under signed scope and therefore also consume tokens.

A full accounting, decomposed into the two computational
actions of Section~\ref{sec:for-as-txn}, is:

\begin{center}
  \begin{tabular}{lccc}
    \textbf{For-comprehension}
      & \textbf{Rendezvous}
      & \textbf{Matching}
      & \textbf{Total} \\
    \midrule
    Orchestrator: $\mathtt{aliceDebit?!}$
      & 1 & 1 & 2 \\
    Orchestrator: $\mathtt{bobCredit?!}$
      & 1 & 1 & 2 \\
    Debit handler: $\forc$
      & 1 & 1 & 2 \\
    Credit handler: $\forc$
      & 1 & 1 & 2 \\
    \midrule
    \textbf{Total} & \textbf{4} & \textbf{4} & \textbf{8} \\
  \end{tabular}
\end{center}

\noindent
The six-token count from the sugar expansion is a
\emph{syntactic} count of explicitly written signed layers;
the eight-token count here is the \emph{semantic} count after
full desugaring of~$\mathtt{?!}$.  The static analysis of
Section~\ref{sec:linear-proof} must use the semantic count,
since it is the desugared form that the runtime executes.

The token stack must supply \emph{all} of these or
\emph{none} of them.  Partial supply leads to partial
execution, which is the vulnerability.

\subsection{Static Analysis as a Linear Resource Proof}
\label{sec:linear-proof}

The partial funding vulnerability is eliminated by a
\emph{static analysis} performed at deployment-acceptance
time --- the moment the validator's fuel gate fires and the
signed deployment enters execution.

\begin{definition}[Token demand]
  \label{def:token-demand}
  The \emph{token demand} of a signed term~$T$ with respect
  to a signature~$s$, written $\Delta_s(T)$, is the total
  number of $\signed{\cdots}{s}$ layers in~$T$ and in all
  processes reachable from~$T$ via synchronous calls ($?!$)
  or dequotation ($\drop{\cdot}$).  Formally:
  \begin{align*}
    \Delta_s(\signed{P}{s})
      &= 1 + \Delta_s(P) \\
    \Delta_s(\signed{P}{s'})
      &= \Delta_s(P)
      \qquad\text{for } s' \neq s \\
    \Delta_s(\recv{y}{x}{T})
      &= \Delta_s(T) \\
    \Delta_s(\send{x}{U})
      &= \Delta_s(U) \\
    \Delta_s(T \Par U)
      &= \Delta_s(T) + \Delta_s(U) \\
    \Delta_s(\drop{x})
      &= \Delta_s^{\mathit{resolve}}(x)
  \end{align*}
  where $\Delta_s^{\mathit{resolve}}(x)$ resolves the
  name~$x$ to its definition (if statically known) and
  computes $\Delta_s$ on the result.
\end{definition}

\begin{definition}[Token supply]
  \label{def:token-supply}
  The \emph{token supply} of a signature~$s$ in a system is
  the depth of $s$-indexed token stacks in the ambient
  parallel composition:
  \begin{align*}
    \Sigma_s(s : S)
      &= 1 + \Sigma_s(S) \\
    \Sigma_s(s' : S)
      &= \Sigma_s(S)
      \qquad\text{for } s' \neq s \\
    \Sigma_s(())
      &= 0 \\
    \Sigma_s(T \Par U)
      &= \Sigma_s(T) + \Sigma_s(U)
  \end{align*}
\end{definition}

\begin{definition}[Funding proof obligation]
  \label{def:funding-proof}
  A deployment is \emph{fully funded} if, for every
  signature~$s$ appearing in the deployment:
  \begin{equation}
    \Sigma_s \;\geq\; \Delta_s
    \label{eq:funding-obligation}
  \end{equation}
  This is a \emph{linear resource inequality}: each token is
  consumed exactly once, and the total supply must meet or
  exceed the total demand.
\end{definition}

\begin{theorem}[Decidability of the funding check]
  \label{thm:decidability}
  For a deployment whose call graph is statically known (no
  higher-order channel passing, no recursive dequotation
  chains), the funding proof
  obligation~\eqref{eq:funding-obligation} is decidable in
  time linear in the size of the deployment's AST.
\end{theorem}

\noindent
\textit{Proof sketch.}  $\Delta_s$ is computed by a single
pass over the AST, incrementing a counter at each
$\signed{\cdots}{s}$ node.  $\Sigma_s$ is computed by
inspecting the token stacks in the ambient parallel
composition.  The comparison is a single integer inequality.
In the presence of higher-order channels or recursive
dequotation, the analysis becomes a conservative
over-approximation: unresolvable $\drop{x}$ terms contribute
an ``unknown'' demand, and the validator rejects unless the
supply exceeds the known lower bound plus a configurable
safety margin.\qed

\subsection{Data-Dependent Interaction: Overcharge and Refund}
\label{sec:data-dependent}

The linear resource proof certifies a \emph{fixed}
consumption: it sums the tokens a deployment will draw and
checks the sum against the supply.  This is sound when the
interactions a deployment performs are determinable
statically.  Under \emph{data-dependent interaction} it is
not immediate.  If a received value selects a branch, or a
name computed at run time determines which $\mathit{near}$
match fires, then which interactions occur --- and so how many
tokens of which signatures are drawn --- depends on the run,
and there is no single fixed quantity for the proof to
sum~\cite{CostEndofunctor}.

The companion paper identifies two strategies for the
data-dependent fragment: make the proof \emph{dependent}
(parameterised by the data, with each branch discharging its
own linear obligation), or fall back to a \emph{conservative
static bound} --- the worst case over all branches --- which
is sound but may over-charge.

\paragraph{We prefer overcharge-and-refund.}
For the F1R3FLY platform we adopt the conservative-bound
strategy, deliberately preferring to \emph{overcharge and
refund} rather than to under-fund.  The reasoning is a direct
consequence of the atomicity analysis:

\begin{itemize}[nosep]
  \item \textbf{Under-funding is a correctness failure.}  If
    the static bound under-estimates the demand --- as a
    naive per-branch or average-case bound might --- then a
    deployment can be accepted yet run out of phlogiston
    mid-execution, stranding funds in a purse
    (Section~\ref{sec:partial-funding}).  This reintroduces
    the partial-funding vulnerability the static analysis
    exists to eliminate.
  \item \textbf{Over-funding is merely an inefficiency.}  If
    the static bound over-estimates the demand --- charging
    the worst case over all branches --- then a deployment is
    accepted with more phlogiston committed than it will
    actually consume.  The unused phlogiston is simply
    refunded after execution completes.  No funds are
    stranded; no transaction stalls.
\end{itemize}

\noindent
Concretely, the validator computes a conservative demand
$\Delta_c^{\max}(D)$ --- the maximum over all reachable
branches of the data-dependent control flow --- and checks
$\Sigma_c \geq \Delta_c^{\max}$.  Execution consumes only the
tokens actually forced; the residual
$\Delta_c^{\max} - (\text{tokens consumed})$ is returned to
the client's token stack as part of the deployment's
settlement.

\begin{definition}[Conservative demand with refund]
  \label{def:conservative-demand}
  For a deployment~$D$ with data-dependent control flow, the
  \emph{conservative demand} is
  \[
    \Delta_c^{\max}(D)
    \;=\;
    \max_{\beta \in \mathrm{Branches}(D)} \Delta_c(\beta),
  \]
  the maximum token demand over all statically reachable
  branches.  Acceptance requires $\Sigma_c \geq
  \Delta_c^{\max}(D)$.  After execution, the refund is
  $\Delta_c^{\max}(D) - \kappa$, where~$\kappa$ is the number
  of tokens actually forced during the run.
\end{definition}

\begin{remark}[Why refund is safe]
  The refund is computed from the \emph{actual} run, which is
  observable: by Proposition~\ref{prop:subst-commute} and the
  subject-reduction lemma for wrapping
  (\cite{CostEndofunctor}, Lemma~3.2), the number of tokens
  consumed equals the number of forced redexes, which the
  runtime counts exactly as it executes.  Over-funding never
  threatens atomicity because the worst-case bound guarantees
  that \emph{every} branch is fully funded; the branch that
  actually runs is therefore funded a fortiori.  The refund
  reconciles the conservative estimate with the realised cost
  without ever risking a stall.
\end{remark}

\begin{remark}[Located purses sharpen the bound]
  The located resource stacks of
  Section~\ref{sec:ocap-authority} let the conservative bound
  be computed \emph{per surface} rather than globally.  If
  data-dependence is confined to a particular interaction
  surface, only that purse's sub-bound need be conservative;
  every other purse retains its exact static demand.  Location
  thus quarantines the over-charge to the surfaces that
  actually require it, minimising the refund overhead.  This
  is the rho-calculus realisation of the local-sufficiency
  decomposition of the companion paper
  (\cite{CostEndofunctor}, Proposition~12.1).
\end{remark}

\subsection{The Validator's Acceptance Protocol}
\label{sec:acceptance-protocol}

The static analysis integrates into the validator's acceptance
protocol (Appendix~\ref{app:example}) as follows:

\begin{enumerate}[nosep]
  \item The validator receives a signed deployment
    $\signed{D}{c}$ and looks up the client's token stack.
  \item \textbf{Before} executing any part of the deployment,
    the validator computes the demand by static analysis of
    the deployment's AST and call graph: the exact demand
    $\Delta_c(D)$ for the data-independent fragment, or the
    conservative bound $\Delta_c^{\max}(D)$
    (Definition~\ref{def:conservative-demand}) where control
    flow is data-dependent.
  \item The validator computes $\Sigma_c$ from the available
    token stack.
  \item If $\Sigma_c \geq \Delta_c^{\max}$, the deployment is
    \emph{accepted}: the validator knows, by the linear
    resource proof, that every for-comprehension in
    \emph{every reachable branch} of the deployment's
    execution will be funded.  Every substitution that begins
    will complete.  No funds will be stranded.  After
    execution, any phlogiston committed beyond what was
    actually forced is refunded.
  \item If $\Sigma_c < \Delta_c^{\max}$, the deployment is
    \emph{rejected}: the validator returns an error
    (``insufficient phlogiston'') without executing any part
    of the deployment.  No state change occurs.  No tokens
    are consumed.
\end{enumerate}

\noindent
The choice of the conservative bound $\Delta_c^{\max}$ over an
optimistic estimate is deliberate: it trades a refund (a
benign inefficiency) for the guarantee that no accepted
deployment can ever stall for lack of phlogiston (a
correctness property).  Over-charge, never under-fund.

\noindent
This protocol converts the database-level
atomicity-by-construction of individual rewrite rules into
\emph{financial-level} atomicity of multi-step transactions:
either the entire transaction is funded and executes to
completion, or it is rejected before any step fires.  The gap
between ``each step is atomic'' and ``the whole transaction is
atomic'' is bridged by the static analysis at acceptance time.

\subsection{Deployment Boundaries and Financial Atomicity}
\label{sec:deploy-boundaries}

The \emph{deployment} is the unit of financial-transaction
atomicity.  A deployment is a signed term submitted to a
validator for execution.  The validator's acceptance protocol
(Section~\ref{sec:acceptance-protocol}) either accepts the
entire deployment (the linear resource proof succeeds) or
rejects it entirely (the proof fails).  No partial acceptance
is possible.

This deployment-boundary semantics resolves several scenarios
that are problematic in conventional transaction processing:

\paragraph{Duplicate deployments.}
Suppose two copies of the same Alice~$\to$~Bob payment
deployment arrive at a validator, and Alice's token stack
contains exactly three tokens --- enough for one payment but
not two.  The validator processes each deployment as a
separate acceptance decision:

\begin{enumerate}[nosep]
  \item The first deployment arrives.  The validator computes
    $\Delta_{\mathit{Alice}} = 3$ and $\Sigma_{\mathit{Alice}}
    = 3$.  The linear proof succeeds: $3 \geq 3$.  The
    deployment is accepted.  Three tokens are committed.
  \item The second deployment arrives.  The validator computes
    $\Delta_{\mathit{Alice}} = 3$ and $\Sigma_{\mathit{Alice}}
    = 0$ (the tokens are already committed to the first
    deployment).  The linear proof fails: $0 < 3$.  The
    deployment is rejected.  No state change occurs.
\end{enumerate}

\noindent
The second deployment is rejected \emph{before any step
executes}.  There is no risk of a partial debit, no stranded
purse, no inconsistent state.  The linear resource proof
ensures that accepted deployments run to completion and
rejected deployments have no effect.

\paragraph{Simultaneous arrival.}
If the two deployments arrive at the same time, the validator
treats them as a \emph{single deployment} consisting of the
parallel composition of both terms:
\[
  \signed{D_1}{\mathit{Alice}} \Par \signed{D_2}{\mathit{Alice}}
\]
The static analysis computes
$\Delta_{\mathit{Alice}} = 3 + 3 = 6$ but
$\Sigma_{\mathit{Alice}} = 3$.  The linear proof fails:
$3 < 6$.  The combined deployment is rejected.  Neither
payment executes.

This is the correct behavior: there are not enough resources
to complete both financial transactions, so neither should
begin.  The linear resource proof detects this automatically,
without locks, priority queues, or rollback mechanisms.

\paragraph{Interleaved racing execution.}
In a na\"{\i}ve runtime that interleaves execution steps from
multiple deployments without checking resource bounds, the
following race is possible: deployment~1's debit fires
(consuming 2 tokens), then deployment~2's debit fires
(consuming the last token), and now neither deployment has
enough tokens for its credit step.  Both payments stall with
funds stranded in purses.

The linear resource proof eliminates this race entirely.
Acceptance is a \emph{gate}: a deployment passes through the
gate only if it carries a proof that all its resources are
available.  Once accepted, those resources are committed and
unavailable to other deployments.  There is no interleaving
of acceptance and execution --- acceptance is a single,
atomic decision that precedes all execution.

\begin{remark}[Deployment acceptance as linear proof search]
  The semantics is precise: a deployment is accepted
  \emph{if and only if} a linear proof of its resource usage
  exists.  The tokens consumed by the proof are
  \emph{linearly} committed --- they cannot be used by any
  other proof.  Two deployments competing for the same tokens
  are two proof obligations competing for the same linear
  hypotheses.  At most one can succeed.  This is not an
  implementation choice or a heuristic --- it is a theorem of
  linear logic.
\end{remark}

\begin{remark}[Connection to linear logic]
  The funding proof obligation~\eqref{eq:funding-obligation}
  is a judgment in the resource-sensitive fragment of linear
  logic.  Each token is a \emph{linear resource} that must
  be consumed exactly once.  The demand $\Delta_s$ is the
  number of linear hypotheses required; the supply $\Sigma_s$
  is the number of linear hypotheses available.  A successful
  funding check is a \emph{proof} in the linear resource
  logic that the deployment is well-resourced.

  This connection is not merely an analogy.  The OSLF functor
  (observer-generated logic) associated with the
  cost-accounted GSLT can be extended to generate a linear
  resource logic whose judgments \emph{are} funding proofs.
  The static analysis is then a proof search in this logic,
  and the validator is a proof checker.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}

We have presented the rho calculus as an instantiation of the
cost endofunctor~$\mathcal{C}$ on the category \textbf{ciGSLT}
of continued interactive graph-structured lambda
theories~\cite{CostEndofunctor}.  The instantiation proceeds
in three phases: identifying the rho interaction cut
($K = \Par$, $K_p = \forc$, $K_e = {!}$, $\mathit{compute}$
= pseudo-application), applying the cost construction to
produce the cost-accounted rho calculus, and exploiting
Turing completeness to translate the result back to the pure
rho calculus via the internalisation adjunction.

Each feature of the cost-accounted calculus is a consequence
of the general construction: signatures as names (the section
adjoined by~$\mathcal{C}$), signed terms pervading the syntax
(continuation slots re-sorted to~$T$), token stacks as
first-class terms (the temporal monoid), polymorphic~$\Par$
($K$ is the only constructor that needs overloading), and the
five rewrite rules (the gated family (R1)--(R3) specialised to
the rho interaction cut).  The disentangling of section and
quotient resolves the apparent need for ``two axes of
reflection'' --- a self-overlap specific to rho, where
$\quot{\cdot}$ serves both roles.

The identification of the for-comprehension as the transaction
primitive --- with rendezvous and matching as the two funded
actions, the substitution as the witness, and application to the
continuation as the state update --- yields a precise theory of
atomicity at two levels.  \emph{Database-transaction atomicity}
(individual communications) is a theorem of the rewrite rules
--- indeed, it is the subject-reduction lemma (Lemma~3.2
of~\cite{CostEndofunctor}) for the wrapping-by-construction
discipline.  \emph{Financial-transaction atomicity} (multi-step
operations like payments) is guaranteed by a decidable static
analysis at deployment-acceptance time: a linear resource proof
that verifies every for-comprehension in the deployment's call
graph is funded before any step executes.  Deployment boundaries
provide the unit of financial atomicity, and the linear proof
serves as the concurrency control mechanism --- competing
deployments are competing proof obligations over the same
linear resources, and at most one can succeed.

The object-capability analysis --- carried by the general
construction's treatment of $K$'s equational theory --- ensures
that the AC nature of~$\Par$ does not introduce ambient
authority: the nearness predicate (channel-name equality) and
located resource stacks confine authority to holders of matching
channel names, with unforgeable channels created by
$\mathbf{new}$ providing genuine capabilities.

The result is a calculus that embodies \emph{valuable friction}
at the syntactic level: every communicating process must engage
the cost layer, and the cost layer is rich enough to guarantee
transactional integrity.  Because the rho calculus is the formal
core of Rholang, this instantiation serves directly as the
design specification for rearchitecting phlogiston accounting in
Rholang~1.4 and the F1R3FLY platform.

\bigskip
\begin{flushright}
  \textit{E Pluribus Potentia}
\end{flushright}

\appendix

%% ══════════════════════════════════════════════════════════════════
\section{Source-to-Source Translation to Pure Rho Calculus}
\label{app:translation}

This appendix defines a compositional source-to-source
translation from the cost-accounted rho calculus of
Section~\ref{sec:calculus} to the \emph{pure} rho calculus.
In the categorical framework of~\cite{CostEndofunctor}, this
translation is an instance of the \emph{internalisation
adjunction} (Adjunction~II): because the rho calculus is
Turing-complete, the entire cost-accounting apparatus of
$\mathcal{C}(G)$ --- signature equality, stack maintenance,
gating --- is computable and hence encodable in~$G$ itself.
The translation realizes the cost-accounted semantics inside
the un-metered but universal base by simulation: tokens become
encoded data, the gated rules become a fuel-acquisition
protocol, and each forced step is simulated by a finite run of
the pure rho calculus.

The translation proceeds in three phases, mirroring the
three-phase structure of the paper:

\begin{enumerate}[label=\textbf{Phase \arabic*.}, nosep]
  \item \textbf{Instantiation $\to$ signatures.}
    The interaction-cut components of
    Section~\ref{sec:instantiation} determine how signatures
    are mapped to pure-rho names ($\Sigma\lsem\cdot\rsem$).
  \item \textbf{Cost construction $\to$ fuel gates.}
    The wrapping-by-construction discipline determines how
    signed terms become fuel-gated processes
    ($\mathcal{T}\lsem\cdot\rsem$), and how token stacks
    become messages on signature channels
    ($\mathcal{K}\lsem\cdot\rsem$).
  \item \textbf{Internalisation $\to$ infrastructure.}
    The splitter and joiner processes that mediate between
    combined and decomposed token representations are the
    ``administrative reductions'' of the internalisation:
    they simulate the token-distribution cases (R2), (R3)
    that the cost endofunctor provides as primitive rules.
\end{enumerate}

\noindent
The object-capability analysis of
Section~\ref{sec:ocap-authority} is preserved by the
translation: in the pure-rho encoding, the \emph{nearness
predicate} $\mathit{near}(I, J)$ becomes channel-name equality
--- only processes holding a matching channel name can consume
tokens from a located stack.  Because the rho calculus creates
unforgeable channel names via $\mathbf{new}$, the translation
automatically inherits the ocap discipline: authority over a
resource stack is conferred by possession of the channel name,
not by adjacency in the parallel-composition bag.

The translation is organized into five mutually recursive
functions:

\begin{center}
  \begin{tabular}{llll}
    \textbf{Function} & \textbf{Domain} & \textbf{Codomain}
      & \textbf{Purpose} \\
    \midrule
    $\mathcal{P}\lsem\cdot\rsem$
      & Processes   & Pure processes
      & Translate processes \\
    $\mathcal{N}\lsem\cdot\rsem$
      & Names       & Pure names
      & Translate names \\
    $\mathcal{T}\lsem\cdot\rsem$
      & Signed terms & Pure processes
      & Translate signed terms \\
    $\mathcal{K}\lsem\cdot\rsem$
      & Token stacks & Pure processes
      & Translate token stacks \\
    $\Sigma\lsem\cdot\rsem$
      & Signatures  & Pure names
      & Translate signatures \\
  \end{tabular}
\end{center}

\subsection{Signature Translation: $\Sigma\lsem\cdot\rsem$}
\label{app:sig-trans}

Signatures are mapped to names in the pure rho calculus.  Each
signature becomes a channel on which fuel tokens will be
communicated.

\begin{align}
  \Sigma\lsem g \rsem
    &= \quot{H_g}
    \quad\text{for a canonical process~$H_g$ encoding ground
      signature~$g \in G$}
    \label{eq:app-sig-ground}
  \\[4pt]
  \Sigma\lsem \hashf P \rsem
    &= \quot{H(\mathcal{P}\lsem P \rsem)}
    \quad\text{where~$H(\cdot)$ is a hash-encoding process
      constructor}
    \label{eq:app-sig-hash}
  \\[4pt]
  \Sigma\lsem s_1 \comp s_2 \rsem
    &= \quot{\bigl(
        \drop{\Sigma\lsem s_1 \rsem}
        \Par
        \drop{\Sigma\lsem s_2 \rsem}
      \bigr)}
    \label{eq:app-sig-comp}
\end{align}

\noindent
Equation~\eqref{eq:app-sig-comp} exploits reflection: the
name for a compound signature is the quote of the parallel
composition of the dequotations of the component names,
ensuring injectivity.

\begin{remark}[Ground-signature encoding]
  The canonical process~$H_g$ must be injective in~$g$: distinct
  ground signatures yield distinct processes.  In practice, $H_g$
  can be a process whose serialized form is the byte sequence of
  the public key or key hash~$g$.  Similarly, $H(\cdot)$ encodes
  the output of the cryptographic hash function applied to the
  serialized process tree.
\end{remark}

\subsection{Name Translation: $\mathcal{N}\lsem\cdot\rsem$}
\label{app:name-trans}

\begin{align}
  \mathcal{N}\lsem \quot{T} \rsem
    &= \quot{\mathcal{T}\lsem T \rsem}
    \label{eq:app-name-quote}
  \\[4pt]
  \mathcal{N}\lsem s \rsem
    &= \Sigma\lsem s \rsem
    \label{eq:app-name-sig}
\end{align}

\noindent
Quoted signed terms become quoted translations.  Signatures
become their signature-channel names.

\subsection{Token-Stack Translation: $\mathcal{K}\lsem\cdot\rsem$}
\label{app:token-trans}

A token stack becomes a chain of sends, one per layer, on the
corresponding signature channels.  Each send carries the
translation of the remaining stack as its payload.

\begin{align}
  \mathcal{K}\lsem () \rsem
    &= \nil
    \label{eq:app-tok-empty}
  \\[4pt]
  \mathcal{K}\lsem s : S \rsem
    &= \send{\Sigma\lsem s \rsem}{\mathcal{K}\lsem S \rsem}
    \label{eq:app-tok-cons}
\end{align}

\noindent
The token $s : S$ becomes a message on channel
$\Sigma\lsem s \rsem$ whose payload is the code of the
remaining balance~$\mathcal{K}\lsem S \rsem$.  This is the
\emph{fuel} that a signed process must consume before it can
act.

\subsection{Signed-Term Translation: $\mathcal{T}\lsem\cdot\rsem$}
\label{app:signed-trans}

Signed terms translate to pure processes.

\begin{align}
  \mathcal{T}\lsem T \Par U \rsem
    &= \mathcal{T}\lsem T \rsem \Par \mathcal{T}\lsem U \rsem
    \label{eq:app-st-par}
  \\[4pt]
  \mathcal{T}\lsem S \rsem
    &= \mathcal{K}\lsem S \rsem
    \label{eq:app-st-stack}
\end{align}

\noindent
For signed processes $\signed{P}{s}$, the translation wraps
the process in a \emph{fuel gate}: a for-comprehension on the
signature channel that must receive a token before the process
can proceed.  The consumed token's payload (the remaining
balance) is released into the continuation via dequotation.

\medskip
\noindent
\textbf{Atomic signature:}
\begin{equation}
  \mathcal{T}\lsem \signed{P}{s} \rsem
  \;=\;
  \recv{t}{\Sigma\lsem s \rsem}{
    \bigl(\drop{t} \Par \mathcal{P}\lsem P \rsem\bigr)
  }
  \label{eq:app-st-signed-atomic}
\end{equation}

\noindent
The process $\mathcal{P}\lsem P \rsem$ is blocked behind the
fuel gate~$\recv{t}{\Sigma\lsem s \rsem}{\cdots}$.  When a
token arrives, $t$ is bound to the code of the remaining
balance; $\drop{t}$ releases it; and the translated process
runs in parallel.

\medskip
\noindent
\textbf{Compound signature:}
\begin{equation}
  \mathcal{T}\lsem \signed{P}{s_1 \comp s_2} \rsem
  \;=\;
  \recv{t_1}{\Sigma\lsem s_1 \rsem}{
    \recv{t_2}{\Sigma\lsem s_2 \rsem}{
      \bigl(
        \drop{t_1} \Par \drop{t_2}
        \Par \mathcal{P}\lsem P \rsem
      \bigr)
    }
  }
  \label{eq:app-st-signed-compound}
\end{equation}

\noindent
Both component channels must supply fuel.  This generalizes
to $n$-ary composition $s_1 \comp s_2 \comp \cdots \comp s_n$
by nesting~$n$ fuel gates.

\subsection{Process Translation: $\mathcal{P}\lsem\cdot\rsem$}
\label{app:proc-trans}

\begin{align}
  \mathcal{P}\lsem \nil \rsem
    &= \nil
    \label{eq:app-p-nil}
  \\[4pt]
  \mathcal{P}\lsem \recv{y}{x}{T} \rsem
    &= \recv{y}{\mathcal{N}\lsem x \rsem}{
        \mathcal{T}\lsem T \rsem
      }
    \label{eq:app-p-recv}
  \\[4pt]
  \mathcal{P}\lsem \send{x}{U} \rsem
    &= \send{\mathcal{N}\lsem x \rsem}{
        \mathcal{T}\lsem U \rsem
      }
    \label{eq:app-p-send}
  \\[4pt]
  \mathcal{P}\lsem P \Par Q \rsem
    &= \mathcal{P}\lsem P \rsem
       \Par \mathcal{P}\lsem Q \rsem
    \label{eq:app-p-par}
  \\[4pt]
  \mathcal{P}\lsem \drop{x} \rsem
    &= \drop{\mathcal{N}\lsem x \rsem}
    \label{eq:app-p-drop}
\end{align}

\subsection{Infrastructure Processes}
\label{app:infrastructure}

When the granularity of token holding (combined vs.\ split)
does not match what the fuel gates expect, infrastructure
processes mediate the conversion.  These are deployed as
persistent system-level contracts.

\begin{definition}[Splitter]
  \label{def:app-splitter}
  Converts a combined token for~$s_1 \comp s_2$ into separate
  tokens for~$s_1$ and~$s_2$:
  \[
    \mathrm{Split}(s_1, s_2)
    \;=\;
    \recv{t}{\Sigma\lsem s_1 \comp s_2 \rsem}{
      \bigl(
        \send{\Sigma\lsem s_1 \rsem}{\nil}
        \Par
        \send{\Sigma\lsem s_2 \rsem}{\drop{t}}
      \bigr)
    }
  \]
\end{definition}

\begin{definition}[Joiner]
  \label{def:app-joiner}
  Converts separate tokens for~$s_1$ and~$s_2$ into a combined
  token for~$s_1 \comp s_2$:
  \[
    \mathrm{Join}(s_1, s_2)
    \;=\;
    \recv{t_1}{\Sigma\lsem s_1 \rsem}{
      \recv{t_2}{\Sigma\lsem s_2 \rsem}{
        \send{\Sigma\lsem s_1 \comp s_2 \rsem}{
          (\drop{t_1} \Par \drop{t_2})
        }
      }
    }
  \]
\end{definition}

\begin{remark}[Persistence]
  In production, splitters and joiners must be
  \emph{replicated} (persistent) processes.  The standard
  rho-calculus encoding of replication via self-referencing
  through reflection applies:
  \[
    !\,P
    \;\triangleq\;
    \recv{y}{\quot{D}}{(\drop{y} \Par P)}
    \Par
    \send{\quot{D}}{D}
    \qquad\text{where } D = \recv{y}{\quot{D}}{(\drop{y} \Par P)}
  \]
\end{remark}

\subsection{Verification Sketch}
\label{app:verification}

We verify that each rewrite rule of the cost-accounted
calculus is faithfully simulated by the translation.  Write
$\Vert\cdot\Vert$ for the system-level translation that
applies $\mathcal{T}\lsem\cdot\rsem$ to every signed-term
component and $\mathcal{K}\lsem\cdot\rsem$ to every token
stack, composing them in parallel.

\bigskip
\noindent
\textbf{Rule~1.}\;
$\signed{\recv{y}{x}{T} \Par \send{x}{U}}{s} \Par s : S
\;\red\; \subst{T}{\quot{U}}{y} \Par S$

\medskip\noindent
\textit{Translation of LHS:}
\begin{align*}
  &\recv{t}{\Sigma\lsem s \rsem}{
    \bigl(
      \drop{t}
      \Par \recv{y}{\mathcal{N}\lsem x \rsem}{
        \mathcal{T}\lsem T \rsem
      }
      \Par \send{\mathcal{N}\lsem x \rsem}{
        \mathcal{T}\lsem U \rsem
      }
    \bigr)
  }
  \\
  &\Par\;
  \send{\Sigma\lsem s \rsem}{\mathcal{K}\lsem S \rsem}
\end{align*}

\noindent
\textit{Step~1} --- fuel consumption on
$\Sigma\lsem s \rsem$, binding
$t \mapsto \quot{\mathcal{K}\lsem S \rsem}$:
\[
  \mathcal{K}\lsem S \rsem
  \Par
  \recv{y}{\mathcal{N}\lsem x \rsem}{
    \mathcal{T}\lsem T \rsem
  }
  \Par
  \send{\mathcal{N}\lsem x \rsem}{
    \mathcal{T}\lsem U \rsem
  }
\]

\noindent
\textit{Step~2} --- communication on
$\mathcal{N}\lsem x \rsem$, binding
$y \mapsto \quot{\mathcal{T}\lsem U \rsem}$:
\[
  \mathcal{K}\lsem S \rsem
  \Par
  \mathcal{T}\lsem T \rsem
    \bigl\{\quot{\mathcal{T}\lsem U \rsem}\big/y\bigr\}
\]

\noindent
By compositionality of the translation (the translation
commutes with substitution, Proposition~\ref{prop:subst-commute}),
this equals
$\mathcal{T}\lsem \subst{T}{\quot{U}}{y} \rsem \Par
\mathcal{K}\lsem S \rsem = \Vert \subst{T}{\quot{U}}{y} \Par S
\Vert$.\quad$\checkmark$

\bigskip
\noindent
\textbf{Rule~2.}\;
$\signed{\recv{y}{x}{T} \Par \send{x}{U}}{s_1 \comp s_2}
\Par s_1 : S \Par s_2 : S'
\;\red\; \subst{T}{\quot{U}}{y} \Par S \Par S'$

\medskip\noindent
The compound-signature fuel gate
\eqref{eq:app-st-signed-compound} nests two
$\mathbf{for}$-comprehensions on $\Sigma\lsem s_1 \rsem$ and
$\Sigma\lsem s_2 \rsem$.  The two token translations
$\mathcal{K}\lsem s_1 : S \rsem$ and
$\mathcal{K}\lsem s_2 : S' \rsem$ supply fuel to each.
After both fire and the data communication on
$\mathcal{N}\lsem x \rsem$ completes, the result is
$\mathcal{K}\lsem S \rsem \Par \mathcal{K}\lsem S' \rsem \Par
\mathcal{T}\lsem \subst{T}{\quot{U}}{y} \rsem
= \Vert \subst{T}{\quot{U}}{y} \Par S \Par S' \Vert$.
\quad$\checkmark$

\bigskip
\noindent
\textbf{Rule~3.}\;
$\signed{\recv{y}{x}{T} \Par \send{x}{U}}{s_1 \comp s_2}
\Par (s_1 \comp s_2) : S
\;\red\; \subst{T}{\quot{U}}{y} \Par S$

\medskip\noindent
The fuel gate listens on $\Sigma\lsem s_1 \rsem$ and
$\Sigma\lsem s_2 \rsem$, but the token lives on
$\Sigma\lsem s_1 \comp s_2 \rsem$.  The splitter
(Definition~\ref{def:app-splitter}) mediates:
\[
  \mathrm{Split}(s_1, s_2)
  \Par
  \send{\Sigma\lsem s_1 \comp s_2 \rsem}{
    \mathcal{K}\lsem S \rsem
  }
  \;\red\;
  \send{\Sigma\lsem s_1 \rsem}{\nil}
  \Par
  \send{\Sigma\lsem s_2 \rsem}{
    \mathcal{K}\lsem S \rsem
  }
\]
Now both fuel channels are populated and the reduction
proceeds as in Rule~2.  The residual $\nil$ from the
$s_1$-token is absorbed by structural equivalence.
\quad$\checkmark$

\bigskip
\noindent
\textbf{Rule~4.}\;
$\signed{\recv{y}{x}{T}}{s_1}
\Par \signed{\send{x}{U}}{s_2}
\Par s_1 : S \Par s_2 : S'
\;\red\; \subst{T}{\quot{U}}{y} \Par S \Par S'$

\medskip\noindent
Each signed process has an atomic fuel gate.  The
separately signed for-comprehension listens on
$\Sigma\lsem s_1 \rsem$; the separately signed send listens
on $\Sigma\lsem s_2 \rsem$.  Both tokens supply fuel
independently.  After fuel consumption:
\[
  \mathcal{K}\lsem S \rsem
  \Par
  \recv{y}{\mathcal{N}\lsem x \rsem}{
    \mathcal{T}\lsem T \rsem
  }
  \Par
  \send{\mathcal{N}\lsem x \rsem}{
    \mathcal{T}\lsem U \rsem
  }
  \Par
  \mathcal{K}\lsem S' \rsem
\]
The data communication on $\mathcal{N}\lsem x \rsem$
completes, yielding
$\Vert \subst{T}{\quot{U}}{y} \Par S \Par S' \Vert$.
\quad$\checkmark$

\bigskip
\noindent
\textbf{Rule~5.}\;
$\signed{\recv{y}{x}{T}}{s_1}
\Par \signed{\send{x}{U}}{s_2}
\Par (s_1 \comp s_2) : S
\;\red\; \subst{T}{\quot{U}}{y} \Par S$

\medskip\noindent
The fuel gates listen on $\Sigma\lsem s_1 \rsem$ and
$\Sigma\lsem s_2 \rsem$ separately, but the token is
combined on $\Sigma\lsem s_1 \comp s_2 \rsem$.  The splitter
mediates exactly as in Rule~3, producing separate tokens.
After fuel consumption and data communication, the result is
$\Vert \subst{T}{\quot{U}}{y} \Par S \Vert$.
\quad$\checkmark$

\subsection{Compositionality}
\label{app:compositionality}

The correctness of the verification rests on the following
property:

\begin{proposition}[Translation commutes with substitution]
  \label{prop:subst-commute}
  For all signed terms~$T$, $U$ and variables~$y$:
  \[
    \mathcal{T}\lsem T \rsem
    \bigl\{
      \quot{\mathcal{T}\lsem U \rsem} \big/ y
    \bigr\}
    \;=\;
    \mathcal{T}\lsem
      \subst{T}{\quot{U}}{y}
    \rsem
  \]
\end{proposition}

\noindent
The proof proceeds by structural induction on~$T$, with the
base case at~$\drop{x}$ using the semantic substitution
mechanism of the rho calculus.  The critical case is:
\[
  \drop{\mathcal{N}\lsem x \rsem}
  \bigl\{
    \quot{\mathcal{T}\lsem U \rsem} \big/ y
  \bigr\}
  \;=\;
  \begin{cases}
    \mathcal{T}\lsem U \rsem
      & \text{if } y \Neq \mathcal{N}\lsem x \rsem \\
    \drop{\mathcal{N}\lsem x \rsem}
      & \text{otherwise}
  \end{cases}
\]
which matches
$\mathcal{P}\lsem (\drop{x})\{\quot{U}/y\} \rsem$
by definition of semantic substitution.

\subsection{Implementation Notes}
\label{app:impl}

The translation is designed as a \textbf{compiler pass} in the
Rholang toolchain, structured to mirror the three phases
above:

\begin{enumerate}[nosep]
  \item \textbf{Phase 1 (Instantiation):}  Parse the extended
    syntax (processes, signed terms, token stacks, parameterized
    signatures) into an extended AST.  The parser recognizes the
    interaction-cut components: $\forc$ as~$K_p$, $!$ as~$K_e$,
    channels as interaction surfaces, and $\signed{\cdot}{s}$ as
    wrapping.
  \item \textbf{Phase 2 (Cost construction):}  Apply
    $\mathcal{P}\lsem\cdot\rsem$ to the top-level process,
    which recursively invokes
    $\mathcal{T}\lsem\cdot\rsem$,
    $\mathcal{K}\lsem\cdot\rsem$,
    $\mathcal{N}\lsem\cdot\rsem$, and
    $\Sigma\lsem\cdot\rsem$.  Each signed term becomes a
    fuel-gated process; each token stack becomes a chain of
    sends on the corresponding signature channel.  The
    ocap discipline is preserved: channel-name equality in
    the output serves as the $\mathit{near}(I, J)$ predicate.
  \item \textbf{Phase 3 (Internalisation):}  Inject persistent
    $\mathrm{Split}$ and $\mathrm{Join}$ infrastructure for
    every compound signature that appears in the source,
    unless the programmer has explicitly provided them.
    These are the administrative reductions of the
    internalisation adjunction: they simulate the
    token-distribution cases (R2), (R3) within the pure rho
    calculus.
  \item \textbf{Emit} pure rho calculus (Rholang~1.1) source.
    The existing runtime, reducer, and RSpace handle
    execution without modification.
\end{enumerate}

\noindent
In the terminology of~\cite{CostEndofunctor}, this compiler
pass is the internalisation map
$I_G : \mathcal{C}(G) \to G$ of Adjunction~II
(Proposition~9.5): it realises the cost-accounted semantics
inside the un-metered but Turing-complete base by simulation,
with each forced step simulated by a finite run of the pure
rho calculus.  The agreement $I_G \circ \eta_G \approx
\mathrm{id}_G$ holds up to weak bisimulation, the
administrative reductions being those of the fuel-acquisition
protocol.

This gives the platform immediate access to cost-accounted
semantics while the native implementation
(Section~\ref{sec:implications}) is developed in parallel.

%% ══════════════════════════════════════════════════════════════════
\section{Worked Example: Validator with Fee Extraction}
\label{app:example}

This appendix develops a complete worked example: a simplified
model of a validator node in the F1R3FLY network, expressed
first in the cost-accounted rho calculus and then translated
to pure rho via the compiler pass of Appendix~\ref{app:translation}.
The example exercises the bootstrap problem, token lookup,
error handling, and conversion of client tokens into validator
fees.

\subsection{Scenario and Conventions}
\label{app:ex-scenario}

Fix a ground signature scheme~$G$.  The actors are:

\begin{center}
  \begin{tabular}{lll}
    \textbf{Actor} & \textbf{Signature} & \textbf{Role} \\
    \midrule
    Validator & $v \in G$ & Accepts and executes deploys \\
    Client    & $c \in G$ & Submits a signed deploy \\
  \end{tabular}
\end{center}

\noindent
Since signatures are names, $v$ and $c$ double as channel
names.  We introduce five auxiliary channels (pure-rho names):

\begin{center}
  \begin{tabular}{ll}
    $\quot{DQ_v}$ & The validator's \emph{deploy queue} \\
    $\quot{F_v}$ & The validator's fee-collection channel \\
    $\quot{E}$   & The error-reporting channel \\
    $\quot{A_c}$ & The address where the client deposits tokens \\
    $\quot{W_v}$ & The validator's \emph{phlogiston wallet} \\
  \end{tabular}
\end{center}

\paragraph{Stake vs.\ phlogiston.}
It is important to distinguish the validator's \emph{stake}
from its \emph{phlogiston}.  The stake is the quantity of
$v$-tokens locked in the staking contract; it determines the
validator's weight in Casper CBC consensus and is subject to
slashing.  Phlogiston is a \emph{renewable resource} that is
\emph{minted} at the time of staking and \emph{freshly
minted} at each epoch boundary, deposited at the wallet
channel~$\quot{W_v}$.  Phlogiston is not drawn from stake
--- it is a separate resource whose continued minting is
contingent on the validator's good behavior.

At each \emph{epoch boundary} (when the set of staked
validators is updated), the protocol mints a fresh supply of
phlogiston for the validator's wallet, provided the validator
has followed the rules of the protocol.  If the validator
commits a slashing offense, all remaining phlogiston is
removed and no further phlogiston is minted.  The validator
is effectively offline: the bootstrap for-comprehension
on~$\quot{W_v}$ finds no message and blocks.

A slashed validator may be \emph{redeemed} --- proven to have
been in the right on the slashing offense, or having reached
some other arrangement with the shard.  Upon redemption,
phlogiston minting resumes at the next epoch boundary.  The
stake and any phlogiston held at the time of slashing are
moved to a private unforgeable channel pending adjudication.
After adjudication, the tokens may be returned, partially
redistributed as a penalty, or burned.

In the example below, we model the phlogiston wallet as a
message on~$\quot{W_v}$ carrying a token stack.  The
validator handler consumes phlogiston from this wallet to
process client deployments.

\subsection{Token Stack Decomposition}
\label{app:ex-decomp}

We adopt the following structural equivalence for token stacks,
which asserts that tokens are \emph{fungible and decomposable}:

\begin{equation}
  s : S
  \;\Seq\;
  (s : ()) \Par S
  \qquad\text{(token decomposition)}
  \label{eq:token-decomp}
\end{equation}

\noindent
Read: a multi-layer token stack is structurally equivalent to
the parallel composition of a single-token stack and the
remaining stack.  Repeated application decomposes
$s : s : s : ()$ into three parallel single-token stacks
$\bigl(s : ()\bigr) \Par \bigl(s : ()\bigr) \Par
\bigl(s : ()\bigr)$.

This equivalence is justified operationally: in the
source-to-source translation
(Appendix~\ref{app:translation}), the token stack
$s : s : s : ()$ translates to a \emph{nested} chain of
sends on $\Sigma\lsem s \rsem$, each carrying the remainder
as payload.  Decomposition corresponds to the standard
property that a communication can peel off one layer,
releasing the rest.

\subsection{Client-Side Assembly}
\label{app:ex-client}

The client prepares a deployment~$D$ (a process containing a
communication) and deposits three tokens at address~$\quot{A_c}$:

\begin{equation}
  \boxed{
    C
    \;\triangleq\;
    \send{\quot{A_c}}{c : c : c : ()}
    \;\Par\;
    \send{\quot{DQ_v}}{\signed{D}{c}}
  }
  \label{eq:client-assembly}
\end{equation}

\noindent
Here $\send{\quot{A_c}}{c : c : c : ()}$ deposits the
three-token stack at the client's token address, and
$\send{\quot{DQ_v}}{\signed{D}{c}}$ sends the signed deployment to
the validator's deploy queue~$\quot{DQ_v}$.

For concreteness, let~$D$ be a simple communication:
\begin{equation}
  D \;\triangleq\;
  \recv{z}{x}{T_{\mathit{app}}} \Par \send{x}{U_{\mathit{app}}}
  \label{eq:deploy-body}
\end{equation}
where $x$, $T_{\mathit{app}}$, and $U_{\mathit{app}}$ are the
client's application-level channel, continuation, and payload.

\subsection{Validator Handler}
\label{app:ex-validator}

The validator handler is a process that:
\begin{enumerate}[nosep]
  \item receives a signed deployment on the deploy
    queue~$\quot{DQ_v}$;
  \item looks up the client's tokens at address~$\quot{A_c}$;
  \item if tokens are present, extracts one token as a fee and
    composes the deployment with the remaining tokens;
  \item if no tokens are present, reports an error.
\end{enumerate}

\noindent
We define the handler using the uniform-signing sugar
(Definition~\ref{def:sugar-uniform}):

\begin{equation}
  \boxed{
    \mathit{VH}
    \;\triangleq\;
    \signed{
      \recv{\mathit{dep}}{\quot{DQ_v}}{
        \recv{\mathit{tok}}{\quot{A_c}}{
          \mathit{FeeExtract}
        }
      }
    }{v}
  }
  \label{eq:validator-handler}
\end{equation}

\noindent
The single outer $\signed{\cdot}{v}$ suffices because the
uniform-signing sugar~\eqref{eq:sugar-uniform} nests
recursively.  Expanding one level:
$\signed{\recv{\mathit{dep}}{\quot{DQ_v}}{\recv{\mathit{tok}}
{\quot{A_c}}{\mathit{FeeExtract}}}}{v}$ becomes
$\signed{\recv{\mathit{dep}}{\quot{DQ_v}}{\signed{\recv{\mathit{tok}}
{\quot{A_c}}{\mathit{FeeExtract}}}{v}}}{v}$, and the inner
$\signed{\recv{\mathit{tok}}{\quot{A_c}}
{\mathit{FeeExtract}}}{v}$ is itself sugar, expanding to
$\signed{\recv{\mathit{tok}}{\quot{A_c}}
{\signed{\mathit{FeeExtract}}{v}}}{v}$.  The fully desugared
form therefore has three $\signed{\cdot}{v}$ layers,
consuming three phlogiston.

The fee-extraction process $\mathit{FeeExtract}$ applies
token decomposition~\eqref{eq:token-decomp} to the received
token stack and splits it into a fee portion and a deployment
portion:

\begin{equation}
  \mathit{FeeExtract}
  \;\triangleq\;
  \send{\quot{F_v}}{c : ()}
  \;\Par\;
  \drop{\mathit{dep}} \;\Par\; (c : c : ())
  \label{eq:fee-extract}
\end{equation}

\noindent
Here:
\begin{itemize}[nosep]
  \item $\send{\quot{F_v}}{c : ()}$ sends one token (the fee)
    to the validator's fee-collection channel~$\quot{F_v}$.
  \item $\drop{\mathit{dep}}$ dequotes the received deployment,
    recovering the signed term~$\signed{D}{c}$.
  \item $(c : c : ())$ is the remaining two-token stack,
    placed in parallel with the deployment so that the
    cost-accounted rewrite rules can fire.
\end{itemize}

\noindent
The token decomposition~\eqref{eq:token-decomp} justifies the
split: when the validator receives the three-token stack via
$\recv{\mathit{tok}}{\quot{A_c}}{\cdots}$, the bound
name~$\mathit{tok}$ carries $\quot{c : c : c : ()}$.
Dequoting and decomposing:
\[
  c : c : c : ()
  \;\Seq\;
  (c : ()) \Par (c : c : ())
\]

\begin{remark}[Error handling]
  \label{rem:error}
  If the client fails to deposit tokens, the lookup
  $\recv{\mathit{tok}}{\quot{A_c}}{\cdots}$ simply
  \emph{blocks}: no message is available on~$\quot{A_c}$,
  and the fuel gate prevents execution.  This is the
  \emph{default} error mode --- the deployment is silently
  rejected by resource starvation.

  For proactive error reporting, the validator can race the
  token lookup against a timeout.  Using the standard
  select encoding in the rho calculus (a pair of
  for-comprehensions on the token channel and a timeout
  channel, with the first to fire pre-empting the other):
  \[
    \recv{\mathit{tok}}{\quot{A_c}}{
      \signed{\mathit{FeeExtract}}{v}
    }
    \;\Par\;
    \recv{\mathit{\_}}{\quot{\mathit{timeout}}}{
      \signed{
        \send{\quot{E}}{
          \signed{\mathit{ErrNoTokens}}{v}
        }
      }{v}
    }
  \]
  where a timeout process concurrently sends on
  $\quot{\mathit{timeout}}$ after a deadline.
\end{remark}

\subsection{Bootstrap: The Full System}
\label{app:ex-bootstrap}

The validator handler~$\mathit{VH}$ is itself a signed
process~$\signed{\cdots}{v}$.  By the cost-accounted rewrite
rules, it cannot execute unless $v$-tokens are present.
These tokens come from the validator's phlogiston
wallet~$\quot{W_v}$, which receives freshly minted
phlogiston at each epoch boundary.

To connect the handler to the wallet, we wrap it in a
for-comprehension that draws phlogiston from
$\quot{W_v}$:

\begin{equation}
  \boxed{
    \mathit{VB}
    \;\triangleq\;
    \recv{\mathit{phlo}}{\quot{W_v}}{
      \bigl(
        \mathit{VH} \Par \drop{\mathit{phlo}}
      \bigr)
    }
  }
  \label{eq:validator-boot}
\end{equation}

\noindent
When a message carrying a token stack arrives on
$\quot{W_v}$, the variable~$\mathit{phlo}$ is bound to
the quoted stack; $\drop{\mathit{phlo}}$ dequotes it,
releasing the tokens into the ambient parallel composition
where they can be consumed by $\mathit{VH}$'s signed layers.

The full bootstrapped system is:

\begin{equation}
  \boxed{
    \Omega
    \;\triangleq\;
    \underbrace{
      \mathit{VB}
    }_{\text{validator bootstrap}}
    \;\Par\;
    \underbrace{
      \send{\quot{W_v}}{v : v : v : ()}
    }_{\text{phlogiston from wallet}}
    \;\Par\;
    \underbrace{
      C
    }_{\text{client assembly}}
  }
  \label{eq:full-system}
\end{equation}

\noindent
Expanding $\mathit{VB}$ and $C$ (with $\mathit{VH}$ in
sugared form per~\eqref{eq:validator-handler}):
\begin{align}
  \Omega
  \;=\;\;
  &\recv{\mathit{phlo}}{\quot{W_v}}{
    \bigl(
      \signed{
        \recv{\mathit{dep}}{\quot{DQ_v}}{
          \recv{\mathit{tok}}{\quot{A_c}}{
            \mathit{FeeExtract}
          }
        }
      }{v}
      \Par \drop{\mathit{phlo}}
    \bigr)
  }
  \notag\\
  &\Par\; \send{\quot{W_v}}{v : v : v : ()}
  \notag\\
  &\Par\; \send{\quot{A_c}}{c : c : c : ()}
  \;\Par\;
  \send{\quot{DQ_v}}{\signed{D}{c}}
  \label{eq:full-expanded}
\end{align}

\noindent
Recall that $\mathit{VH}$ uses the recursive uniform-signing
sugar (Definition~\ref{def:sugar-uniform}): the single outer
$\signed{\cdot}{v}$ expands recursively to three
$\signed{\cdot}{v}$ layers in the fully desugared form, one
per for-comprehension plus one for $\mathit{FeeExtract}$.

The validator's stake (locked in the staking contract) is
\emph{not} consumed directly.  The phlogiston
$v : v : v : ()$ on~$\quot{W_v}$ is a renewable resource
minted at epoch boundaries, matched exactly to the three
desugared signed layers of the handler.

\subsection{Reduction in the Cost-Accounted Calculus}
\label{app:ex-reduction-ca}

We trace the reduction of~$\Omega$ using the cost-accounted
rewrite rules of Section~\ref{sec:rewrites}.  At each step
we identify which rule applies.

\bigskip
\noindent
\textbf{Step~0} --- The bootstrap process draws execution
tokens from the wallet (\textsc{Comm} on~$\quot{W_v}$):

The for-comprehension $\recv{\mathit{phlo}}{\quot{W_v}}{\cdots}$
communicates with $\send{\quot{W_v}}{v : v : v : ()}$, binding
$\mathit{phlo} \mapsto \quot{v : v : v : ()}$.  After
dequotation of $\drop{\mathit{phlo}}$, the system becomes:
\begin{align*}
  &\signed{
    \recv{\mathit{dep}}{\quot{DQ_v}}{
      \signed{
        \recv{\mathit{tok}}{\quot{A_c}}{
          \signed{\mathit{FeeExtract}}{v}
        }
      }{v}
    }
  }{v}
  \;\Par\; v : v : v : ()
  \\
  &\Par\; \send{\quot{A_c}}{c : c : c : ()}
  \;\Par\;
  \send{\quot{DQ_v}}{\signed{D}{c}}
\end{align*}

\noindent
The phlogiston are now in the ambient parallel
composition, available to the handler's signed layers.  Note
that this step is \emph{not} a cost-accounted rewrite: the
bootstrap for-comprehension on~$\quot{W_v}$ is an unsigned
process (a pure-rho communication), not a signed term.  It
is free infrastructure, analogous to the operating system's
process scheduler.

\bigskip
\noindent
\textbf{Step~1} --- The outermost signed handler consumes one
phlogiston and receives the signed deployment
(\textbf{Rule~1}):

The signed handler's outermost layer is
$\signed{\recv{\mathit{dep}}{\quot{DQ_v}}{\cdots}}{v}$, and the send
$\send{\quot{DQ_v}}{\signed{D}{c}}$ provides a message on~$\quot{DQ_v}$.  With
token~$v : v : v : ()$, decomposed as
$(v : ()) \Par (v : v : ())$:
\begin{align*}
  &\signed{
    \recv{\mathit{dep}}{\quot{DQ_v}}{\cdots} \Par \send{\quot{DQ_v}}{\signed{D}{c}}
  }{v}
  \;\Par\; v : ()
  \\
  &\quad\red\quad
  \cdots\bigl\{\quot{\signed{D}{c}}\big/\mathit{dep}\bigr\}
\end{align*}
After this step, $\mathit{dep}$ is bound to
$\quot{\signed{D}{c}}$ and one phlogiston is consumed.
The system becomes:
\begin{align*}
  &\signed{
    \recv{\mathit{tok}}{\quot{A_c}}{
      \signed{\mathit{FeeExtract}}{v}
    }
  }{v}
  \;\Par\; v : v : ()
  \\
  &\Par\; \send{\quot{A_c}}{c : c : c : ()}
\end{align*}

\noindent
(with $\mathit{dep} = \quot{\signed{D}{c}}$ captured in
$\mathit{FeeExtract}$).

\bigskip
\noindent
\textbf{Step~2} --- The second layer consumes one execution
token and looks up the client's tokens (\textbf{Rule~1}):
\begin{align*}
  &\signed{
    \recv{\mathit{tok}}{\quot{A_c}}{\cdots}
    \Par \send{\quot{A_c}}{c : c : c : ()}
  }{v}
  \;\Par\; v : ()
  \\
  &\quad\red\quad
  \signed{\mathit{FeeExtract}}{v}
    \bigl\{\quot{c : c : c : ()}\big/\mathit{tok}\bigr\}
  \;\Par\; v : ()
\end{align*}

\noindent
Now $\mathit{tok} = \quot{c : c : c : ()}$ and one more
phlogiston is consumed.  One phlogiston remains.

\bigskip
\noindent
\textbf{Step~3} --- The innermost layer consumes the last
phlogiston and executes fee extraction
(\textbf{Rule~1}):
\[
  \signed{\mathit{FeeExtract}}{v}
  \;\Par\; v : ()
  \;\;\red\;\;
  \mathit{FeeExtract} \Par ()
\]

\noindent
Substituting the bound values into~$\mathit{FeeExtract}$
\eqref{eq:fee-extract} and applying token decomposition:
\begin{equation}
  \send{\quot{F_v}}{c : ()}
  \;\Par\;
  \signed{D}{c}
  \;\Par\;
  (c : c : ())
  \label{eq:after-fee}
\end{equation}

\noindent
The system now contains the signed deployment~$\signed{D}{c}$
in parallel with two client tokens~$(c : c : ())$, plus a
fee token~$(c : ())$ en route to the validator's fee channel.

\bigskip
\noindent
\textbf{Step~4} --- The client deployment fires
(\textbf{Rule~1}):

Expanding $D = \recv{z}{x}{T_{\mathit{app}}} \Par
\send{x}{U_{\mathit{app}}}$:
\begin{align*}
  &\signed{
    \recv{z}{x}{T_{\mathit{app}}}
    \Par \send{x}{U_{\mathit{app}}}
  }{c}
  \;\Par\; c : c : ()
  \\[4pt]
  &\quad\red\quad
  \subst{T_{\mathit{app}}}
    {\quot{U_{\mathit{app}}}}{z}
  \;\Par\; c : ()
\end{align*}

\noindent
One client token is consumed.  The application continuation
runs with one remaining client token.

\bigskip
\noindent
\textbf{Final state:}
\[
  \send{\quot{F_v}}{c : ()}
  \;\Par\;
  \subst{T_{\mathit{app}}}
    {\quot{U_{\mathit{app}}}}{z}
  \;\Par\;
  (c : ())
\]

\noindent
The fee token is deposited at the validator's fee channel.
The application continuation runs.  One client token remains
for further computation.

\subsection{Source-to-Source Translation}
\label{app:ex-translation}

We now apply the translation of
Appendix~\ref{app:translation} to the system~$\Omega$.
Write $n_v = \Sigma\lsem v \rsem$ and
$n_c = \Sigma\lsem c \rsem$ for the translated signature
channels.

\medskip
\noindent
\textbf{Validator tokens:}
\begin{align*}
  \mathcal{K}\lsem v : v : v : () \rsem
  &= \send{n_v}{\send{n_v}{\send{n_v}{\nil}}}
\end{align*}

\noindent
A triple-nested send on~$n_v$.

\medskip
\noindent
\textbf{Client tokens:}
\begin{align*}
  \mathcal{K}\lsem c : c : c : () \rsem
  &= \send{n_c}{\send{n_c}{\send{n_c}{\nil}}}
\end{align*}

\medskip
\noindent
\textbf{Validator handler} (outermost fuel gate only,
for brevity):
\begin{align*}
  \mathcal{T}\lsem \mathit{VH} \rsem
  \;=\;\;
  &\recv{t_1}{n_v}{\bigl(
    \drop{t_1} \Par
    \recv{\mathit{dep}}{n_v}{\mathcal{T}\lsem\text{inner}\rsem}
  \bigr)}
\end{align*}

\noindent
where $\mathcal{T}\lsem\text{inner}\rsem$ nests two more
fuel gates on~$n_v$ (one per remaining $\signed{\cdots}{v}$
layer), followed by the token lookup on~$\quot{A_c}$, the
fee extraction, and the translated deployment.

\medskip
\noindent
\textbf{Fee extraction in the translated system:}

The critical observation is that fee extraction corresponds to
an \emph{additional fuel gate} on the client's signature
channel~$n_c$.  After the validator handler runs, the
translated system contains (among other terms):
\begin{equation}
  \underbrace{
    \recv{t_{\mathit{fee}}}{n_c}{
      \bigl(
        \send{\quot{F_v}}{\nil}
        \Par \drop{t_{\mathit{fee}}}
      \bigr)
    }
  }_{\text{fee interceptor}}
  \;\Par\;
  \underbrace{
    \send{n_c}{\send{n_c}{\send{n_c}{\nil}}}
  }_{\text{client's 3 tokens}}
  \label{eq:fee-interceptor}
\end{equation}

\noindent
The fee interceptor consumes the outermost send on~$n_c$
(one token).  Its payload is
$\send{n_c}{\send{n_c}{\nil}}$ --- the remaining two
tokens --- which is released via $\drop{t_{\mathit{fee}}}$.
The fee is recorded as a send on~$\quot{F_v}$.

After this reduction:
\[
  \send{\quot{F_v}}{\nil}
  \;\Par\;
  \send{n_c}{\send{n_c}{\nil}}
  \;\Par\;
  \underbrace{
    \recv{t}{n_c}{
      \bigl(\drop{t} \Par
        \mathcal{P}\lsem D \rsem
      \bigr)
    }
  }_{\text{deployment fuel gate}}
\]

\noindent
The deployment's own fuel gate now consumes from the
remaining two-token chain on~$n_c$, exactly as verified in
Appendix~\ref{app:verification}.

\subsection{Reduction Trace in Pure Rho}
\label{app:ex-trace-rho}

We trace the key reductions in the translated system,
suppressing intermediate structural equivalence steps.

\begin{enumerate}[label=\textbf{Step \arabic*.}, leftmargin=3em, start=0]
  \item \textbf{Wallet draw} on~$\quot{W_v}$.  The bootstrap
    for-comprehension receives the phlogiston stack.
    $\drop{\mathit{phlo}}$ releases three sends on~$n_v$
    into the ambient parallel composition.

  \item \textbf{Validator fuel gate~1} fires on~$n_v$.
    The outermost send in the phlogiston chain is
    consumed.  The handler's first $\forc$ unblocks.
    Remainder: two sends on~$n_v$.

  \item \textbf{Deployment received} on~$n_v$.  The
    translated $\send{n_v}{\mathcal{T}\lsem\signed{D}{c}\rsem}$
    communicates with the handler's inner $\forc$ on~$n_v$.
    $\mathit{dep}$ is bound.

  \item \textbf{Validator fuel gate~2} fires on~$n_v$.
    One more phlogiston consumed.

  \item \textbf{Token lookup} on~$\quot{A_c}$.  The
    client's token deposit communicates with the handler's
    $\forc$ on~$\quot{A_c}$.  $\mathit{tok}$ is bound
    to the quoted three-token chain.

  \item \textbf{Validator fuel gate~3} fires on~$n_v$.
    Last phlogiston consumed.  $\mathit{FeeExtract}$
    is now active.

  \item \textbf{Fee interceptor} fires on~$n_c$
    \eqref{eq:fee-interceptor}.  One client token consumed
    as fee.  Two-token chain released.

  \item \textbf{Deployment fuel gate} fires on~$n_c$.
    One client token consumed.  Application continuation
    $\mathcal{P}\lsem T_{\mathit{app}} \rsem$ unblocks.
    One-token chain released.

  \item \textbf{Application communication} on
    $\mathcal{N}\lsem x \rsem$.  The translated
    $\recv{z}{\mathcal{N}\lsem x \rsem}{\cdots}$ and
    $\send{\mathcal{N}\lsem x \rsem}{\cdots}$
    communicate.  The application payload is delivered.
\end{enumerate}

\noindent
\textbf{Final state} (in pure rho):
\[
  \underbrace{
    \send{\quot{F_v}}{\nil}
  }_{\text{fee deposited}}
  \;\Par\;
  \underbrace{
    \mathcal{T}\lsem
      \subst{T_{\mathit{app}}}
        {\quot{U_{\mathit{app}}}}{z}
    \rsem
  }_{\text{application result}}
  \;\Par\;
  \underbrace{
    \send{n_c}{\nil}
  }_{\text{1 client token remaining}}
\]

\subsection{Discussion}
\label{app:ex-discussion}

\paragraph{Stake, phlogiston, and epochs.}
The validator's \emph{stake} --- the quantity of $v$-tokens
locked in the staking contract --- is not consumed directly
by the handler.  Instead, the protocol \emph{mints} a fresh
supply of phlogiston at the wallet channel~$\quot{W_v}$ at
the time of staking and again at each epoch boundary (the
point at which the set of staked validators is updated),
provided the validator has been following the protocol rules.
The handler draws from this wallet.

The phlogiston wallet must supply at least as many
tokens as the handler's nesting depth per deployment.  In the
example, three $\signed{\cdots}{v}$ layers require three
units of phlogiston per deploy.  A persistent validator
($!\,\mathit{VB}$) re-draws from the wallet after each
deployment, consuming phlogiston at a rate proportional
to the throughput of client deploys.

\paragraph{Slashing.}
In the current F1R3FLY slashing algorithm, any slashing
offense results in the removal of the validator's entire
stake.  In the cost-accounted model, slashing has two
effects: (i)~all remaining phlogiston is removed and no
further phlogiston is minted, and (ii)~the validator's stake
is moved to a private unforgeable channel pending
adjudication.

Because the handler draws phlogiston from
$\quot{W_v}$, removing the wallet's contents
immediately halts the validator: the bootstrap
for-comprehension $\recv{\mathit{phlo}}{\quot{W_v}}{\cdots}$
finds no message and blocks.  No further deployments can be
processed.  The validator is effectively offline.

A slashed validator may be \emph{redeemed} --- proven to
have been in the right, or having reached an arrangement
with the shard.  Upon redemption, phlogiston minting resumes
at the next epoch boundary.  The stake held at the private
channel may be:
\begin{itemize}[nosep]
  \item \textbf{Returned} to the validator (if the
    offense was spurious or the result of a network fault).
  \item \textbf{Partially redistributed} to affected parties
    (as a penalty proportional to the severity of the
    offense).
  \item \textbf{Burned} (removed from circulation, reducing
    the total token supply).
\end{itemize}
The adjudication contract is itself a Rholang program,
expressible in the cost-accounted calculus, and subject to
the same formal verification tools.

\paragraph{Fee conversion.}
The fee token $(c : ())$ is a \emph{client-signature}
token.  For the validator to use it as fuel for its own
operations (which require $v$-signature tokens), the fee
must be \emph{converted}.  This conversion is a
market-making operation, expressible as a Rholang contract:
\[
  \mathrm{Exchange}(c, v)
  \;=\;
  \recv{t_c}{n_c}{
    \recv{t_v}{n_v}{
      \bigl(
        \send{n_c}{\drop{t_v}}
        \Par
        \send{n_v}{\drop{t_c}}
      \bigr)
    }
  }
\]
This persistent exchange contract swaps one $c$-token for
one $v$-token (and vice versa), implementing a 1:1 peg.
More sophisticated exchange contracts can implement
variable rates, order books, or automated market makers ---
all within Rholang.

\paragraph{Persistent validator.}
The example bootstrap processes a single deployment and
halts.  A production validator runs persistently,
re-drawing phlogiston from the wallet after each
deployment.  Using the standard reflective replication
encoding:
\[
  !\, \mathit{VB}
\]
Each iteration draws phlogiston from~$\quot{W_v}$,
processes a deployment, and collects fees.  Via the exchange
contract, fees can be converted to replenish the
phlogiston supply between epoch-boundary minting events,
establishing the economic feedback loop that sustains the
validator.

\paragraph{Connection to Casper CBC.}
In the F1R3FLY platform, the validator's stake
($v$-tokens locked in the staking contract) determines
the validator's weight in Casper CBC consensus.  The
cost-accounted calculus makes this relationship
\emph{intrinsic}: the validator literally cannot act
without phlogiston, whose minting is contingent on
continued good behavior.  The rate of phlogiston
consumption is determined by the structure of the
handler --- the same structure that the consensus layer
inspects.  Slashing, epoch-boundary minting, redemption,
and stake-weighted voting are all expressible as
cost-accounted Rholang contracts, unifying the consensus
and execution layers under a single formal semantics.

%% ══════════════════════════════════════════════════════════════════
\section{Concrete Syntax for Rholang 1.2}
\label{app:concrete}

The abstract syntax used throughout this paper --- in particular
the wrapper $\signed{P}{s}$ and the compound-signature
operator~$\comp$ --- is convenient for mathematical exposition
but cannot be used as concrete syntax: the brace-subscript form
$\{\cdot\}_s$ has no linear ASCII rendering, and~$\comp$
collides with the heavily overloaded \texttt{*}
(dereference~\texttt{PEval}, multiplication~\texttt{PMult}) of
Rholang.  This appendix specifies a concrete syntax that
extends the Rholang~1.2 BNFC grammar, chosen to avoid all such
collisions.

\subsection{The Wrapper and Signature Notation}
\label{app:concrete-wrapper}

The signed-term wrapper $\signed{P}{s}$ is written
\begin{center}
  \verb|{% P %}[s]|
\end{center}
The \verb|{%|\,\ldots\,\verb|%}| delimiters are unambiguous
against the existing block-brace \verb|{|\,\ldots\,\verb|}|
(\texttt{Proc16}) and map-collection
\verb|{|\,\ldots\,\verb|}| (\texttt{CollectMap}), and the
signature rides in a trailing \verb|[s]| bracket.  Within that
bracket:

\begin{center}
  \begin{tabular}{lll}
    \textbf{Abstract} & \textbf{Concrete} & \textbf{Reading} \\
    \midrule
    $g$        & \verb|g|          & ground signature (a name) \\
    $\hashf P$ & \verb|# P|        & cryptographic section \\
    $s_1 \comp s_2$ & \verb|s1 (*) s2| & compound signature \\
    $s_1 \multimap s_2$ & \verb|s1 -o s2| & linear transfer \\
  \end{tabular}
\end{center}

\noindent
The bracketed operator \verb|(*)| renders~$\comp$ without
colliding with \texttt{*}: the parentheses make it a distinct
lexeme available only in signature position.  The
lollipop~$\multimap$ is the ASCII \verb|-o|.  The
section~$\hashf$ uses \verb|#|, which is absent from the base
grammar.

\subsection{BNFC Productions}
\label{app:concrete-bnfc}

The following productions extend the Rholang~1.2 grammar.  They
introduce a new syntactic category \texttt{SignedProc} for
wrapped terms, polymorphic forms of send, receive, and parallel
composition that range over signed terms, a \texttt{Sig}
category for signatures, and a \texttt{Stk} category for token
stacks.

\begin{lstlisting}[basicstyle=\small\ttfamily,frame=single,framesep=6pt,xleftmargin=1em,xrightmargin=1em]
-- Signed processes (the wrapper {P}_s).
-- A new syntactic category sitting alongside Proc.
SignedProcWrap.   SignedProc ::= "{%" Proc "%}" "[" Sig "]" ;
SignedProcStk.    SignedProc ::= Stk ;
SignedProcPar.    SignedProc ::= SignedProc "|" SignedProc1 ;
_.                SignedProc1 ::= SignedProc2 ;
_.                SignedProc2 ::= "(" SignedProc ")" ;

-- Signatures (the sort Sig).
SigGround.        Sig2 ::= Name ;             -- ground signature g
SigHash.          Sig2 ::= "#" Proc12 ;       -- section  # P
SigCompound.      Sig1 ::= Sig1 "(*)" Sig2 ;  -- compound s1 (*) s2
SigTransfer.      Sig  ::= Sig1 "-o" Sig ;    -- transfer s1 -o s2
_.                Sig  ::= Sig1 ;
_.                Sig1 ::= Sig2 ;
_.                Sig2 ::= "(" Sig ")" ;

-- Token stacks (the sort Stk).
StkEmpty.         Stk2 ::= "(" ")" ;          -- empty stack ()
StkCons.          Stk1 ::= Sig "::" Stk1 ;    -- cons   s :: S
StkPar.           Stk  ::= Stk "|" Stk1 ;     -- located/parallel stacks
_.                Stk  ::= Stk1 ;
_.                Stk1 ::= Stk2 ;
_.                Stk2 ::= "(" Stk ")" ;

-- Polymorphic send/receive over signed terms.
-- The payload and continuation range over SignedProc.
PSignedSend.      SignedProc2 ::= Name Send "(" SignedProc ")" ;
PSignedInput.     SignedProc2 ::= "for" "(" [SignedReceipt] ")"
                                  "{%" SignedProc "%}" ;

-- Per-clause signed binds (Axis C of the join schema).
SignedBindImpl.   SignedBind ::= "{%" [Name] NameRemainder
                                 "<-" NameSource "%}" "[" Sig "]" ;
PlainBindImpl.    SignedBind ::= [Name] NameRemainder
                                 "<-" NameSource ;
separator nonempty SignedBind "&" ;

SignedReceiptImpl. SignedReceipt ::= [SignedBind] ;
separator nonempty SignedReceipt ";" ;
\end{lstlisting}

\noindent
Three points on the design:

\begin{enumerate}[nosep]
  \item \textbf{Precedence of signatures.}  \texttt{Sig} is
    stratified so that \verb|-o| binds loosest and \verb|(*)|
    tighter, matching the abstract reading
    $s_1 \comp s_2 \multimap s_3 = (s_1 \comp s_2) \multimap s_3$
    and the curry iso of Remark~\ref{rem:tensor-hom}.  Ground
    signatures and sections sit at the tightest level
    \texttt{Sig2}.
  \item \textbf{Token stacks use \texttt{::}.}  The cons
    operator is the double colon \verb|::|, distinct from the
    single \verb|:| of map key--value pairs
    (\texttt{KeyValuePairImpl}).  The \texttt{StkPar}
    production accommodates located stacks
    (Section~\ref{sec:ocap-authority}): several stacks parked
    at different surfaces, composed with~\verb"|".
  \item \textbf{Polymorphic \texttt{|}.}  Parallel composition
    appears at three levels --- \texttt{PPar} (processes),
    \texttt{SignedProcPar} (signed terms), and \texttt{StkPar}
    (stacks) --- realising the parametric polymorphism
    of~$K = \Par$ (Remark~2.2 of~\cite{CostEndofunctor}: $K$ is
    the only constructor that needs overloading).
\end{enumerate}

\subsection{Worked Examples in Concrete Syntax}
\label{app:concrete-examples}

The uniform-signing sugar (Definition~\ref{def:sugar-uniform})
$\signed{\recv{y}{x}{P}}{s}$, whose desugaring nests
recursively, is written:
\begin{lstlisting}[style=rholang]
{% for( y <- x ) P %}[ s ]
\end{lstlisting}

\noindent
The linear-transfer form
$\signed{\recv{y}{x}{P}}{s_1 \multimap s_2}$ of
Example~\ref{ex:cross-party} (client~$c$ funds the call,
provider~$p$ funds the continuation):
\begin{lstlisting}[style=rholang]
{% for( result <- service ) handleResult %}[ c -o p ]
\end{lstlisting}

\noindent
The multi-signature transfer contract
$\signed{\mathit{transfer}}{s_{\mathit{sender}} \comp
s_{\mathit{compliance}}}$ of
Section~\ref{sec:multisig-motivation}:
\begin{lstlisting}[style=rholang]
{% transfer %}[ sender (*) compliance ]
\end{lstlisting}

\noindent
A per-clause-signed join (Axis~C,
Section~\ref{sec:joins}) in which each rendezvous surface is
funded independently:
\begin{lstlisting}[style=rholang]
{% for( {% y1 <- x1 %}[s1] & {% y2 <- x2 %}[s2] ) P %}[ s0 ]
\end{lstlisting}

\noindent
A token stack of three units keyed to signature~\verb|c|,
located on the client's address (a single stack at one
surface):
\begin{lstlisting}[style=rholang]
c :: c :: c :: ()
\end{lstlisting}

\noindent
The validator's deploy-queue handler of
Appendix~\ref{app:example}
$\signed{\recv{\mathit{dep}}{\quot{DQ_v}}{
\recv{\mathit{tok}}{\quot{A_c}}{\mathit{FeeExtract}}}}{v}$:
\begin{lstlisting}[style=rholang]
{% for( dep <- DQv ) for( tok <- Ac ) FeeExtract %}[ v ]
\end{lstlisting}

\subsection{Abstract-to-Concrete Correspondence}
\label{app:concrete-table}

\begin{center}
  \begin{tabular}{lll}
    \textbf{Construct} & \textbf{Abstract} & \textbf{Concrete} \\
    \midrule
    Signed process & $\signed{P}{s}$
      & \verb|{% P %}[s]| \\
    Ground signature & $g$
      & \verb|g| \\
    Section & $\hashf P$
      & \verb|# P| \\
    Compound signature & $s_1 \comp s_2$
      & \verb|s1 (*) s2| \\
    Linear transfer & $s_1 \multimap s_2$
      & \verb|s1 -o s2| \\
    Empty stack & $()$
      & \verb|()| \\
    Token cons & $s : S$
      & \verb|s :: S| \\
    Located/parallel stacks & $S_1 \Par S_2$
      & \verb#S1 | S2# \\
    Signed send & $\send{x}{U}$
      & \verb|x!( U )| \\
    Signed receive & $\recv{y}{x}{T}$
      & \verb|for( y <- x ) T| \\
    Per-clause bind & $\signed{y \leftarrow x}{s}$
      & \verb|{% y <- x %}[s]| \\
  \end{tabular}
\end{center}

\noindent
The mapping is purely notational: a front-end parser reads the
concrete forms into the abstract syntax of
Section~\ref{sec:calculus}, after which the source-to-source
translation of Appendix~\ref{app:translation} compiles to pure
Rholang~1.1.  Because \verb|?!| and \verb|!?| are already part
of the Rholang~1.2 \texttt{NameSource} and \texttt{PSendSynch}
productions, the synchronous-call patterns of
Section~\ref{sec:rholang-illustration} require no new syntax;
they appear inside \texttt{NameSource} positions of signed
binds unchanged.

%% ══════════════════════════════════════════════════════════════════
\section*{Acknowledgements}

The author thanks Mike Stay for extensive discussions on the
connections between linear logic and resource accounting,
Dylon Edwards for detailed feedback on the operational
semantics and implementation path, and Kevin Machiels for
insightful comments on the validator economics and slashing
model.

%% ══════════════════════════════════════════════════════════════════
\begin{thebibliography}{9}

\bibitem{MeredithRadestock2005}
  L.~G.~Meredith and M.~Radestock,
  ``A reflective higher-order calculus,''
  \textit{Electronic Notes in Theoretical Computer Science},
  vol.~141, no.~5, pp.~49--67, 2005.

\bibitem{Milner1999}
  R.~Milner,
  \textit{Communicating and Mobile Systems: the $\pi$-Calculus},
  Cambridge University Press, 1999.

\bibitem{RholangSpec}
  L.~G.~Meredith \textit{et al.},
  ``Rholang Specification,''
  F1R3FLY.io / RChain Cooperative, 2017--2026.

\bibitem{MeTTaIL}
  L.~G.~Meredith,
  ``MeTTaIL (Rholang 1.4): A language for defining languages,''
  F1R3FLY.io, 2025--2026.

\bibitem{CostEndofunctor}
  L.~G.~Meredith,
  ``Continued interactive GSLTs and the cost endofunctor:
  A construction one level up from cost-accounted Rholang,''
  F1R3FLY.io, 2026.

\bibitem{MillerYeeShapiro2003}
  M.~S.~Miller, K.-P.~Yee, and J.~Shapiro,
  ``Capability myths demolished,''
  Technical Report SRL2003-02, Systems Research Laboratory,
  Johns Hopkins University, 2003.

\bibitem{MillerPhD2006}
  M.~S.~Miller,
  \textit{Robust Composition: Towards a Unified Approach to
  Access Control and Concurrency Control},
  PhD thesis, Johns Hopkins University, 2006.

\bibitem{VACRADA}
  L.~G.~Meredith \textit{et al.},
  ``Secure Blockchain-Based Health Data Platform: Cooperative
  Research and Development Agreement,''
  F1R3FLY Industries / U.S.\ Department of Veterans Affairs,
  2025--2026.

\end{thebibliography}

\end{document}

