\documentclass[11pt]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools,stmaryrd}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{enumitem}
\usepackage[expansion=false]{microtype}
\usepackage[hidelinks]{hyperref}
\hypersetup{pdftitle={Splitting the Mind of Zeus},pdfauthor={L. G. Meredith}}

% ---- theorem environments ----
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{example}[definition]{Example}
\newtheorem{remark}[definition]{Remark}
\newtheorem{construction}[definition]{Construction}

\theoremstyle{plain}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{principle}[definition]{Principle}
\newtheorem{observation}[definition]{Observation}

% ---- notation ----
\newcommand{\p}{\partial}
\newcommand{\T}{T}
\newcommand{\Coll}{S}
\newcommand{\List}{\mathsf{List}}
\newcommand{\Mset}{\mathsf{Bag}}
\newcommand{\Pfin}{\mathsf{P}_{\mathrm{fin}}}
\newcommand{\Dist}{\mathsf{D}}
\newcommand{\plug}{\mathsf{plug}}
\newcommand{\Split}{\mathsf{Split}}
\newcommand{\Pack}{\mathsf{PACK}}
\newcommand{\pack}{\mathsf{pack}}
\newcommand{\Proto}{\Pi}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\real}{\Vdash}
\newcommand{\id}{\mathrm{id}}
\newcommand{\catC}{\mathcal{C}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\op}[1]{\widehat{#1}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\qmap}{q}
\newcommand{\Pos}{\mathsf{Pos}}

\title{\bfseries Splitting the Mind of Zeus\\[4pt]
\large How crucial aspects of logic arise, like Pallas Athena,\\
from splitting the focus of attention from its context\\[6pt]
\normalsize\normalfont\itshape Realizability, collection monads, and the choice
content of the McBride derivative in a finite-limits metalanguage}
\author{L.\,G.\ Meredith\\\small F1R3FLY.io}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
We set out to make visible the seams along which a logic is assembled. The
commitment that makes the seams visible is a realizability semantics:
a formula denotes the collection of computations that witness its
computational content. Under that commitment a logic factors into three
components --- a \emph{collection} component (how witnesses are aggregated),
a \emph{structural} component (term constructors lifted to predicates on
witnesses), and a \emph{behavioral} component (contexts labeling
transitions lifted to modalities). The collection component sits awkwardly
with categorical logic, because the subobject classifier bakes in
\emph{set}-shaped collection: there is no list of all lists. We reconcile the
two by going syntactic. Treating every collection as a term type, we use the
McBride derivative to form the object of all splittings $(\p\T,\T)$ --- the
\emph{proto power object} --- and then ask what must be added to the ambient
metalanguage to \emph{package} the splittings back into the collection type.
Committing the metalanguage to finite limits (cartesian logic) is what keeps
the whole apparatus syntactic: there is no subobject classifier to lean on,
so the derivative route is the only route; and choice cannot be asserted as a
proposition, so it can enter only as a named section operation. The central
claim is that packaging is exactly a section of the quotient presenting the
collection monad, that its choice content equals the assertion that the atom
object admits a linear order, and that going syntactic discharges this because
term positions are already ordered. The construction is kept parametric in the
order on atoms, isolating the choice residue at the atom boundary --- which is
where later reflective (knot-tying) constructions will act. Throughout we note
that the equations one quotients by --- commutativity and idempotence --- are
exactly the equations the no-go theorems for distributive laws turn on, so the
distributive-law obstruction and the packaging cost are one datum read twice.
\end{abstract}

\tableofcontents

\section{Why realizability, and why this matters for the argument to the
AI community}
\label{sec:motivation}

Most accounts take a logic as given: a fixed vocabulary of connectives, a
notion of model, and a truth predicate that floats above computation. Truth
values arrive from a Boolean or Heyting algebra that is postulated, not built.
For a discipline that wants its reasoning systems to be inspectable ---
traceable to constructions whose behavior can be run, audited, and held
accountable --- this is exactly the wrong starting point. The glue that binds
truth to the algebra is opaque, and once it is in place one can no longer see
what the logic is \emph{made of}.

The stakes here are specific to artificial intelligence. AI advances a value
proposition: that important and useful aspects of (human) intelligence can be
faithfully rendered as computation. Logic has long been cast as the calculus
of reason --- the very paradigm of the rational faculty. To exhibit logic
itself as \emph{arising from} computation is therefore direct support for that
proposition, since the calculus of reason is thereby shown to be, at bottom,
computational. But the philosophical vindication is the smaller part. The
larger part is that a computational account of logic yields \emph{effective
procedures}: procedures for \emph{generating} logics and for
\emph{generalizing} logic-based algorithms. If the connectives are operations
of three named kinds (\S\ref{sec:three}), then a new logic is obtained by
turning the dials, and an algorithm defined over one logic can be transported,
mechanically, to another by re-instantiating the same three components. The
decomposition is not a description of logic arrived at after the fact; it is a
construction kit. That generative, effective character --- not the slogan that
reason is computation --- is what this research programme offers in support of
AI's value proposition.

We want the opposite. We want a logic whose every connective is an operation on
computations, so that the entire edifice is a construction and nothing is
postulated except the constructions themselves. The commitment that delivers
this is realizability. Rather than asking whether a formula is true, we ask
which computations \emph{witness} it, and we let the formula denote the
collection of its witnesses:
\[
  \sem{\varphi} \;=\; \{\, e : e \real \varphi \,\}.
\]
This is the Kleene move \cite{Kleene45,vanOosten08}, but with the aggregating
container promoted from an incidental detail to a \emph{parameter} of the
construction. Once witnesses are computations and formulae are collections of
them, every connective is forced to be an operation of one of exactly three
kinds, corresponding to the three things one can vary: how witnesses are
collected, how they are built, and how they move. This is the decomposition
we develop in \S\ref{sec:three}.

There is a second, sharper commitment, and it is the one we most want to
defend to the AI community. We insist that the ambient metalanguage --- the
language in which the collections, the constructions, and the packaging all
live --- be a \emph{finite-limits} (cartesian, essentially-algebraic) theory
\cite{Freyd72,GabrielUlmer71,PalmgrenVickers07}. The reason is not
minimalism for its own sake. It is that a finite-limits metalanguage is the
maximal commitment to ``everything is a construction'':
\begin{itemize}[leftmargin=1.4em]
\item there are no truth values floating in from a Boolean algebra, because
      there is in general no subobject classifier (a lex category need not be
      a topos);
\item there is no non-constructive existential, because the only existentials
      one may assert are provably \emph{unique} ones, carved out as partial
      operations on equationally-defined subdomains;
\item and --- the point that will do the real work in
      \S\ref{sec:finlim} and \S\ref{sec:packaging} --- choice cannot be
      \emph{stated} as a proposition at all. There is no $\forall$, and $\exists$
      is available only when the witness is unique. So the only way to
      ``add choice'' is to add a \emph{function symbol} to the signature,
      together with equations making it a section. In a finite-limits
      metalanguage, choice is not a truth one assumes; it is an operation one
      names.
\end{itemize}
This is exactly the inspectability we are after. When choice enters, it enters
as a symbol you can point at, with a documented arity and documented
equations, and you can localize precisely where in the construction it was
spent. Nothing is hidden in a semantic ether. The seams are visible because
we have refused the glue --- the subobject classifier --- that ordinarily
hides them.

The remainder of the note carries this out. \S\ref{sec:three} gives the
three-component decomposition. \S\ref{sec:dial} shows that the collection
component is a substructural dial whose settings are governed by two
equations, commutativity and idempotence. \S\ref{sec:disconnect} states the
disconnect with categorical logic, and \S\ref{sec:nogo} shows that the same
two equations are what the no-go theorems for distributive laws turn on ---
so the obstruction to a uniform collection component and the cost of packaging
are, as we will see, one datum read twice. \S\ref{sec:finlim} cashes out the
finite-limits commitment. \S\ref{sec:derivative} builds the proto power
object from the McBride derivative, and \S\ref{sec:packaging} defines
packaging as a section and states the choice conjecture, parametric in the
order on atoms. \S\ref{sec:grading} places the conjecture in the Weihrauch
lattice, \S\ref{sec:instances} works the three instances, and
\S\ref{sec:discussion} draws out the consequences, including the absence of
any classicality collapse and the reflective use of the atom-order parameter.

\section{Three components of a logic under realizability}
\label{sec:three}

Fix a representation of computation. Throughout we take it to be a term
rewrite system: terms are computational states, and rewrites are steps. Write
$\T$ for the type of terms. Fix also a way of aggregating witnesses: a
\emph{collection type} $\Coll$, which will always carry the structure of a
monad (\S\ref{sec:dial}). A formula denotes an object $\sem{\varphi} :
\Coll(\T)$ --- a collection of terms.

Under these two fixings, three families of connectives arise, and every
connective is of one of the three kinds.

\paragraph{Collection connectives.}
Any operation on $\Coll$ lifts to a connective. If witnesses are collected in
sets, the semilattice operations of $\Coll = \Pfin$ give the familiar
connectives:
\[
  \sem{\varphi \wedge \psi} = \sem{\varphi} \cap \sem{\psi},
  \qquad
  \sem{\varphi \vee \psi} = \sem{\varphi} \cup \sem{\psi}.
\]
This is where the choice of container is felt: change $\Coll$ from sets to
multisets, lists, trees, or distributions, and these connectives change with
it. The collection component is the parameter we most want to vary, and
\S\ref{sec:dial} explains what varying it does.

\paragraph{Structural connectives.}
Each term constructor of $\T$ lifts to a predicate transformer. If
$f : \T^{n} \to \T$ is an $n$-ary constructor, define
\[
  \sem{\op{f}(\varphi_{1},\dots,\varphi_{n})}
  \;=\;
  \{\, f(e_{1},\dots,e_{n}) : e_{i} \real \varphi_{i} \,\}.
\]
The lifted constructor $\op{f}$ is a \emph{structural} predicate: it speaks
about the \emph{shape} of a witness, asserting that the witnessing computation
was built by $f$ from sub-witnesses of the $\varphi_{i}$. In the reflective
setting these are the constructors of the rho calculus lifted to namespace
predicates \cite{MeredithRadestock05a,MeredithRadestock05b}; in general they
are the introduction half of an Operational-Semantics-in-Logical-Form (OSLF)
presentation.

\paragraph{Behavioral connectives.}
Each rewrite is labeled by the context in which it fires. A step
$C[l] \to C[r]$ carries the one-hole context $C$, an element of the
\emph{derivative} $\p\T$ (\S\ref{sec:derivative}). Lifting the labels to
modalities gives
\[
  \sem{\langle C\rangle \varphi}
  = \{\, e : e \xrightarrow{\,C\,} e',\ e' \real \varphi \,\},
  \qquad
  \sem{[C]\varphi}
  = \{\, e : \forall e'.\ e \xrightarrow{\,C\,} e' \Rightarrow e' \real \varphi \,\},
\]
the context-indexed Hennessy--Milner modalities. This is the elimination half
of OSLF.

\begin{remark}
The structural and behavioral components are the two halves of OSLF, generated
respectively from constructors and from labeled transitions. They are already
well-understood in the reflective setting. The component that resists the
categorical-logic packaging, and that this note is about, is the first one ---
collection --- together with the derivative machinery ($\p\T$) that turns out
to glue it to the third.
\end{remark}

\section{The collection component is a substructural dial}
\label{sec:dial}

The collection type $\Coll$ is a monad, and the algebraic theory presenting
$\Coll$ is not incidental: \emph{its equations are the structural rules of the
propositional logic the collection component induces.} This is the first fact
that makes the decomposition worth having, because it says the collection
component is not ``logic plus a data-structure flavor''; it \emph{is} the
substructural hierarchy, indexed by how far $\Coll$ sits from set-shaped.

Write the binary aggregation of $\Coll$ as $\ast$ with unit $\varepsilon$.
The correspondence is:
\[
\begin{array}{lll}
  \text{commutativity of } \ast & \longleftrightarrow & \text{exchange},\\[2pt]
  \text{idempotence of } \ast    & \longleftrightarrow & \text{contraction},\\[2pt]
  \text{unit/absorption laws}    & \longleftrightarrow & \text{weakening}.
\end{array}
\]
Reading the standard collection monads through this correspondence:
\begin{itemize}[leftmargin=1.4em]
\item $\Pfin$, the finite-powerset monad --- a commutative, idempotent monoid
      (a join-semilattice) --- gives exchange and contraction (and, with the
      unit, weakening): the full structural regime, hence Heyting/Boolean
      propositional logic.
\item $\Mset$, the finite-multiset monad --- the free commutative monoid,
      commutative but not idempotent --- gives exchange without contraction:
      the multiplicative-linear regime with $\otimes$.
\item $\List$, the list monad --- the free monoid, neither commutative nor
      idempotent --- gives neither exchange nor contraction: a
      non-commutative, Lambek-style regime.
\item $\Dist$, the distribution monad --- commutative with a convex (barycentric)
      structure in place of idempotence --- gives the probabilistic regime.
\end{itemize}

\begin{principle}[Collection is substructure]
\label{prin:dial}
The setting of the collection dial is the pair of Booleans
$(\text{commutative?}, \text{idempotent?})$ of $\Coll$, and this pair is
exactly the presence or absence of exchange and contraction in the induced
logic.
\end{principle}

Keep the two coordinates of Principle~\ref{prin:dial} in view. They are the
same two coordinates on which the no-go theorems of \S\ref{sec:nogo} turn,
and the same two equations one quotients by when packaging in
\S\ref{sec:packaging}. Everything in this note is, at bottom, bookkeeping on
commutativity and idempotence.

\section{The disconnect with categorical logic}
\label{sec:disconnect}

Categorical logic packages the collection component through the subobject
classifier $\Omega$: the power object of $A$ is $\Omega^{A}$, the object of
subobjects of $A$. This is a beautiful and total solution --- for one setting
of the dial. It hard-wires $\Coll = $ powerset. Two things go wrong when we
try to honor Principle~\ref{prin:dial} and turn the dial.

First, the classifier fixes the collection shape. There is no
$\Omega$-parametrized-by-$\Coll$; the power object is subsets, and only
subsets. One cannot, inside the $\Omega$ story, ask for the multiset or list
of witnesses.

Second, and more tellingly, even the operation of forming ``the collection of
all subcollections \emph{as an object of the same shape}'' fails away from
sets. The slogan is: \emph{there is no list of all lists.} To assemble all
sublists of a given list into a single \emph{list}, one must linearly order
the sublists; there is no canonical order on them, and choosing one is choice.
For sets the analogous assembly is free of order --- the subsets form a set
with nothing to arrange --- which is exactly why the $\Omega$ story works for
the powerset setting and only there.

\begin{remark}
The disconnect is not a defect of categorical logic to be repaired by a
cleverer classifier. It is a signal that the packaging of the collection
component into an object of the same shape is \emph{order-laden}, and that the
$\Omega$ story succeeds precisely by living at the one setting of the dial
(powerset) where the order requirement is vacuous. The reconciliation in
\S\ref{sec:derivative}--\S\ref{sec:packaging} takes the order requirement
seriously and localizes it, rather than wishing it away.
\end{remark}

\section{The same two equations: distributive laws and their no-go theorems}
\label{sec:nogo}

Before the reconciliation, we want to record why the collection dial cannot be
turned freely if one insists on the standard monadic route to the structural
connectives, because it is the same two equations again.

The structural connectives of \S\ref{sec:three}, viewed monadically, ask the
collection monad $\Coll$ to commute past the term monad $\T$. Lifting a
constructor $f : \T^{n} \to \T$ to collections requires a coherent map
$(\Coll\,A)^{n} \to \Coll(A^{n}) \to \Coll\,A$, and for this to be coherent
across \emph{all} the constructors of $\T$ simultaneously one wants a
distributive law
\[
  d : \Coll \circ \T \;\Rightarrow\; \T \circ \Coll,
\]
in the sense of Beck \cite{Beck69}. A distributive law is exactly what lets
$\Coll$-collections be pushed through $\T$-structure functorially, yielding
the lifted constructors.

The trouble is that such laws are scarce, and the scarcity is governed by
commutativity and idempotence. The covariant powerset monad does not
distribute over itself \cite{KlinSalamanca18}; the distribution monad does not
distribute over powerset \cite{VaraccaWinskel06}; and the general no-go
theorems of Zwart and Marsden \cite{ZwartMarsden19} show that once the two
theories being combined carry equational shapes of the wrong kind --- an
idempotence or an absorption clashing with the other theory --- \emph{no}
distributive law can exist. The upshot is that, along the distributive-law
route, the only collection monads that commute nicely past term-like $\T$ are
the set/multiset/distribution-shaped ones: exactly the ones near the
top-and-commutative corner of the dial.

\begin{principle}[One datum, read twice]
\label{prin:twice}
The equations that obstruct the distributive law $d : \Coll\T \Rightarrow
\T\Coll$ --- commutativity and idempotence --- are the same equations that,
in \S\ref{sec:packaging}, one must quotient $\List$ by to present $\Coll$, and
hence the same equations whose section must be postulated to package. The
no-go obstruction and the packaging cost are one datum read twice: once as a
law that fails, once as an operation that must be added.
\end{principle}

The reconciliation of \S\ref{sec:derivative}--\S\ref{sec:packaging} is, in
effect, a third way around the no-go theorems. The two known escapes are to
\emph{quotient} (convex powerset, indexed valuations --- adjust $\Coll$ until
the equations agree) or to \emph{restrict} (finite, ordered --- shrink until
the clash cannot arise). The route here is to \emph{add a global structuring
operation} --- the section --- and pay for it in a named, localized axiom.
\S\ref{sec:discussion} argues these are dual dodges.

\section{The finite-limits metalanguage does exactly what we need}
\label{sec:finlim}

Now the metalanguage commitment earns its keep. Take the ambient metalanguage
to be a finite-limits theory: the cartesian fragment, with $\top$, $\wedge$,
$=$, and existentials only where the witness is provably unique
\cite{Freyd72,GabrielUlmer71,PalmgrenVickers07}. Absent are $\vee$, $\bot$,
$\Rightarrow$, $\neg$, unrestricted $\exists$ and $\forall$, and --- in
general --- a subobject classifier. Three consequences, each of which turns a
would-be obstruction into a feature.

\paragraph{(1) There is no $\Omega$ to lean on, so the syntactic route is the
only route.} In a topos one might hope to repair \S\ref{sec:disconnect} with a
cleverer use of the classifier. In a finite-limits metalanguage that hope is
simply unavailable: a lex category need not be a topos, and the power object
is not on the table. The McBride-derivative construction of
\S\ref{sec:derivative} is therefore not \emph{an} alternative to the
$\Omega$-story; it is the \emph{only} story. The commitment vindicates going
syntactic instead of competing with it.

\paragraph{(2) Choice cannot be a proposition, only an operation.} In
cartesian logic one cannot write the choice principle in $\forall\exists$
form: there is no $\forall$, and $\exists$ may be asserted only under
uniqueness. So ``adding choice'' cannot mean asserting a sentence. It can only
mean enlarging the signature with a function symbol and equations that make it
a section of some quotient. This is precisely the shape that packaging will
take in \S\ref{sec:packaging}: $\pack$ is a new operation, and the equation
$\qmap \circ \pack = \id$ is choice-as-structure. The metalanguage forces the
conjecture into its most honest and most inspectable form.

\paragraph{(3) No classicality collapse.} In a topos, internal choice forces
Boolean logic (Diaconescu \cite{Diaconescu75,MacLaneMoerdijk92}). That
argument needs $\vee$, comprehension via the power object, and effective
quotients to run --- \emph{none} of which are present in the cartesian
fragment. Consequently one may add packaging/choice as pure structure without
collapsing the object logic to classical. Choosing finite limits is exactly
what keeps choice \emph{structural} rather than \emph{logical}. This is the
technical heart of the inspectability argument of \S\ref{sec:motivation}: the
choice we add is a nameable operation and it does not silently classicalize
everything downstream.

\section{The McBride derivative and the proto power object}
\label{sec:derivative}

We now build the object that replaces the power object. The move is to treat
every collection \emph{syntactically} --- as a term type --- and to use the
McBride derivative to enumerate its splittings.

The choice of the McBride derivative is not opportunistic. Defining or
understanding an object by the totality of its splittings is an intuition that
recurs across mathematics, and the McBride derivative is its syntactic,
symbolic form. A Dedekind cut defines a real number as a splitting of the
rationals into a lower and an upper part; the number \emph{is} the totality of
its cuts, an object pinned down entirely by how it comes apart. The covariant
derivative in geometry understands a field by splitting its variation along
directions, comparing a value against its infinitesimally displaced neighbours;
again the object is grasped through its controlled decompositions. In each case
an analytic or order-theoretic or differential intuition does the same work:
know the whole by the family of its splittings. The McBride derivative
\cite{McBride01,AAGM05} performs precisely this move for datatypes --- it
splits a term into a one-hole context and a focus, and the type of one-hole
contexts, the derivative, collects all such splittings --- but it performs it
on \emph{syntax}, so the splittings are enumerable, computable, and
inspectable. That is exactly what the finite-limits commitment of
\S\ref{sec:finlim} demands. We adopt the derivative not by loose analogy to
cuts and connections but because it is the constructive, symbolic rendering of
a move that mathematics has repeatedly found indispensable, and rendering that
move syntactically is the whole point.

\subsection{Terms, derivatives, and the zipper}

Let $F$ be a polynomial (container) endofunctor and let $\T = \mu X.\,F X$ be
its initial algebra, the type of terms. Write $\p F$ for the derivative of
$F$: the functor of one-hole $F$-contexts \cite{McBride01,AAGM05,Huet97}. The
derivative of the fixpoint is a list of layers up the spine:
\[
  \p\T \;\cong\; \bigl(\p F[\T]\bigr)^{\!\ast}
       \;=\; \List\!\bigl(\p F[\T]\bigr),
\]
each layer an $F$-node with one hole and its other children filled by
subtrees. Plugging fills the hole:
\[
  \plug : \p\T \times \T \longrightarrow \T,
  \qquad
  \plug(\mathsf{nil}, t) = t, \quad
  \plug(\ell :: c, t) = \plug\bigl(c, \mathsf{fill}(\ell, t)\bigr).
\]

\begin{definition}[Splittings]
The object of \emph{splittings} of $\T$ is
\[
  \Split(\T) \;=\; \{\, (c,t) \in \p\T \times \T \,\}
             \;\cong\; \textstyle\sum_{c:\p\T} \T,
\]
the pointed terms, with the whole recovered by $\plug(c,t)$. A splitting is a
context together with a focus.
\end{definition}

Two facts about splittings are load-bearing.

\begin{proposition}[Positions are ordered by syntax]
\label{prop:dewey}
The positions of a term --- equivalently, the contexts $c$ appearing in its
splittings --- carry a canonical linear order, the traversal (Dewey) order,
definable from the term's own structure by a fold. Hence $\p\T$ is linearly
ordered \emph{per term}, with no appeal to any order on the atoms.
\end{proposition}

\begin{proof}[Sketch]
A depth-first left-to-right traversal assigns to each position a Dewey path,
and Dewey paths are linearly ordered lexicographically. The assignment is a
catamorphism over $\T$, so it is definable once $\T$ is equipped with its
recursor (\S\ref{sec:packaging}); no comparison on atoms is used, because
positions are indexed by constructor-argument slots, which are ordered by the
arity, not by the atoms sitting in them.
\end{proof}

\begin{proposition}[The aperture is the modal label]
\label{prop:glue}
The contexts that index splittings are the same objects that label rewrites in
the behavioral component of \S\ref{sec:three}: both are elements of $\p\T$.
Thus the collection component (via its splittings) and the behavioral
component (via its modal labels) are glued at the derivative. In the agency
vocabulary, the aperture --- the locus at which a computation may be acted
upon --- is simultaneously the modal label and the raw material of the proto
power object.
\end{proposition}

Proposition~\ref{prop:glue} is the structural dividend of going through $\p\T$:
components (1) and (3) of \S\ref{sec:three} are not merely analogous, they
share their carrier.

This is the point at which the title earns its keep. The splitting
$\plug : \p\T\times\T \to \T$, run backwards, is the \emph{axe-stroke}: it
cleaves a term into a focus and the context around it, and the aperture $\p\T$
is the opening the stroke leaves. What springs from that opening --- fully
armed, in the sense that it arrives already carrying the connectives of
\S\ref{sec:three} --- is the logic itself. We will call the backward splitting
the axe-stroke throughout, and $\p\T$ the aperture it opens; the crucial
aspects of logic arise, like Pallas Athena, precisely from separating the
focus of attention from its context.

\subsection{Collections are term types too}

The syntactic stance says: present the collection type itself as a term type,
so the same derivative machinery applies to it. The paradigm is the list
monad,
\[
  \List\,A = \mu X.\, 1 + A \times X,
\]
a genuine $\mu$-type. Its derivative in the element parameter is the pair of a
prefix and a suffix,
\[
  \p_{A}(\List\,A) \;\cong\; \List\,A \times \List\,A,
\]
so a splitting of a list is a triple $(\text{prefix}, \text{focus},
\text{suffix})$ --- exactly a choice of one element to keep in view, with its
surroundings. Gathering foci across a family of splittings and re-collecting
rebuilds a \emph{sub}collection. This is the concrete content of ``no list of
all lists'': each individual sublist is free to form, but assembling all of
them into one list requires ordering them.

\begin{definition}[Proto power object]
\label{def:proto}
For a term type $\T$, the \emph{proto power object} $\Proto(\T)$ is the
container of splittings, written $(\p\T, \T)$: shapes are contexts $c : \p\T$,
and the position at each shape is the focus of type $\T$. Its extension is
$\Proto(\T)(X) = \sum_{c:\p\T} X$, and $\Proto(\T)(\T) = \Split(\T)$.
\end{definition}

$\Proto(\T)$ is ``proto'' because it enumerates one-step focusings; it is not
yet closed under gathering foci into a subcollection of the same shape as
$\T$. Closing it is packaging.

\section{Packaging as a section, and the choice conjecture}
\label{sec:packaging}

\subsection{Presenting collections as quotients of the free case}

Present each collection monad $\Coll$ syntactically, as a quotient of the
free (list) collection:
\[
  \List\,A \xrightarrow{\ \qmap_{\Coll}\ } \Coll\,A,
\]
where
\[
  \qmap_{\List} = \id, \qquad
  \qmap_{\Mset} = (-)/\text{perm}, \qquad
  \qmap_{\Pfin} = (-)/(\text{perm} + \text{idem}).
\]
$\List$ is the free monoid --- the syntactic collection, paying nothing.
$\Mset$ adds commutativity (permutation). $\Pfin$ adds commutativity and
idempotence. These are precisely the two coordinates of the dial
(Principle~\ref{prin:dial}) and the two culprits of the no-go theorems
(Principle~\ref{prin:twice}).

\subsection{Two additions before choice}

Two things must be added to the bare finite-limits base before packaging can
even be discussed, and honesty requires naming them.

\begin{remark}[The recursor is the first non-Horn addition]
\label{rem:recursor}
Having ``the type of terms'' means an initial algebra $\mu F$, and using it
--- defining any fold, hence $\pack$, hence the Dewey order of
Proposition~\ref{prop:dewey} --- requires its \emph{recursor}. Formation and
introduction for a $\mu$-type are algebraic; elimination (induction/recursion)
is not cartesian. So the recursor is the first thing added to the bare base,
and we add the \emph{weakest} one that still defines $\pack$: a
non-dependent catamorphism. Everything downstream, including packaging, is a
fold and inherits exactly this recursor and no more.
\end{remark}

\begin{remark}[Quotients before sections]
For non-free $\Coll$, the map $\qmap_{\Coll}$ is a coequalizer, which a bare
lex category may lack. Presenting $\Coll$ therefore already commits one to
effective quotients (pretopos-flavored structure) \emph{before} asking them to
split. Packaging then splits what one was forced to construct.
\end{remark}

So the layering is: cartesian base $\;\to\;$ weakest recursor for $\T$
(Remark~\ref{rem:recursor}) $\;\to\;$ effective quotients for non-free $\Coll$
$\;\to\;$ the section below. Choice sits only at the top layer.

\subsection{The packaging axiom}

\begin{definition}[Packaging]
\label{def:pack}
A \emph{packaging} for $\Coll$ is a signature operation
\[
  \pack_{\Coll} : \Coll\,A \longrightarrow \List\,A
  \qquad\text{with}\qquad
  \qmap_{\Coll} \circ \pack_{\Coll} = \id_{\Coll A},
\]
a section of the presenting quotient: a choice, for each $\Coll$-collection, of
a list representative. The axiom asserting the existence of $\pack_{\Coll}$ is
$\Pack_{\Coll}$. In the finite-limits metalanguage $\Pack_{\Coll}$ is not a
proposition but the enlargement of the signature by $\pack_{\Coll}$ and the
above equation --- choice-as-structure, per \S\ref{sec:finlim}(2).
\end{definition}

Definition~\ref{def:pack} is the syntactic form of the closure operation on
the proto power object: to assemble the splittings of $\Proto(\T)$
(Definition~\ref{def:proto}) into a $\Coll$-collection of the same shape is to
apply $\qmap_{\Coll}$, and to make that assembly canonical --- to build
$\T[(\p\T,\T)]$, the collection whose atoms are splittings --- is to choose the
section $\pack_{\Coll}$.

\subsection{The choice conjecture, parametric in the atom order}

\begin{theorem}[Derivability of packaging, per instance]
\label{thm:instances}
Modulo the weakest recursor of Remark~\ref{rem:recursor}, and for the section
taken uniformly (natural in $A$):
\begin{enumerate}[label=\textup{(\roman*)},leftmargin=2.2em]
\item $\Pack_{\List}$ is derivable outright; $\pack_{\List} = \id$.
\item $\Pack_{\Mset}$ is derivable if the atom object $A$ carries a linear
      order $\le$ definable in the theory --- take $\pack_{\Mset} =
      \mathsf{sort}_{\le}$ --- and, conversely, a uniform section of
      $\qmap_{\Mset}$ determines a linear order on $A$ (read off from its
      action on two-element bags). So $\Pack_{\Mset}$ is inter-derivable with
      ``$A$ is linearly ordered.''
\item $\Pack_{\Pfin}$ is derivable if $A$ is linearly ordered \emph{and}
      equality on $A$ is available to deduplicate --- take
      $\pack_{\Pfin} = \mathsf{dedup} \circ \mathsf{sort}_{\le}$; conversely a
      uniform section determines both an order and a canonical
      representative.
\end{enumerate}
\end{theorem}

\begin{proof}[Sketch]
(i) is immediate. For (ii), $\mathsf{sort}_{\le}$ is an insertion sort, a fold
over $\List\,A$ using $\le$, hence definable from the weakest recursor and
$\le$; it lands in a canonical (sorted) representative of each permutation
class, so $\qmap_{\Mset}\circ\mathsf{sort}_{\le} = \id$. Conversely, given a
uniform section $s$, its value on the bag $\{a,b\}$ is one of the lists
$[a,b]$ or $[b,a]$; uniformity (naturality in $A$) forces this decision to
cohere into a total order, and the monoid equations force transitivity.
(iii) adds idempotence: a section of the combined quotient must additionally
select, from each finite subset, a duplicate-free list, i.e.\ a decision of
equality plus the order of (ii). \end{proof}

Theorem~\ref{thm:instances} localizes the choice content exactly. Read
against the earlier, topos-flavored heuristic in which ordered collections
were the expensive case (linearize an unordered carrier $=$ well-order $=$
full choice), it \emph{inverts} the cost: here $\List$ is free and the
quotients are dear. The two framings concern different maps, and the
finite-limits, syntactic stance is what reconciles them:

\begin{principle}[The McBride move trades and discharges]
\label{prin:trade}
Going through the derivative trades the \emph{well-ordering flavor} of choice
for the \emph{quotient-section flavor}, and then discharges the former for
free. The foci of a term-splitting are \emph{positions}, and positions are
pre-ordered by the term's own structure (Proposition~\ref{prop:dewey}). Syntax
hands you the linear order. So over a term rewrite system, packaging is
derivable from the positional order, and choice re-enters only through the
quotient equations, and only when the atom object $A$ carries no
cartesian-definable linear order.
\end{principle}

\begin{conjecture}[Choice content of packaging]
\label{conj:main}
The choice content of $\Pack_{\Coll}$ equals the assertion that the atom
object $A$ admits a linear order. Equivalently: with an order on $A$ supplied,
$\Pack_{\Coll}$ is derivable for every finitary $\Coll$ presented as a quotient
of $\List$; without one, $\Pack_{\Coll}$ is the added choice operation, and its
strength is graded by the equations $\qmap_{\Coll}$ quotients $\List$ by. In
particular the construction is kept \emph{parametric in an order on atoms}:
supplying the parameter makes packaging free; withholding it isolates the
entire choice residue at the atom object.
\end{conjecture}

The parametricity is deliberate and is the hook for later work
(\S\ref{sec:discussion}): in a reflective setting the atoms are quoted
processes --- names --- and an order on names is exactly a place where a knot
may be tied. Leaving the order parametric localizes the choice content at the
reflection boundary, where we will want to act on it.

\section{Grading the conjecture in the Weihrauch lattice}
\label{sec:grading}

Conjecture~\ref{conj:main} is not a switch but a dial, and the natural home
for its grading is the Weihrauch lattice \cite{BrattkaGherardi11}, which
measures the uniform strength of a choice principle. We state the intended
grading as a program; the finitary cases are within reach of the sketch of
Theorem~\ref{thm:instances}, the infinitary ones are open.

\begin{conjecture}[Weihrauch grading]
\label{conj:weihrauch}
The Weihrauch degree of $\Pack_{\Coll}$ over an atom object $A$ is monotone in
the substructural strength of $\Coll$ --- i.e.\ in the equations
$\qmap_{\Coll}$ adds to $\List$ --- and is bounded as follows:
\begin{itemize}[leftmargin=1.4em]
\item $\Pack_{\List}$: the identity; the bottom of the lattice.
\item $\Pack_{\Mset}$ over finite bags on discrete $A$: the finite/bounded
      choice needed to order each finite support uniformly; low in the
      lattice, below $\mathsf{LPO}$ when $A$ is discrete.
\item $\Pack_{\Mset}$ over an arbitrary (possibly infinite) atom object: a
      closed-choice principle $\mathsf{C}_{A}$ on the carrier, climbing toward
      well-ordering as $A$ grows; the well-ordering flavor of
      Principle~\ref{prin:trade} re-emerges exactly here, when the syntactic
      order is unavailable.
\item $\Pack_{\Pfin}$: the degree of $\Pack_{\Mset}$ joined with the
      equality/deduplication decision, adding an $\mathsf{LPO}$-flavored
      component for the idempotence quotient.
\end{itemize}
\end{conjecture}

The value of the Weihrauch reading is that it reframes the no-go theorems of
\S\ref{sec:nogo} not as yes/no obstructions but as \emph{choice-strength
measurements}. Where the classical no-go statement says ``no distributive law
exists,'' the graded statement says ``the section that stands in for the
missing law has degree $d$,'' and $d$ is a monotone invariant of how far
$\Coll$ sits from the free case. That is the stronger thesis hiding inside the
bare conjecture.

\section{The three instances, concretely}
\label{sec:instances}

We collect the worked instances. In each, the theory is the cartesian base
extended by the weakest recursor for the relevant $\mu$-type
(Remark~\ref{rem:recursor}); the atom order is the parameter of
Conjecture~\ref{conj:main}.

\begin{center}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{@{}llll@{}}
\hline
$\Coll$ & equations added to $\List$ & $\pack_{\Coll}$ & choice residue \\
\hline
$\List$  & none                       & $\id$                                   & none \\
$\Mset$  & commutativity              & $\mathsf{sort}_{\le}$                   & order on $A$ \\
$\Pfin$  & commutativity, idempotence & $\mathsf{dedup}\circ\mathsf{sort}_{\le}$ & order on $A$, $=$ on $A$ \\
\hline
\end{tabular}
\end{center}

\begin{example}[List: nothing is spent]
$\List\,A = \mu X.\,1 + A\times X$; $\qmap_{\List}=\id$; the section is the
identity. The proto power object $\Proto(\List\,A)$ has $\p(\List\,A) \cong
\List\,A \times \List\,A$ (prefix, suffix), and a splitting is
$(\text{prefix},\text{focus},\text{suffix})$. Packaging is free because there
is nothing to reassemble past the given list order.
\end{example}

\begin{example}[Bag: order on atoms is spent]
$\Mset\,A = \List\,A/\text{perm}$. With $\le$ on $A$, insertion sort is a fold
selecting the sorted representative, and $\qmap_{\Mset}\circ\mathsf{sort}_{\le}
= \id$. Withhold $\le$ and no cartesian term produces a representative
uniformly; $\pack_{\Mset}$ must be postulated, and by Theorem~\ref{thm:instances}(ii)
its postulation is inter-derivable with linearly ordering $A$. This is the
cleanest witness for Conjecture~\ref{conj:main}.
\end{example}

\begin{example}[Finite powerset: order and equality]
$\Pfin\,A = \List\,A/(\text{perm}+\text{idem})$. Packaging needs the sorted
representative of the previous example \emph{and} a deduplication, hence
decidable equality on $A$ in addition to the order. The two additions are
exactly the two quotient equations, confirming Principle~\ref{prin:twice}: the
idempotence that obstructs $\Pfin$'s distributive laws is the same idempotence
that must be paid for, as $\mathsf{dedup}$, at packaging time.
\end{example}

\section{A payoff: the collection component of Gabbay's coalition logic}
\label{sec:semitopology}

We now give the reader a payoff for the formalism. We take a logic that arose
independently, for reasons that have nothing to do with realizability ---
Gabbay's logic of \emph{actionable coalitions}, whose model theory is
\emph{semitopology} \cite{Gabbay25,GabbayLosa23,GabbaySemiframes,GabbayBook}
--- and we show that its most distinctive features fall out of the
collection-component analysis of \S\ref{sec:three}--\S\ref{sec:disconnect} as
consequences of the shape of a single container. We treat \emph{only} the
collection component here; the structural and behavioral components, and the
identification of the underlying points with terms of the rho calculus, are
the subject of a companion paper.

\subsection{Semitopologies, and why they are already a realizability
semantics}

A \emph{semitopology} \cite{Gabbay25} is a pair $(P,\mathsf{Open})$ of a set
$P$ of points and a set $\mathsf{Open}\subseteq \mathcal{P}(P)$ of open sets
with $\emptyset,P\in\mathsf{Open}$ and $\mathsf{Open}$ closed under arbitrary
unions --- but \emph{not} necessarily under intersection. The intended reading
is distributed: points are participants, and an open set is an
\emph{actionable coalition}, a set of participants with the resources to take
a collaborative action without the permission of anyone outside it. A union of
actionable coalitions is actionable; an intersection need not be (two coalitions
each strong enough to act may overlap in a single participant, who is not).
This one dropped axiom is the whole of the difference from topology.

What matters for us is that Gabbay motivates the definition in exactly the
terms of \S\ref{sec:motivation}. Following the observational tradition in which
opens model affirmations, or semidecidable properties \cite{Vickers89}, he
reads an open set as an \emph{actionable affirmation}: an outcome that can be
agreed and enacted within the coalition. Topology, on this reading, is about
what can be positively \emph{said}; semitopology is about what can be
positively \emph{done}. An open is thus already a collection of computational
witnesses --- participants who can jointly realize an outcome --- which is to
say that semitopology is a realizability semantics whose container is the opens
of a semitopology. This is precisely the setting our collection-component
analysis was built to read.

\subsection{The connectives fall out of the container}

Fix a semitopology $(P,\mathsf{Open})$ and take the collection container to be
$\Coll = \mathsf{Open}(P)$, ordered by $\subseteq$, with a formula denoting the
actionable coalition of participant-witnesses that realize it. Running the
collection component of \S\ref{sec:three} --- connectives are exactly the
operations under which the container is closed --- gives the following, with
nothing added by hand.

\begin{proposition}[Connectives of the semitopological collection component]
\label{prop:semi-conn}
Let $\Coll = \mathsf{Open}(P)$ for a semitopology $(P,\mathsf{Open})$. Then:
\begin{enumerate}[label=\textup{(\roman*)},leftmargin=2.2em]
\item $\Coll$ is closed under arbitrary unions, so disjunction
      $\bigvee = \bigcup$ is a total, arbitrary (geometric) collection
      operation: a genuine collection connective.
\item $\Coll$ is in general \emph{not} closed under binary intersection, so
      conjunction is \emph{not} interpretable as a collection operation on
      $\Coll$. The largest coalition below two opens,
      $\mathrm{int}(A\cap B)=\bigcup\{U\in\Coll : U\subseteq A\cap B\}$, exists
      by union-closure but is a \emph{derived} operation --- an interior, i.e.\
      a closure --- not the container's intersection.
\item The primitive relational residue of conjunction is therefore the
      \emph{compatibility relation}
      \[
        A * B \quad\Longleftrightarrow\quad A\cap B \neq \emptyset,
      \]
      which records \emph{co-realizability} --- the existence of a witness
      realizing both --- without producing a witness-collection.
\end{enumerate}
\end{proposition}

\begin{proof}
(i) is closure under arbitrary unions, an axiom of semitopology. (ii): if
$\mathsf{Open}$ were closed under $\cap$ it would be a topology; a semitopology
is exactly one that need not be, and $\mathrm{int}(A\cap B)$ is the greatest
open below $A\cap B$ by union-closure (the union of all opens contained in
$A\cap B$). (iii) is immediate: $*$ is definable but is a relation, not an
operation into $\Coll$. \end{proof}

The content of Proposition~\ref{prop:semi-conn} is the asymmetry it exposes.
On the substructural dial of Principle~\ref{prin:dial}, the aggregation of this
container is union, which is commutative and idempotent, so on the \emph{join}
the logic looks fully structural. But the dial's two coordinates do not see the
phenomenon that actually distinguishes this setting, which is that
\emph{disjunction is container-primitive while conjunction is not}. This is a
finer axis than the two Booleans of \S\ref{sec:dial}:

\begin{principle}[Which lattice operations are container-primitive]
\label{prin:primitive}
Beyond \emph{which structural rules hold} (Principle~\ref{prin:dial}), a
container is graded by \emph{which lattice operations it makes primitive}. A
topology makes both arbitrary $\bigvee$ and finite $\wedge$ primitive: its
opens form a frame, and the collection component yields geometric/intuitionistic
logic. A semitopology makes only $\bigvee$ primitive: its opens form a
\emph{semiframe}, and the collection component yields a weaker logic of
actionable coalitions in which conjunction survives only relationally, as $*$,
or derivatively, as $\mathrm{int}(-\cap-)$.
\end{principle}

\subsection{The framework predicts the semiframe}

Here is the payoff proper. The algebraic dual of a topology is a frame: a
complete lattice in which finite meets distribute over arbitrary joins. The
algebraic dual of a semitopology cannot be a frame, because
Proposition~\ref{prop:semi-conn}(ii) has removed the meet. What must replace
it? Exactly the relational residue of Proposition~\ref{prop:semi-conn}(iii).

\begin{observation}[The compatibility relation is forced]
\label{obs:semiframe}
The collection-component analysis predicts, from the closure properties of the
container alone, that the algebraic dual of a semitopology is a complete
join-semilattice equipped not with a meet operation but with a symmetric,
join-distributive compatibility relation $*$ abstracting nonempty intersection.
This is exactly Gabbay's \emph{semiframe} \cite{GabbaySemiframes}: a complete
join-semilattice $(X,\le)$ with a relation ${*}\subseteq X\times X$ that is
symmetric, properly reflexive ($x * x \Leftrightarrow x\neq\bot$), and
completely distributive over joins
($x * \bigvee X' \Leftrightarrow \exists x'\!\in X'.\,x * x'$). Gabbay arrived
at $*$ independently, as the structure needed to make the
semitopology--semiframe duality go through and to express the anti-separation
(intertwinedness) conditions that govern consensus; the collection component
recovers the same relation as the forced residue of a meet-free container. Two
routes, one relation.
\end{observation}

Observation~\ref{obs:semiframe} is the concrete cash value of the abstract
disconnect of \S\ref{sec:disconnect}. There we said that categorical logic
packages the collection component through a device --- the subobject classifier,
equivalently the frame of subobjects --- that bakes in set-shaped, meet-closed
collection, and that the packaging fails once the container is not meet-closed.
Semitopologies are exactly such a container, and the failure is not pathology
but the whole point: in a decentralised system the intersection of two
actionable coalitions is not actionable, so the logic of the system \emph{must}
be the meet-free one. The seam that \S\ref{sec:disconnect} exposed abstractly
is, here, the seam along which distributed consensus is actually cut.

\subsection{What the residual structure means under realizability}

Reading Gabbay's constructions back through the realizability lens gives them a
computational gloss worth recording, since it is what the companion paper will
build on.

The compatibility relation $A * B$ says that $\sem{\varphi}$ and $\sem{\psi}$
share a witness: $\varphi$ and $\psi$ are co-realizable, even when
$\varphi\wedge\psi$ has no direct reading as a coalition. This is strictly
weaker than demanding that the meet be a nonempty collection, and it is the
right primitive: in a distributed setting one wants to know whether two actions
\emph{can} be taken by an overlapping party, not to name the coalition of all
parties who take both.

A \emph{topen} --- a transitive open set, the pivotal notion in Gabbay's
consensus theory --- is an open $T$ such that any two opens meeting $T$ meet
each other. Under realizability this is a witness-collection with internal
coherence: on a topen, continuous value assignments are forced to agree
\cite{Gabbay25}, so a topen is a collection of witnesses \emph{guaranteed to
decide alike}. A \emph{regular} point is a participant lying in such a coherent
community. In the companion paper, where points are rho-calculus processes, a
topen becomes a set of processes guaranteed to reach consensus, and the
collection component developed here becomes the propositional logic of
actionable coalitions of processes.

\subsection{Why the finite-limits commitment is repaid here}

There is a final dividend, and it ties directly to \S\ref{sec:finlim}. Moving
from a topology to a semitopology moves conjunction from an \emph{operation}
$\wedge : X\times X\to X$ subject to the frame distributive law, to a
\emph{relation} $*\subseteq X\times X$ subject to three Horn-shaped axioms. A
relation is, in a finite-limits metalanguage, the most native thing there is: a
subobject ${*}\rightarrowtail X\times X$ carved out by equations, with the
join-distributivity axiom a geometric implication of exactly the kind cartesian
logic is built to state. So dropping intersection-closure does not merely model
distributed systems; it lands the collection component in a more purely
relational, more cartesian presentation than the topological case. The
semitopological setting is \emph{more} syntactic than the topological one, not
less --- precisely the direction the finite-limits commitment of
\S\ref{sec:motivation} and \S\ref{sec:finlim} was pushing us.

\begin{remark}[Compatibility is a proto-conjunction]
The parallel with the proto power object (\S\ref{sec:derivative}) is worth one
line. The compatibility relation $*$ is to conjunction what the object of
splittings is to the power object: the un-packaged, relational form that
records possibility without producing the packaged object. Recovering an actual
conjunction from $*$ means applying the interior --- a closure, a global
operation --- just as recovering a subcollection from splittings meant applying
$\pack$. Conjunction-as-relation versus conjunction-as-operation is one more
instance of proto versus packaged.
\end{remark}

\section{Discussion}
\label{sec:discussion}

\paragraph{No classicality collapse.} The absence of Diaconescu's argument in
the cartesian fragment (\S\ref{sec:finlim}(3)) is what makes this entire
program viable. We add a choice operation --- $\pack_{\Coll}$ --- and the
object logic does not thereby go Boolean, because the collapse needs $\vee$,
comprehension, and effective-quotient machinery that the metalanguage does not
provide. Choice remains structural. A logic assembled this way can have any
setting of the substructural dial (Principle~\ref{prin:dial}) and still admit
packaging, without being forced to classicalize. This is the payoff we most
want the AI community to register: inspectable choice that does not
contaminate the logic it serves.

\paragraph{Quotient versus choice as dual dodges.} \S\ref{sec:nogo} noted two
standard escapes from the no-go theorems --- quotient (adjust $\Coll$ until the
equations agree) and restrict (shrink until they cannot clash) --- and offered
a third, adding a global section. We conjecture these are dual: the
quotient-dodge builds a coequalizer, the section-dodge splits it; the
restrict-dodge builds a well-founded (finite, ordered) carrier, the
section-dodge well-orders it. In slogan form: \emph{coequalizer versus Zorn}.
Making this duality precise --- exhibiting the section-dodge as the formal dual
of the quotient-dodge in the appropriate fibration --- is the natural next
technical step, and it would situate the construction cleanly against the
existing distributive-law literature.

\paragraph{Graphs are the genuine frontier.} Everything here rests on $\T$
being a polynomial/container functor, so that $\p\T$ exists and
Definition~\ref{def:proto} is available. Lists and trees are fine. Graphs are
not free/regular in the naive sense, the McBride derivative does not directly
apply, and the proto power object is not defined. We bracket the graph case
explicitly rather than let it ride along with lists and trees; adapting the
derivative to non-regular carriers (or replacing it by a suitable
comonadic/traced structure) is the honest open problem for extending the
collection dial past the container-shaped settings.

\paragraph{The atom-order parameter and knot-tying.} Conjecture~\ref{conj:main}
keeps the order on atoms parametric on purpose. In the reflective (rho) setting
the atoms are quoted processes --- names --- and the boundary between process
and name is exactly the reflection boundary. An order on names is therefore a
structure imposed \emph{at} that boundary, and it is precisely the datum a
knot-tying construction will manipulate. By isolating the entire choice residue
of packaging in the atom order, the present construction leaves the knot
exactly one place to be tied: supply the order (and packaging trivializes; the
knot is untied) or withhold it (and the choice residue is the knot). This is
the sense in which the finite-limits, derivative-based account is not merely a
reconciliation of the collection component with categorical logic, but a setup
for the reflective constructions to come.

\paragraph{Summary.} The three-component decomposition makes the seams of a
logic visible under realizability; the collection component is a substructural
dial governed by commutativity and idempotence; those same two equations
obstruct the distributive laws and set the cost of packaging; the
finite-limits metalanguage removes the subobject classifier (forcing the
syntactic route) and forbids stating choice as a proposition (forcing it to be
a named section), while blocking the classicality collapse that would
otherwise make this untenable; the McBride derivative supplies the proto power
object and, through the positional order on splittings, discharges the
well-ordering flavor of choice; and what remains --- the choice content of
packaging --- is exactly the order on atoms, kept parametric for the knot to
act on. That is the construction this programme set out to formalize, and the shape of the
program it opens.

\section{Coda: collections have computational content}
\label{sec:coda}

One point remains, and it is the one we most want to leave with the reader,
because it dissolves a confusion that has dogged program extraction for
decades. A collection is not a neutral container. A collection type carries
computational content in and of itself, and so the realizability content
$\sem{\varphi} : \Coll(\T)$ has its computation \emph{parsed} between the
collection $\Coll$ and the explicit computational ontology $\T$. Neither
component alone fixes the computational model, and until the parse is
acknowledged, extraction and its dual are ill-posed.

\subsection{The collection's equations are a computational ontology}

Read the quotient tower of \S\ref{sec:packaging} once more, now for its
computational content rather than its choice content:
\begin{itemize}[leftmargin=1.4em]
\item $\List$ (ordered, multiplicities retained) is \emph{sequential}
      computation: traces, iteration, the fold as the machine.
\item $\Mset$ (commutativity added) is \emph{concurrent, resource-sensitive}
      computation: multiset rewriting, the chemical abstract machine
      \cite{BerryBoudol92}, the commutative parallel composition of a process
      calculus.
\item $\Pfin$ (idempotence added) is \emph{nondeterministic, relational}
      computation: choice with neither order nor multiplicity.
\item $\Dist$ (convexity in place of idempotence) is \emph{probabilistic}
      computation.
\end{itemize}
Each quotient in the tower forgets a layer of computational content:
permutation forgets sequencing, idempotence forgets multiplicity. The
substructural dial of \S\ref{sec:dial} is therefore, read from this side, a
dial on \emph{how much computation the collection itself carries}, and the
packaging quotients of \S\ref{sec:packaging} are exactly the maps that discard
it.

\begin{principle}[The dial is a computational ontology]
\label{prin:onto}
The algebraic theory of $\Coll$ is not merely a set of structural rules; it is
a computational ontology. The tower $\List \twoheadrightarrow \Mset
\twoheadrightarrow \Pfin$ is a spectrum of progressive forgetting ---
sequential, then concurrent, then nondeterministic --- and where a collection
sits on it fixes the computational model it can natively express.
\end{principle}

\subsection{Computation is conserved and parsed across $(\Coll,\T)$}

A domain-specific computational model must be recovered from the whole of
$\sem{\varphi}:\Coll(\T)$, and its content is apportioned between the two
components. Coarsen $\Coll$ --- move up the tower, forget --- and the forgotten
content must reappear in $\T$ if the same model is to be recovered; enrich
$\Coll$ and $\T$ may be left bare. The extremes and the middle are all
inhabited:
\begin{itemize}[leftmargin=1.4em]
\item \emph{All in $\Coll$}: $\T$ degenerate, the model \emph{is} the
      collection's native one --- pure nondeterminism is $\Pfin$ over trivial
      terms.
\item \emph{All in $\T$}: $\Coll$ a bare set, every domain-specific feature
      carried explicitly by the rewrite system --- the default posture of
      classical realizability, and the source of the confusion below.
\item \emph{The middle}, where real systems live: a process calculus places
      its concurrency in a bag-shaped collection (parallel composition is
      commutative) and its reflection and higher-order structure in $\T$. This
      is the parse the companion paper adopts, with $\T$ the rho calculus
      \cite{MeredithRadestock05a}.
\end{itemize}

\begin{principle}[Conservation and parsing]
\label{prin:parse}
The computational content of $\sem{\varphi}:\Coll(\T)$ is conserved and parsed
across the pair $(\Coll,\T)$. Content forgotten by $\Coll$ must be supplied by
$\T$, and content carried by $\Coll$ need not be duplicated in $\T$. No single
component fixes the model; the pair does.
\end{principle}

\subsection{Why extraction was confused: the Olympian's headaches}

Now the confusion --- the headaches that, in the myth, precede the stroke. The
proofs-as-programs tradition typically fixes $\Coll$ at
the set/powerset end --- propositions as sets of realizers --- and then expects
the extracted program to inhabit an arbitrary, often sequential and
Turing-complete, computational model. But a set or powerset collection has
\emph{already} quotiented out order and multiplicity
(Principle~\ref{prin:onto}). Extraction through it cannot recover sequencing or
resource-sensitivity; it recovers only the nondeterministic-relational shadow
that survives the quotient.

\begin{observation}[Collections do not recover arbitrary models]
\label{obs:extract}
A collection, without additional information, cannot recover an arbitrary
computational model. It recovers precisely the models native to its ontology.
To extract a domain-specific program one must either move $\Coll$ down the
tower until it retains the needed content, or carry that content explicitly in
$\T$; expecting a fixed collection to deliver an arbitrary model is the
category error at the root of the extraction confusion.
\end{observation}

\subsection{Extraction and classification are dual}

The parse governs not one procedure but two, dual to one another. Write the
realizability relation as ${\real}\subseteq \mathrm{Comp}\times\mathrm{Form}$.
Its two transposes are:
\[
  \text{extraction:}\quad \varphi \longmapsto \sem{\varphi}
  \qquad\text{(synthesis; formula to its witnesses),}
\]
\[
  \text{classification:}\quad e \longmapsto \{\varphi : e \real \varphi\}
  \qquad\text{(analysis; computation to the formulae it realizes),}
\]
adjoint in the manner of a Galois connection. The parse $(\Coll,\T)$ sets the
resolution of both at once: coarsening $\Coll$ lowers the resolution of
extracted programs and, in the same motion, coarsens classification, so that
formulae separate computations less finely. The granularity is shared because
it is one relation read in two directions. Consequently, fixing the parse makes
both procedures effective: one knows which model the extracted program inhabits
(the native model of $\Coll$, refined by $\T$), and at what granularity
witnesses are classified. That is what it means to understand extraction and
witness classification effectively.

\begin{observation}[Semitopology, once more]
\label{obs:coda-semi}
The payoff of \S\ref{sec:semitopology} is an instance. The open-set collection
has semidecidability --- affirmability, actionability --- as its native
ontology, so extraction from an open is a semidecision procedure and
classification is the assignment to a participant of the affirmations it
realizes, i.e.\ the coalitions it joins. A topen is then a classification
guaranteed to be consensus-coherent: a family of witnesses that must be
classified alike. The semitopology is simply the choice to parse observational
content into $\Coll$.
\end{observation}

\subsection{The last seam}

There is a reason this could only be said cleanly from a finite-limits
metalanguage. Keeping $\T$ explicit and syntactic (\S\ref{sec:finlim}) is what
makes the parse legible; hide $\T$ inside an ambient model --- a topos, a fixed
$\Omega$, an assumed Turing-complete substrate --- and the division of
computational content between collection and terms becomes invisible, which is
exactly how the extraction confusion took hold. The three components of
\S\ref{sec:three} were the seams of a logic. The parse is the seam within
computation itself: between what the collection already computes and what the
terms make explicit. Making that seam visible was the whole point of the
programme, and it is what turns program extraction and witness classification
from standing sources of confusion into effective, dual procedures. The
headaches lift when the stroke falls: it is the splitting of focus from
context that lets the logic spring out whole.

\begin{thebibliography}{99}

\bibitem{Kleene45} S.~C.\ Kleene. On the interpretation of intuitionistic
number theory. \emph{Journal of Symbolic Logic}, 10:109--124, 1945.

\bibitem{vanOosten08} J.\ van Oosten. \emph{Realizability: An Introduction to
its Categorical Side}. Studies in Logic and the Foundations of Mathematics
152, Elsevier, 2008.

\bibitem{MeredithRadestock05a} L.~G.\ Meredith and M.\ Radestock. A reflective
higher-order calculus. \emph{Electronic Notes in Theoretical Computer
Science}, 141(5):49--67, 2005.

\bibitem{MeredithRadestock05b} L.~G.\ Meredith and M.\ Radestock. Namespace
logic: A logic for a reflective higher-order calculus. In \emph{Trustworthy
Global Computing (TGC)}, LNCS 3705, 2005.

\bibitem{Beck69} J.\ Beck. Distributive laws. In \emph{Seminar on Triples and
Categorical Homology Theory}, LNM 80, pp.\ 119--140, Springer, 1969.

\bibitem{KlinSalamanca18} B.\ Klin and J.\ Salamanca. Iterated covariant
powerset is not a monad. In \emph{Mathematical Foundations of Programming
Semantics (MFPS)}, ENTCS 341, 2018.

\bibitem{VaraccaWinskel06} D.\ Varacca and G.\ Winskel. Distributing
probability over non-determinism. \emph{Mathematical Structures in Computer
Science}, 16(1):87--113, 2006.

\bibitem{ZwartMarsden19} M.\ Zwart and D.\ Marsden. No-go theorems for
distributive laws. In \emph{Logic in Computer Science (LICS)}, 2019.
(Extended version: \emph{Logical Methods in Computer Science}, 2022.)

\bibitem{McBride01} C.\ McBride. The derivative of a regular type is its type
of one-hole contexts. Unpublished manuscript, 2001.

\bibitem{AAGM05} M.\ Abbott, T.\ Altenkirch, N.\ Ghani, and C.\ McBride.
$\partial$ for data: Differentiating data structures. \emph{Fundamenta
Informaticae}, 65(1--2):1--28, 2005.

\bibitem{Huet97} G.\ Huet. The zipper. \emph{Journal of Functional
Programming}, 7(5):549--554, 1997.

\bibitem{Freyd72} P.\ Freyd. Aspects of topoi. \emph{Bulletin of the
Australian Mathematical Society}, 7:1--76, 1972.

\bibitem{GabrielUlmer71} P.\ Gabriel and F.\ Ulmer. \emph{Lokal
pr\"asentierbare Kategorien}. LNM 221, Springer, 1971.

\bibitem{PalmgrenVickers07} E.\ Palmgren and S.~J.\ Vickers. Partial Horn
logic and cartesian categories. \emph{Annals of Pure and Applied Logic},
145(3):314--353, 2007.

\bibitem{Diaconescu75} R.\ Diaconescu. Axiom of choice and complementation.
\emph{Proceedings of the American Mathematical Society}, 51:176--178, 1975.

\bibitem{MacLaneMoerdijk92} S.\ Mac Lane and I.\ Moerdijk. \emph{Sheaves in
Geometry and Logic: A First Introduction to Topos Theory}. Springer, 1992.

\bibitem{BrattkaGherardi11} V.\ Brattka and G.\ Gherardi. Weihrauch degrees,
omniscience principles and weak computability. \emph{Journal of Symbolic
Logic}, 76(1):143--176, 2011.

\bibitem{Gabbay25} M.\ J.\ Gabbay. Semitopology: a topological approach to
decentralised collaborative action. \emph{Journal of Logic and Computation},
35(5):exae050, 2025. (arXiv:2303.09287.)

\bibitem{GabbayLosa23} M.\ J.\ Gabbay and G.\ Losa. Semitopology: a survey of
a topological approach to distributed collaborative action.
arXiv:2402.03253, 2023--2024.

\bibitem{GabbaySemiframes} M.\ J.\ Gabbay. Semiframes: the algebra of
semitopologies and actionable coalitions. \emph{Mathematical Structures in
Computer Science}, to appear. (arXiv:2310.00956.)

\bibitem{GabbayBook} M.\ J.\ Gabbay. \emph{Semitopology: Decentralised
Collaborative Action via Topology, Algebra, and Logic}. College Publications,
2024.

\bibitem{Vickers89} S.\ Vickers. \emph{Topology via Logic}. Cambridge Tracts
in Theoretical Computer Science 5, Cambridge University Press, 1989.

\bibitem{Pauly02} M.\ Pauly. A modal logic for coalitional power in games.
\emph{Journal of Logic and Computation}, 12(1):149--166, 2002.

\bibitem{BerryBoudol92} G.\ Berry and G.\ Boudol. The chemical abstract
machine. \emph{Theoretical Computer Science}, 96(1):217--248, 1992.

\end{thebibliography}

\end{document}
