\documentclass[11pt]{article}

\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\definecolor{f1r3}{HTML}{C0392B}
\definecolor{ink}{HTML}{1A1A1A}

\hypersetup{colorlinks=true,linkcolor=f1r3,urlcolor=f1r3,citecolor=f1r3}

\theoremstyle{plain}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

% notation
\newcommand{\G}{\mathcal{G}}
\newcommand{\Cost}{\mathsf{Cost}}
\newcommand{\C}{\mathsf{C}}
\newcommand{\T}{\mathsf{T}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\sigof}{\mathsf{sig\text{-}of}}
\newcommand{\quo}[1]{@#1}
\newcommand{\deq}[1]{{\ast}#1}
\newcommand{\bisim}{\;\approx\;}
\newcommand{\cmaint}{c_{\mathrm{maint}}}
\newcommand{\crep}{c_{\mathrm{rep}}}
\newcommand{\cmint}{c_{\mathrm{mint}}}
\newcommand{\ccopy}{c_{\mathrm{copy}}}
\newcommand{\R}{\mathbb{R}}

\newcommand{\hrulethick}{\noindent\textcolor{f1r3}{\rule{\linewidth}{1.2pt}}}

\title{\vspace{-3em}
  {\color{f1r3}\Large\bfseries Mortal Computation}\\[0.35em]
  {\large Mortality, the Genome, and Life-History in Cost-Accounted GSLTs}}
\author{L. G. Meredith \\ {\small F1R3FLY.io}}
\date{\small 31 May 2026 --- working note}

\begin{document}
\maketitle
\vspace{-1.5em}
\hrulethick
\vspace{1em}

\begin{abstract}
\noindent
We read the cost-accounting (metering) construction on graph-structured lambda
theories (GSLTs) as a biology rather than as an economics. The token-gated
rewrite that meters computation is, on the nose, a mechanism for making
computation \emph{mortal}: death is deadlock-by-starvation, and metabolism is
the splitter/joiner token economy that postpones it. For Turing-complete
continued-interactive GSLTs the metered calculus translates faithfully back
into the original model, and the image of the round trip is a sublanguage in
which mortality and metabolism are implemented from native parts. Turing
completeness further yields, by composing the two universality maps, a
meta-circular interpreter; when the model hosts reflection natively this
interpreter is internalised, giving the quote/dequote pair the dual-use
character von~Neumann required of a self-reproducing description. Quoted
description behaves as a potentially immortal germ line; the dequoted running
process is the mortal soma. With reproduction thereby isolated as a metabolic
cost separable from a computation's other functions, we recover life-history
theory: the logistic law $\dot N = rN(1-N/K)$ emerges as the mean-field of a
population of cost-accounted replicators on a bounded-inflow substrate, with
$r$ and $K$ as closed forms in the construction's primitives, and $r$- versus
$K$-selection appears as the two extreme vertices of a continuous allocation
simplex.
\end{abstract}

\vspace{0.5em}

\section{Setup: the cost endofunctor and its round trip}
\label{sec:setup}

Write an interactive GSLT as $\G=(\Sigma,E,R)$ with a distinguished interacting
sort carrying a labelled transition system, structural equations $E$, and
rewrite rules $R$. The cost construction is an endofunctor $\C$ on the
subcategory of interactive GSLTs and bisimulation-preserving morphisms. On
objects $\C$ adjoins a signature sort $\Sig$ as a commutative monoid, a
token-stack sort $S ::= () \mid s\!:\!S$ as the free $\Sig$-graded structure, a
signed-term wrapper $\{\cdot\}_s$, and a constructor $\sigof$ that mints a fresh
signature from runtime data. It replaces each rule $\ell\to r$ of $R$ by its
\emph{gated family}: the redex fires only when matching tokens are present and
consuming them. The five rules that govern signed communication are not five
facts; they are the image of the single communication rule under $\C$.

We assume throughout that $\G$ is Turing-complete and
\emph{continued-interactive}: it can express persistent processes that continue
to interact after producing output. Under this assumption the metered calculus
$\C\G$ admits a compositional translation
\[
  \T:\C\G\longrightarrow\G,
\]
which is a faithful simulation up to bisimulation: signatures map to channel
names (via reflection where available, or to fresh names otherwise), tokens
become messages on those channels carrying continuation payloads, and signed
processes become \emph{fuel-gates}---guard patterns that must consume a token
message before the underlying redex proceeds. Splitters and joiners reconcile
combined and decomposed token granularities. The object of study here is the
\emph{round trip}
\[
  \T\circ\C:\G\longrightarrow\G,
\]
an endomorphism of the original model. Its image $\operatorname{im}(\T\circ\C)
\subseteq\G$ is a distinguished sublanguage built entirely from native
constructs. The claim of this note is that the biology---mortality, the genome,
and life-history---lives in that image, and that all of it is therefore native
to any Turing-complete continued-interactive $\G$, the cost construction merely
locating where.

\section{Mortality and metabolism}
\label{sec:mortality}

\paragraph{Death.}
In $\C\G$ a redex fires only by consuming a matching token; a process whose
stack is $()$ has no enabled rewrite and blocks permanently. Pushing this
through $\T$, the token message on the relevant signature-channel simply stops
arriving and the fuel-gate never completes. In $\G$ this is an ordinary
configuration: a guarded process whose guard is never discharged.

\begin{definition}[Starvation death]
A process $P$ in $\operatorname{im}(\T\circ\C)$ is \emph{dead} at a
configuration if every redex of $P$ is fuel-gated and no token message is
reachable on any of its gating channels. Death is permanent deadlock by
fuel exhaustion.
\end{definition}

Mortality is thus not a new primitive but a fragment of $\G$ that $\G$ admitted
all along; $\C$ exhibits which fragment.

\paragraph{Metabolism.}
The token economy that postpones death is metabolic in a precise sense. A
joiner assembles a high-denomination token from components (anabolism); a
splitter cleaves one into spendable units (catabolism); and the gated rewrite
couples each unit of computational work to a unit of fuel consumption---no step
without an accompanying expenditure. This is the same coupling that defines a
metabolism thermodynamically. Read against a Ruelle--Bowen
potential~\cite{ruelle1978,bowen1975} on the
transition graph, the splitter/joiner dynamics are the flow and the zero-fuel
configurations are the absorbing set: metabolism is motion on the landscape,
death is its absorbing boundary.

\paragraph{The round-trip image and the Buss argument.}
A parallel composition of metered processes is exactly the kind of population
that supports selection: copies are autonomous ($P\mid P\not\equiv P$ once
embedded in distinct local contexts), and selection exploits inexactness rather
than collapsing it. In the un-metered $\G$ an immortal replicator
$\,! P$ with a faster-cycling $P$ always invades a parallel composition without
bound; there is no configuration in which a higher-level whole is stable
against its own fastest subterm. This is Buss's cell-lineage
problem~\cite{buss1987} in its
rawest form. In $\operatorname{im}(\T\circ\C)$ the defector is fuel-bounded: it
out-cycles its neighbours only until its stack is spent, after which it
deadlocks.

\begin{proposition}[Mortality caps invasion]
\label{prop:invasion}
Let $D$ be a subterm-variant with per-cycle fuel draw strictly exceeding its
per-cycle fuel income on a shared bounded substrate. In $\C\G$, $D$ executes for
a number of cycles bounded by its initial endowment and then dies, so the
cumulative interactions it contributes are finite. In the un-metered $\G$ the
same $D$ contributes unboundedly. Hence configurations in
$\operatorname{im}(\T\circ\C)$ that sequester the minting constructor $\sigof$
admit no heritable take-over by a high-draw somatic defector, whereas no
configuration of un-metered $\G$ resists one.
\end{proposition}

The biological reading is Buss's: somatic mortality together with germ
sequestration aligns the levels of selection by denying a within-individual
defector any future. Mortality is selected not in spite of bounding individual
runs but because a bounded run is the precondition for a stable higher-level
individual that can itself be a durable unit of selection.

\section{Reflection and the genome}
\label{sec:genome}

\paragraph{A self-interpreter from universality.}
Turing completeness furnishes two universality maps. The first,
$e:\mathrm{TM}\to\G$, is completeness proper: $\G$ simulates any machine. The
second, $i:\G\to\mathrm{TM}$, is the Church--Turing direction~\cite{turing1936}: any effectively
presented reduction relation can be run by a machine. Their composite
\[
  i\circ e:\G\longrightarrow\G
\]
is a program \emph{in} $\G$ that takes a $\G$-program as data and runs it,
correct up to bisimulation: a meta-circular interpreter. To feed a program to
this interpreter one must first represent it as a value, and the TM detour
manufactures such a code-as-data representation with no reflective primitive in
sight. \emph{Reflection appears as a consequence of universality}---a theorem,
not a term. This is exogenous reflection.

\paragraph{What native reflection changes.}
When $\G$ hosts reflection natively---as the rho calculus does~\cite{meredith2005},
a reflective higher-order extension of Milner's $\pi$-calculus~\cite{milner1992},
where $\quo P$
is already a value in the same algebra as the running process---the
meta-circular interpreter is internalised, and the gain is representational, not
merely a saved detour. A TM-tape encoding has a brutal mutation neighbourhood:
almost any edit to the tape is catastrophic because the tape does not respect
the term structure it encodes. A quoted term $\quo P$ has a benign one: mutation
is a structural edit of a subterm and recombination is the splicing of quoted
subprocesses, both staying inside the model's own grammar. Native reflection is
therefore the difference between a genome that is evolvable and one that is not.

\begin{proposition}[Two readings agree on behaviour, differ on the description]
\label{prop:fidelity}
The exogenous interpreter $i\circ e$ and the internalised interpreter agree up
to bisimulation on the behaviour of every interpreted program, but differ on
the algebra of the code-as-data representation: the former carries a foreign
encoding closed under tape edits, the latter the model's own term algebra closed
under structural edits and splices. Evolvability is a property of the latter
algebra, not of the interpreted behaviour, and so distinguishes the two readings
even though they are behaviourally equivalent.
\end{proposition}

\paragraph{Von Neumann's dual use, recovered.}
A self-reproducing description $\varphi$~\cite{vonneumann1966} must support two
distinct operations:
it is \emph{copied uninterpreted} (replication) and \emph{interpreted to build
the machinery} (expression). Native reflection supplies both on a single
object. A name $\quo P$ can be transmitted on a channel and received
\emph{without} dequoting---replication, the description copied as inert
data---or dequoted by $\deq{}$ and run---expression, the description read as
instructions. Crick's transcription/translation split~\cite{crick1970} and
Weismann's two modes
of the germ-line description~\cite{weismann1893} are these same two operations on $\quo P$. The
meta-circular structure is what closes the loop: the reader is encoded in what
it reads.

\paragraph{Germ and soma from quote and dequote.}
Refining \S\ref{sec:mortality}: the germ line is the sublanguage that retains
$\sigof$ \emph{and} carries the quoted description $\quo P$---together exactly
what is needed to fuel and to instantiate an offspring. The soma is the dequoted
running process $\deq{\quo P}$. Weismann's mortality asymmetry then falls out of
the quote/dequote distinction rather than being postulated:
\[
  \underbrace{\deq{\quo P}}_{\text{soma}}\ \text{metabolises and starves;}
  \qquad
  \underbrace{\quo P}_{\text{germ}}\ \text{is inert, cannot starve, is re-instantiable.}
\]
\begin{center}
\textit{Mortal soma $=$ the dequoted process. Immortal germ $=$ the quoted description.}
\end{center}
Finally, applying a fuel-gate not to a bare communication but to the
\emph{dequoting} operation is regulated expression: the gene $\quo P$ is read and
run only when the requisite tokens are present. This is operon logic---expression
coupled to metabolic state. The genome's full operation set is thus native:
replication is copying the name, expression is token-gated dequoting, mutation
and recombination are edits and splices of the quoted term, and regulation is
the fuel-gate deciding \emph{when} the description is read.

\section{Life-history theory}
\label{sec:lifehistory}

Metering makes ``the cost of reproduction'' a measurable quantity rather than a
category error. In un-metered $\G$, $\,!P$ replicates for free and nothing is
separable from anything else. Once every operation is token-gated, the
reproduction operations---minting via $\sigof$, copying $\quo P$, endowing the
offspring's initial stack, and the gated dequoting that instantiates it---each
carry an identifiable charge in the same currency as every other charge the
process incurs.

\subsection{The metabolic ledger}

Per metabolic cycle a replicating computation has realised income $B$ and
partitions it into maintenance and reproduction,
\[
  B \;=\; \cmaint \;+\; \crep,
  \qquad
  \crep \;\approx\; n\,(\cmint + \ccopy + e),
\]
where $n$ is the number of offspring per cycle and $e$ is the
\emph{endowment}---the initial token stack that travels with each copied
description $\quo P$. The endowment is literally metabolic: tokens removed from
the parent's budget and placed in the offspring's. The partition (maintenance
versus reproduction, and within reproduction $n$ versus $e$) is the life-history
allocation problem stated in the construction's own primitives.

\subsection{Carrying capacity as a fuel-balance fixed point and the emergence of the logistic law}

Let a bounded token inflow $\Phi$ (tokens per unit time) feed the substrate
through the splitter/joiner network, shared among $N$ identical replicators.
Realised per-process income is density-dependent through contention on the
bounded inflow. To leading order we take the contention response to be linear,
\begin{equation}
  B(N) \;=\; B_0\!\left(1 - \frac{N}{N_{\mathrm{sat}}}\right),
  \qquad
  N_{\mathrm{sat}} = \frac{\Phi}{B_0},
  \label{eq:contention}
\end{equation}
the first-order term of any saturating response that vanishes when aggregate
demand $N B_0$ exhausts the inflow. Per-capita birth rate is the reproductive
surplus divided by the unit cost of reproduction, and death has a baseline
density-independent rate $d_0$ (accident and exogenous shock) on top of
starvation:
\begin{equation}
  b(N) = \frac{\max\!\big(0,\; B(N)-\cmaint\big)}{\crep},
  \qquad d(N) = d_0 .
  \label{eq:rates}
\end{equation}

\begin{proposition}[Emergence of the logistic law]
\label{prop:logistic}
Under \eqref{eq:contention}--\eqref{eq:rates}, the mean-field population
dynamics of a set of cost-accounted replicating computations on a bounded-inflow
substrate, in the regime $B(N)\ge \cmaint$, obey the logistic
law~\cite{verhulst1838}
\[
  \frac{dN}{dt} \;=\; \big(b(N)-d_0\big)\,N \;=\; rN\!\left(1 - \frac{N}{K}\right),
\]
with intrinsic rate and carrying capacity given in closed form by the ledger
primitives,
\[
  r \;=\; \frac{B_0 - \cmaint}{\crep} - d_0,
  \qquad
  K \;=\; N_{\mathrm{sat}}\left(1 - \frac{\cmaint + d_0\,\crep}{B_0}\right)
        \;=\; \frac{\Phi}{B_0}\left(1 - \frac{\cmaint + d_0\,\crep}{B_0}\right).
\]
\end{proposition}

\begin{proof}[Derivation]
Substituting \eqref{eq:contention} into \eqref{eq:rates} and dropping the clamp
in the stated regime,
\[
  b(N) - d_0
  = \frac{B_0(1-N/N_{\mathrm{sat}}) - \cmaint}{\crep} - d_0
  = \underbrace{\left(\frac{B_0-\cmaint}{\crep}-d_0\right)}_{r}
    \;-\; \frac{B_0}{\crep\,N_{\mathrm{sat}}}\,N .
\]
Multiplying by $N$ gives $\dot N = rN - \big(B_0/(\crep N_{\mathrm{sat}})\big)N^2$,
which is $rN(1-N/K)$ with $K = r\,\crep N_{\mathrm{sat}}/B_0$. Expanding $r$ and
using $N_{\mathrm{sat}}=\Phi/B_0$ yields the stated closed form.
\end{proof}

\begin{remark}
The quadratic term---hence the entire logistic shape---comes from the
density-dependence of realised income, i.e.\ from contention on the bounded
inflow. The law is therefore \emph{derived}, not assumed. The qualitative
content---carrying capacity set by inflow over per-capita draw, intrinsic rate
set by surplus over reproductive unit cost---is robust to the precise contention
ansatz; the exact coefficient in $K$ depends on it. In the maintenance-dominated
small-$d_0$ limit the carrying capacity reduces to the bare fuel balance
$N^\ast\cmaint=\Phi$, i.e.\ $K\to\Phi/\cmaint$: as many processes as the inflow
can keep from starving. Density-dependent mortality is not imposed; it emerges
as the rising starvation rate when $N\cmaint$ approaches $\Phi$ and the splitters
can no longer top every stack.
\end{remark}

\subsection{\texorpdfstring{$r$}{r}- and \texorpdfstring{$K$}{K}-selection as corners of an allocation simplex}

Selection acts on the allocation, and which regime is favoured is set by which
term in the ledger binds; the $r$/$K$
dichotomy~\cite{macarthur1967,pianka1970} is the coarsest reading of it. The
criterion is the invasion analysis of
\S\ref{sec:mortality}; only the binding constraint differs.

\begin{itemize}[leftmargin=1.4em,itemsep=0.3em]
\item \textbf{$r$-corner.} Uncrowded or volatile substrate ($N\ll K$, death
dominated by exogenous supply shocks, density-independent) rewards maximising
$r=(B_0-\cmaint)/\crep-d_0$: push $n$ up, push $e$ and $\cmaint$ down. Many
minimal-endowment offspring, fire-and-forget, short-lived soma. This is the
immortal-replicator defector of \S\ref{sec:mortality}, now adaptive rather than
pathological, because in an empty substrate raw instantiation rate is what
invasion rewards.

\item \textbf{$K$-corner.} Saturated, stable substrate ($N\approx K$, death
dominated by competition, density-dependent) rewards competitive fuel capture:
push $e$ and $\cmaint$ up, push $n$ down. Richly-endowed offspring,
long-lived competitive soma; per-offspring capture beats raw count.
\end{itemize}

\paragraph{Corollaries in the primitives.}
Parental care is a persistent process that keeps feeding tokens to the
offspring's signature-channel after instantiation---a continued-interactive
contract, which is what ``continued'' is for; $r$-strategists cut the channel at
birth, $K$-strategists hold it open. Semelparity is the parent dumping its
residual budget into one reproductive burst and then deadlocking by design
($\cmaint\to 0$ post-burst); iteroparity retains maintenance to survive further
cycles. Body size maps to the soma's standing token stack---the capital a
process keeps on hand---and the familiar $r/K$ body-size correlation appears
because large standing capital repays its maintenance draw only when competition,
not throughput, binds.

\begin{remark}[The construction gives more than the dichotomy]
$r$ and $K$ are the two extreme vertices of a continuous allocation simplex over
$(\cmaint, n, e, \text{timing})$. That continuum is exactly modern
age-structured / disposable-soma life-history
theory~\cite{stearns1992,kirkwood1977}, which
$r/K$ coarsens into a binary. The construction natively produces the continuum,
so the right computational taxonomy of reproduction schemes is a classification
of points in the simplex with $r$ and $K$ as named corners---truer to the
biology and more natural in the construction than a dichotomy.
\end{remark}

\section{What to make rigorous next}
\label{sec:next}

Three load-bearing claims are stated above at the level a working note allows
and deserve full treatment.

\begin{enumerate}[leftmargin=1.6em,itemsep=0.35em]
\item \textbf{Idempotence of the round trip up to bisimulation.} If
$(\T\circ\C)^2 \bisim \T\circ\C$, then $\operatorname{im}(\T\circ\C)$ is exactly
the set of bisimulation fixed points of the metabolism comonad---the processes
already mortal and metabolising---and ``the organisms inside $\G$'' acquires a
crisp definition. The entire biological reading rests on this image being a
well-defined sublanguage rather than a construction artifact.

\item \textbf{The fidelity comparison (Prop.~\ref{prop:fidelity}).} Make precise
the sense in which the exogenous and internalised self-interpreters agree on
behaviour up to bisimulation while differing on the description algebra, and
prove that evolvability---benign mutation neighbourhood under structural
edits---is a property of the description algebra alone.

\item \textbf{The logistic-emergence proposition (Prop.~\ref{prop:logistic}).}
Promote the mean-field derivation to a theorem: specify the stochastic
token-flow process whose hydrodynamic limit is the stated ODE, identify the
conditions on the splitter/joiner network under which the contention response
\eqref{eq:contention} is the correct leading order, and characterise the
corrections that distinguish the exact $K$ from the bare fuel balance
$\Phi/\cmaint$. This is the technical spine of the programme.
\end{enumerate}

\vspace{0.5em}
\hrulethick

\begin{thebibliography}{99}
\itemsep0.35em

\bibitem{bowen1975}
R.~Bowen.
\newblock \emph{Equilibrium States and the Ergodic Theory of Anosov
  Diffeomorphisms}.
\newblock Lecture Notes in Mathematics, vol.~470. Springer, Berlin, 1975.

\bibitem{buss1987}
L.~W. Buss.
\newblock \emph{The Evolution of Individuality}.
\newblock Princeton University Press, Princeton, NJ, 1987.

\bibitem{crick1970}
F.~Crick.
\newblock Central dogma of molecular biology.
\newblock \emph{Nature}, 227(5258):561--563, 1970.

\bibitem{kirkwood1977}
T.~B.~L. Kirkwood.
\newblock Evolution of ageing.
\newblock \emph{Nature}, 270(5635):301--304, 1977.

\bibitem{macarthur1967}
R.~H. MacArthur and E.~O. Wilson.
\newblock \emph{The Theory of Island Biogeography}.
\newblock Princeton University Press, Princeton, NJ, 1967.

\bibitem{meredith2005}
L.~G. Meredith and M.~Radestock.
\newblock A reflective higher-order calculus.
\newblock \emph{Electronic Notes in Theoretical Computer Science},
  141(5):49--67, 2005.

\bibitem{milner1992}
R.~Milner, J.~Parrow, and D.~Walker.
\newblock A calculus of mobile processes, I.
\newblock \emph{Information and Computation}, 100(1):1--40, 1992.

\bibitem{pianka1970}
E.~R. Pianka.
\newblock On $r$- and $K$-selection.
\newblock \emph{The American Naturalist}, 104(940):592--597, 1970.

\bibitem{ruelle1978}
D.~Ruelle.
\newblock \emph{Thermodynamic Formalism}.
\newblock Encyclopedia of Mathematics and its Applications, vol.~5.
  Addison-Wesley, Reading, MA, 1978.

\bibitem{stearns1992}
S.~C. Stearns.
\newblock \emph{The Evolution of Life Histories}.
\newblock Oxford University Press, Oxford, 1992.

\bibitem{turing1936}
A.~M. Turing.
\newblock On computable numbers, with an application to the
  Entscheidungsproblem.
\newblock \emph{Proceedings of the London Mathematical Society},
  s2-42(1):230--265, 1937.

\bibitem{vonneumann1966}
J.~von Neumann.
\newblock \emph{Theory of Self-Reproducing Automata}.
\newblock Edited and completed by A.~W. Burks. University of Illinois Press,
  Urbana, IL, 1966.

\bibitem{verhulst1838}
P.-F. Verhulst.
\newblock Notice sur la loi que la population suit dans son accroissement.
\newblock \emph{Correspondance Math\'ematique et Physique}, 10:113--121, 1838.

\bibitem{weismann1893}
A.~Weismann.
\newblock \emph{The Germ-Plasm: A Theory of Heredity}.
\newblock Translated by W.~N. Parker and H.~R\"onnfeldt. Charles Scribner's
  Sons, New York, 1893.

\end{thebibliography}

\end{document}
