\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper,margin=1.1in}
\hypersetup{colorlinks=true,linkcolor=blue!55!black,citecolor=blue!55!black,urlcolor=blue!55!black}

%% --- notation ---------------------------------------------------------------
\newcommand{\sem}[1]{[\![#1]\!]}
\newcommand{\refl}[1]{\ulcorner #1 \urcorner}
\newcommand{\Proc}{\mathsf{Proc}}
\newcommand{\Name}{\mathsf{Name}}
\newcommand{\dT}{\partial T}
\newcommand{\Top}{\mathfrak{Top}}
\newcommand{\GSLT}{\mathbf{GSLT}}
\newcommand{\GSLTsem}{\mathbf{GSLT}_{\!\sim}}
\newcommand{\GSLTfp}{\mathbf{GSLT}^{\mathrm{fp}}_{\!\sim}}
\newcommand{\MeTTaIL}{\mathbf{MeTTaIL}}
\newcommand{\Cl}{\mathrm{Cl}}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\Ps}{\mathcal{P}}
\newcommand{\sH}{\widehat{\mathsf{H}}}
\newcommand{\Bfun}{\mathsf{B}}
\newcommand{\sB}{\widehat{\mathsf{B}}}
\newcommand{\swap}{\mathsf{s}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\Sym}{\mathrm{Sym}}
\newcommand{\sorts}{\mathcal{G}}
\newcommand{\bEs}{\overline{\mathcal{E}}}
\newcommand{\Ks}{\mathcal{K}}
\newcommand{\Av}{\Omega}
\newcommand{\for}[2]{\mathsf{for}(#1 \Leftarrow #2)}
\newcommand{\out}[2]{#1\,!\,(#2)}
\newcommand{\quo}[1]{@#1}
\newcommand{\drf}[1]{{*}#1}

%% --- colour coding of the knotted copies -----------------------------------
\definecolor{redcopy}{rgb}{0.78,0.08,0.08}
\newcommand{\rc}[1]{\textcolor{redcopy}{#1}}   % red copy
\newcommand{\rSet}{\rc{\mathsf{Set}}}          % red set theory
\newcommand{\bSet}{\overline{\mathsf{Set}}}    % black set theory
\newcommand{\Es}{\rc{\mathcal{E}}}             % red copy topos
\newcommand{\rX}{\rc{X}}
\newcommand{\Ss}{\rc{S}}
\newcommand{\As}{\rc{A}}
\newcommand{\bSs}{\overline{S}}
\newcommand{\bAs}{\overline{A}}
\newcommand{\St}{\rc{\mathcal S}}
\newcommand{\At}{\rc{\mathcal A}}
\newcommand{\bSt}{\overline{\mathcal S}}
\newcommand{\bAt}{\overline{\mathcal A}}

%% --- theorem environments ---------------------------------------------------
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{obligation}{Obligation}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\title{Knotted Topoi:\\
the lift of the knotted set-theoretic universe, and\\
fully abstract denotational semantics for the\\
category of graph-structured lambda theories}
\author{Lucius Gregory Meredith\\
CEO, F1R3FLY.io, and Chief Science Officer, F1R3FLY Industries\\
\texttt{lgreg.meredith@gmail.com}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
We lift the knotted set-theoretic universe of \emph{A Knotted Universe} --- two mutually recursive
copies of set theory tied so that each colour's atoms are the other colour's sets --- one
categorical level, to a \emph{knotted topos}: four sort-topoi (red sets and atoms, black sets and
atoms) tied by two geometric knots and an involutive colour-swap, with the reflection law
$\drf{}\quo{}P\equiv P$ become the involutivity of an equivalence of topoi. This is the principal
result. As an application we transport the operational view of computation to a denotational one.
The MeTTaIL language presents the category of finitely presentable graph-structured lambda theories
(GSLTs); desugaring MeTTaIL into the rho calculus, and composing with the fully abstract rho
semantics of \emph{Quoting is Colour-Swap} lifted into the knotted topos, gives a compositional
denotation $\sem{-}_{\Ks}$ that is fully abstract for context bisimulation:
$\sem{P}_{\Ks}=\sem{Q}_{\Ks}\iff P\sim Q$. Because GSLTs subsume every known classical model of
computation --- the $\lambda$-calculus and hence PCF, CCS, the $\pi$-calculus~\cite{sangiorgiwalker}, the ambient calculus,
SKI, Turing machines, and term rewriting, with universal algebra as the dynamics-free slice --- the
construction supplies a fully abstract denotational semantics for all of them at once, at least in
the finitely presentable case. When Milner found the full abstraction problem for PCF he is said to
have quipped, ``so much for denotational semantics''; the linear-logic and game-semantics programme
answered him for PCF, and the knotted topos answers him uniformly: the model is the bisimulation
quotient itself --- the final coalgebra of a behaviour functor internal to the topos --- so it is
fully abstract by construction, compositionality being the theorem and the context-labelled
(Leifer--Milner) presentation being what makes bisimulation a congruence. The keystone is that the
OSLF-generated Hennessy--Milner logic is the internal logic of the knotted topos: adequacy is the
statement that the subobject classifier classifies bisimilarity. The risk ledger of the source
papers transfers: finitely presentable theories cost no infinite risk, recursion costs one $\omega$
--- black infinity --- and not one primitive name.
\end{abstract}

\tableofcontents

\section{Introduction}\label{sec:intro}

\subsection{Milner's lament}

In the nineteen-seventies Milner and Plotkin gave PCF --- a typed $\lambda$-calculus with arithmetic
and recursion --- both an operational and a denotational semantics, the latter in Scott's domains,
and asked whether the two agree exactly: whether two terms receive the same denotation precisely when
no program context can tell them apart. They do not. The Scott model is not \emph{fully abstract};
the celebrated witness is ``parallel or'', a function definable in the model but not in the language,
so the model distinguishes terms the language equates~\cite{milnerfa,plotkinlcf,scottdomains}. The
\emph{full abstraction problem} for PCF --- to build a model that is neither too coarse nor too fine
--- proved deep and durable, and Milner is said to have summarised the discovery with a quip: ``so
much for denotational semantics.''

The linear-logic and game-semantics programme answered him for PCF two decades later. Reading a type
as a game and a program as a strategy, and cutting the model down to the strategies that are
\emph{innocent} and \emph{history-free}, Abramsky--Jagadeesan--Malacaria and Hyland--Ong built
genuinely fully abstract models of PCF~\cite{ajm,ho}, in the spirit of Girard's linear
logic~\cite{girardll}: definability is recovered exactly, the parallel-or is excluded by the
discipline on plays, and observational equivalence is, on the nose, equality of strategies modulo the
game's own equivalence. The lesson of that programme is that the right denotational object is an
\emph{interaction} structure --- plays between a player and an opponent --- rather than a space of
values.

\subsection{The broader result}

This paper makes the game-semantic moral uniform, and in doing so answers Milner not for one language
but for a category that contains them all. A \emph{graph-structured lambda theory}
(GSLT)~\cite{gsltnotes} is a triple (grammar, equations, rewrites): freely generated terms as states,
equations identifying states, rewrites animating them. The class of GSLTs subsumes every known
classical model of computation. The $\lambda$-calculus is the GSLT with $\beta$ as its base rewrite,
and PCF is that GSLT enriched with constants, conditionals and a fixpoint; CCS, the $\pi$-calculus,
the ambient calculus and the rho calculus are GSLTs whose base rewrite is communication; SKI is a
GSLT of combinator contraction; a Turing machine is a GSLT of tape rewrites; and a term-rewriting
system is a GSLT outright~\cite{baadernipkow}. Universal algebra and the Lawvere theories sit inside as the
dynamics-free slice, where the rewrite component is empty~\cite{gsltnotes}.

For all of these --- at least the finitely presentable ones --- a single construction supplies a
fully abstract denotational semantics. The construction does not search for a model that happens to
be neither too coarse nor too fine; it takes the model to \emph{be} the bisimulation quotient, the
final coalgebra of a behaviour functor, and proves that this quotient is a compositional, internally
topos-theoretic object. Full abstraction is then by construction: the failure that defeated the
Scott model --- a model element with no program realising it --- cannot occur, because every element
of the final coalgebra is the behaviour of a process, and two processes are identified exactly when
bisimilar. What had to be \emph{proved} is the other half, compositionality: that the denotation is a
homomorphism and that bisimulation is a congruence. The second source paper supplies precisely this,
by labelling transitions with contexts after Leifer and Milner~\cite{leifermilner,sewell}, so that
the bisimulation game already ranges over every environment and congruence holds with no closure
step~\cite{knotrho}. The interaction structure the game semanticists found necessary is, here, built
in from the ground: the knot reads red as player and black as opponent~\cite{rset}, and the
denotation lives in their interaction.

\subsection{The two parts}

This is the third in a sequence. \emph{A Knotted Universe}~\cite{rset} builds a universe of reflective
set theories from $\varnothing$ alone, as a red copy $\rSet[\rX]$ and a black copy
$\bSet[\overline X]$ tied by $\rX=\bSet[\overline X]$, $\overline X=\rSet[\rX]$, each colour's atoms
the other's sets. \emph{Quoting is Colour-Swap}~\cite{knotrho} gives a fully abstract denotation of
the rho calculus~\cite{rho} there, refining the knot into four sorts with quoting realised as a
colour-swap and proving full abstraction for the resulting final-coalgebra semantics. The present
paper has a principal result and an application.

\paragraph{Principal result (\S\ref{sec:lift}): the lift to knotted topoi.} We categorify the
four-sorted model. The four sorts become four topoi
$\St,\At,\bSt,\bAt$; the two knots become geometric equivalences; the colour-swap becomes an
involutive auto-equivalence with $\swap\circ\swap\cong\id$; quoting and dereference are derived,
$\quo{}=\kappa\circ\swap$, $\drf{}=\swap\circ\kappa^{-1}$, and $\drf{}\quo{}P\equiv P$ becomes the
involutivity of an equivalence of topoi. The knotted topos $\Ks$ is the (bi)fixpoint of the
two-sorted endofunctor of~\cite[\S8.1]{rset} read in the $2$-category $\Top$ of Grothendieck topoi.
Its finitary core is set-sized, its full version class-sized, exactly along the ledger line of the
source papers. The behaviour functor lifts to an internal endofunctor whose final coalgebra is the
process object inside $\Ks$, where colour-blind non-well-foundedness supplies recursion and finality
makes bisimilar processes internally equal. The OSLF-generated logic is the internal logic of $\Ks$.

\paragraph{Application (\S\ref{sec:app}): operational to denotational.} MeTTaIL (rholang
$1.4$)~\cite{mettail} --- a language of types, equations-on-types and rewrites-on-types --- presents
the category $\GSLTfp$ of finitely presentable GSLTs: a MeTTaIL program \emph{is} a finite GSLT, and
a MeTTaIL language map is a GSLT morphism. Desugaring MeTTaIL into core rho (\S\ref{sec:desugar},
Appendix~\ref{app:desugar}) installs each base rewrite $L\Rightarrow R$ as a guarded receiver at the
channel naming the redex's location,
\[
\sem{L\Rightarrow R}(c)\;=\;\for{\sem{L}}{c}\bigl\{\;\out{c}{\sem{R}}\;\bigr\},
\]
and composing with the rho denotation lifted into $\Ks$ yields a compositional
$\sem{-}_{\Ks}\colon\GSLTfp\to\Ks$ that is fully abstract for context bisimulation. This is the
transport of the operational view to the denotational view, and the fully abstract model promised
above.

\subsection{Two standing conventions}

MeTTaIL is used as the legible surface for writing rules and as the presentation of $\GSLTfp$; it is
\emph{eliminable sugar} over core rho, with its equation component factored through Church encodings,
and the topos-theoretic recapitulation is carried out in core rho after the sugar is compiled away.
Nothing in the four-topos picture depends on a MeTTaIL primitive. As to optimality: the channel
$c$ above is the sound, non-optimal reflection of the location, not the optimal set-automaton state of
\cite{optimal}; optimality is in tension with what a channel must do here (keep distinct runtime
locations distinct), is unneeded for the lift, and is deferred together with the channel-naming
correction noted in~\cite{optimal}. The optimal and the present scheme induce the \emph{same}
context-labelled transition system, so the choice is invisible to everything downstream.

We write in the programmatic-with-obligations style of the source papers~\cite{rset,knotrho}:
definitions and constructions are pinned down, the load-bearing existence, coherence and calibration
facts are collected as numbered obligations (\S\ref{sec:obligations}), and for the finitary core each
reduces to a statement about hereditarily finite data.

\section{Recap}\label{sec:recap}

We fix notation; \cite{rset,knotrho,gsltnotes} carry the details.

\subsection{The knotted universe and its four sorts}

The red and black copies of set theory are tied by $\rX=\bSet[\overline X]$,
$\overline X=\rSet[\rX]$. \cite[\S8.1]{rset} exhibits the tie as the fixpoint of the two-sorted
endofunctor on a category of classes
\begin{equation}\label{eq:H}
\mathsf{H}(R,B)\;=\;\bigl(\,B + \Ps R,\;\; R + \Ps B\,\bigr),
\end{equation}
whose initial algebra $(\rX,\overline X)$ satisfies $\rX\cong\overline X+\Ps \rX$,
$\overline X\cong \rX+\Ps\overline X$. The initial algebra is the well-founded reading (ZFA/FM per
colour); the final coalgebra is the non-well-founded reading, in which colour-blind membership
$\in_\star$ admits alternating-colour infinite descent~\cite[\S7]{rset}. The model
$M$ of~\cite[Thm.~1]{rset} is a nominal set~\cite{gabbaypitts,pitts}, hence lives in the Schanuel topos~\cite[\S6.1]{rset}.

\cite{knotrho} keeps four sorts distinct: red sets $\Ss$, red atoms $\As$, black sets $\bSs$, black
atoms $\bAs$, with knots $\kappa\colon\bSs\xrightarrow{\cong}\As$,
$\overline\kappa\colon\Ss\xrightarrow{\cong}\bAs$, and a polymorphic involution $\swap$ with
$\swap\circ\swap=\id$, homomorphic for the set operations and coherent,
$\swap\circ\kappa=\overline\kappa\circ\swap$. With processes red sets and names red atoms,
\begin{equation}\label{eq:quote}
\quo{}:=\kappa\circ\swap\colon\Ss\to\As,\qquad \drf{}:=\swap\circ\kappa^{-1}\colon\As\to\Ss,
\end{equation}
and $\drf{}\quo{}P\equiv P$ is the calculation
$\swap\circ\kappa^{-1}\circ\kappa\circ\swap=\swap\circ\swap=\id$~\cite[Prop.~6]{knotrho}.

\subsection{The rho model}

The rho grammar is $P::=0\mid P\mid Q\mid \for{y}{x}P\mid \out{x}{Q}\mid \drf{}x$, $x::=\quo{}P$;
there are no primitive names and no restriction $\nu$, and recursion enters through reflection rather
than replication~\cite[\S2]{knotrho}. Transitions are labelled by contexts $P\xrightarrow{F}P'$,
$F$ a minimal environment (an idem-pushout after Leifer--Milner) in which $P$ completes a reaction.
The behaviour functor is $\Bfun(R)=\Ps(\dT_K\times R)$, $\dT_K$ the Fire context-labels; the process
universe is $\Proc=\nu\Bfun$, and bisimilar processes are equal in it. Full abstraction
$\sem{P}=\sem{Q}\iff P\sim Q$ holds for a congruence, read through OSLF adequacy as agreement of
$\dT_K$-indexed Hennessy--Milner formulae~\cite[\S6]{knotrho}.

\subsection{GSLTs, the K-family, and the three views of \texorpdfstring{$\dT$}{dT}}

A GSLT is (grammar, equations, rewrites)~\cite{gsltnotes}. An \emph{interactive} GSLT carries a
distinguished family $K$ of constructors --- the heads of the hypothesis-free (\emph{base}) rewrites,
which are the interactions; rules with hypotheses are propagation discipline. A \emph{location} is a
splitting $(k,t)$ of a term into a context $k$ and a subterm $t$, with the cut between them as the
place; the splittings $\dT\times T$ fibre over $T$ by plugging, and the sub-bundle \emph{Fire}
consists of those whose subterm matches a base LHS. The one object $\dT_K$ is read three ways: as
Firing labels, as Witnessing indices of OSLF modalities $\langle F\rangle\varphi$, and as Placing of
interactions in term structure. The two categories of GSLTs differ in what morphisms preserve: the
\emph{syntactic} category preserves term structure, the \emph{semantic} category $\GSLTsem$ preserves
bisimulation. We work throughout with the semantic category, and write $\GSLTfp$ for its full
subcategory of finitely presentable objects.

\section{Principal result: the knotted topoi}\label{sec:lift}

Throughout $\Top$ is the $2$-category of Grothendieck topoi, geometric morphisms, and geometric
transformations; $f^*\dashv f_*$ is a geometric morphism and $\simeq$ equivalence of
topoi~\cite{maclane,elephant}.

\subsection{Four topoi, two knots, one swap}

\begin{definition}[The sort topoi]\label{def:sorttopoi}
The four sorts lift to four topoi --- $\St$ (red sets), $\At$ (red atoms), $\bSt$ (black sets),
$\bAt$ (black atoms) --- organised as a fibration $p\colon\Ks\to\sorts$ over the four-object index
category $\sorts$ whose objects are the sorts and whose non-identity arrows are the kind-coercions
$\At\to\St$, $\bAt\to\bSt$ together with the colour edges. The fibre over a sort is that sort's topos;
the total category $\Ks$ is the \emph{knotted topos}, and $p$ is its \emph{grading}.
\end{definition}

\begin{definition}[Knots as geometric coercions]\label{def:knots}
The two knots lift to geometric equivalences
\[
\kappa\colon \bSt\ \xrightarrow{\ \simeq\ }\ \At,\qquad
\overline\kappa\colon \St\ \xrightarrow{\ \simeq\ }\ \bAt,
\]
presenting a black-sets topos as the red-atoms topos and vice versa. The presentation is
\emph{opaque}: the geometric inclusion of an atom topos into the ambient sheaf topos forgets the other
colour's interior, exactly the forgetfulness by which the atoms of~\cite[\S8.1]{rset} ``acquire no
members''. Opacity is colour-opacity~\cite[Rem.~2]{rset}.
\end{definition}

\begin{definition}[The swap as an involutive auto-equivalence]\label{def:swap}
The colour symmetry lifts to a polymorphic, sort-respecting equivalence
\[
\swap\colon\St\simeq\bSt,\qquad \swap\colon\At\simeq\bAt,\qquad \swap\circ\swap\cong\id,
\]
homomorphic for the geometric structure (initial object, coproducts, subobject classifier), with the
coherence $2$-cell
\begin{equation}\label{eq:cohere}
\swap\circ\kappa\ \cong\ \overline\kappa\circ\swap \qquad(\text{both }\bSt\to\bAt).
\end{equation}
\end{definition}

\begin{proposition}[Reflection, lifted; sketch]\label{prop:reflift}
Defining $\quo{}:=\kappa\circ\swap\colon\St\to\At$ and $\drf{}:=\swap\circ\kappa^{-1}\colon\At\to\St$
as geometric morphisms, one has $\drf{}\circ\quo{}\cong\swap\circ\swap\cong\id$ and
$\quo{}\circ\drf{}\cong\id$, so the model validates $\drf{}\quo{}P\equiv P$ and $\quo{}\drf{}x\equiv_N
x$ not by an imposed equation but as the involutivity of an equivalence of topoi.
\end{proposition}

\begin{proof}[Proof sketch]
The middle isomorphisms are the unit/counit of the knot equivalence (Definition~\ref{def:knots}) and
the involutivity $2$-cell (Definition~\ref{def:swap}); coherence is~\eqref{eq:cohere}. This is
\cite[Prop.~6]{knotrho} one level up, equality of maps replaced by canonical isomorphism; the triangle
and pentagon coherences are Obligation~\ref{ob:cohere}.
\end{proof}

\subsection{The knotted topos as a bifixpoint}

\begin{construction}[The knot $2$-functor]\label{con:knot2}
Let $\Ps\colon\Top\to\Top$ be the $2$-functor sending a topos to its topos of internal sheaves on the
canonical (power) site --- the topos-level analogue of the powerclass operator of~\eqref{eq:H}
\cite{jm,maclane} --- and let $\ltimes$ denote the atom-coproduct adjoining one topos to another as a
topos of opaque urelements (the geometric analogue of $A+(-)$, Definition~\ref{def:knots}). Define the
endo-$2$-functor on $\Top\times\Top$
\begin{equation}\label{eq:H2}
\sH(\mathcal R,\mathcal B)\;=\;\bigl(\,\mathcal B\ltimes\Ps\mathcal R,\;\;\mathcal R\ltimes\Ps\mathcal B\,\bigr).
\end{equation}
A \emph{red/black pair of knotted topoi} is a (pseudo-)fixpoint $(\Es,\bEs)$ of $\sH$:
\[
\Es\ \simeq\ \bEs\ltimes\Ps\Es,\qquad \bEs\ \simeq\ \Es\ltimes\Ps\bEs,
\]
each colour's topos its own sheaves with the other's as opaque atoms. The four-sort fibration of
Definition~\ref{def:sorttopoi} is read off this pair: $\St$ and $\At=\kappa\bSt$ from $\Es$, and
$\bSt$, $\bAt=\overline\kappa\St$ from $\bEs$.
\end{construction}

\begin{remark}[Which fixpoint: the knob is the source papers']\label{rem:knob}
The \emph{initial} algebra of $\sH$ is the well-founded reading; each colour's topos models ZFA/FM
internally, the picture of~\cite[\S5]{rset} lifted. The \emph{final} coalgebra is the
non-well-founded reading, admitting the alternating-colour descent of~\cite[\S7]{rset}, and is the
reading the process semantics needs (\S\ref{sec:proc}). For the \emph{finitary} core --- $\Ps$
replaced by its finite-power analogue --- the fixpoint exists at the level of topoi as the
colimit/limit of the evident $\omega$-chain, set-sized; for the full $\Ps$ it lives in the class-sized
framework, along the finitary/infinitary ledger line of~\cite[\S5.5--5.6, \S8.1]{rset}. Existence is
Obligation~\ref{ob:exist}.
\end{remark}

\begin{remark}[The grading is the modality, internalised]\label{rem:grading}
\cite[\S6.5]{rset} and~\cite[Rem.~8]{knotrho} ask for a graded modality separating colour-respecting
(equivariant) from reflective computation. The fibration $p\colon\Ks\to\sorts$ \emph{is} that grading:
a computation over a single colour $\{\St,\At\}$ using only $\quo{},\drf{}$ never observes black
structure --- the equivariant fragment, ordinary nominal names with every guarantee intact --- while
applying $\swap$ to a name's interior crosses a colour edge, and reindexing along that edge records the
reflective surplus. Decidable structural identity of names is the decidability of equality in $\At$,
settled out of red's sight by black equality of swapped objects.
\end{remark}

\begin{proposition}[Two dual rho calculi in one universe; sketch]\label{prop:dual}
With the black reflection $\quo{}_{\!\overline{\phantom{a}}}:=\overline\kappa\circ\swap$, coherence
\eqref{eq:cohere} gives $\quo{}_{\!\overline{\phantom{a}}}\cong\swap\circ\quo{}\circ\swap$: the two
colours' quotes are swap-conjugate, one geometric reflection determining the other. The knotted topos
carries two dual rho calculi at once, recovering the player/opponent duality of~\cite[Cor.~2]{rset} at
the level of topoi --- the interaction structure the game-semantic full abstraction of
\cite{ajm,ho} required.
\end{proposition}

\subsection{The behaviour endofunctor and the process object}\label{sec:proc}

\begin{definition}[Internal behaviour endofunctor and process object]\label{def:Blift}
Let $\dT_K\in\Ks$ be the object of Fire context-labels (\S\ref{sec:logic} identifies it with the
site). Define $\sB\colon\Ks\to\Ks$, $\sB(R)=\Ps(\dT_K\times R)$, with $\Ps$ the power object; its
commitment refinement $\sB^\flat$ --- finite powerobject, products, coproducts, the name-abstraction
functor $[\,\mathcal N\,](-)$ --- presents the same behaviour finitely, after Milner~\cite{milnerpoly,knotrho}. The
\emph{process object} is $\Proc:=\nu\sB$, the final $\sB$-coalgebra internal to $\Ks$, with
$\Name:=\swap(\Proc)$ presented as $\At$ via $\kappa$. For $\sB^\flat$ the final coalgebra exists at
set size on the finitary core; for the full $\sB$ it is the class-sized, colour-blind non-well-founded
object of Remark~\ref{rem:knob}, whose alternating-colour descent supplies the recursion of process
behaviour for free, where Aczel's non-well-founded model imports it by axiom~\cite{aczel}.
\end{definition}

\begin{proposition}[Finality internalised; sketch]\label{prop:final}
For the weak-pullback-preserving $\sB^\flat$, internal coalgebraic bisimilarity is the kernel of the
unique morphism to $\Proc$; hence on $\Proc$ itself two elements are context-bisimilar iff internally
equal. Bisimulation is the equality of the process object in $\Ks$.
\end{proposition}

\begin{proof}[Proof sketch]
$\sB^\flat$ is assembled from constants, products, coproducts, finite powerobject and the abstraction
functor, each preserving weak pullbacks internally; universal coalgebra~\cite{rutten} identifies
behavioural equivalence with the kernel of $!\colon C\to\nu\sB^\flat$. This is
\cite[Lem.~17, Cor.~20]{knotrho} read internally to $\Ks$; the internal weak-pullback preservation of
the abstraction functor and the agreement of $\nu\sB^\flat$ with the class-sized $\nu\sB$ on equality
are part of Obligation~\ref{ob:classify}.
\end{proof}

\subsection{The lift, as a theorem}

\begin{theorem}[The lift]\label{thm:lift}
The four-sorted model of~\cite{knotrho} lifts to $\Top$: there is a knotted topos $\Ks$, fibred over
the four-sort index category $\sorts$, carrying geometric knots $\kappa,\overline\kappa$ and an
involutive colour-swap $\swap$ with $\swap\circ\swap\cong\id$, hence derived reflection
$\quo{}=\kappa\circ\swap$ and $\drf{}=\swap\circ\kappa^{-1}$ with $\drf{}\quo{}\cong\id$
(Proposition~\ref{prop:reflift}); presented as the $(\text{bi})$fixpoint of $\sH$
$(\text{Construction~\ref{con:knot2}})$, set-sized on the finitary core and class-sized in general
$(\text{Remark~\ref{rem:knob}})$; equipped with an internal behaviour endofunctor $\sB$ whose final
coalgebra $\Proc$ is an object of $\Ks$ in which context bisimulation is internal equality
$(\text{Proposition~\ref{prop:final}})$. Modulo Obligations~\ref{ob:exist}--\ref{ob:classify}, $\Ks$
is the topos-level knotted universe.
\end{theorem}

\section{Application: from operational to denotational semantics}\label{sec:app}

We now use Theorem~\ref{thm:lift} to transport the operational view of the finitely presentable GSLTs
to a denotational one, fully abstractly.

\subsection{MeTTaIL presents the finitely presentable GSLTs}

\begin{definition}[MeTTaIL as a presentation]\label{def:mettail}
A MeTTaIL program~\cite{mettail} is a triple of a finite collection of syntactic types (a grammar), a
finite collection of equations on types, and a finite collection of rewrites on types. This is exactly
the data of a finitely presentable GSLT. Sending a program to the GSLT it specifies, and a MeTTaIL
language map (a translation of types respecting equations and rewrites) to the induced GSLT morphism,
gives a functor $\MeTTaIL\to\GSLTfp$ that is essentially surjective: \emph{MeTTaIL presents the
category of finitely presentable GSLTs}. We work with GSLTs through this presentation, writing rules
in MeTTaIL and compiling them to core rho.
\end{definition}

\subsection{Desugaring MeTTaIL into rho}\label{sec:desugar}

The schema for a base rewrite is~\eqref{eq:quote}'s companion,
\begin{equation}\label{eq:base}
\sem{L\Rightarrow R}(c)\;=\;\for{\sem{L}}{c}\bigl\{\;\out{c}{\sem{R}}\;\bigr\},
\end{equation}
a receiver consuming a match of $\sem L$ at channel $c$ and re-emitting $\sem R$ there. The MeTTaIL
pattern-receive $\for{\sem L}{c}$ is the sugar; under the embedding into core rho it unfolds to nested
single-name receives with name-equality guards~\cite[\S2.2]{optimal}, and the equation component of
the source program compiles to Church-encoded normalisation, i.e.\ to structural congruence ---
colour-respecting, cost-free, \emph{iso} in the target rather than motion. The channel $c$ names the
runtime location.

\begin{definition}[Location channels]\label{def:loc}
A \emph{location} is a finite path $\ell=\langle(f_1,i_1),\dots,(f_m,i_m)\rangle$ from the root of a
term; step $(f,i)$ descends into the $i$-th argument of an $f$-node, $\varepsilon$ is the root, and
$\ell\cdot(f,i)$ the child. The channel is the reflection $c(\ell):=\refl{\ell}\in\As$; distinct
locations give distinct channels by injectivity of $\refl{\cdot}$.
\end{definition}

\begin{remark}[Freshness by quoting]\label{rem:fresh}
Distinct injected terms must not share a root, lest $\out{a}{0}\mid\out{a}{0}$ collapse to
$\out{a}{0}$ --- the identification $\Proc$'s parallel is built to avoid~\cite[\S5]{knotrho}.
Freshness is supplied as rho supplies all freshness, by quoting~\cite[\S2]{knotrho}: an injection
publishes on a root $\refl{\rho}$ for a quoted nonce $\rho$, and locations within it are
$\refl{(\rho,\ell)}$. No $\nu$, no central allocator. We suppress the root below.
\end{remark}

\begin{construction}[The desugaring functor]\label{con:functor}
For a finitely presentable GSLT $G$ presented in MeTTaIL,
$\sem{G}=\bigm|_{(L\Rightarrow R)\in\mathcal R}\sem{L\Rightarrow R}$ is the parallel composition of the
persistent installations of~\eqref{eq:base} at all matching locations (Appendix~\ref{app:desugar}),
with equations compiled to structural congruence. A $\GSLTfp$-morphism goes to the induced rho-context
map on installed locations.
\end{construction}

\begin{proposition}[Operational correspondence; sketch]\label{prop:opcorr}
The context-labelled transition system of $\sem{t}$ in $\sem{G}$ is bisimilar to the rewrite
transition system of $t$ in $G$: each base-rewrite firing of $t$ at location $\ell$ is matched by a
$\sB$-transition of $\sem{t}$ labelled $c(\ell)$, and conversely. Hence $\sem{-}$ is a functor
$\GSLTfp\to\sB\text{-}\mathbf{Coalg}(\Ks)$ factoring through the bisimulation quotient.
\end{proposition}

\begin{proof}[Proof sketch]
$\Leftarrow$ is~\eqref{eq:base}: a fire at $\ell$ is a rendezvous on $c(\ell)$. $\Rightarrow$ uses
injectivity and plugging-stability of $c(\cdot)$ (Definition~\ref{def:loc}), so no spurious rendezvous
occurs and the minimal enabling contexts are exactly the location channels; the labels coincide with
$\dT_K$, and the context-bisimulation game~\cite[Def.~3]{knotrho} is won by the identity on locations.
Full statement: Obligation~\ref{ob:opcorr}.
\end{proof}

\begin{remark}[Non-optimality]\label{rem:nonopt}
This desugaring re-inspects symbols enclosing a redex, failing the symbol-once condition (O1)
of~\cite{optimal} while satisfying (O2),(O3) trivially --- it is the ``reflect the context verbatim''
choice of~\cite[\S3.2]{optimal} lifted to the full runtime location. The optimal set-automaton scheme~\cite{bouwmanerkens} recovers (O1);
its deferred channel-naming correction and the present scheme induce the \emph{same} context-labelled
transition system, so Proposition~\ref{prop:opcorr} and all below are indifferent to the choice.
\end{remark}

\subsection{Lifting the rho denotation, and full abstraction}\label{sec:fullabs}

By Proposition~\ref{prop:opcorr} a presented GSLT denotes a $\sB$-coalgebra; by finality
(Proposition~\ref{prop:final}) it denotes a sub-coalgebra of $\Proc=\nu\sB$ in $\Ks$. Write
$\sem{-}_{\Ks}$ for the composite.

\begin{theorem}[Fully abstract denotation]\label{thm:fullabs}
For every finitely presentable GSLT $G$ and processes $P,Q$ of $G$,
\[
\sem{P}_{\Ks}=\sem{Q}_{\Ks}\ \text{ in }\Ks \qquad\Longleftrightarrow\qquad P\sim Q,
\]
where $\sim$ is context bisimulation. The denotation is compositional --- a homomorphism for the
GSLT's constructors --- and $\sim$ is a congruence by construction
$(\text{the context labels are idem-pushouts~\cite{leifermilner,sewell,knotrho}})$, so this is full
abstraction with respect to a congruence, with no closure step.
\end{theorem}

\begin{proof}[Proof sketch]
$\sem{-}_{\Ks}$ is the unique morphism to the final coalgebra (Proposition~\ref{prop:final}); its
kernel is internal coalgebraic bisimilarity, which by~\cite[Lem.~16]{knotrho} read internally is
context bisimilarity. Compositionality is the homomorphism property of the rho denotation
\cite[Thm.~15]{knotrho} transported along the desugaring (Proposition~\ref{prop:opcorr}); congruence
is~\cite[Prop.~4]{knotrho}. The whole is~\cite[Thm.~18]{knotrho} internalised to $\Ks$.
\end{proof}

\begin{corollary}[Milner answered, uniformly]\label{cor:milner}
Every finitely presentable classical model of computation receives a fully abstract denotational
semantics in the knotted topos: the $\lambda$-calculus and hence PCF, CCS, the $\pi$-calculus, the
ambient calculus, SKI, Turing machines, and finite term-rewriting systems, with universal algebra as
the dynamics-free slice~\cite{gsltnotes}. The model is the bisimulation quotient itself, so it is fully
abstract by construction; the failure that defeats the Scott model of PCF~\cite{milnerfa,plotkinlcf}
--- a model element with no realising program --- cannot occur. The residual work per language is the
identification of context bisimulation with that language's own observational equivalence
$(\text{e.g.\ contextual equivalence for PCF})$: this is the barbed-congruence calibration of
Obligation~\ref{ob:barbed}, and recasts the classical full-abstraction problem from a
model-\emph{construction} problem into a model-\emph{calibration} one. The game-semantic models of
\cite{ajm,ho} are the equivariant, colour-respecting fragment $(\text{Remark~\ref{rem:grading}})$,
their player/opponent structure the red/black knot.
\end{corollary}

\subsection{The keystone: OSLF is the internal logic}\label{sec:logic}

\begin{definition}[OSLF classifying topos and internal modalities]\label{def:cl}
For a presented GSLT $G$, let $\dT_K(G)$ be its Fire context-labels, a site under the minimal-context
(idem-pushout) covers; put $\Cl(G):=\Sh(\dT_K(G))$. For each label $F$ let
$\langle F\rangle\colon\Av\to\Av$ be the internal operator on the subobject classifier sending
$\varphi$ to ``some $F$-labelled transition reaches a state satisfying $\varphi$'', defined by image
along the $F$-component of the coalgebra structure map. The family $\{\langle F\rangle\}_{F\in\dT_K}$
is the internal Hennessy--Milner logic of $G$. The assignment $\Cl$ extends to a $2$-functor
$\GSLTsem\to\Top$, sending a bisimulation-preserving translation to a geometric morphism: the
\emph{topos-theoretic presentation of the semantic category of GSLTs}, with $\Cl(\text{rho})=\Ks$ the
universal target through which every desugaring factors.
\end{definition}

\begin{theorem}[Adequacy is classification; sketch]\label{thm:keystone}
The internal equality of $\Proc$ coincides with agreement under all internal modal formulae:
\[
\sem{P}_{\Ks}=\sem{Q}_{\Ks}\iff P\sim Q\iff \forall\varphi.\ (P\models\varphi\Leftrightarrow Q\models\varphi).
\]
Equivalently the subobject classifier $\Av$ of $\Ks$, with the modalities of Definition~\ref{def:cl},
\emph{classifies bisimilarity}: the OSLF-generated Hennessy--Milner logic is the internal logic of the
knotted topos. The three readings of $\dT_K$ become site (Placing), labels (Firing), and modal indices
(Witnessing) --- three roles of one object of the topos. ``Topos-theoretic presentation of GSLTs'' and
``OSLF adequacy'' are one statement.
\end{theorem}

\begin{proof}[Proof sketch]
The first equivalence is Theorem~\ref{thm:fullabs}; the second is OSLF
adequacy~\cite[Cor.~19]{knotrho} read internally. That the internal modal theory is the
subobject-classifier logic of $\Sh(\dT_K)$ is Obligation~\ref{ob:classify}.
\end{proof}

\subsection{The risk ledger, lifted}\label{sec:ledger}

A finitely presentable GSLT with finite synchronisation trees is a hereditarily finite object of the
finitary-core topos, where $\nu\sB^\flat$ is literal and set-sized; its commitment is the black empty
set and one successor, no infinite risk. Recursion --- reached through reflection, with no primitive
replication --- is a non-well-founded inhabitant of $\Proc$, available only in the infinitary reading,
which by Remark~\ref{rem:knob} costs exactly one $\omega$, black infinity, and lives in the class-sized
framework. No primitive name is ever reintroduced. ``Finitely presentable'' is therefore not a
blemish on Corollary~\ref{cor:milner} but the exact regime in which the accounting is cleanest: the
finite presentation is the finitary core, and only genuine recursion crosses into the one-$\omega$
ledger.

\section{Toward a rigorous construction}\label{sec:obligations}

In the manner of~\cite[\S8]{rset} and~\cite[\S8]{knotrho}. For the finitary core each obligation
reduces to a statement about hereditarily finite data.

\begin{obligation}[Existence of the knotted topos]\label{ob:exist}
Exhibit the (bi)fixpoint of $\sH$ of~\eqref{eq:H2} in $\Top$: for the finitary $\Ps$ as the set-sized
colimit/limit of the $\omega$-chain, each $\sH^n(\varnothing)$ a topos; for the full $\Ps$ in the
class-sized framework. The natural route is a bicategorical algebraic set theory~\cite{jm,awodey,elephant}
run two-sorted on $\sH$, mirroring~\cite[\S8.1]{rset}; an explicit alternative is the \emph{knotted
site}, two sites each of whose atom-objects are the other's representables.
\end{obligation}

\begin{obligation}[Coherence of swap and knots]\label{ob:cohere}
Verify the $2$-level equational theory: $\swap$ an involution and a homomorphism up to coherent
isomorphism, the coherence $2$-cell~\eqref{eq:cohere} satisfying the triangle and pentagon identities,
$\kappa,\overline\kappa$ equivalences. Then $\quo{},\drf{}$ are derived and Proposition~\ref{prop:reflift}
is a calculation; the source papers' collapse (knots literal identities) is a localisation under which
the reflection law survives.
\end{obligation}

\begin{obligation}[Operational correspondence of the desugaring]\label{ob:opcorr}
Promote Proposition~\ref{prop:opcorr} to a theorem: define the desugaring of
Appendix~\ref{app:desugar} precisely, including persistence and the Church-encoded equation component,
and prove bisimilarity of the two transition systems with labels matched by location channels; confirm
independence from the channel intension (Remark~\ref{rem:nonopt}).
\end{obligation}

\begin{obligation}[The classifier classifies bisimilarity]\label{ob:classify}
Prove Theorem~\ref{thm:keystone}: that the internal modal theory of Definition~\ref{def:cl} coincides
with the subobject-classifier logic of $\Sh(\dT_K)$, that $\nu\sB^\flat$ agrees with the class-sized
$\nu\sB$ on equality, and hence that $\Av$ with its modalities classifies bisimilarity --- internal
OSLF adequacy, the keystone of Theorem~\ref{thm:keystone}.
\end{obligation}

\begin{obligation}[Barbed-congruence calibration]\label{ob:barbed}
For each presented GSLT, calibrate context bisimulation $\sim$ against the language's own observational
equivalence: reduction-barbed congruence for the process calculi, contextual equivalence at ground
type for PCF, $\beta\eta$ or applicative bisimilarity for the $\lambda$-calculus. This is the residual
of Corollary~\ref{cor:milner} --- the classical full-abstraction question recast as calibration ---
and is expected tight where names are quoted structure and there is no $\nu$~\cite[Obl.~6]{knotrho}.
\end{obligation}

\begin{obligation}[Functoriality on the encoded regime]\label{ob:functor}
Prove $\Cl$ functorial on the \emph{encoded} regime of $\GSLTsem$-morphisms --- a source $K$
implemented by target $K$'s plus structural context ($\pi$-into-rho, $\lambda$-into-SKI) --- so that
operational-correspondence encodings induce geometric morphisms~\cite{gsltnotes}.
\end{obligation}

\begin{obligation}[Size calibration]\label{ob:size}
State the finitary/infinitary calibration: the finitary-core $\Ks$ set-sized and interpretable in
hereditarily finite data (equiconsistent with arithmetic~\cite[\S8.2]{rset}); the full $\Ks$ raising
the target by one $\omega$. The well-founded vs non-well-founded knob is the initial- vs final-fixpoint
choice of Remark~\ref{rem:knob}.
\end{obligation}

\begin{obligation}[The ultrafilter metric, internally]\label{ob:metric}
Internalise the ultrafilter metric on bisimulation classes~\cite{gsltnotes}: construct the points of
the internal modal algebra of $\Cl(G)$, equip them with the metric induced by the site's spatial
reach, and recover the real/complex-weight dichotomy as classical/Born-rule structure. This is the
door for the physics reading.
\end{obligation}

\begin{obligation}[Placing the construction]\label{ob:place}
Locate the knotted-topos presentation among its neighbours: classifying topoi of geometric theories
and the topos-as-bridge methodology~\cite{caramello}; the game-semantic models of
PCF~\cite{ajm,ho,girardll}, of which the equivariant fragment is an instance; the presheaf and nominal
coalgebraic models of name-passing (Fiore--Turi~\cite{fioreturi}, Stark~\cite{stark}), of which this is
the reflective, $\nu$-free instance with $\Name\simeq\Proc$; the bialgebraic semantics of
Turi--Plotkin~\cite{turiplotkin}; and the nominal/Schanuel topos~\cite{pitts}. The feature absent from
all of these --- $\Name\simeq\Proc$ via $\swap$, the second knot --- is what the lift adds.
\end{obligation}

\section{Conclusion}\label{sec:conclusion}

The principal result is the lift: the knotted set-theoretic universe rises one categorical level to a
knotted topos, four sort-topoi tied by two geometric knots and an involutive colour-swap, the
reflection law become the swap's involutivity, the behaviour functor and its final coalgebra
recapitulated internally where colour-blind non-well-foundedness gives recursion and finality makes
bisimilar processes equal. The application transports the operational semantics of the finitely
presentable GSLTs to a denotational one: MeTTaIL presents that category, desugars to rho, and the rho
denotation lifted into the knotted topos is fully abstract for context bisimulation. Because GSLTs
subsume the classical models of computation, this is a fully abstract denotational semantics for all of
them at once. The model is the bisimulation quotient itself, so full abstraction is by construction;
what is proved is compositionality, and the context-labelled presentation makes bisimulation a
congruence with no closure step. The keystone is that the OSLF-generated Hennessy--Milner logic is the
internal logic of the knotted topos, adequacy the statement that the subobject classifier classifies
bisimilarity, and the three readings of $\dT_K$ --- Firing, Witnessing, Placing --- its labels,
modalities and site. Milner's lament was that a denotational model can be too coarse or too fine for
its language; the knotted topos answers that the right denotational object is neither chosen nor
guessed but constructed as the interaction quotient, and that one such construction serves every
finitely presentable model of computation. The risk ledger is unchanged: finite presentations cost
nothing beyond the empty set, recursion costs one $\omega$, and not one primitive name is ever spent.

\appendix

\section{The desugaring, in clauses}\label{app:desugar}

We record $\sem{\cdot}$ used in \S\ref{sec:desugar}, threading the location $\ell$
(Definition~\ref{def:loc}); $c(\ell)=\refl{\ell}$ is its channel and $\ell\cdot(f,i)$ the child
location. The target is core rho; MeTTaIL pattern-receives abbreviate nested single-name receives with
name-equality guards~\cite[\S2.2]{optimal}, and equations are Church-encoded normalisations suppressed
here.

\paragraph{Terms.} For a constructor $f$ of arity $n$,
\[
\sem{f(t_1,\dots,t_n)}_\ell
\;=\;
\out{c(\ell)}{\,\underline{f}\,}
\ \big|\
\textstyle\bigm|_{i=1}^{n}\ \sem{t_i}_{\,\ell\cdot(f,i)},
\]
publishing the head tag $\underline f$ at the node's channel and installing each argument at its child
location. A nullary constructor publishes only its tag; a schema variable installs nothing.

\paragraph{Base rewrites.} For $L\Rightarrow R$ with $\mathrm{hd}(L)=f$,
\[
\sem{L\Rightarrow R}
\;=\;
\textstyle\bigm|_{\ell\,:\,\mathrm{hd}=f}\
\for{\sem{L}}{c(\ell)}\bigl\{\ \out{c(\ell)}{\sem{R}_\ell}\ \big|\ \sem{L\Rightarrow R}\ \bigr\},
\]
the inner copy re-installing the listener after each fire by the reflection idiom
of~\cite[\S2.1]{knotrho} (no replication). Bound names of $\sem{L}$ are the hole-fillers, delivered on
$c(\ell)$.

\paragraph{Contextual rewrites.} For
$L_1\Rightarrow R_1,\dots,L_n\Rightarrow R_n\Rightarrow K\Rightarrow K'$ with outer context $K$ at
$\ell$,
\[
\mathsf{let}\ c=c(\ell)\ \mathsf{in}\
\for{(\sem{L_1},\dots,\sem{L_n})}{c}\bigl\{\ \out{c}{\sem{K'}(\sem{R_1},\dots,\sem{R_n})}\ \bigr\},
\]
blocking until the $n$ hole-fillers arrive, one per inner location $\ell\cdot(K,i)$, each a distinct
channel by Definition~\ref{def:loc}; the send emits the rewritten outer right-hand side at $\ell$.

\paragraph{A whole GSLT.} $\sem{G}=\bigm|_{(L\Rightarrow R)\in\mathcal R}\sem{L\Rightarrow R}$, with
equations compiled to structural congruence. A term $t$ is run by injecting $\sem{t}_\varepsilon$ on a
fresh root (Remark~\ref{rem:fresh}).

\begin{thebibliography}{99}

\bibitem{rset} L.~G.~Meredith. \emph{A Knotted Universe: a new notion of reflective set theories.}
Manuscript, 2026.

\bibitem{knotrho} L.~G.~Meredith. \emph{Quoting is Colour-Swap: a model of the rho calculus in the
knotted universe.} Manuscript, 2026.

\bibitem{optimal} L.~G.~Meredith. \emph{Optimal Channel Naming for Compositional Rewrite Translations
via Set Automaton Partial Evaluation.} F1R3FLY.io, 2026.

\bibitem{gsltnotes} L.~G.~Meredith et al. \emph{Graph-Structured Lambda Theories: conversation
context.} Working notes, 2026.

\bibitem{rho} L.~G.~Meredith and M.~Radestock. A reflective higher-order calculus. \emph{Electr.
Notes Theor. Comput. Sci.}, 141(5):49--67, 2005.

\bibitem{mettail} L.~G.~Meredith. \emph{The MeTTaIL language: defining languages as triples of
syntactic types, equations, and rewrites.} F1R3FLY.io technical report and prototype,
\url{https://github.com/F1R3FLY-io/mettail-rust}, 2025.

\bibitem{milnerfa} R.~Milner. Fully abstract models of typed $\lambda$-calculi. \emph{Theoret.
Comput. Sci.}, 4(1):1--22, 1977.

\bibitem{plotkinlcf} G.~D.~Plotkin. LCF considered as a programming language. \emph{Theoret. Comput.
Sci.}, 5(3):223--255, 1977.

\bibitem{scottdomains} D.~S.~Scott. A type-theoretical alternative to ISWIM, CUCH, OWHY.
\emph{Theoret. Comput. Sci.}, 121(1--2):411--440, 1993. (Manuscript, 1969.)

\bibitem{ajm} S.~Abramsky, R.~Jagadeesan, and P.~Malacaria. Full abstraction for PCF. \emph{Inform.
and Comput.}, 163(2):409--470, 2000.

\bibitem{ho} J.~M.~E.~Hyland and C.-H.~L.~Ong. On full abstraction for PCF: I, II, and III.
\emph{Inform. and Comput.}, 163(2):285--408, 2000.

\bibitem{girardll} J.-Y.~Girard. Linear logic. \emph{Theoret. Comput. Sci.}, 50(1):1--101, 1987.

\bibitem{leifermilner} J.~J.~Leifer and R.~Milner. Deriving bisimulation congruences for reactive
systems. In \emph{CONCUR 2000}, LNCS 1877, pages 243--258. Springer, 2000.

\bibitem{sewell} P.~Sewell. From rewrite rules to bisimulation congruences. \emph{Theoret. Comput.
Sci.}, 274(1--2):183--230, 2002.

\bibitem{milnerpoly} R.~Milner. \emph{The Polyadic $\pi$-Calculus: A Tutorial.} In F.~L.~Bauer,
W.~Brauer, and H.~Schwichtenberg, eds., Logic and Algebra of Specification, NATO ASI Series F~94,
pages 203--246. Springer, 1993.

\bibitem{sangiorgiwalker} D.~Sangiorgi and D.~Walker. \emph{The $\pi$-Calculus: A Theory of Mobile
Processes.} Cambridge University Press, 2001.

\bibitem{rutten} J.~J.~M.~M.~Rutten. Universal coalgebra: a theory of systems. \emph{Theoret. Comput.
Sci.}, 249(1):3--80, 2000.

\bibitem{turiplotkin} D.~Turi and G.~Plotkin. Towards a mathematical operational semantics. In
\emph{LICS 1997}, pages 280--291. IEEE, 1997.

\bibitem{jm} A.~Joyal and I.~Moerdijk. \emph{Algebraic Set Theory.} London Math. Soc. Lecture Note
Series 220, Cambridge University Press, 1995.

\bibitem{awodey} S.~Awodey. A brief introduction to algebraic set theory. \emph{Bull. Symbolic
Logic}, 14(3):281--298, 2008.

\bibitem{maclane} S.~Mac~Lane and I.~Moerdijk. \emph{Sheaves in Geometry and Logic: A First
Introduction to Topos Theory.} Springer, 1992.

\bibitem{elephant} P.~T.~Johnstone. \emph{Sketches of an Elephant: A Topos Theory Compendium.} Oxford
University Press, 2002.

\bibitem{caramello} O.~Caramello. \emph{Theories, Sites, Toposes: Relating and Studying Mathematical
Theories Through Topos-Theoretic `Bridges'.} Oxford University Press, 2018.

\bibitem{fioreturi} M.~Fiore and D.~Turi. Semantics of name and value passing. In \emph{LICS 2001},
pages 93--104. IEEE, 2001.

\bibitem{stark} I.~Stark. A fully abstract domain model for the $\pi$-calculus. In \emph{LICS 1996},
pages 36--42. IEEE, 1996.

\bibitem{pitts} A.~M.~Pitts. \emph{Nominal Sets: Names and Symmetry in Computer Science.} Cambridge
Tracts in Theoretical Computer Science 57, Cambridge University Press, 2013.

\bibitem{gabbaypitts} M.~J.~Gabbay and A.~M.~Pitts. A new approach to abstract syntax with variable
binding. \emph{Formal Asp. Comput.}, 13(3--5):341--363, 2002.

\bibitem{aczel} P.~Aczel. \emph{Non-Well-Founded Sets.} CSLI Lecture Notes 14, CSLI Publications,
1988.

\bibitem{bouwmanerkens} M.~Bouwman and R.~Erkens. Term rewriting based on set automaton matching.
\emph{arXiv:2202.08687}, 2022.

\bibitem{baadernipkow} F.~Baader and T.~Nipkow. \emph{Term Rewriting and All That.} Cambridge
University Press, 1998.

\end{thebibliography}

\end{document}
