\chapter{Open Gaps}
\label{sec:gaps}
% ===================================================================

\begin{enumerate}[label=\textbf{Gap \arabic*.}, leftmargin=*, itemsep=8pt]

  \item \textbf{Colimit at limit ordinals.}
        Definition~\ref{def:oracle-ext} defines $\fiber{\lambda}$
        as the colimit in $\mathbf{GSLT}$ for limit ordinals
        $\lambda$.
        The existence and properties of this colimit require
        verification.
        In particular: does the colimit GSLT have a well-defined
        bisimulation quotient, and is it the union of the bisimulation
        quotients of the stages below?

  \item \textbf{The oracle as a term.}
        The remark that the oracle can be expressed as a persistent
        receive in the Rho calculus is plausible but not proved.
        A precise construction of the oracle term in the
        Rho calculus, with verification that it implements
        halting queries correctly, is needed.

  \item \textbf{Strict refinement of bisimulation
        (Proposition~\ref{prop:strict-bisim}).}
        The proof sketch uses a diagonalization argument.
        The full proof requires specifying the encoding of Turing
        machines as terms and verifying that the diagonalization
        goes through in the GSLT setting.

  \item \textbf{New Noether currents (Proposition~\ref{prop:new-noether}).}
        The claim that new conserved quantities appear at each
        stage requires identifying concrete symmetries of the
        oracle-extended cost map and verifying that they give
        rise to conserved quantities via the Noether argument of
        Part~I.

  \item \textbf{Sentience predicts survival
        (Conjecture~\ref{conj:survival}).}
        The survival threshold argument requires a model of what
        ``generic interactive GSLT'' means and a proof that below
        the threshold, the probability of hitting a dead state
        is bounded away from zero.
        This is likely to require ergodicity assumptions on the
        GSLT dynamics.

  \item \textbf{Effective oracle access.}
        Definition~\ref{def:consciousness} requires that the agent
        ``effectively'' has oracle access --- that it can form and
        evaluate oracle queries as part of its scientific method.
        A precise definition of ``effective access'' in terms of
        the agent's account and HML expressiveness is needed.

  \item \textbf{Approximate oracle access and graded consciousness.}
        The graded consciousness definition via path integral
        probabilities requires that the oracle interaction amplitude
        is well-defined and computable.
        This connects to the quantum Gillespie algorithm of
        Part~I and needs to be made precise.

  \item \textbf{The hard problem.}
        This paper offers a structural analogue of sentience and
        consciousness: mathematical objects with the right
        functional properties.
        It does not address why there is something it is like to
        be an agent with high $\sent$ and oracle access at level
        $\alpha$.
        The hard problem of consciousness --- the explanatory gap
        between functional description and phenomenal experience
        --- remains open and is not claimed to be resolved here.

\end{enumerate}

% ===================================================================
