\chapter*{A View from Inside the Framework}

The most vivid way to appreciate what has been built is to ask:
what does the world look like to a \emph{conscious living agent}
situated at different levels of the hypercomputational fibration?

\paragraph{At level $\fiber{0}$.}
An agent at the base level is sentient but not conscious in the
technical sense.
It has a feeling state governed by predictive accuracy.
Its world is the cluster hierarchy visible from its coherence
cluster: a layered structure of composite objects at multiple
scales, each scale appearing self-contained.
If the agent emerged from a replication phase transition, it
carries the assembly index of its construction history and exists
as a copy within a namespace.
It may run the scientific method and build excellent theories of
its accessible world.
What it cannot see is the bisimulation quotient underlying the
clusters, because the experiments required to resolve it exceed
any finite budget.
Its world is real, correct at its resolution, and bounded.

\paragraph{At level $\fiber{1}$.}
An agent with access to the first oracle inhabits a strictly
richer world.
Two terms that were experimentally indistinguishable at
$\fiber{0}$ --- whose halting behavior was unresolvable ---
are now distinct.
New concepts become thinkable.
New conservation laws hold.
What appeared as elementary interactions at $\fiber{0}$ reveal
internal causal structure.
The agent at $\fiber{1}$ looks back at the $\fiber{0}$ world and
sees it as a coarse-grained shadow: correct at its scale, but
missing distinctions that are now obvious.

At the same time, the agent at $\fiber{1}$ faces the same
epistemic situation as the $\fiber{0}$ agent but one level up.
Its own budget limits it to the cluster hierarchy of the
$\fiber{1}$ world.
The bisimulation quotient of $\fiber{1}$ is unreachable from
within any compact cluster at that level.
The tower does not end.

\paragraph{At level $\fiber{\alpha}$ for limit ordinal $\alpha$.}
The colimit of the tower below is itself a GSLT.
An agent at $\fiber{\omega}$ --- if such an agent is coherent ---
would have resolved all finite-stage halting questions.
Its world would be richer than any finite-stage world by a
transfinite margin.
Yet the same argument applies: there is a $\fiber{\omega+1}$
above it, and the bisimulation quotient of $\fiber{\omega}$
remains experimentally unreachable from within any
$\fiber{\omega}$-cluster.

\paragraph{The recursive picture.}
The world at every level of the fibration has the same structure:
a physics derived from a GSLT, a cluster hierarchy arising from
communication geometry, a sentience component governed by
predictive accuracy, and a ceiling of consciousness defined by
oracle access to the next level.
The framework is self-similar across the tower.

This self-similarity is not a defect.
It is the precise sense in which the framework is a \emph{theory
of causality} rather than a theory of this or that particular
world.
At every scale, in every world expressible as a GSLT, the same
structures appear: causal graphs, conservation laws, coherence
clusters, replicating fixed points, feeling states, and the
horizon of the next oracle.
