\chapter{Complex Amplitudes and Quantum Dynamics}
\label{sec:quantum}
% ===================================================================

\section{Lifting to Complex Amplitudes}

So far the weight map could take real values; we now fully embrace complex
values.

\begin{definition}[Amplitude-Weighted GSLT]
An \emph{amplitude-weighted GSLT} is a weighted GSLT $(\GSLT, \fwd)$ in
which $\fwd_r : \HML(\ctx) \to \CC$ for all $r \in \rules$, with the
CPT constraint $\bwd_r(\phi) = \overline{\fwd_r(\phi)}$ for all $\phi$.
\end{definition}

\section{The Path Integral}

\begin{definition}[Transition Amplitude]
The \emph{transition amplitude} from term $P$ to term $Q$ is:
\[
  \langle Q \mid P \rangle \;=\; \sum_{\gamma : P \to Q} W(\gamma)
\]
where the sum is over all finite rewrite paths from $P$ to $Q$.
The \emph{transition probability} is $|\langle Q \mid P \rangle|^2$.
\end{definition}

\begin{remark}[Interference]
Because amplitudes are complex, two paths $\gamma_1, \gamma_2$ from $P$ to $Q$
with $W(\gamma_1) = -W(\gamma_2)$ \emph{cancel}: the transition
$P \to Q$ has zero probability despite having two distinct causal histories.
This is the GSLT analogue of quantum interference.  It arises not from external
imposition of quantum mechanics, but from the complex-valuedness of the weight
map.
\end{remark}

\section{The Density Matrix and the Account}

When the account $\vect{A}$ is allowed to become a positive semidefinite
matrix $\rho$ (a density matrix), conservation of the account becomes:
\[
  \mathrm{tr}(\rho) = 1 \quad \text{(conservation of total probability)}
\]
together with conservation of coherences (off-diagonal elements of $\rho$).
This is the quantum analogue of the classical resource conservation
Theorem~\ref{thm:conservation}.

\section{The Quantum Gillespie Algorithm}

The classical Gillespie algorithm~\cite{gillespie1977exact} simulates a
continuous-time Markov chain (CTMC) by:
\begin{enumerate}[itemsep=2pt]
  \item computing exit rates $\lambda_r$ from the current state for each rule $r$;
  \item sampling the waiting time $\Delta t \sim \mathrm{Exp}(\sum_r \lambda_r)$;
  \item selecting the next rule proportionally to rates.
\end{enumerate}
As of 2021~\cite{cao2021quantum}, quantum analogues of the Gillespie algorithm
have been developed for open quantum systems described by Lindblad master
equations.  The key modification: the exit rates $\lambda_r$ become complex
decay parameters $\gamma_r \in \CC$, encoding both the decay rate
$\mathrm{Re}(\gamma_r)$ and the oscillation frequency $\mathrm{Im}(\gamma_r)$
of quantum coherences.

\begin{definition}[Quantum-Gillespie-Weighted GSLT]
An amplitude-weighted GSLT $(\GSLT, \fwd)$ is \emph{Gillespie-compatible} if
for each term $P$ and rule $r$, the weight $\fwd_r(\phi_P)$ (at the most
refined formula $\phi_P$ of $P$) is interpretable as the complex decay
parameter $\gamma_r$ of the quantum Gillespie algorithm applied at $P$.
\end{definition}

\begin{remark}[Simulation]
A Gillespie-compatible amplitude-weighted GSLT can be directly simulated by
the quantum Gillespie algorithm: sample the next rule and waiting time from the
complex exit rates, advance the term, update the trace and account.  This is
the connection between the abstract construction and running code.  In
particular, the \texttt{gillespie} crate extending MeTTaIL's rewrite rules with
stochastic and quantum simulation capabilities implements precisely this loop:
the base rewrite rules of a MeTTaIL language definition carry weight maps, and
the \texttt{gillespie} crate evaluates the quantum Gillespie step at each state.
\end{remark}

% ===================================================================
