\chapter*{What Has Been Built}

This work set out to answer a single question: what is the minimum
mathematical structure needed to ground physics, knowledge, life,
and mind?
The answer, developed across four parts, is that the structure of a
graph-structured lambda theory --- three ingredients: a grammar,
equations, and rewrite rules --- suffices, provided it is equipped
with two maps (weights and costs) indexed by a logic it generates
itself.

Part~I established that this structure already contains a full
physics.
The reversibility construction gives time-symmetry.
The open and closed synchronization trees give two notions of
causal depth.
The weight map gives a path integral with interference.
The cost map gives conservation laws as theorems.
The extended modal operators internalize the Lagrangian and
Hamiltonian as logical facts.
These are constructions, not analogies.
They are stated precisely enough to be wrong, and we believe they
are right.

Part~II established that agents embedded in a massive population
of processes will, under any finite experimental budget, converge
to a theory of the cluster hierarchy rather than the bisimulation
quotient.
This is not a failure of the scientific method.
The cluster hierarchy is objectively present, arising from the
communication geometry of the GSLT.
A budget-limited agent does good science and arrives at correct
theories at its scale, unaware that the scale is not the bottom.
The compactness argument --- that compact populations support
consensus, and that coherence clusters are maximal compact
subpopulations --- is new to our knowledge and is the sharpest
result of Part~II.

Part~III proposed, more speculatively, that sentience and
consciousness have precise structural analogues in the GSLT
framework.
Sentience is the value of a distinguished resource component
governed by predictive accuracy.
Consciousness is oracle access to a higher stage of the
hypercomputational fibration.
The distinction is not terminological: sentience is continuous
and lives within a stage; consciousness is the capacity to
inhabit a strictly richer world.
The virtuous and vicious cycles linking the two are structural
consequences of the resource framework.

Part~IV grounded the origins of life.
Assembly index is depth in the open synchronization tree.
Copy number is count within a namespace.
Life emerges when a prebiotic population wanders into the basin
of attraction of a replicating fixed point, which is dense in
the logical topology.
The phase transition to life is the crossing of that basin
boundary, and what emerges is a compact coherence cluster
centered on the fixed point --- the new geosphere.
