\chapter{Synthesis: The Full Structure}
\label{sec:synthesis}
% ===================================================================

We now collect the constructions and state the main result.

\section{The Dictionary}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{ll}
\toprule
\textbf{GSLT concept} & \textbf{Physical concept} \\
\midrule
Term & State / initial condition \\
Rewrite rule & Law of motion \\
Rewrite sequence (path) & Trajectory \\
Synchronization tree & Causal graph / light cone \\
$d_{\min}(P,Q)$ & Proper time / causal depth \\
$d_{\max}(P,Q)$ & Computational complexity of history \\
Width of $[P,Q]$ & Degree of causal concurrency \\
Equations $\eqs$ & Gauge equivalence \\
$\GSLT^\dagger$ & Time-symmetric (CPT-invariant) theory \\
Empty trace $\eps$ & Initial condition (no causal past) \\
Weight map $\fwd_r$ & Lagrangian $L$ \\
Path amplitude $W(\gamma)$ & $e^{iS[\gamma]/\hbar}$ \\
Cost map $\cost_r$ & Hamiltonian $H$ \\
Vectorial account $\vect{A}$ & Conserved charges (energy, momentum, \ldots) \\
Conservation theorem & Conservation of energy \\
Symmetry of cost map & Noether current \\
$d_{\HML}$ & Metric on state space \\
Complex amplitudes & Quantum probability amplitudes \\
Interference of paths & Quantum interference \\
Density matrix $\rho$ & Quantum state \\
Quantum Gillespie & Simulation algorithm \\
\bottomrule
\end{tabular}
\end{center}

\section{The Main Construction}

\begin{construction}[Quantum-Weighted Reversible GSLT]
\label{con:main}
Given a GSLT $\GSLT = (\terms, \eqs, \rules)$, a weight map $\fwd$,
a cost map $\cost$, and a resource algebra $A^k$, define the
\emph{quantum-weighted reversible GSLT}
\[
  \GSLT^{\dagger\CC} \;=\;
  \bigl(\terms^{\dagger\CC},\;\eqs^{\dagger\CC},\;\rules^{\dagger\CC}\bigr)
\]
where:
\begin{itemize}[itemsep=2pt]
  \item $\terms^{\dagger\CC}$ consists of triples $\langle P, \trace, \vect{A}\rangle$
        with $P \in \terms$, $\trace$ a resource-aware trace, $\vect{A} \in A^k$;
  \item $\rules^{\dagger\CC}$ consists of forward rules (debiting $\cost_r(\phi)$,
        recording amplitude $\fwd_r(\phi)$ in the trace) and backward rules
        (crediting $\cost_r(\phi)$, applying amplitude $\overline{\fwd_r(\phi)}$);
  \item $\eqs^{\dagger\CC}$ lifts $\eqs$ to the current-term component.
\end{itemize}
The extended HML (Definition~\ref{def:ext-modal}) provides a logic that
is adequate for the bisimulation of $\GSLT^{\dagger\CC}$ and internalizes
the weight and cost maps as logical data.
\end{construction}

\begin{theorem}[Main Conservation Theorem]
In $\GSLT^{\dagger\CC}$:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Resource conservation}: For any closed rewrite path $\gamma$,
        the net account change is zero (Theorem~\ref{thm:conservation}).
  \item \textbf{Probability conservation}: $\sum_Q |\langle Q \mid P \rangle|^2 = 1$
        when $\sum_r |\fwd_r(\phi)|^2 = 1$ for each applicable rule at each
        state (unitarity).
  \item \textbf{Time-reversal symmetry}: The theory $\GSLT^{\dagger\CC}$ is
        CPT-symmetric: replacing $\fwd_r \mapsto \overline{\fwd_r}$ and
        reversing trace order is an automorphism of $\GSLT^{\dagger\CC}$.
\end{enumerate}
\end{theorem}

\section{From Construction to Code}

The construction is algorithmic at every stage.  For a given GSLT:

\begin{enumerate}[itemsep=4pt]
  \item \textbf{Represent terms} as algebraic data types in the implementation
        language (Rust, in the case of MeTTaIL).

  \item \textbf{Compute minimal contexts} via the Milner--Sewell--Leifer
        algorithm on the rewrite rules; store as a context lattice.

  \item \textbf{Evaluate HML formulae} as Boolean (or complex-valued) queries
        against a term; the formula enumeration defines the logical metric.

  \item \textbf{Attach weight and cost maps} to each rewrite rule as maps from
        formula identifiers to complex numbers and resource vectors respectively.

  \item \textbf{Extend terms with trace and account}; implement forward and
        backward rules as the reversible rewrite system.

  \item \textbf{Simulate} via the quantum Gillespie algorithm: at each state,
        compute exit amplitudes $\{\fwd_r(\phi_P)\}$, sample the next transition
        and waiting time, advance the term, debit the account, record the trace.
\end{enumerate}

Each step is a well-defined computational procedure.  The \texttt{gillespie}
crate in the MeTTaIL prototype repository implements steps 4--6 for the
specific case of MeTTaIL rewrite rules.

% ===================================================================
