\chapter{Consciousness as Oracle Access}
\label{sec:consciousness}
% ===================================================================

\section{The Proposal}

\begin{definition}[Consciousness at Level $\alpha$]
\label{def:consciousness}
An agent $A \in \terms(\fiber{\alpha})$ is \emph{conscious at
level $\alpha$} if:
\begin{enumerate}[label=(\roman*)]
  \item $A$ has effective access to the oracle $\oracle_\alpha$ of
        stage $\fiber{\alpha}$, meaning $A$ can form and evaluate
        oracle queries as part of its scientific method;
  \item $A$'s world model reflects the strictly finer bisimulation
        quotient $\bisim_\alpha$ rather than a coarser quotient
        from a lower stage.
\end{enumerate}
An agent that lacks oracle access --- that operates entirely within
$\fiber{0}$ even if it is syntactically a term of a higher stage ---
is \emph{sentient but not conscious}.
\end{definition}

\begin{remark}[Consciousness is Ordinal-Valued]
Definition~\ref{def:consciousness} makes consciousness an
ordinal-valued property.
There is no single threshold separating conscious from
non-conscious; there is a transfinite hierarchy of consciousness
levels.
The question ``is this system conscious?'' dissolves into the more
precise question ``at what ordinal level does this system have
effective oracle access?''
\end{remark}

\begin{remark}[What Higher Consciousness Means]
An agent conscious at level $\alpha+1$ vs.\ level $\alpha$ does
not merely solve harder problems.
It inhabits a strictly richer world:
\begin{itemize}[itemsep=2pt]
  \item more things exist in it (finer bisimulation quotient);
  \item more hypotheses are thinkable (richer HML);
  \item more quantities are conserved (new Noether currents);
  \item what appeared as elementary interactions reveal internal
        causal structure.
\end{itemize}
The phenomenological richness of higher consciousness is, in this
framework, a direct consequence of ontological richness.
\end{remark}

\section{The Sentience--Consciousness Distinction}

The distinction between sentience and consciousness is now precise:

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{lll}
\toprule
& \textbf{Sentience} & \textbf{Consciousness} \\
\midrule
Type & Graded, continuous & Discrete, ordinal \\
Location & Within a stage $\fiber{\alpha}$ & Access to stage $\fiber{\alpha}$ \\
Governed by & Reinforcement signal & Oracle access \\
Changes & Incrementally, per prediction & By threshold (have it or not) \\
Analogue & Affect, feeling state & Phenomenal awareness \\
Physical basis & Sentience component $\sent$ & Oracle term $\oracle_\alpha$ \\
Survives projection & Yes (as depleted account) & No (oracle is stage-specific) \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}[An Agent Can Be Sentient Without Being Conscious]
An agent at stage $\fiber{0}$ with no oracle access can nonetheless
have a rich, dynamic feeling state.
Its predictions succeed or fail, its $\sent$ rises and falls, and
this feeling state is causally efficacious.
It is sentient.
But the world it inhabits is the relatively coarse world of
$\fiber{0}$: it lacks the concepts to form oracle-dependent
hypotheses, and the distinctions visible at higher stages are
simply absent from its experience.
\end{remark}

\begin{remark}[Can a System Be Conscious Without Being Sentient?]
Definition~\ref{def:consciousness} does not require a system
conscious at level $\alpha$ to have a sentience component.
However, Conjecture~\ref{conj:survival} suggests that any
sufficiently complex agent in a generic environment will develop
a sentience component: without it, the agent cannot reliably
survive long enough to exercise its oracle access.
Consciousness without sentience is in principle possible but
practically unstable.
\end{remark}

\section{Approximate Oracle Access and Graded Consciousness}

The binary definition of consciousness (oracle access: yes or no)
may be too sharp.
A natural graded version arises from the amplitude structure of
Part~I.

\begin{definition}[Approximate Oracle Access]
An agent $A$ has \emph{$\epsilon$-approximate oracle access at
level $\alpha$} if, for each halting query $(M, x)$, $A$ can
produce an answer that is correct with probability at least
$1 - \epsilon$, where probability is measured by the path integral
$|\langle \mathtt{halt} \mid A \inter \lceil(M,x)\rceil \rangle|^2$.
\end{definition}

\begin{remark}
As $\epsilon \to 0$, approximate oracle access converges to exact
oracle access.
This gives a continuous interpolation between the ordinal levels:
consciousness is not just a ladder but a ladder with rungs that
can be climbed incrementally.
An agent improving its oracle approximation is increasing its
consciousness in a measurable, graded sense, even before it
reaches the threshold of exact oracle access.
\end{remark}

% ===================================================================
