\chapter{Discussion and Future Directions}
% ===================================================================

\section{Relation to Existing Work}

The reversibility construction is closely related to the work of Danos and
Krivine on reversible CCS~\cite{danos2004reversible} and to Phillips and
Ulidowski's reversible process calculi.  The present contribution is to
show that this construction lifts uniformly to all GSLTs and, crucially,
that it is the first step in a sequence of constructions that terminates at
quantum dynamics.

The causal set interpretation connects to the causal set programme in quantum
gravity~\cite{bombelli1987space}.  The present framework gives a \emph{generative}
presentation of causal sets (via rewrite rules) rather than the axiomatic
presentation standard in the physics literature.

The context-decorated HML and the Milner--Sewell--Leifer construction are
established; what is new is their use as the index set for the weight and
cost maps, and the internalization of dynamical data into the modal operators.

\section{Open Questions}

Several natural questions remain:

\begin{enumerate}[itemsep=4pt]
  \item \textbf{Universality of the reversibility envelope}: Is $\GSLT^\dagger$
        the \emph{minimal} reversible GSLT receiving a morphism from $\GSLT$?
        That is, does $\eta : \GSLT \to \GSLT^\dagger$ satisfy a universal
        property in $\mathbf{GSLT}$?

  \item \textbf{Symplectic structure}: Does the space of bisimulation classes
        of $\GSLT^{\dagger\CC}$ carry a natural symplectic form, making the
        phase-space analogy with $\langle P, \trace \rangle$ (current term
        + history = position + momentum) precise?

  \item \textbf{Legendre transform}: Under what convexity condition on the
        HML formula space is the Legendre transform between the weight map
        (Lagrangian) and the cost map (Hamiltonian) invertible?

  \item \textbf{The Rho calculus case}: The involution ${*}{@}P = P$
        already suggests the Rho calculus is closer to self-dual than most
        GSLTs.  Can the reversibility envelope the Rho calculus reversibility envelope be constructed
        with less added structure, exploiting this symmetry?

  \item \textbf{Conserved quantities from $\eqs$}: Can Noether's theorem
        be made fully precise in this setting --- that is, can every symmetry
        of the equations $\eqs$ be shown to correspond to a conserved component
        of the vectorial account?
\end{enumerate}

% ===================================================================
