\chapter{Ontological Isolation}
\label{sec:isolation}
% ===================================================================

\begin{definition}[Ontological Isolation]
A computational system $\GSLT$ is \emph{ontologically isolated} if,
for any term $P \in \terms(\GSLT)$ and any entity $X$ outside
$\terms(\GSLT)$, $P$ can interact with $X$ only via an encoding
$\lceil X \rceil \in \terms(\GSLT)$.
No term can refer to, branch on, or be influenced by unencoded
external entities.
\end{definition}

\begin{remark}
This is not an assumption but a structural consequence of what it
means to be a term in a GSLT.
The rewrite rules operate on terms; they cannot pattern-match on
entities that have no term representation.
The JVM sensor example makes this vivid: a stream of hardware
events arriving at a JVM instance has no effect on any Java program
until those events are encoded as Java data structures.
The encoding step is where the world outside enters; without it,
the outside world is simply absent from the computation.
\end{remark}

\begin{proposition}[Isolation Licenses Closure]
For the purposes of studying the epistemic limits of agents in $\GSLT$,
it is without loss of generality to assume that the agent's world
consists entirely of $\terms(\GSLT)/{\bisim}$.
The outside world can be forgotten.
\end{proposition}

\begin{proof}[Proof sketch]
Any influence of the outside world on an agent $P$ must pass through
an encoding.
The encoding is itself a term.
The effect of the outside world on $P$ is therefore indistinguishable,
from $P$'s perspective, from the effect of the encoding term acting
on $P$ from within $\GSLT$.
\end{proof}

\begin{remark}
Ontological isolation is what licenses the move to the category of
GSLTs as the natural setting for studying AI.
If AI's value proposition is that useful aspects of intelligence
are faithfully rendered as computations, and if computations are
ontologically isolated, then the category of GSLTs captures
\emph{everything} that matters for the limits of AI --- without
needing to appeal to an external world.
\end{remark}

% ===================================================================
