\chapter{The Reversibility Construction}
\label{sec:reversibility}
% ===================================================================

In a bare GSLT, rewrite rules are directed: once a term $P$ reduces to $Q$,
there is in general no canonical way to recover $P$ from $Q$.  This
\emph{irreversibility} is the computational analogue of thermodynamic
time-asymmetry.  We now construct, for any GSLT $\GSLT$, a reversible envelope
$\GSLT^\dagger$ in which every forward step has an explicit inverse.

\section{Traces}

\begin{definition}[Trace]
Let $\GSLT = (\terms, \eqs, \rules)$.  A \emph{trace} is a finite sequence
\[
  \trace \;=\; r_1(l_1) \cdot r_2(l_2) \cdots r_n(l_n)
\]
where each $r_i \in \rules$ and $l_i \in \terms(\GSLT)$ is the specific
left-hand-side term matched at step $i$.  The empty trace is $\eps$.
Traces are quotiented by $\equiv$ acting on each $l_i$ component.
\end{definition}

\begin{definition}[Extended Terms]
The \emph{extended grammar} $\terms^\dagger$ consists of pairs $\langle P, \trace \rangle$
where $P \in \terms(\GSLT)$ is the \emph{current term} and $\trace$ is the
\emph{causal history} of how $P$ was reached.
\end{definition}

\section{The Reversible Rewrite System}

\begin{definition}[Reversible Rules]
For each rule $r : l_r \rewrite r_r$ in $\rules$, define:
\begin{align}
  \text{Forward rule } r^+ &:\quad
    \langle l_r,\, \trace \rangle \;\rewrite\;
    \langle r_r,\, r(l_r) \cdot \trace \rangle \\[4pt]
  \text{Backward rule } r^- &:\quad
    \langle r_r,\, r(l_r) \cdot \trace \rangle \;\rewrite\;
    \langle l_r,\, \trace \rangle
\end{align}
The set of reversible rules is $\rules^\dagger = \{r^+ \mid r \in \rules\} \cup
\{r^- \mid r \in \rules\}$.
\end{definition}

\begin{definition}[Reversible Envelope]
The \emph{reversible envelope} of $\GSLT$ is the GSLT
\[
  \GSLT^\dagger \;=\; \bigl(\terms^\dagger,\; \eqs^\dagger,\; \rules^\dagger\bigr)
\]
where $\eqs^\dagger$ lifts $\equiv$ to act on the current-term component of
extended terms, leaving traces invariant.
\end{definition}

\begin{proposition}
There exist morphisms in $\mathbf{GSLT}$:
\[
  \eta : \GSLT \to \GSLT^\dagger, \quad P \mapsto \langle P, \eps \rangle
\]
\[
  \pi : \GSLT^\dagger \to \GSLT, \quad \langle P, \trace \rangle \mapsto P
\]
satisfying $\pi \circ \eta = \mathrm{id}_{\GSLT}$.
\end{proposition}

\begin{remark}[Physical Interpretation]
A term $\langle P, \eps \rangle$ is an \emph{initial condition}: it has no causal
past.  A term $\langle P, \trace \rangle$ with $\trace \neq \eps$ is a term with a
recorded history.  The time-asymmetry present in $\GSLT$ (forward rules only)
disappears in $\GSLT^\dagger$: the laws are symmetric, and the arrow of time is
entirely encoded in the boundary condition ($\trace = \eps$ for the initial
state).  This is precisely the modern understanding of the thermodynamic arrow
of time.
\end{remark}

% ===================================================================
