\documentclass[12pt,openright,twoside]{book}

%% ================================================================
%%  PACKAGES
%% ================================================================
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, calc}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tocloft}
\setcounter{tocdepth}{2}
\usepackage{fancyhdr}
\usepackage{emptypage}
\usepackage{epigraph}
\usepackage{setspace}

\geometry{
  paper=letterpaper,
  top=1.25in, bottom=1.25in,
  left=1.25in, right=1.25in,
  headheight=15pt
}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black,
  pdftitle={Finding Mind},
  pdfauthor={Lucius Gregory Meredith}
}

%% ================================================================
%%  COLOURS
%% ================================================================
\definecolor{defblue}{RGB}{220,235,250}
\definecolor{thmgreen}{RGB}{220,245,225}
\definecolor{remarkgray}{RGB}{245,245,245}
\definecolor{accentblue}{RGB}{30,90,160}
\definecolor{partcolor}{RGB}{15,60,120}

%% ================================================================
%%  THEOREM ENVIRONMENTS
%% ================================================================
\newmdtheoremenv[
  backgroundcolor=defblue,
  linecolor=accentblue,
  linewidth=1.5pt,
  topline=false, rightline=false, bottomline=false
]{definition}{Definition}[chapter]

\newmdtheoremenv[
  backgroundcolor=thmgreen,
  linecolor=green!50!black,
  linewidth=1.5pt,
  topline=false, rightline=false, bottomline=false
]{theorem}{Theorem}[chapter]

\newmdtheoremenv[
  backgroundcolor=thmgreen,
  linecolor=green!50!black,
  linewidth=1.5pt,
  topline=false, rightline=false, bottomline=false
]{proposition}{Proposition}[chapter]

\theoremstyle{remark}
\newmdtheoremenv[
  backgroundcolor=remarkgray,
  linecolor=gray!60,
  linewidth=1pt,
  topline=false, rightline=false, bottomline=false
]{remark}{Remark}[chapter]

\newtheorem{example}{Example}[chapter]
\newtheorem{construction}{Construction}[chapter]
\newtheorem{conjecture}{Conjecture}[chapter]

%% ================================================================
%%  NOTATION
%% ================================================================
\newcommand{\GSLT}{\mathcal{S}}
\newcommand{\terms}{\mathcal{T}}
\newcommand{\rules}{\mathcal{R}}
\newcommand{\eqs}{\mathcal{E}}
\newcommand{\ctx}{\mathcal{K}}
\newcommand{\HML}{\mathrm{HML}}
\newcommand{\rewrite}{\longrightarrow}
\newcommand{\bisim}{\sim}
\newcommand{\fwd}{w^{+}}
\newcommand{\bwd}{w^{-}}
\newcommand{\cost}{c}
\newcommand{\acct}{A}
\newcommand{\sat}{\models}
\newcommand{\llb}{\llbracket}
\newcommand{\rrb}{\rrbracket}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\Ord}{\mathbf{Ord}}
\newcommand{\inter}{\otimes}
\newcommand{\pop}{\mathcal{P}}
\newcommand{\nerve}{\mathcal{N}}
\newcommand{\comb}{\mathcal{C}}
\newcommand{\sent}{A_{\mathrm{s}}}
\newcommand{\oracle}{\mathcal{O}}
\newcommand{\fiber}[1]{\GSLT^{(#1)}}
\newcommand{\proj}[1]{\pi_{#1}}
\newcommand{\lts}{\mathit{LTS}}
\newcommand{\Rho}{\mathbf{Rho}}
\newcommand{\vect}[1]{\mathbf{#1}}
\newcommand{\trace}{\tau}
\newcommand{\eps}{\varepsilon}

%% ================================================================
%%  SECTIONING STYLES
%% ================================================================
\titleformat{\chapter}[display]
  {\normalfont\huge\bfseries\color{accentblue}}
  {\chaptertitlename\ \thechapter}{18pt}{\Huge}

\titleformat{\section}{\large\bfseries\color{accentblue}}{{\thesection}}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{{\thesubsection}}{1em}{}

\titleformat{\part}[display]
  {\centering\huge\bfseries\color{partcolor}}
  {\Large\scshape Part \thepart}
  {0.5em}{}
  [\vspace{0.5em}\hrule height 1.5pt\vspace{1em}]

%% ================================================================
%%  RUNNING HEADERS
%% ================================================================
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.4pt}
\fancyhead[LE,RO]{\small\thepage}
\fancyhead[LO]{\small\itshape\rightmark}
\fancyhead[RE]{\small\itshape Meredith --- Finding Mind}

%% ================================================================
%%  TOC SPACING
%% ================================================================
\setlength{\cftbeforesecskip}{4pt}
\setlength{\cftbeforesubsecskip}{1pt}

%% ================================================================
%%  TITLE
%% ================================================================
\title{%
  \vspace{-0.8in}
  {\large\scshape Working Draft}\\[0.6em]
  {\Huge\bfseries Finding Mind}\\[0.5em]
  {\large\itshape The Pledge, the Turn, and the Prestige}
}
\author{%
  Lucius Gregory Meredith\\[0.3em]
  {\small Chief Executive Officer, F1R3FLY.io}\\
  {\small Chief Science Officer, F1R3FLY Industries}
}
\date{Revision: \today}

%% ================================================================
\begin{document}
%% ================================================================

\maketitle
\thispagestyle{empty}

%% ----------------------------------------------------------------
%%  EPIGRAPH
%% ----------------------------------------------------------------
\newpage
\thispagestyle{empty}
\vspace*{\fill}
\begin{center}
  \large\itshape
  ``The magician takes the ordinary something and makes it do\\
  something extraordinary. Now you're looking for the secret\ldots{}\\
  but you won't find it, because of course you're not really looking.\\
  You don't really want to know.\\
  You want to be fooled.''
  \medskip

  \normalsize\upshape --- Christopher Priest, \textit{The Prestige}
\end{center}
\vspace*{\fill}
\newpage

%% ----------------------------------------------------------------
%%  TABLE OF CONTENTS
%% ----------------------------------------------------------------
\tableofcontents
\newpage

%% ================================================================
%%  PART I: THE PLEDGE
%% ================================================================
\part{The Pledge}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
The Pledge shows you something ordinary.
A mind. Your mind.
The one that right now is reading these words and asking whether it
understands them.
It asks you to look at what you already know about compositionality,
agency, information, and causation --- and to notice the blind spot
hiding in plain sight.
The extraordinary is already here.
You just haven't been shown where to look.
\end{mdframed}

\bigskip

%% --- Pledge chapters ---
\input{ThePledge/pledge_ch01}  % Prolegomena to Finding Mind
\input{ThePledge/pledge_ch02}  % The Ordinary World
\input{ThePledge/pledge_ch03}  % Information, Causation, and the Blind Spot
\input{ThePledge/pledge_ch04}  % Finding the Shape

%% ================================================================
%%  PART II: THE TURN
%% ================================================================
\part{The Turn}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
The Turn takes the ordinary something --- computation --- and makes it do
something extraordinary.
A single mathematical structure, a \emph{graph-structured lambda theory},
is shown to be sufficient to ground physics, the limits of scientific
knowledge, the origin of life, and the phenomenon of mind.
None of these correspondences are metaphors.
Each is a mathematical construction, stated precisely enough to be wrong.
\end{mdframed}

\bigskip

\phantomsection
\addcontentsline{toc}{section}{Overarching Introduction}
{\large\bfseries\color{accentblue} Overarching Introduction}
\vspace{0.8em}

\noindent
This work grows from a single question: \emph{what is the minimum
mathematical structure needed to ground physics, knowledge, life,
and mind?}

The answer developed across four parts is that the structure of a
\emph{graph-structured lambda theory} (GSLT) --- a grammar of terms,
equations on those terms, and rewrite rules --- is already sufficient,
provided it is equipped with modest additional data.
Physics emerges from weights and costs on the rewrite rules.
The limits of knowledge emerge from the topology of the formula
logic induced by the GSLT and the communication geometry of a
massive agent population.
The hypercomputational tower above any Turing-complete GSLT
provides a precise setting for sentience and consciousness.
Life emerges from the density of replicating fixed points in the
space of processes.

None of these correspondences are metaphors.
Each is a mathematical construction, stated precisely enough to
be wrong.
Part~I is presented with considerable confidence.
Parts~II and~III are speculative but grounded.
Part~IV is avowedly exploratory, reaching toward the hardest
questions.
All four parts share a common architecture and notation, and a
common conviction: that computation, properly understood, is not
merely a tool for doing science and philosophy but a substrate
in which science and philosophy can be expressed with full
mathematical precision.

\bigskip

\subsection*{Part I\quad Towards a Theory of Causality}

Part~I asks: what is a \emph{physics}?
The answer is that a physics is a GSLT equipped with two maps ---
a \emph{weight map} assigning complex amplitudes to rewrite steps,
and a \emph{cost map} assigning resource consumption --- indexed by
a context-decorated Hennessy--Milner logic derived canonically from
the GSLT itself.

The construction proceeds in stages.
Any GSLT is first lifted to a \emph{time-symmetric envelope}
$\GSLT^\dagger$ in which every rewrite step carries an explicit
inverse.
The arrow of time appears not in the laws but entirely in the
boundary condition: a term with empty history has no causal past.

Two kinds of synchronization tree are distinguished.
The \emph{closed} tree records autonomous evolution (rewrite-rule
labeled edges); only redexes have nontrivial closed trees.
The \emph{open} tree records interactive evolution (minimal-context
labeled edges); every term has a potentially rich open tree.
Both are causal sets in the sense of Bombelli--Lee--Sorkin.

The weight map turns rewrite paths into complex amplitudes,
giving a path integral whose squared modulus is a transition
probability.
The cost map turns rewrite steps into resource debits across a
vectorial account; reversibility ensures these debits balance
over closed loops, making energy conservation a theorem.
The modal operators of the HML logic are extended to carry the full
dynamical state of each transition, internalizing the Lagrangian
and the Hamiltonian as logical facts.
The construction is completed by lifting to complex amplitudes and
connecting to the quantum analogue of the Gillespie algorithm.

\bigskip

\subsection*{Part II\quad Why Scientists Get Misled}

Part~II asks: given that the true ontology of a GSLT is its
bisimulation quotient, why would a situated agent ever settle
for anything less?

The scientific method is sound: given unlimited resources, an
agent converges to the bisimulation quotient.
The obstruction is not the method but the phenomena the method
encounters.
In a population of quark-scale size, communication geometry and
message degradation produce a real, objective structure: the
population fragments into compact coherence clusters whose overlap
forms a simplicial nerve.
This hierarchy is as real as the agents themselves.

A budget-limited agent running the scientific method encounters
the cluster hierarchy as genuine experimental evidence.
Its theory of the hierarchy is correct at the scale it describes.
What the agent cannot suspect --- without resources that grow with
the full population --- is that the clusters are themselves
composite, built over a finer bisimulation structure that is
experimentally unreachable within any finite budget.

The compactness argument is central: a population is compact in
the logical topology if and only if it supports consensus on every
Boolean question.
The coherence clusters are precisely the maximal compact
subpopulations, and compactness is simultaneously the condition
for functioning as a coherent community and the condition for
being an closed epistemic horizon.

\bigskip

\subsection*{Part III\quad Sentience, Consciousness, and the
Hypercomputational Fibration}

Part~III asks whether the GSLT framework can give a precise account
of sentience and consciousness.

Following Turing's construction of a hypercomputational tower above
the Turing-computable, a \emph{fibration} is constructed over the
subcategory of Turing-complete GSLTs: a tower
$\fiber{0} \hookrightarrow \fiber{1} \hookrightarrow \cdots$
in which each $\fiber{\alpha}$ extends the one below with an oracle
for halting.
The physics at each stage is \emph{strictly richer} than below:
the bisimulation quotient is finer, new HML formulae become
expressible, new Noether currents appear, and what appeared as
elementary interactions resolve into structured causal histories.
A higher-stage agent does not merely compute faster; it inhabits
a world with more things in it.

\emph{Sentience} is defined as a graded, continuous property: the
value of a distinguished sentience component $\sent$ of the
agent's vectorial account, governed by a reinforcement signal
that rewards accurate prediction and punishes inaccuracy.
The feeling state is causally efficacious --- it gates which
transitions the agent can afford --- and predictive of survival.

\emph{Consciousness} is defined as oracle access: an agent is
conscious at level $\alpha$ if it can form and evaluate queries
to $\oracle_\alpha$, inhabiting the strictly richer physics of
$\fiber{\alpha}$.
Sentience is real-valued and continuous; consciousness is
ordinal-valued and discrete.
The virtuous and vicious cycles linking the two are examined:
oracle access, good predictions, and high feeling state reinforce
each other, and the collapse of any one can drive the collapse
of the others.

\bigskip

\subsection*{Part IV\quad Origins of Life: Assembly Theory,
Phase Transitions, and Replication}

Part~IV connects the framework to two of the most mathematically
serious proposals about the origins of life.

Assembly Theory (Cronin and Walker) assigns to each object an
assembly index and a copy number, claiming their joint magnitude
is a biosignature.
The assembly index is identified precisely with minimum path length
in the \emph{open} synchronization tree: each edge is a minimal
context representing one environmental joining step, and nesting
depth is exactly the accumulated computational history that
Cronin and Walker call ``time as an object.''
Two gaps in Assembly Theory are filled: copy number is given a
rigorous definition using the Rho calculus namespace structure
(channels are locations; a namespace is a region; a copy is a
bisimilar process at a channel), and the identity criterion for
copies is bisimulation.

The connection to Eric Smith's fourth geosphere runs through the
density of replicating fixed points.
The Rho calculus concurrent Y combinator --- the canonical fixed
point of the duplication map --- is the abstract core of hereditary
replication.
The comm rule is one definitional step from a genetic crossover
operator, placing replication, recombination, and variation all
within reach of a single primitive.
If replicating terms are dense in the logical topology, then any
prebiotic population wandering under selection pressure inevitably
enters the basin of attraction of a replicating fixed point.
The phase transition to life is the crossing of that basin
boundary; the new geosphere is the compact coherence cluster
centered on the fixed point.

\bigskip

\subsection*{The Architecture of the Work}

The four parts form a directed dependency structure:

\begin{center}
\begin{tikzpicture}[
  node distance=1.8cm,
  box/.style={draw=accentblue, thick, rounded corners=4pt,
               fill=defblue, minimum width=4.0cm,
               minimum height=1.0cm, align=center, font=\small}
]
\node[box] (p1) {Part I\\Physics from GSLTs};
\node[box, right=2.6cm of p1] (p2) {Part II\\Emergent Ontology};
\node[box, below=1.6cm of p1] (p3) {Part III\\Sentience \&
  Consciousness};
\node[box, right=2.6cm of p3] (p4) {Part IV\\Origins of Life};
\draw[-{Stealth[length=6pt]}, thick, accentblue] (p1) -- (p2)
  node[midway, above, font=\footnotesize\itshape] {logical topology};
\draw[-{Stealth[length=6pt]}, thick, accentblue] (p1) -- (p3)
  node[midway, left, font=\footnotesize\itshape, xshift=-2pt]
  {account / fibration};
\draw[-{Stealth[length=6pt]}, thick, accentblue] (p2) -- (p4)
  node[midway, right, font=\footnotesize\itshape, xshift=2pt]
  {basin topology};
\draw[-{Stealth[length=6pt]}, thick, accentblue] (p1) -- (p4)
  node[midway, font=\footnotesize\itshape, yshift=-7pt]
  {open sync tree};
\draw[-{Stealth[length=6pt]}, thick, accentblue] (p3) -- (p4)
  node[midway, below, font=\footnotesize\itshape]
  {conscious living agents};
\end{tikzpicture}
\end{center}

Part~I provides the physical and logical machinery on which all
subsequent parts depend.
Part~II uses the logical topology to argue that budget-limited
agents converge to the cluster hierarchy rather than the
bisimulation quotient.
Part~III uses the account and fibration structure to ground
sentience and consciousness.
Part~IV uses the open synchronization tree from Part~I, the basin
topology from Part~II, and the conscious living agents of Part~III
to ground the origins of life.

A reader primarily interested in the physics may read Part~I
independently.
The epistemological argument requires Parts~I and~II.
The account of sentience and consciousness requires Parts~I
and~III.
The origins of life argument draws on all four parts, and the
conclusion --- which examines what a conscious living agent at
different levels of the fibration sees --- assumes the full
framework.

\bigskip
\noindent\hrule
\bigskip

\bigskip
\noindent\hrule
\bigskip


%% ----------------------------------------------------------------
%%  The Turn, Part I: Towards a Theory of Causality
%% ----------------------------------------------------------------
\part{Towards a Theory of Causality}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
This part establishes the physical framework: a sequence of
constructions taking any graph-structured lambda theory to a
quantum-weighted reversible dynamical system, with Lagrangian,
Hamiltonian, path integral, and conservation laws emerging as theorems.
It is the most technically developed of the parts and is presented
with the most confidence.
\end{mdframed}

\bigskip

\input{TheTurn/turn_causality_ch01}  % Introduction and Motivation
\input{TheTurn/turn_causality_ch02}  % Graph-Structured Lambda Theories
\input{TheTurn/turn_causality_ch03}  % The Reversibility Construction
\input{TheTurn/turn_causality_ch04}  % Causal Structure and Time
\input{TheTurn/turn_causality_ch05}  % The Labeled Transition System and Context-Decorated HML
\input{TheTurn/turn_causality_ch06}  % The Weight Map and Lagrangian Structure
\input{TheTurn/turn_causality_ch07}  % The Cost Map and Hamiltonian Structure
\input{TheTurn/turn_causality_ch08}  % Extended Modal Operators
\input{TheTurn/turn_causality_ch09}  % Complex Amplitudes and Quantum Dynamics
\input{TheTurn/turn_causality_ch10}  % Synthesis: The Full Structure
\input{TheTurn/turn_causality_ch11}  % Discussion and Future Directions

%% ----------------------------------------------------------------
%%  The Turn, Part II: Why Scientists Get Misled
%% ----------------------------------------------------------------
\part{Why Scientists Get Misled: Emergent Ontology in Massive Agent Populations}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
This part argues that agents in a massive computational population
are structurally prevented from discovering the true ontology of
their GSLT.
The argument rests on ontological isolation, communication
degradation at scale, and compactness as the topological signature
of consensus.
The key claims are stated as conjectures; the gaps are identified
explicitly.
\end{mdframed}

\bigskip

\input{TheTurn/turn_sciencebot_ch01}  % Introduction
\input{TheTurn/turn_sciencebot_ch02}  % Ontological Isolation
\input{TheTurn/turn_sciencebot_ch03}  % Agents and the Scientific Method
\input{TheTurn/turn_sciencebot_ch04}  % Interactive GSLTs and Behavioral Primes
\input{TheTurn/turn_sciencebot_ch05}  % Massive Populations and Communication Degradation
\input{TheTurn/turn_sciencebot_ch06}  % The Logical Topology and Compactness
\input{TheTurn/turn_sciencebot_ch07}  % Coherence Clusters and Their Nerve
\input{TheTurn/turn_sciencebot_ch08}  % The Apparent Ontology
\input{TheTurn/turn_sciencebot_ch09}  % Open Gaps and Future Work
\input{TheTurn/turn_sciencebot_ch10}  % Discussion

%% ----------------------------------------------------------------
%%  The Turn, Part III: Origins of Life
%% ----------------------------------------------------------------
\part{Origins of Life: Assembly Theory, Phase Transitions, and Replication}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
This part connects the framework to Assembly Theory and Eric
Smith's fourth geosphere.
Assembly index is identified with depth in the open synchronization
tree; copy number is given a rigorous definition via the Rho calculus
namespace structure; and the emergence of life is identified with
the crossing of a basin boundary in the logical topology, driven
by the density of replicating fixed points.
\end{mdframed}

\bigskip

\input{TheTurn/turn_origins_ch01}  % Introduction
\input{TheTurn/turn_origins_ch02}  % Assembly Theory: Causal Depth and Copy Number
\input{TheTurn/turn_origins_ch03}  % Replication as a Fixed Point
\input{TheTurn/turn_origins_ch04}  % Smith's Phase Transition and the Density of Attractors
\input{TheTurn/turn_origins_ch05}  % Metabolism, Energy Gradients, and the Ecosystem
\input{TheTurn/turn_origins_ch06}  % Open Gaps
\input{TheTurn/turn_origins_ch07}  % Discussion

%% ----------------------------------------------------------------
%%  The Turn, Part IV: Sentience, Consciousness, and the Hypercomputational Fibration
%% ----------------------------------------------------------------
\part{Sentience, Consciousness, and the Hypercomputational Fibration}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
This part constructs a fibration of GSLTs over the Turing-complete
subcategory indexed by ordinals.
The physics is strictly richer at each stage.
Sentience is a graded feeling state within a stage;
consciousness is oracle access to a higher stage.
Several arguments are frankly conjectural and are labeled as such.
\end{mdframed}

\bigskip

\input{TheTurn/turn_hypercomp_ch01}  % Introduction
\input{TheTurn/turn_hypercomp_ch02}  % The Hypercomputational Fibration
\input{TheTurn/turn_hypercomp_ch03}  % Strictly Richer Physics at Each Stage
\input{TheTurn/turn_hypercomp_ch04}  % Sentience as a Reinforced Feeling State
\input{TheTurn/turn_hypercomp_ch05}  % Consciousness as Oracle Access
\input{TheTurn/turn_hypercomp_ch06}  % Sentience, Consciousness, and Survival
\input{TheTurn/turn_hypercomp_ch07}  % Open Gaps
\input{TheTurn/turn_hypercomp_ch08}  % Discussion

%% ----------------------------------------------------------------
%%  The Turn: Conclusion
%% ----------------------------------------------------------------
\part*{Conclusion}
\addcontentsline{toc}{part}{Conclusion}
\markright{Conclusion}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
The conclusion draws the four parts into a single narrative,
assesses what has been established versus conjectured, and
sketches the view from inside the framework: what a conscious
living agent at different levels of the hypercomputational
fibration sees.
\end{mdframed}

\bigskip

\input{TheTurn/turn_conclusion_ch01}  % What Has Been Built
\input{TheTurn/turn_conclusion_ch02}  % A View from Inside the Framework
\input{TheTurn/turn_conclusion_ch03}  % What Remains to Be Done

%% ================================================================
%%  PART III: THE PRESTIGE
%% ================================================================
\part{The Prestige}

\vspace{1em}
\begin{mdframed}[backgroundcolor=remarkgray, linecolor=gray!60,
                 linewidth=1pt, topline=false, rightline=false,
                 bottomline=false]
\small\itshape
\textbf{[Placeholder]} The Prestige brings it back.
The impossible is revealed to have been here all along.
\end{mdframed}

\bigskip

\input{ThePrestige/prestige_placeholder}  % The Prestige (placeholder)

\end{document}
