\documentclass[11pt]{article}

%% ── packages ──────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{array}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage[colorlinks=true,linkcolor=NavyBlue,citecolor=NavyBlue,urlcolor=SkyBlue]{hyperref}

%% ── F1R3FLY brand colours ─────────────────────────────────────────
\definecolor{NavyBlue}{HTML}{1A1A2E}
\definecolor{SkyBlue}{HTML}{3FA9F5}
\definecolor{Yellow}{HTML}{F3D630}
\definecolor{Sage}{HTML}{8BB999}

%% ── theorem environments ──────────────────────────────────────────
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{construction}[definition]{Construction}
\newtheorem{example}[definition]{Example}
\newtheorem{remark}[definition]{Remark}

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}

%% ── notation macros ───────────────────────────────────────────────
\newcommand{\GSLT}{\mathrm{GSLT}}
\newcommand{\iGSLT}{\mathbf{iGSLT}}
\newcommand{\ciGSLT}{\mathbf{ciGSLT}}
\newcommand{\ciGSLTtc}{\mathbf{ciGSLT}_{\mathrm{TC}}}
\newcommand{\CostC}{\mathbf{Cost}\text{-}\ciGSLT}
\newcommand{\Free}{\mathsf{Free}}
\newcommand{\Forget}{\mathsf{Forget}}
\newcommand{\Imp}{\mathcal{I}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\near}{\mathsf{near}}
\newcommand{\lstk}[2]{S(#1,\, #2)}        % located stack S(I, s:S')
\newcommand{\Kctx}[1]{K([\,],\, #1)}
\newcommand{\sep}{\mathbin{*}}            % separating-style spatial conn (reuse * visually)
\newcommand{\dia}[1]{\langle #1 \rangle}  % modal diamond
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\Tok}{\mathsf{Stk}}
\newcommand{\WT}{\mathbb{T}}                  % wrapped-term sort
\newcommand{\Cost}{\mathfrak{C}}              % the cost / continuation endofunctor
\newcommand{\sgn}[2]{\{#1\}_{#2}}             % signed term  {T}_s
\newcommand{\cons}{\mathbin{:}}               % temporal cons
\newcommand{\nil}{(\,)}
\newcommand{\red}{\longrightarrow}
\newcommand{\Seq}{\equiv}
\newcommand{\hashf}{\#}
\newcommand{\quot}[1]{@#1}
\newcommand{\drop}[1]{{\ast}#1}
\newcommand{\subst}[3]{#1\{#2/#3\}}
\newcommand{\compute}{\mathsf{compute}}
\newcommand{\cf}{\mathsf{cf}}
\newcommand{\Term}{\mathcal{T}}
\newcommand{\cTerm}{\widehat{\mathcal{T}}}
\newcommand{\Kp}{K_{\mathrm{p}}}
\newcommand{\Ke}{K_{\mathrm{e}}}
\newcommand{\defeq}{\;\mathrel{:=}\;}
\newcommand{\Par}{\mathbin{|}}
\newcommand{\Card}[1]{\lvert #1 \rvert}

\title{\textbf{Continued Interactive GSLTs and the\\[2pt]
Cost Endofunctor}\\[6pt]
\large A construction one level up from cost-accounted Rholang\\[2pt]
\normalsize\textit{(revised: wrapping by construction)}}

\author{L.~G.~Meredith\thanks{\texttt{lgreg.meredith@f1r3fly.io}}
\quad\textit{with}\quad Claude (Anthropic)\\[2pt]
\small F1R3FLY.io}

\date{May 2026}

\begin{document}
\maketitle

\begin{abstract}
We lift the cost-accounting construction of the rho calculus from a
single calculus to an endofunctor on a category of interactive
graph-structured lambda theories (GSLTs). The earlier work folded
phlogiston accounting into the grammar of Rholang by signing terms and
replacing the communication rule by a family of rules that rewrite
\emph{signed} terms while consuming \emph{tokens}. Here we observe that
this is a generic procedure governed by a single shape of dynamics, the
\emph{interaction cut}: an interaction constructor $K$ brings two
introductions $\Kp(I,C_{\mathrm p})$ and $\Ke(J,C_{\mathrm e})$ into
contact along matching \emph{interaction surfaces} $I, J$, and a
contraction $\compute(I,J,C_{\mathrm p},C_{\mathrm e})$ --- Milner's
pseudo-application in the process-calculus instances --- combines their
\emph{program} and \emph{environment} continuations
$C_{\mathrm p}, C_{\mathrm e}$. Binding is one mode by which information
migrates between environment and program under this contraction, not an
intrinsic feature of the cut: CCS realises the same shape with no
migration at all, and is in this sense the paradigmatic instance. An
interactive GSLT whose dynamics
admits this presentation, together with a computable section of its
structural-congruence quotient and a contraction compatible with
wrapping, we call a \emph{continued} interactive GSLT. The point of the
adjective is not that such a theory is already metered or wrapped --- a
continued interactive GSLT is an ordinary, un-metered calculus --- but
that it carries exactly the structure needed for the cost construction
to \emph{build} a metered version from it.

That construction is the endofunctor $\Cost$. Given a continued
interactive GSLT, $\Cost$ produces a calculus in which the continuation
slots are \emph{wrapped terms by construction}: every continuation is a
metered thunk in the grammar of the image. The decisive design move is
that this wrapping lives in the \emph{output}, so that ``every redex sits
inside a wrapper'' is a sorting invariant of $\Cost(G)$ preserved by
reduction --- not a property that must be re-established by traversing
each contractum. The source theory's contraction need not be lifted; in
the image it already maps wrapped terms to wrapped terms, and the only
run-time interaction with tokens is \emph{unwrapping}: forcing a thunk by
spending a matching token. Metering is thereby \emph{lazy} --- work is
charged when performed, not when exposed --- and the temporal token stack
is consumed exactly in step with the reductions, realising the modulus of
the underlying cut elimination as the run length itself.

We show that $\Cost$ is an endofunctor on the category $\ciGSLT$ of
continued interactive GSLTs with bisimulation-preserving, quote-faithful
morphisms. Throughout we keep $\ciGSLT$ formally distinct from the
ambient category $\iGSLT$ of interactive GSLTs: a forgetful functor
$U : \ciGSLT \to \iGSLT$ exhibits ``continued'' as a proper structural
enrichment of ``interactive,'' so that the endofunctor's domain pins down
exactly which interactive theories admit sound, lazy cost accounting.
Finally, iterated cost accounting flattens --- nested signatures multiply
and stacks-of-stacks concatenate --- so $\Cost$ is a (non-idempotent)
monad, resolving through two adjunctions with opposite embeddings: a
structural free--forgetful adjunction that detaches the apparatus, and,
over a Turing-complete base, an internalisation adjunction that dissolves
the apparatus into the base's own computation up to weak bisimulation.
We close with refinements that make the metered calculus a usable resource
discipline --- object capabilities, an opt-in type discipline for token
usage, and resource-sufficiency proofs --- and with an analogy, offered
for further investigation, in which $K$ is space and $\cons$ is time, cut
elimination evolves an information distribution and the nearness structure
together in the manner of Einstein's field equations, and object-capability
authority entangles space and time into a spacetime.
\end{abstract}

\tableofcontents
\bigskip

%% ══════════════════════════════════════════════════════════════════
\section{Introduction}
\label{sec:intro}

In previous work we showed that the cost-accounting layer of Rholang ---
the phlogiston (phlo) system --- need not be a privileged runtime
extension. By treating signatures as a species of name, token stacks as
first-class terms, and the for/send operators as ranging over
\emph{signed} terms, we folded cost accounting directly into the grammar
and operational semantics of the rho calculus~\cite{MeredithRadestock2005}.
The single communication
rule \textsc{Comm} became a family of rules that gate communication on
the consumption of a matching token. Because the rho calculus is the
formal core of Rholang, this revised calculus serves as the design
specification for rearchitecting phlo accounting in Rholang~1.4. The rho
calculus is a member of the family of mobile process
calculi~\cite{Milner1999,SangiorgiWalker2001}, distinguished by its
reflective treatment of names as quoted
processes~\cite{MeredithRadestock2005,SmithdesRivieres1984}.

This note takes the construction \emph{one level up}, asking what the
move requires of the underlying theory and to which theories it applies.
Two observations drive the development. First, the dynamics must be
presented as an \emph{interaction cut}: two introductions meet along
matching interaction surfaces, each carrying its own continuation --- a
program continuation and an environment continuation --- which a
contraction then combines. This is visible already in Milner's polyadic
$\pi$-calculus tutorial~\cite{MilnerPolyadic1993}, where the input
$\mathrm{for}(y\leftarrow x)P$ is the pair $(x,\lambda y.P)$ and the
output $x!(u).Q$ is the pair $(x,[u,Q])$: the subjects $x$ are the
surfaces that must match, the objects $\lambda y.P$ and $[u,Q]$ are the
continuations, and the contraction is Milner's \emph{pseudo-application},
returning $P\{u/y\} \mathbin{|} Q$. Second --- and this is the heart of the
revision --- a theory presented in this form carries enough structure for
the cost endofunctor to build a metered version in which the continuation
slots are \emph{wrapped terms by construction}. We are careful to keep
the two apart: the underlying theory (a \emph{continued} interactive
GSLT) is an ordinary un-metered calculus that merely \emph{admits} the
construction; the wrapping is a feature of the \emph{image} $\Cost(G)$,
not of $G$. In that image every continuation is a metered thunk in the
grammar, so the contractum is born guarded: no traversal re-wraps it, the
contraction is not lifted, and the absence of leaks is a sorting
invariant of $\Cost(G)$ rather than a dynamic obligation. The run-time
semantics of the metered calculus only ever \emph{unwraps}, forcing a
thunk by spending a token.

The construction spans calculi with and without binding, and we use
several as foils. CCS~\cite{MilnerCCS1980} is the paradigmatic continued
interactive GSLT: its interaction $a.P \mathbin{|} \bar a.Q \to P
\mathbin{|} Q$ is an interaction cut with \emph{no} binding and no
information migration --- the surfaces are complementary labels, and
pseudo-application degenerates to releasing the two continuations. The
interaction categories of Abramsky, Gay, and
Nagarajan~\cite{AbramskyGayNagarajan1996} generalise this further, with
interaction as composition along a shared interface. The rho calculus and
the $\lambda$-calculus sit at the binding end of the spectrum, where
pseudo-application substitutes. In the rho calculus
several distinct structures are fused into $\Par$ and $\quot{\cdot}$;
pulling them apart in $\lambda$, where they are visibly separate, exposes
the construction as a genuine endofunctor and identifies precisely what
an interactive theory must supply. Binding, on this view, is one way
information migrates between environment and program --- not what makes an
interaction a cut.

Section~\ref{sec:metered-cut} isolates the interaction cut.
Section~\ref{sec:wrapped} introduces wrapping by construction, derives
no-leak as subject reduction, and contrasts lazy with eager metering.
Section~\ref{sec:two-monoids} identifies the spatial monoid $K$ and the
temporal monoid $\cons$, with the stack as a reified clock and lazily
extracted modulus. Section~\ref{sec:signatures} treats signature
construction as a section of the structural-congruence quotient, with
de~Bruijn canonicalisation supplying it for $\lambda$.
Section~\ref{sec:functor} assembles the endofunctor $\Cost$ on $\ciGSLT$.
Section~\ref{sec:adequacy} records the graded adequacy statement.
Section~\ref{sec:examples} works the two anchoring instances.
Section~\ref{sec:monad} observes that iterated cost accounting flattens,
so that $\Cost$ is a monad, and exhibits its two resolutions: a
structural free--forgetful adjunction and, over Turing-complete bases, a
computational internalisation adjunction. Section~\ref{sec:capabilities}
relates the construction to object capabilities, showing that whether
proximity to a resource stack confers authority depends on the
equational theory of $K$, and refining the stack to be \emph{located} at
an interaction surface. Section~\ref{sec:types} uses OSLF on the
cost-accounted theory as an opt-in type discipline for token usage,
expressed through spatial and modal connectives.
Section~\ref{sec:linear} lifts the linear-proof guarantee of resource
sufficiency, examines its fate under data-dependent interaction, and
shows that located purses make the guarantee local and compositional.

%% ══════════════════════════════════════════════════════════════════
\section{The interaction cut}
\label{sec:metered-cut}

\subsection{Interactive GSLTs}

A \emph{graph-structured lambda theory} is a multi-sorted term theory
$G = (\Sigma, E, R)$: a signature $\Sigma$ of sorts and operators, a set
$E$ of equations imposing a structural congruence $\Seq$, and a set $R$
of rewrite rules. The framing of operational semantics in this
algebraic, graph-enriched style follows Stay and Meredith and
Baez and Williams~\cite{StayMeredith2017,BaezWilliams2019}. It is
\emph{interactive} when it carries a
distinguished interacting sort $P$ with a labelled transition system,
generated by $R$ modulo $\Seq$, on which bisimulation is the intended
equivalence. CCS~\cite{MilnerCCS1980} ($P$ = processes, interaction by
complementary actions), the rho calculus ($P$ = processes, $\Seq$ the
$\Par$-monoid congruence with $\alpha$, $R$ = \textsc{Comm}), the
$\lambda$-calculus ($P$ = terms, $\Seq$ = $\alpha$, $R$ = $\beta$), and
the interaction categories of Abramsky, Gay, and
Nagarajan~\cite{AbramskyGayNagarajan1996} are all motivating examples ---
spanning, as we will see, calculi with and without binding.

\begin{definition}[The category $\iGSLT$]
\label{def:iGSLT}
$\iGSLT$ is the category whose objects are interactive GSLTs and whose
morphisms are \emph{theory maps} --- sort- and operator-preserving
translations that respect $\Seq$ and $R$ --- that are
\textbf{bisimulation-preserving} on the interacting sort.
\end{definition}

This is the base category of the present note. The cost construction
does \emph{not} act on all of $\iGSLT$; it acts on the objects that
carry enough additional structure to be metered, which we isolate as a
subcategory in Section~\ref{sec:functor}. Keeping the two apart is
deliberate: ``interactive'' is the ambient setting, while ``continued''
(Definition~\ref{def:ciGSLT}) is the structural condition under which
sound, lazy cost accounting exists. The distinction is what the
endofunctor's domain makes precise --- an interactive GSLT is a place
where processes interact; a continued interactive GSLT is one whose
interaction is presented as a meterable interaction cut.

\subsection{The interaction-cut presentation}

The cost construction acts on dynamics presented in a particular shape.
Write the generating rule of $R$ as
%
\begin{equation}
  \frac{(C_{\mathrm p}',\,C_{\mathrm e}') \;=\;
        \compute(I,\,J,\,C_{\mathrm p},\,C_{\mathrm e})}
       {\,K\bigl(\Kp(I,\,C_{\mathrm p}),\; \Ke(J,\,C_{\mathrm e})\bigr)
        \ \red\
        K'\bigl(C_{\mathrm p}',\,C_{\mathrm e}'\bigr)\,}
  \tag{$\dagger$}
  \label{eq:metered-cut}
\end{equation}
%
distinguishing:
%
\begin{description}[leftmargin=2.2em,itemsep=2pt]
  \item[the interaction constructor $K$,] which brings two operands into
    \emph{contact}. In rho, $K = {\Par}$; in $\lambda$, $K = \mathrm{App}$;
    in CCS, $K = {\Par}$.
  \item[two introductions $\Kp(I,C_{\mathrm p})$ and
    $\Ke(J,C_{\mathrm e})$,] each a constructor separating an
    \emph{interaction surface} ($I$, $J$) --- the part that must match for
    the cut to fire --- from a \emph{continuation}: $C_{\mathrm p}$ the
    \emph{program continuation} (the rest of the computation) and
    $C_{\mathrm e}$ the \emph{environment continuation} (the rest of the
    environment). A surface may be carried \emph{nominally}, as an
    explicit $I$ or $J$ that must match (as the subject $x$ does in rho),
    or \emph{structurally}, by the rigidity of $K$: when $K$ is non-AC,
    position itself is unforgeable and the context $K([\,],-)$ \emph{is}
    the surface, so a nominally \emph{degenerate} (absent) surface is not
    a missing match condition but one carried by the geometry of $K$
    instead. $\lambda$ is exactly this case on the environment side --- its
    rigidity supplies the surface that no explicit $J$ records. The two
    carriers are developed in Section~\ref{sec:capabilities}. The factoring is
    Milner's~\cite{MilnerPolyadic1993}: in
    rho and $\pi$, the input $\mathrm{for}(y\leftarrow x)P$ is the pair
    $(x,\lambda y.P)$ and the output $x!(u).Q$ is the pair $(x,[u,Q])$, so
    $I = x = J$ are the matching subjects while $C_{\mathrm p} = \lambda
    y.P$ and $C_{\mathrm e} = [u,Q]$ are the continuations. In CCS the
    surfaces are complementary actions $a, \bar a$ and the continuations
    carry no data.
  \item[the contraction $\compute$,] a partial operation that, on a
    surface match, combines the two continuations into residuals
    $C_{\mathrm p}', C_{\mathrm e}'$ repackaged by $K'$. This is Milner's
    \emph{pseudo-application}~\cite{MilnerPolyadic1993}: applied to
    $\lambda y.P$ and $[u,Q]$ it returns $P\{u/y\} \mathbin{|} Q$. It
    factors into two parts --- the substitution $P\{u/y\}$, by which the
    datum $u$ \emph{migrates} from the environment continuation into the
    program continuation through the binder $y$, and the release
    $\mathbin{|}\,Q$ of the environment's residual.
\end{description}
%
We call \eqref{eq:metered-cut} an \emph{interaction cut}: two
introductions meet along matching surfaces and a contraction combines
their continuations. Both sides are structured --- the eliminand is not a
bare term but an introduction carrying its own continuation, exactly as
$x!(u).Q$ in rho is the pair $(x,[u,Q])$ rather than a bare datum. We
avoid the word \emph{symmetric} for this: the cut need not be commutative
(in $\lambda$ it is not), and whether it is commutative is a separate
property of the equational theory of $K$, treated in
Section~\ref{sec:capabilities}. \emph{Binding is one mode of information
migration}, not a feature of the cut as such: it is the substitution half
of pseudo-application, present in rho, $\pi$, and $\lambda$, absent in
CCS, where pseudo-application degenerates to pure release. The
$\lambda$-calculus is the further degenerate case in which the environment
introduction $\Ke$ has a degenerate (absent) surface and the environment
continuation carries no residual, so that the argument is its own
continuation and pseudo-application is ordinary application
(Section~\ref{sec:examples}).

\begin{remark}[$K$ is the only constructor that needs overloading]
The cut isolates one contact site. Every cost rule interposes a token
between the operands of $K$ and drains it on contact. There is one $K$
per cut, hence one constructor to make polymorphic. ``Meter the cut'' is
the generic content of ``place valuable friction at every interaction.''
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Wrapping by construction}
\label{sec:wrapped}

This section describes what the cost endofunctor \emph{builds}. Given a
continued interactive GSLT $G$ --- an un-metered calculus presented as a
interaction cut, with a section and a wrapping-compatible contraction
(Section~\ref{sec:functor}) --- the endofunctor $\Cost$ produces a
metered calculus $\Cost(G)$ whose grammar makes every continuation a
metered thunk. The wrapping discipline below is therefore a property of
the image $\Cost(G)$; the source $G$ supplies only the structure that
makes the discipline installable.

\subsection{Continuations are metered thunks in the image}

In $\Cost(G)$ the cost endofunctor adjoins a sort $\Sig$ of signatures
(Section~\ref{sec:signatures}), a \emph{wrapped-term} sort $\WT$ with the
wrapper $\sgn{\cdot}{s} : P \times \Sig \to \WT$, and a sort $\Tok$ of
token stacks
%
\begin{equation}
  S \;::=\; \nil \;\mid\; s \cons S, \qquad s \in \Sig .
  \label{eq:stack}
\end{equation}
%
The pivotal discipline is grammatical, and it is imposed by $\Cost$ on
its image: \textbf{in $\Cost(G)$ every continuation slot of $\Kp$ and
$\Ke$ has sort $\WT$.} Structural composition (parallel composition in
rho) lives at the $\WT$ level, between wrappers, so that each
redex-bearing thread is individually wrapped. Thus a continuation in the
metered calculus is not a bare process awaiting metering; it is a
\emph{metered thunk} --- $\sgn{P}{s}$ --- whose activation is gated by a
token keyed to $s$, and whose own continuations are again thunks. (In
the source $G$ these slots carried ordinary terms; re-sorting them to
$\WT$ is part of what $\Cost$ does, Construction~\ref{con:object}.)
Reading $C_{\mathrm p} = \sgn{P}{s_{\mathrm p}}$ and
$C_{\mathrm e} = \sgn{Q}{s_{\mathrm e}}$, the contraction of the image has
the type
%
\[
  \compute \;:\; \mathit{Surf}\times\mathit{Surf}\times\WT\times\WT
            \;\longrightarrow\; \WT\times\WT,
\]
%
mapping wrapped continuations to wrapped continuations. It need not know
that wrapping exists: substituting a wrapped term into a position yields a
wrapped term, so $\compute$ preserves the sort $\WT$ for free.

\subsection{The gated rule: force by unwrapping}

A wrapped redex is forced by consuming a matching token. The
single-signature rule is
%
\begin{equation}
  \frac{(C_{\mathrm p}',C_{\mathrm e}') = \compute(I,J,C_{\mathrm p},C_{\mathrm e})}
       {\,K\!\bigl(\sgn{K(\Kp(I,C_{\mathrm p}),\,\Ke(J,C_{\mathrm e}))}{s},
        \; s \cons S\bigr)
        \ \red\
        K\!\bigl(K'(C_{\mathrm p}',C_{\mathrm e}'),\; S\bigr)\,}
  \tag{R1}
  \label{eq:R1}
\end{equation}
%
in which the outer polymorphic $K$ combines the wrapped redex with the
token stack, the head cell $s$ is consumed (the act of \emph{unwrapping}
the redex), and the result $K'(C_{\mathrm p}',C_{\mathrm e}')$ is --- by
the typing of $\compute$ --- already built from wrapped terms. There is
no re-wrapping step and no lifted contraction. The remaining
token-distribution cases vary how the redex is split across $K$ and how
tokens are presented:
%
\begin{align}
  &K\!\bigl(K(\sgn{\Kp(I,C_{\mathrm p})}{s_1},\,\sgn{\Ke(J,C_{\mathrm e})}{s_2}),\;
        s_1 \cons s_2 \cons S\bigr)
   \ \red\
   K\!\bigl(K'(C_{\mathrm p}',C_{\mathrm e}'),\; S\bigr),
   \tag{R2}\label{eq:R2}\\[3pt]
  &K\!\bigl(K(\sgn{\Kp(I,C_{\mathrm p})}{s_1},\,\sgn{\Ke(J,C_{\mathrm e})}{s_2}),\;
        K(s_1\cons S_1,\, s_2\cons S_2)\bigr)
   \ \red\
   K\!\bigl(K'(C_{\mathrm p}',C_{\mathrm e}'),\; K(S_1,S_2)\bigr),
   \tag{R3}\label{eq:R3}
\end{align}
%
each under the same premise. Here the continuations $C_{\mathrm p},
C_{\mathrm e}$ are themselves wrapped (suppressed inner signatures), so
the residuals $C_{\mathrm p}', C_{\mathrm e}'$ remain wrapped and the next
layer of redexes is guarded without any action by the rule.

\subsection{No leak, as a sorting invariant}

\begin{definition}[Well-wrapped configuration]
A configuration of $\Cost(G)$ is \emph{well-wrapped} when it is
well-sorted in the grammar of Construction~\ref{con:object} --- in
particular, every continuation occupies a slot of sort $\WT$, so every
redex occurrence lies inside a wrapper $\sgn{\cdot}{s}$.
\end{definition}

\begin{lemma}[Subject reduction for wrapping]
\label{lem:subject}
Well-wrapping is preserved by \eqref{eq:R1}--\eqref{eq:R3}.
\end{lemma}

\begin{proof}[Proof sketch]
Each rule replaces a wrapped redex by $K'(C_{\mathrm p}',C_{\mathrm e}')$
with $(C_{\mathrm p}',C_{\mathrm e}') = \compute(\dots)$. Since
$\compute : \cdots \to \WT\times\WT$ and $K'$ packages terms of sort
$\WT$, the residual is well-sorted; the consumed cell leaves a
well-sorted stack. Redexes elsewhere are untouched. Hence the result is
well-wrapped.
\end{proof}

\begin{corollary}[No leak, by construction]
\label{cor:noleak}
In a well-wrapped configuration no redex can fire without being unwrapped,
i.e.\ without consuming a token. There is no dynamic wrapping operation;
the cost-accounting steps only ever unwrap.
\end{corollary}

The contrast with the earlier formulation is the substance of this
revision.

\begin{remark}[Eager vs.\ lazy metering]
\label{rem:lazy}
An alternative discipline re-wraps the contractum at contraction time,
traversing it to sign each exposed redex --- charging \emph{eagerly} for
every redex the step exposes. Wrapping by construction charges
\emph{lazily}: each redex is born wrapped and is charged only when
actually forced. Lazy metering pays for work performed, not work merely
exposed; redexes beneath a discarded binder are never charged. The token
stack drains exactly in step with the reductions performed, which is also
the operational reality of phlogiston consumption.
\end{remark}

\begin{remark}[Duplication needs no fresh signatures]
\label{rem:dup}
Because multiplicity is carried by the \emph{stack} (linear token
consumption) and not by key distinctness, copying a wrapped term
$\sgn{Q}{s}$ copies its key $s$ but not its tokens. Each copy must be
forced separately and so funded separately from the stack: $n$ copies of
an $s$-redex require $n$ cells keyed to $s$. Shared keys with a linear
token supply already make duplication cost, so no fresh-per-copy minting
is required. Created redexes --- where a substituted wrapped term lands in
head position --- are guarded for the same reason: the substituted
material carries its own wrapper. Thus a single mechanism (wrapped
continuations $+$ linear stack) handles duplicated and created redexes
alike, where the eager formulation needed a separate fresh-signature
discipline.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Two monoids: space and time}
\label{sec:two-monoids}

The construction places two combination operators side by side, of
different character. The interaction constructor $K$ is \emph{spatial}:
it records what is in contact with what, and may carry equations ---
associative--commutative in rho (an unordered bag of processes), free in
$\lambda$ (a positional application spine). The cons operator $\cons$ of
\eqref{eq:stack} is \emph{temporal}: it records what is spent next, then
next. It is a free monoid (a list), \emph{never} commutative, because
reduction is sequential even when interaction is parallel. Each forcing
pops a cell; the order of popping is the order of steps. The stack is the
time axis of the computation reified as a term: $\Seq$ leaves it
invariant (spatial congruence consumes no time), $\red$ pops it (a step
is a tick), and its ordering is the arrow of the computation.

\begin{proposition}[Stack consumption is the modulus, lazily realised]
\label{prop:modulus}
For a terminating reduction in $\Cost(G)$, the number of cells consumed
equals the number of forced redexes, which equals the number of cuts
eliminated --- including those introduced by duplication, each charged
when forced. The consumed prefix of the temporal stack is therefore an
exact operational modulus for the cut elimination, realised by running
the reduction rather than by a separate traversal.
\end{proposition}

Where eager metering pre-computes a bound by traversing each contractum,
wrapping by construction lets the reduction \emph{be} the bound
extraction: the consumed stack is the realised cost, tight by
construction and lazy. Evaluation is cut elimination; the token stack is
the extracted modulus; metering at the granularity of forcing makes the
two coincide.

%% ══════════════════════════════════════════════════════════════════
\section{Signature construction as a section of the quotient}
\label{sec:signatures}

Tokenisation derives its expressiveness from minting signatures from
runtime data, so the construction adjoins a constructor turning terms
into signatures. We isolate what the base theory must provide.

\subsection{Quotient and section}

Let $\Term$ be the free (pre-congruence) syntax of the interacting sort
and $\cTerm = \Term/{\Seq}$ its quotient by structural congruence. Two
maps relate them, differing precisely in their relationship to $\Seq$:
%
\begin{itemize}[leftmargin=1.6em,itemsep=2pt]
  \item the \textbf{quotient} $\Term \twoheadrightarrow \cTerm$,
    \emph{respecting} $\Seq$, invertible up to congruence --- the
    structure a name constructor exposes when channels are taken up to
    $\Seq$ (in rho, $\quot{P}$ with $\quot{P}=\quot{Q}\iff P\Seq Q$). It
    is not a constructor the cost functor requires; it is the congruence
    the cut is gated modulo.
  \item a \textbf{section} $\cf : \cTerm \to \Term$, a choice of
    canonical representative per class. Composing with a
    collision-resistant digest gives the signature constructor
    $\hashf = \mathrm{digest}\circ\cf : \cTerm \to \Sig$, a one-way
    commitment to a representative. It does \emph{not} respect $\Seq$ ---
    and must not, since a commitment identifying congruent-but-distinct
    representatives would be no commitment.
\end{itemize}
%
$\Sig$ carries a commutative monoid $(\Sig,*,\nil)$ compounding
signatures, matching the compound-signature gating of \eqref{eq:R2},
\eqref{eq:R3}. Signatures never reduce: they are a rewrite-free
sub-theory meeting the behavioural theory only at the matching guard.
Hence the $\Seq$-incoherence of $\hashf$ is harmless for bisimulation,
which factors as ``$P$-bisimulation gated by signature-key matching.''

\begin{remark}[The two axes, disentangled]
In rho the section and the communication-quotient are the \emph{same}
operator $\quot{\cdot}$ viewed two ways, which is why the cost story there
appeared to need ``two axes of reflection.'' They are not two signature
schemes. One axis ($\quot{\cdot}$ up to $\Seq$) is the communication
quotient the LTS already carries; the other
($\hashf=\mathrm{digest}\circ\cf$) is the authentication section the cost
functor adds. The signature is $\hashf$ alone; the functor needs both
because metering tracks \emph{what a process does} (quotient) and
\emph{who committed to it} (section).
\end{remark}

\subsection{The \texorpdfstring{$\lambda$}{lambda}-calculus supplies a
section without a channel constructor}

The $\lambda$-calculus has no analogue of $\quot{\cdot}$. This is the
sharp test, and it passes, because the functor never required a
\emph{channel} constructor; it required a \emph{section}. For $\lambda$
the congruence is $\alpha$ and a section is a canonical representative of
each $\alpha$-class: the de~Bruijn encoding~\cite{deBruijn1972}. Thus
$\hashf_\lambda = \mathrm{digest}\circ(\text{de~Bruijn})$, and a signature
commits to the de~Bruijn form. The communication-quotient role that
$\quot{\cdot}$ played in rho is, in $\lambda$, filled not by a constructor
but by $\alpha$-equality of redexes in the guard. Rho fused section and
quotient; $\lambda$ pulls them apart, and the construction is identical.

\begin{definition}[Section-equipped]
\label{def:section}
An interactive GSLT is \emph{section-equipped} when it comes with a
computable section $\cf : \cTerm \to \Term$ of its $\Seq$-quotient
(equivalently, a computable canonical-form function on the interacting
sort).
\end{definition}

%% ══════════════════════════════════════════════════════════════════
\section{The cost endofunctor}
\label{sec:functor}

\subsection{Continued interactive GSLTs}

\begin{definition}[Continued interactive GSLT]
\label{def:ciGSLT}
A \emph{continued interactive GSLT} is an interactive GSLT (an object of
$\iGSLT$) \emph{equipped with} the following additional structure:
\begin{enumerate}[label=(\roman*),itemsep=1pt]
  \item a presentation of its dynamics in interaction-cut form
    \eqref{eq:metered-cut}, naming the constructors $K, \Kp, \Ke, K'$
    and the partial contraction $\compute$;
  \item a section (Definition~\ref{def:section}), i.e.\ it is
    section-equipped; and
  \item \emph{wrappability}: the contraction $\compute$ is well-sorted as
    an operation on wrapped continuations,
    $\compute : \cdots \to \WT\times\WT$ --- equivalently, $\compute$ is
    definable on metered thunks, mapping wrapped inputs to wrapped
    outputs.
\end{enumerate}
We write $\ciGSLT$ for the category whose objects are continued
interactive GSLTs and whose morphisms are theory maps that are
\textbf{bisimulation-preserving} on the interacting sort and
\textbf{quote-faithful}: injective on the reachable signature keys.
\end{definition}

The adjective \emph{continued} thus names a strict enrichment of
\emph{interactive}: clauses (i)--(iii) are data and conditions added on
top of an $\iGSLT$-object, not a restriction of it. Forgetting them
recovers the underlying interactive GSLT.

\begin{definition}[The forgetful functor]
\label{def:forget}
Let $U : \ciGSLT \to \iGSLT$ send a continued interactive GSLT to its
underlying interactive GSLT --- discarding the metered-cut presentation,
the section, and the wrappability witness --- and act as the identity on
the underlying theory map of a morphism.
\end{definition}

\begin{proposition}[The distinction is proper]
\label{prop:proper}
$U$ is faithful but neither full nor essentially surjective. Hence
$\ciGSLT$ is a genuine, proper subcategory of $\iGSLT$: strictly more
structure on objects, strictly fewer morphisms between them.
\end{proposition}

\begin{proof}[Discussion]
\emph{Faithful, not full.} $U$ forgets no information about a morphism's
action, so it is faithful. It is not full because a $\ciGSLT$-morphism
must additionally be quote-faithful: an $\iGSLT$-morphism that is
bisimulation-preserving yet collapses two distinct signature keys is a
map between the underlying objects with no lift to $\ciGSLT$. (This
refines the naive expectation that the inclusion be full: the section
structure on objects induces a real constraint on morphisms, so the
inclusion is faithful-not-full rather than full.)

\emph{Not essentially surjective.} Not every interactive GSLT admits the
added structure. Clause (i) requires the dynamics to factor as an
interaction cut --- contact $K$ then contraction $\compute$ --- which a
theory with non-local or unfactorable rewrites need not; clause (ii)
requires a \emph{computable} section of the $\Seq$-quotient, which a
theory with undecidable structural congruence lacks; clause (iii)
requires the contraction to preserve wrapping. An interactive GSLT
failing any of these is an object of $\iGSLT$ with no preimage under
$U$.
\end{proof}

\begin{remark}[$\lambda$ exhibits the distinction]
\label{rem:lambda-distinction}
The $\lambda$-calculus is an object of $\iGSLT$ \emph{as such}: terms,
$\alpha$, $\beta$. It becomes an object of $\ciGSLT$ only once one
\emph{chooses} the added structure --- the de~Bruijn section for (ii),
the degenerate-$\Ke$ interaction-cut presentation of $\beta$ for (i),
substitution-on-thunks for (iii) (Section~\ref{sec:examples}). The same
underlying calculus thus sits at both levels, witnessing that
``continued'' is added structure and not a property read off the
interactive GSLT alone. This is exactly the role $\lambda$ plays as a
control: each clause is a real question about it, with a real answer.
\end{remark}

The morphism conditions are dual and jointly necessary.
Bisimulation-preservation forbids distinguishing too much
(behaviourally); quote-faithfulness forbids collapsing too much
(representationally). A morphism merging two signatures would let a term
signed by one drain a token minted for the other, so the target would
gain transitions the source lacked. $\Cost$ lives exactly where both are
jointly satisfiable.

\begin{remark}[Wrappability replaces traversability]
The condition (iii) is the wrapped-by-construction successor of the
earlier ``traversable contractum'' requirement. We no longer need to walk
a contractum to re-sign its exposed redexes; we need only that the
contraction be a well-sorted operation on thunks. An opaque contraction
that destroyed wrapping --- returning bare redexes --- would violate
(iii) and is excluded, which is the same exclusion as a contraction whose
cost could not be accounted.
\end{remark}

\subsection{The functor on objects}

\begin{construction}[$\Cost$ on objects]
\label{con:object}
For $G = (\Sigma,E,R,K,\Kp,\Ke,K',\compute,\cf) \in \ciGSLT$, let
$\Cost(G)$ be the continued interactive GSLT with:
\begin{itemize}[leftmargin=1.6em,itemsep=2pt]
  \item \textbf{signature} $\Sigma$ extended by $\Sig$ (with
    $\hashf=\mathrm{digest}\circ\cf$, monoid $*,\nil$); the wrapped-term
    sort $\WT$ with $\sgn{\cdot}{s}$; the token-stack sort $\Tok$ with
    $\cons$; the polymorphic typing of $K$ over wrapped-term/stack and
    stack/stack pairs; and, crucially, the re-sorting of every
    continuation slot of $\Kp,\Ke$ to $\WT$;
  \item \textbf{equations} $E$ with the commutative-monoid equations on
    $(\Sig,*,\nil)$ and the inertness of $\Sig$ and $\Tok$ under $\red$;
  \item \textbf{rewrites} the gated family
    \eqref{eq:R1}--\eqref{eq:R3}, with the \emph{unchanged} contraction
    $\compute$ (now read at sort $\WT\times\WT$);
  \item \textbf{section} the evident lift of $\cf$ ignoring inert sorts.
\end{itemize}
\end{construction}

\begin{proposition}[Closure]
\label{prop:closure}
$\Cost(G)$ is again a continued interactive GSLT.
\end{proposition}

\begin{proof}[Proof sketch]
The gated rules are themselves in interaction-cut form: the outer
polymorphic $K$ is contact, the signed redex and the token stack are the
two operands, and $\compute$ (with the stack tail) is the contraction.
$\Cost(G)$ is section-equipped by the lifted $\cf$, and wrappable because
$\compute$ preserves $\WT$ by Lemma~\ref{lem:subject}. All three clauses
of Definition~\ref{def:ciGSLT} hold.
\end{proof}

\subsection{The functor on morphisms}

\begin{construction}[$\Cost$ on morphisms]
For $f : G \to H$ in $\ciGSLT$, let $\Cost(f)$ act as $f$ on the
interacting sort and transport the adjoined structure along $f$:
signatures by $\hashf_H\circ f$ on canonical forms, wrappers and stacks
componentwise, the polymorphic $K$ structurally.
\end{construction}

\begin{theorem}[$\Cost$ is an endofunctor on $\ciGSLT$]
\label{thm:functor}
$\Cost : \ciGSLT \to \ciGSLT$ preserves identities and composition, and
$\Cost(f)$ is a $\ciGSLT$-morphism whenever $f$ is.
\end{theorem}

\begin{proof}[Proof sketch]
Object closure is Proposition~\ref{prop:closure}; identity and
composition preservation are immediate from the componentwise definition.
By the factoring of bisimulation as ``$P$-bisimulation gated by key
matching,'' and since $f$ is $P$-bisimulation-preserving and
quote-faithful, $\Cost(f)$ preserves a gated transition iff $f$ preserves
the underlying $P$-transition and respects the key match --- both hold.
Quote-faithfulness transports along the injective $\hashf_H$.
\end{proof}

%% ══════════════════════════════════════════════════════════════════
\section{Graded adequacy}
\label{sec:adequacy}

The token stack grades transitions of $\Cost(G)$ by the signature monoid
$(\Sig,*,\nil)$: a forced step is labelled by the signature(s) it
consumes. Applying OSLF --- which auto-generates a Hennessy--Milner
logic~\cite{HennessyMilner1985} from a GSLT with an adequacy theorem, in
the lineage of the namespace logic for the rho
calculus~\cite{NamespaceLogic} --- to the graded transition system
of $\Cost(G)$ yields a \emph{graded} Hennessy--Milner logic with
modalities $\langle a\rangle_s$ indexed by the consumed signature.

\begin{theorem}[Graded adequacy, schematic]
\label{thm:adequacy}
For $\Cost(G)$ the graded Hennessy--Milner logic generated by OSLF is
adequate for quote-faithful bisimulation: two configurations satisfy the
same graded-HML formulae iff they are quote-faithfully bisimilar.
\end{theorem}

This also explains the throughput improvement mechanically. In the
translation presentation, fuel acquisition was an encoded multi-step
protocol on auxiliary channels; under $\Cost$ it is a single primitive
graded transition --- the grading absorbs the sub-protocol into the step
relation (\emph{protocol fusion}). With wrapping by construction the
fusion is total: there is no re-wrapping sub-step either, only forcing.
Together with Proposition~\ref{prop:modulus}, phlogiston is the conserved
grade, invariant under $\Seq$ (signatures inert), strictly consumed along
$\red$ (forcing drains the temporal stack), in the order recorded by
$\cons$.

%% ══════════════════════════════════════════════════════════════════
\section{Anchoring instances across the migration spectrum}
\label{sec:examples}

The construction applies uniformly across calculi whose interaction cut
carries no information migration, some, or migration by binding. We work
four instances, ordered by increasing migration.
Table~\ref{tab:spectrum} summarises the interaction surfaces,
continuations, and contraction for each.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.5}
\setlength{\tabcolsep}{6pt}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{2.3cm}
                  >{\centering\arraybackslash}p{1.9cm}
                  >{\centering\arraybackslash}p{1.9cm}
                  >{\centering\arraybackslash}p{2.0cm}
                  >{\raggedright\arraybackslash}p{4.2cm}@{}}
\toprule
\textbf{Instance} & \textbf{Surfaces} $I, J$ & \textbf{Prog.\ cont.}
  $C_{\mathrm p}$ & \textbf{Env.\ cont.} $C_{\mathrm e}$ &
  \textbf{Contraction} $\compute$ \\
\midrule
CCS
  & $a,\ \bar a$ \newline (complementary)
  & $P$
  & $Q$
  & pure release: $P \mathbin{|} Q$ \quad(no migration) \\
rho / $\pi$
  & $x,\ x$ \newline (name-equal)
  & $\lambda y.P$
  & $[u,Q]$
  & pseudo-app: $P\{u/y\} \mathbin{|} Q$ \quad(migration by binding) \\
$\lambda$
  & $x,\ {-}$ \newline ($\Ke$ degenerate)
  & $\lambda x.M$
  & $N$ \newline (no residual)
  & application: $M\{N/x\}$ \quad(migration, no release) \\
Interaction \newline categories
  & shared interface \newline (agreement)
  & process
  & process
  & interface composition \quad(no binding) \\
\bottomrule
\end{tabular}
\caption{The four anchoring instances across the migration spectrum.
A surface entry of ${-}$ denotes a nominally degenerate surface, carried
structurally by the rigidity of $K$ (Section~\ref{sec:capabilities})
rather than absent. Migration
is the substitution half of the contraction: absent in CCS and the
interaction categories, present by binding in rho, $\pi$, and $\lambda$.}
\label{tab:spectrum}
\end{table}

\subsection{CCS: the paradigmatic instance, null migration}

CCS~\cite{MilnerCCS1980} is the paradigmatic continued interactive GSLT
precisely because it exhibits the interaction cut with binding and
migration stripped away. Instantiate: $K = {\Par}$
(associative--commutative); the introductions are prefixes $a.P$ and
$\bar a.Q$ with interaction surfaces $I = a$, $J = \bar a$ and
continuations $C_{\mathrm p} = P$, $C_{\mathrm e} = Q$; the surface match
$\near(a,\bar a)$ is complementarity; and pseudo-application carries no
data, so $\compute$ degenerates to pure release, $C_{\mathrm p}' = P$,
$C_{\mathrm e}' = Q$. The gated rule \eqref{eq:R1} reads
%
\[
  \sgn{a.\sgn{P}{s_{\mathrm p}} \,\Par\, \bar a.\sgn{Q}{s_{\mathrm e}}}{s}
  \;\Par\; s\cons S
  \ \red\
  \sgn{P}{s_{\mathrm p}} \,\Par\, \sgn{Q}{s_{\mathrm e}} \;\Par\; S.
\]
%
No substitution occurs --- nothing migrates from environment to program ---
yet the construction goes through unchanged: the continuations are
wrapped thunks, forcing drains a token, and the AC structure of $\Par$
supplies the split cases. That CCS, with no binding anywhere, is a clean
continued interactive GSLT is the evidence that binding is incidental to
the construction and migration is optional.

\subsection{Communication (rho): migration by binding}

Instantiate at the rho calculus: $K = {\Par}$ (associative--commutative);
$\Kp = \mathrm{for}$ with surface $I = x$ and program continuation
$C_{\mathrm p} = \lambda y.T = \sgn{P}{s_{\mathrm p}}$ (under the abstraction
on $y$);
$\Ke = {!}$ with surface $J = x$ and environment continuation
$C_{\mathrm e} = [\,u, U\,] = \sgn{Q}{s_{\mathrm e}}$;
$\compute$ is pseudo-application, i.e.\ semantic substitution; $\cf$ a
normal form for the $\Par$-congruence. The surfaces $I = x = J$ match by
name-equality, and pseudo-application migrates the datum into the program
continuation through $y$. Rule \eqref{eq:R1} reads
%
\[
  \sgn{\mathrm{for}(y\!\leftarrow\! x)\,T \,\Par\, x!(U)}{s}
  \;\Par\; s\cons S
  \ \red\
  \subst{T}{\quot{U}}{y} \;\Par\; S,
\]
%
and since $T = \sgn{P}{s_{\mathrm p}}$ and $U = \sgn{Q}{s_{\mathrm e}}$,
the residual $\subst{T}{\quot{U}}{y} =
\sgn{\subst{P}{\quot{\sgn{Q}{s_{\mathrm e}}}}{y}}{s_{\mathrm p}}$ is again
a wrapped thunk, to be forced later by a token keyed to $s_{\mathrm p}$.
No re-wrapping occurs; the AC structure of $\Par$ supplies the split-redex
and split/combined-token cases \eqref{eq:R2}, \eqref{eq:R3} up to $\Seq$,
recovering the earlier five-rule lattice. Against CCS, the only difference
is that pseudo-application here has a non-trivial substitution half.

\subsection{\texorpdfstring{$\beta$}{beta}-reduction: migration, no environment residual}

Instantiate at the $\lambda$-calculus: $K = \mathrm{App}$ (free, no
equations); $\Kp = \lambda$ with surface $I = x$ and wrapped body
$C_{\mathrm p} = \sgn{M}{s_{\mathrm p}}$. The argument has a nominally
degenerate surface and no residual, so $\Ke$ is \emph{degenerate}: the
match condition is carried structurally by the rigid $\mathrm{App}$ rather
than by an explicit $J$ (Section~\ref{sec:capabilities}), and the
argument is its own wrapped environment continuation
$C_{\mathrm e} = \sgn{N}{s_{\mathrm e}}$. Pseudo-application is then
ordinary application --- substitution with empty release,
$\compute(x,-,\sgn{M}{s_{\mathrm p}},\sgn{N}{s_{\mathrm e}}) =
(\subst{\sgn{M}{s_{\mathrm p}}}{\sgn{N}{s_{\mathrm e}}}{x},\ \nil)$. Rule
\eqref{eq:R1} reads
%
\[
  \mathrm{App}\!\bigl(
    \sgn{\mathrm{App}(\lambda x.\sgn{M}{s_{\mathrm p}},\,
                      \sgn{N}{s_{\mathrm e}})}{s},
    \; s\cons S\bigr)
  \ \red\
  \mathrm{App}\!\bigl(\subst{\sgn{M}{s_{\mathrm p}}}{\sgn{N}{s_{\mathrm e}}}{x},\; S\bigr),
\]
%
with $s = \hashf_\lambda$ of the de~Bruijn form of the redex. Every copy
of $\sgn{N}{s_{\mathrm e}}$ that substitution places into $M$ carries its
wrapper, so duplicated redexes are guarded; and where $x$ stood in head
position, the landed $\sgn{N}{s_{\mathrm e}}$ guards the newly created
application --- both by construction, with no traversal
(Remark~\ref{rem:dup}). Token supply is \emph{not} function application;
the temptation to write $(\sgn{\mathrm{App}(\lambda x.M,N)}{s}\; s)$ feeds
the key as a value and gates nothing. The token sits at the contact site
and is consumed by forcing.

\subsection{Interaction categories: migration as interface composition}

The interaction categories of Abramsky, Gay, and
Nagarajan~\cite{AbramskyGayNagarajan1996} present interaction as
composition of processes along a shared typed interface. In the present
terms the interaction surfaces $I, J$ are the matching interface --- the
surface match $\near(I,J)$ is interface agreement --- and the contraction
is composition rather than substitution, with no binding. They occupy a
distinct point on the spectrum from CCS: still binding-free, but with
$\compute$ a genuine combination of the two continuations along their
shared boundary rather than a pure release. That they too are continued
interactive GSLTs shows the construction is indifferent to whether
migration is by substitution, by interface composition, or absent
entirely; what it requires is the interaction-cut shape, a section, and a
wrappable contraction.

%% ══════════════════════════════════════════════════════════════════
\section{The cost monad and two adjunctions}
\label{sec:monad}

Three further facts organise the construction: iterated cost accounting
flattens, so $\Cost$ carries a monad structure; and that monad relates to
its base through two different adjunctions --- one structural, one
computational --- whose embeddings have opposite characteristics.

\subsection{Flattening: \texorpdfstring{$\Cost$}{Cost} is a monad}

Applying the construction twice yields $\Cost^2(G) = \Cost(\Cost(G))$: a
calculus that meters the metering of $G$. Its signatures are signatures
of signed terms, its stacks are stacks of stacks, and its wrappers nest as
$\sgn{\sgn{P}{s_2}}{s_1}$. There is a canonical way to collapse the two
layers into one, and --- tellingly --- it is assembled entirely from the
two monoids already present in the construction.

\begin{construction}[Unit and multiplication]
\label{con:monad}
Define the \emph{unit} $\eta_G : G \to \Cost(G)$ to wrap each term with
the unit signature and present it against the freely available unit
token,
\[
  \eta_G(P) \;=\; \sgn{P}{\nil}.
\]
Since $\nil$ is the unit of $(\Sig,*,\nil)$, an $\nil$-keyed redex fires
without net resource; thus $\eta_G$ embeds $G$ as the \emph{cost-free
fragment} of $\Cost(G)$ --- the metered calculus restricted to
trivially-funded interactions. (This is consistent with the source/image
discipline of Section~\ref{sec:wrapped}: $\eta$ is precisely the wrapping
that installs apparatus at zero charge.)

Define the \emph{multiplication} $\mu_G : \Cost^2(G) \to \Cost(G)$ to
flatten the two layers by
\begin{enumerate}[label=(\roman*),itemsep=1pt]
  \item \textbf{multiplying nested signatures} in $\Sig$,
    $\sgn{\sgn{P}{s_2}}{s_1} \mapsto \sgn{P}{\,s_1 * s_2\,}$; and
  \item \textbf{concatenating the stack-of-stacks} into a single stack ---
    the free-monoid (list) multiplication on $\Tok$.
\end{enumerate}
The outer account is thereby merged into the inner: the grade of the
flattened redex is the product of the two grades, and its temporal stack
is the concatenation, in the order the two layers would have drained them.
\end{construction}

\begin{proposition}[The cost monad]
\label{prop:monad}
$(\Cost, \eta, \mu)$ is a monad~\cite{MacLane1998} on $\ciGSLT$. Its laws descend from the
laws of the two constituent monoids: the unit laws from
$s * \nil = s = \nil * s$ together with the singleton/empty laws of the
list monad, and associativity from associativity of $*$ and of list
concatenation.
\end{proposition}

\begin{proof}[Proof sketch]
$\eta$ and $\mu$ are $\ciGSLT$-morphisms. Each is quote-faithful, since
$*$ is injective in each argument up to the collision-resistance of the
digest, and bisimulation-preserving, since flattening neither creates nor
destroys gated transitions --- it relabels each by the product grade and
consolidates the (spatially commutative, temporally sequential)
consumption order consistently. Naturality is the componentwise action on
the adjoined sorts. The monad identities
$\mu \circ \Cost\eta = \id = \mu \circ \eta_{\Cost}$ reduce to the unit
laws of $*$ and of concatenation; $\mu \circ \Cost\mu = \mu \circ
\mu_{\Cost}$ reduces to their associativity.
\end{proof}

\begin{remark}[A genuine, non-idempotent resource monad]
$\Cost$ is not idempotent: $\mu_G$ is no isomorphism, because
concatenating two stacks forgets the boundary between them and
multiplying two grades forgets the factorisation. Flattening is a true
\emph{merge} of two resource accounts, not the collapse of a redundant
copy. This is the expected signature of a resource monad --- re-metering a
metered calculus yields a finer account, which $\mu$ then consolidates ---
and it distinguishes $\Cost$ from idempotent ``closure'' monads.
\end{remark}

The Eilenberg--Moore algebras of $\Cost$ are continued interactive GSLTs
equipped with a coherent \emph{discharge} map $\alpha : \Cost(A) \to A$
interpreting tokens --- a calculus that knows how to ``pay'' --- and the
Kleisli category is that of cost-accounted translations between calculi.
The structural adjunction below is the free resolution generating the
monad; the Turing-complete adjunction is a separate phenomenon.

\subsection{Adjunction I: install \texorpdfstring{$\dashv$}{-|} forget}

Let $\CostC$ be the category of \emph{cost-accounted} continued
interactive GSLTs: objects carry the installed apparatus
($\Sig$, $\Tok$, the wrapper, the gated family), morphisms preserve it.
A forgetful functor $\Forget : \CostC \to \ciGSLT$ strips the apparatus,
returning the underlying un-metered calculus; a free functor
$\Free : \ciGSLT \to \CostC$ installs it --- exactly
Construction~\ref{con:object}.

\begin{proposition}[Free--forgetful resolution]
\label{prop:adj1}
$\Free \dashv \Forget$, with unit $\eta$ of Construction~\ref{con:monad},
and the induced monad $\Forget \circ \Free$ on $\ciGSLT$ is $\Cost$.
\end{proposition}

The embedding is \emph{structural and strict}: $\Free$ adjoins sorts and
rules, $\Forget$ erases them on the nose. Crucially, forgetting is
\emph{not} behaviour-preserving in the metering sense --- erasing the
apparatus removes the gating, so $\Forget\,\Free\,G$ runs $G$'s
interactions \emph{without friction}. The structural resolution answers
``what apparatus was added?'' and lets us remove it; it does not claim the
metered and un-metered dynamics agree. Its character is
\textbf{structure-preserving, behaviour-altering}.

\subsection{Adjunction II: internalise \texorpdfstring{$\dashv$}{-|} include, over a universal base}

Restrict to the full subcategory $\ciGSLTtc \hookrightarrow \ciGSLT$ of
\emph{Turing-complete} continued interactive GSLTs. For such a $G$, the
entire metering apparatus of $\Cost(G)$ --- signature equality, stack
maintenance, gating --- is computable, hence \emph{encodable in $G$
itself}. This yields an \emph{internalisation}
\[
  \Imp_G : \Cost(G) \longrightarrow G,
\]
realising the cost-accounted semantics inside the un-metered but
universal base by simulation: tokens become encoded data, the gated rules
become an interpreter loop, and each forced step is simulated by a finite
run of $G$.

\begin{proposition}[Internalisation as an adjoint retraction]
\label{prop:adj2}
For $G \in \ciGSLTtc$ the internalisation satisfies
$\Imp_G \circ \eta_G \approx \id_G$ up to weak bisimulation, exhibiting
$\Cost(G)$ as a behavioural retract of $G$. In the bicategory of
continued interactive GSLTs, simulations, and simulation transformations,
$\eta_G \dashv \Imp_G$ is an adjoint retraction: the unit
$\eta_G$ is an (iso-up-to-bisimulation) section and the counit is the
collapsing simulation $\eta_G \circ \Imp_G \Rightarrow \id_{\Cost(G)}$.
\end{proposition}

\begin{proof}[Proof sketch]
Turing completeness of $G$ gives a computable encoding of the finite data
and decidable guards of $\Cost(G)$; standard interpreter constructions
make $\Imp_G$ a weak simulation that reflects every gated transition.
Composing with $\eta_G$ (which installs apparatus at unit cost) returns
$G$'s dynamics up to the administrative reductions of the interpreter,
i.e.\ up to weak bisimulation. The triangle identities hold up to the
2-cells witnessing these weak bisimulations, which is the content of an
adjunction internal to the simulation bicategory.
\end{proof}

The embedding is the mirror image of Adjunction~I. It is
\emph{behaviour-preserving} --- the simulation faithfully reproduces every
gated step --- but \emph{structure-encoding rather than
structure-preserving}: the clean sorts $\Sig, \Tok$ dissolve into encoded
data, and the agreement holds only up to weak bisimulation, since the
interpreter interposes administrative reductions. Its character is
\textbf{behaviour-preserving, structure-dissolving}, exactly opposite to
Adjunction~I.

\begin{remark}[Two senses of conservativity, and where they hold]
The two adjunctions express two senses in which cost accounting is
conservative over its base. Structurally (Adjunction~I, available for
every object) the apparatus is a detachable layer. Computationally
(Adjunction~II) the apparatus is already expressible, so metering is a
definitional extension up to weak bisimulation --- but this holds
\emph{only} when the base is Turing complete. The second is therefore a
genuine property of the object: universal interactive theories are
exactly those whose cost-accounted versions fold back into them, a
distinction invisible to the purely structural resolution.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{Capabilities: located authority and the theory of \texorpdfstring{$K$}{K}}
\label{sec:capabilities}

A cost-accounted calculus is a calculus in which the right to draw on a
resource stack is itself an authority. We must therefore ask what confers
that authority, and ensure the construction does not silently grant it ---
that it does not introduce \emph{ambient
authority}~\cite{MillerYeeShapiro2003,Miller2006}, the antithesis of an
object-capability discipline. The answer turns entirely on the equational
theory of the interaction constructor $K$, and it is most cleanly stated
through the interaction surface of Section~\ref{sec:metered-cut}. The
surface --- the part that must match for a cut to fire --- can be carried
in two ways. It may be carried \emph{structurally}, by the rigidity of
$K$: when $K$ is non-AC, position is unforgeable and the context
$K([\,],-)$ itself is the surface. Or it may be carried \emph{nominally},
by an explicit $I, J$ matched by a predicate, which becomes necessary
precisely when $K$ is AC and position is no longer exclusive. A
nominally degenerate surface, as on $\lambda$'s environment side, is thus
not a missing surface but a structurally carried one. These are the two
faces of a single notion, and which one a calculus uses is dictated by
$K$.

\subsection{The structural surface: position as capability when \texorpdfstring{$K$}{K} is rigid}

When $K$ carries no equations, position is unforgeable: nothing can drift
into contact with a stack it does not already neighbour. The rigidity of
$K$ thereby \emph{supplies} the interaction surface --- the surface is not
recorded by any explicit $I, J$ but by the structural context. A
capability to draw on a stack $S$ is then literally that context
\[
  \Kctx{S},
\]
the right to occupy the hole adjacent to $S$ --- to be ``next to the
cookie jar.'' Because $K$ is rigid, occupying that position is the whole
of the authority: no other program can manoeuvre itself into the same
position, so confinement is automatic and the capability is exactly the
context. This is the $\lambda$-calculus regime, and it is the
object-capability ideal: authority is structural, held by reference, and
unforgeable by construction. This is also what the degenerate environment
surface of $\beta$ (Section~\ref{sec:metered-cut}) amounts to --- not the
absence of a surface but its structural provision by the rigid
$\mathrm{App}$.

\subsection{The leak of ambient authority when \texorpdfstring{$K$}{K} is AC}

When $K$ is associative--commutative --- the rho regime, $K = {\Par}$ ---
position is no longer exclusive. The commutative bag lets arbitrary
programs float into adjacency with a stack and compete for it. Concretely,
because $\Par$ is AC the configuration
\[
  K\bigl(\Kp^1(I^1,C_{\mathrm p}^1),\;
         \Kp^2(I^2,C_{\mathrm p}^2),\;
         \Ke(J,C_{\mathrm e})\bigr)
\]
is well-formed, and \emph{both} introductions $\Kp^1, \Kp^2$ sit adjacent
to the eliminand. Mere adjacency can no longer be the capability, because
adjacency has become cheap and non-exclusive: every program in the bag is
``next to'' every stack in it. Taking proximity as authority here is
exactly ambient authority, and it is a leak.

\subsection{The nominal surface: the nearness predicate when \texorpdfstring{$K$}{K} is AC}

To recover an object-capability discipline under an AC $K$ --- where the
structural surface is unavailable because position is non-exclusive --- we
carry the surface nominally instead. The interaction surfaces $I$ (of the
receiving introduction) and $J$ (of the eliminand) are the parts that must
match, and we demand a nearness predicate
\[
  \near(I, J)
\]
as a precondition of interaction: the gated rules of
Section~\ref{sec:wrapped} fire only between an $I$ and a $J$ with
$\near(I,J)$. Interaction is then licensed not by position but by the
matchability of the two surfaces; $I$ and $J$ become a \emph{proxy for
capability}. The predicate specialises across the spectrum: it is
complementarity of actions in CCS, name-equality of subjects in
rho and $\pi$, and interface agreement in the interaction categories.

The two carriers are now one notion seen twice. In the rigid case the
surface is the context $\Kctx{S}$ and $\near$ is discharged for free ---
being in the hole \emph{is} being near. In the AC case the rigid context
is no longer exclusive, so the same surface must be carried nominally by
$I, J$ and checked by $\near$. The match condition has not changed; only
its carrier has, from the geometry of $K$ to an explicit predicate.
Equivalently: a rigid $K$ supplies the surface structurally and needs no
gate, while an AC $K$ has dissolved the structural surface into the bag
and must reintroduce it nominally to avoid ambient authority.

\begin{remark}[Nearness bounds but does not eliminate competition]
$\near(I,J)$ does not by itself make access exclusive: two receivers
$I^1, I^2$ may both be near the same $J$, so competition for a stack
remains possible. What the predicate guarantees is that only holders of a
\emph{matching surface} may compete at all --- it replaces ``anyone
adjacent'' with ``anyone holding a matching surface,'' which is precisely
the recovery of capability discipline. Exclusivity, where required, is an
additional matter for the type discipline of Section~\ref{sec:types}.
\end{remark}

\subsection{Located resource stacks}

The nearness analysis motivates refining the stack constructor itself.
Rather than a bare list of signatures floating in the AC bag, a stack is
\emph{located} at an interaction surface:
\[
  \lstk{I}{s \cons S'},
\]
a stack indexed by the surface $I$ that may draw on it. The cookie jar now
carries a label naming which hands may reach in. Locating the stack turns
``reachable by whatever is adjacent'' into ``reachable by a matching
surface,'' so authority over a resource is named in the term rather than
conferred by the geometry of the bag. This is the form on which the type
discipline and the resource-sufficiency proofs of the next two sections
operate; in the rigid regime the location is implicit in the context
$\Kctx{S}$, while under an AC $K$ it is the explicit carrier of authority.

\begin{remark}[Naming coincidences in the rho instance]
In rho the receiver's surface is $I \sim (y,x)$ and the eliminand's is
$J \sim x$, so $\near(I,J)$ partly coincides with channel-name matching:
the channel already does surface-like work. The general construction
separates the two notions --- name matching selects \emph{which token} is
spent (the signature guard), while $\near$ governs \emph{who may approach}
the stack at all. Rho fuses them on the channel name, as it fused quotient
and section (Section~\ref{sec:signatures}); the located form keeps them
distinct.
\end{remark}

%% ══════════════════════════════════════════════════════════════════
\section{A type discipline for token usage}
\label{sec:types}

Linearity of tokens should not be imposed by fiat in the operational
semantics, for some token types ought to be \emph{copyable} --- freely
duplicable authorisations --- while others are \emph{linear} --- spent
exactly once. Rather than hardwire a single choice, we let the discipline
be \emph{checked}: a programmer asserts the usage discipline a term
respects, and the assertion is verified against a type system. The type
system is not new machinery; it is the logic OSLF already generates.

\subsection{OSLF as a checkable discipline}

Running OSLF on the cost-accounted theory $\Cost(G)$ --- not on the bare
$G$ --- yields a Hennessy--Milner-style logic for the metered calculus,
adequate for quote-faithful bisimulation (Theorem~\ref{thm:adequacy}).
We read this logic as a type system: a term is \emph{typed} by checking
that it satisfies a formula expressing the intended token discipline. The
discipline is opt-in and per-term: the calculus does not force linearity,
but a programmer who requires it writes a term and checks it against the
linear formula. Copyable tokens are simply those whose types impose no
single-use constraint.

\subsection{Spatial and modal connectives suffice}

Two connective families generated by OSLF are enough to express a variety
of token disciplines.

\emph{Spatial types} $K(\varphi_1, \varphi_2)$ read the interaction
constructor as a separating-style connective: a term satisfies
$K(\varphi_1,\varphi_2)$ when it splits across $K$ into a part satisfying
$\varphi_1$ and a part satisfying $\varphi_2$. This is the bookkeeping of
\emph{duplication and partition} --- whether token-bearing sub-terms are
shared or held disjointly, and how many copies are present. When $K$ is
AC this is a genuine separating conjunction in the sense of spatial logic;
when $K$ is rigid it is a positional split. The discipline of linearity is
thus, in its static aspect, spatial. The reading of an interaction
constructor as a separating conjunction is that of spatial and separation
logics~\cite{CairesCardelli2003,OHearnReynolds2001,Reynolds2002}.

\emph{Modal types} $\dia{K(\dots)}\varphi$ are the graded modalities of
Section~\ref{sec:adequacy}, indexed by the interaction: a term satisfies
$\dia{K(\dots)}\varphi$ when it can take a $K$-interaction --- a draw on
the stack --- after which $\varphi$ holds. This is the dynamics of
\emph{consumption}: that a spend is available and what the post-state
looks like.

Combining the two expresses the usual lattice of disciplines. Writing
$\ell$ for a token-bearing predicate, linearity is a spatial constraint
forbidding a second copy ($\neg\, K(\ell, K(\ell, \top))$, no split
exposing two live copies) together with a modal obligation that the single
copy be spent; copyability drops the spatial constraint; relevance keeps
the modal obligation while permitting copies. The point is not a fixed
type system but that the OSLF-generated logic of $\Cost(G)$ is already
expressive enough to host them.

\subsection{Finiteness and location make checking tractable}

Two properties of the located stack of Section~\ref{sec:capabilities} turn
this from specifiable into statically decidable.

\emph{Finiteness}: if every token stack is finitely bounded --- of
statically known, finite depth --- the spatial connective accounts over a
bounded budget, so ``is there enough'' and ``is there at most one copy''
are finite checks rather than reasoning about an unbounded resource.

\emph{Location}: because each stack sits at a named surface $\lstk{I}{-}$
rather than floating in the bag, the analysis is local --- consumption is
reasoned about per surface, not globally over all adjacent competitors.
Finiteness bounds the quantity; location bounds the scope. Together they
make the spatial/modal discipline decidable, and they are exactly the
refinements the capability analysis already forced.

%% ══════════════════════════════════════════════════════════════════
\section{Resource sufficiency: linear proofs and located purses}
\label{sec:linear}

In the rho-specific development, the guarantee that a deployment carries
enough phlogiston to run to completion was discharged by a \emph{linear
proof}: a derivation in a linear logic~\cite{Girard1987} witnessing that the available
tokens suffice to cover every interaction the term will perform, checked
once at deploy time. We ask whether this lifts to the general $\Cost(G)$
setting, and whether the located stacks of
Section~\ref{sec:capabilities} improve it.

\subsection{Data-dependent interaction}

A linear proof certifies a \emph{fixed} consumption: it sums the tokens a
term will draw and checks the sum against the supply. This is sound when
the interactions a term performs are determinable statically. Under
\emph{data-dependent} interaction it is not immediate. If a received value
selects a branch, or a name computed at run time determines a $\near$
match, then which interactions fire --- and so how many tokens of which
signatures are drawn --- depends on the run, and there is no single fixed
quantity for the proof to sum.

Two honest responses follow. For the data-\emph{independent} fragment the
rho-era guarantee lifts verbatim: the consumption is fixed and the linear
proof bounds it. For the data-\emph{dependent} fragment the proof must
either become \emph{dependent} --- parameterised by the data, with the
branch condition carried in the modal type $\dia{K(\dots)}\varphi$ of
Section~\ref{sec:types}, so that each branch discharges its own linear
obligation --- or fall back to a \emph{conservative} static bound, the
worst case over branches, which is sound but may over-charge. The linear
proof therefore suffices unconditionally only for the static fragment;
data-dependence requires either dependency in the proof or a conservative
envelope.

\subsection{Located purses as an optimisation}

The located stacks $\lstk{I}{-}$ --- ``purses'' parked at interaction
surfaces --- optimise the linear proof regardless of which response above
is taken. Instead of one global linear derivation accounting for the whole
term's draw against a single shared pool, locating the resources
decomposes the obligation into \emph{one local linear proof per purse},
each bounding only the consumption at its surface. The spatial connective
$K(\varphi_1,\varphi_2)$ is precisely the proof-theoretic seam along which
the global proof factors: a separating split of the term induces a split
of its resource obligation into independent conjuncts.

\begin{proposition}[Local sufficiency composes]
\label{prop:local-suff}
If each purse $\lstk{I}{-}$ admits a local linear proof that its supply
covers the consumption at surface $I$, and the surfaces partition the
term's interactions, then the conjunction of the local proofs is a proof
of global resource sufficiency. The global linear derivation factors as
the separating conjunction of the per-purse derivations.
\end{proposition}

This is the lifted and sharpened form of the rho-era guarantee. Beyond
mere economy of proof, location \emph{quarantines} the hard case: if
data-dependence is confined to a particular surface, only that purse's
sub-proof needs the dependent or conservative treatment, while every other
purse retains a simple static linear proof. Locality thus does double
duty --- it makes the sufficiency proof compositional, and it contains the
data-dependent reasoning to the surfaces that actually require it.

%% ══════════════════════════════════════════════════════════════════
\section{Conclusion}
\label{sec:conclusion}

The cost-accounting move of the rho calculus is an instance of a generic
endofunctor on the category $\ciGSLT$ of \emph{continued} interactive
GSLTs: interactive theories presented as an interaction cut
(contact $K$, two introductions $\Kp(I,C_{\mathrm p})$,
$\Ke(J,C_{\mathrm e})$ meeting along matching interaction surfaces $I,J$,
contraction $\compute$), equipped with a
computable section of their structural-congruence quotient, and with
continuation slots sorted as \emph{wrapped terms by construction}. The
functor $\Cost$ signs terms, adjoins a token stack, makes the contact
constructor $K$ polymorphic, and replaces the cut by a gated family. Its
soundness is a sorting invariant: every redex is born inside a wrapper,
reduction preserves wrapping (Lemma~\ref{lem:subject}), so no computation
proceeds for free and the contraction need not be lifted --- the
cost-accounting steps only ever unwrap. Metering is thereby lazy, paying
for work performed rather than exposed, and the temporal token stack is
consumed exactly in step with reduction, realising the modulus of cut
elimination as the run length. The cons operator is extension in time,
dual to the spatial extension of $K$; OSLF lifts $\Cost(G)$ to a graded
Hennessy--Milner logic with adequacy ``graded-HML equivalence equals
quote-faithful bisimulation.'' The contraction is Milner's
pseudo-application, and \emph{binding is one mode of information migration}
under it, not a feature of the cut: CCS instantiates the same shape with
no migration and is in that sense paradigmatic, the interaction categories
of Abramsky, Gay, and Nagarajan migrate by interface composition, and the
rho and $\lambda$ calculi migrate by substitution. The $\lambda$-calculus,
with $K$ and the
section visibly distinct and $\Ke$ degenerate, shows the construction
requires a section --- de~Bruijn canonicalisation --- not a channel
constructor; rho's apparent need for two axes of reflection was the
self-overlap of one operator serving as both quotient and section.

Finally, the construction is deliberately sited one rung below the
ambient category: $\Cost$ is an endofunctor on $\ciGSLT$, not on
$\iGSLT$, and the forgetful functor $U : \ciGSLT \to \iGSLT$
(Definition~\ref{def:forget}, Proposition~\ref{prop:proper}) records that
``continued'' is a proper structural enrichment of ``interactive.'' The
content of the note is therefore as much about \emph{where} cost
accounting lives as about what it does: an interactive GSLT is a setting
for interaction; a continued interactive GSLT is one whose interaction is
presented as a meterable interaction cut, and it is precisely on these that the cost
endofunctor exists --- indeed, as a monad
(Section~\ref{sec:monad}), since iterated metering flattens. Its two
resolutions sharpen the picture: the structural free--forgetful
adjunction shows the apparatus is detachable, while over a Turing-complete
base the internalisation adjunction shows it is already expressible, so
that cost accounting is a definitional extension up to weak bisimulation.
That second adjunction holds only for universal bases, and so quietly
begins to sort interactive theories by a structural property --- a thread
we leave for the sequel, where calculi whose contraction is not
substitution but a spatial rewrite, such as the mobile ambient
calculus~\cite{CardelliGordon2000}, will test which of the present
phenomena depend on substitution specifically and which on any
traversable contraction.

Three further refinements turn the metered calculus into a usable
resource discipline, and each is governed by the same structural feature
--- the equational theory of $K$ --- that organised the construction. The
authority to draw on a resource is a capability, and whether the
construction grants it ambiently depends on whether $K$ mixes: a rigid
$K$ makes position an unforgeable capability $\Kctx{S}$, while an AC $K$
demands an explicit nearness predicate $\near(I,J)$ and a stack
\emph{located} at its surface, $\lstk{I}{-}$, to keep authority from
becoming ambient. On the metered calculus, OSLF generates not merely a
behavioural logic but an opt-in type discipline: spatial types
$K(\varphi_1,\varphi_2)$ and modal types $\dia{K(\dots)}\varphi$ together
express linear, copyable, and relevant token disciplines, made statically
decidable by the finiteness and location of stacks. And the rho-era
guarantee of resource sufficiency by linear proof lifts: verbatim for the
data-independent fragment, with dependent or conservative treatment for
data-dependent interaction, and --- crucially --- factored by located
purses into a separating conjunction of per-surface proofs, which both
makes sufficiency compositional and quarantines the data-dependent case
to the surfaces that require it. Authority, usage, and sufficiency thus
form three layers on the one cost-accounted object: who may reach a stack,
how a holder may spend what they draw, and whether the supply suffices ---
each a checkable discipline, each turning on the structure of $K$.

\subsection*{A spacetime reading}

We close with an analogy that, while not a theorem of this note, is
detailed enough in its correspondence to invite further investigation,
and that connects the construction to the geometric reading of GSLTs.
Recall the two monoids of Section~\ref{sec:two-monoids}: $K$ is spatial,
recording what is near what, and $\cons$ is temporal, recording what is
spent next. Read a $K$-tree as an initial distribution of matter and
energy --- the information carried by the program and environment
continuations, arranged in space. Cut elimination evolves this
distribution: each interaction consumes two introductions, migrates
information through the contraction, and leaves a contractum whose redexes
have a \emph{new} adjacency. What is near what is therefore not fixed but
\emph{dynamical}, changing as the computation runs.

The shape of this dependence is the shape of Einstein's intuition. In
general relativity the distribution of matter and energy determines the
metric --- what is near what --- and the metric in turn governs how matter
moves. Here the distribution of information in the $K$-tree determines the
nearness structure (which cuts are enabled), and the nearness structure
governs how information flows at the next step, which redistributes the
information. The analog of the stress--energy source is the information
distribution; the analog of the metric is the nearness structure; and cut
elimination is the coupling between them, each step letting the
distribution reshape the geometry and the geometry select the next
redistribution.

The analogy sharpens, rather than dissolves, when object capabilities
enter. Without an authority discipline, space ($K$) and time ($\cons$) are
separable: nearness and consumption order are independent coordinates.
But the located stack $\lstk{I}{-}$ of Section~\ref{sec:capabilities} ties
a temporal resource --- what may be spent next --- to a spatial surface
--- who is near. A draw then requires \emph{both} spatial proximity (a
matching surface, $\near(I,J)$) and temporal availability (a token at the
head of the stack), so the spatial and temporal structures can no longer
be varied independently. Authority \emph{entangles} them. What was a
product of a space and a time becomes a single coupled structure gating
interaction --- a spacetime. That the same structural feature, the theory
of $K$, governs both the geometry and its entanglement with time is the
part of the correspondence we find most worth pursuing, and we leave its
formal development --- whether the nearness evolution obeys anything like a
field equation --- to future work.

\begin{thebibliography}{99}
\addcontentsline{toc}{section}{References}

\bibitem{MeredithRadestock2005}
L.~G.~Meredith and M.~Radestock.
\newblock A reflective higher-order calculus.
\newblock In M.~Viroli, editor, \emph{ETAPS 2005 Satellite Events},
  Electronic Notes in Theoretical Computer Science, vol.~141(5),
  pp.~49--67. Elsevier, 2005.

\bibitem{NamespaceLogic}
L.~G.~Meredith and M.~Radestock.
\newblock Namespace logic: A logic for a reflective higher-order calculus.
\newblock In \emph{Trustworthy Global Computing (TGC 2005)}, Lecture Notes
  in Computer Science, vol.~3705, pp.~353--369. Springer, 2005.

\bibitem{StayMeredith2017}
M.~Stay and L.~G.~Meredith.
\newblock Representing operational semantics with enriched Lawvere theories.
\newblock In \emph{Proceedings of the Higher-Dimensional Rewriting and
  Applications workshop (HDRA)}, 2017.
  \newblock arXiv:1704.03080.

\bibitem{BaezWilliams2019}
J.~C.~Baez and C.~Williams.
\newblock Enriched Lawvere theories for operational semantics.
\newblock \emph{Applied Category Theory (ACT)}, 2019.
  \newblock arXiv:1905.05636.

\bibitem{Milner1999}
R.~Milner.
\newblock \emph{Communicating and Mobile Systems: the $\pi$-Calculus}.
\newblock Cambridge University Press, 1999.

\bibitem{MilnerPolyadic1993}
R.~Milner.
\newblock The polyadic $\pi$-calculus: a tutorial.
\newblock In F.~L.~Bauer, W.~Brauer, and H.~Schwichtenberg, editors,
  \emph{Logic and Algebra of Specification}, NATO ASI Series, vol.~94,
  pp.~203--246. Springer, 1993.

\bibitem{MilnerCCS1980}
R.~Milner.
\newblock \emph{A Calculus of Communicating Systems}.
\newblock Lecture Notes in Computer Science, vol.~92. Springer, 1980.

\bibitem{AbramskyGayNagarajan1996}
S.~Abramsky, S.~J.~Gay, and R.~Nagarajan.
\newblock Interaction categories and the foundations of typed concurrent
  programming.
\newblock In M.~Broy, editor, \emph{Deductive Program Design:
  Proceedings of the 1994 Marktoberdorf Summer School}, NATO ASI Series,
  pp.~35--113. Springer, 1996.

\bibitem{SangiorgiWalker2001}
D.~Sangiorgi and D.~Walker.
\newblock \emph{The $\pi$-Calculus: A Theory of Mobile Processes}.
\newblock Cambridge University Press, 2001.

\bibitem{HennessyMilner1985}
M.~Hennessy and R.~Milner.
\newblock Algebraic laws for nondeterminism and concurrency.
\newblock \emph{Journal of the ACM}, 32(1):137--161, 1985.

\bibitem{MillerYeeShapiro2003}
M.~S.~Miller, K.-P.~Yee, and J.~Shapiro.
\newblock Capability myths demolished.
\newblock Technical Report SRL2003-02, Systems Research Laboratory,
  Johns Hopkins University, 2003.

\bibitem{Miller2006}
M.~S.~Miller.
\newblock \emph{Robust Composition: Towards a Unified Approach to Access
  Control and Concurrency Control}.
\newblock PhD thesis, Johns Hopkins University, 2006.

\bibitem{OHearnReynolds2001}
P.~W.~O'Hearn, J.~C.~Reynolds, and H.~Yang.
\newblock Local reasoning about programs that alter data structures.
\newblock In \emph{Computer Science Logic (CSL 2001)}, Lecture Notes in
  Computer Science, vol.~2142, pp.~1--19. Springer, 2001.

\bibitem{Reynolds2002}
J.~C.~Reynolds.
\newblock Separation logic: A logic for shared mutable data structures.
\newblock In \emph{Proceedings of the 17th Annual IEEE Symposium on Logic
  in Computer Science (LICS 2002)}, pp.~55--74. IEEE, 2002.

\bibitem{CairesCardelli2003}
L.~Caires and L.~Cardelli.
\newblock A spatial logic for concurrency (part I).
\newblock \emph{Information and Computation}, 186(2):194--235, 2003.

\bibitem{Girard1987}
J.-Y.~Girard.
\newblock Linear logic.
\newblock \emph{Theoretical Computer Science}, 50(1):1--101, 1987.

\bibitem{deBruijn1972}
N.~G.~de Bruijn.
\newblock Lambda calculus notation with nameless dummies, a tool for
  automatic formula manipulation, with application to the Church--Rosser
  theorem.
\newblock \emph{Indagationes Mathematicae}, 34(5):381--392, 1972.

\bibitem{SmithdesRivieres1984}
B.~C.~Smith and J.~des Rivi\`eres.
\newblock Interim 3-LISP reference manual.
\newblock Technical report, Xerox PARC, 1984.

\bibitem{CardelliGordon2000}
L.~Cardelli and A.~D.~Gordon.
\newblock Mobile ambients.
\newblock \emph{Theoretical Computer Science}, 240(1):177--213, 2000.

\bibitem{MacLane1998}
S.~Mac Lane.
\newblock \emph{Categories for the Working Mathematician}, 2nd edition.
\newblock Graduate Texts in Mathematics, vol.~5. Springer, 1998.

\end{thebibliography}

\end{document}
