\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{array}
\usepackage[table]{xcolor}
\usepackage{siunitx}
\usepackage{enumitem}
\usepackage[expansion=false,protrusion=true]{microtype}
\DisableLigatures[f]{encoding = *}
\usepackage[hidelinks]{hyperref}
\usepackage{titlesec}
\usepackage{caption}

% --- F1R3FLY brand accent ---
\definecolor{f1r3red}{HTML}{D7263D}
\definecolor{f1r3ink}{HTML}{1A1A1A}
\definecolor{f1r3ash}{HTML}{6B6B6B}
\definecolor{coolrow}{HTML}{EAF5EC}
\definecolor{hotrow}{HTML}{FBE9EC}
\definecolor{headrow}{HTML}{F4E1D2}

\titleformat{\section}{\color{f1r3red}\large\bfseries}{\thesection}{0.6em}{}
\titleformat{\subsection}{\color{f1r3ink}\normalsize\bfseries}{\thesubsection}{0.6em}{}

\hypersetup{colorlinks=true,linkcolor=f1r3red,citecolor=f1r3red,urlcolor=f1r3red}

\newcommand{\near}{\operatorname{near}}
\sisetup{group-separator={,},group-minimum-digits=4}

\newtheorem{remark}{Remark}
\newtheorem{observation}{Observation}

\title{\color{f1r3red}\bfseries Sector Economics and Go-to-Market Viability\\[2pt]
\large\color{f1r3ink}\mdseries When a F1R3FLY shard pays for itself, and which markets get it there\\[2pt]
\normalsize\color{f1r3ash}\mdseries A companion to ``Rent and Shard Splitting''}
\author{\normalsize F1R3FLY Industries --- working note}
\date{\normalsize \today}

\begin{document}
\maketitle

\noindent\textit{A companion note. It reuses the cost model, the dense-packing result, and the rent lifecycle of ``Rent and Shard Splitting''~\cite{rentnote} and asks the next question: per market sector, how much monthly traffic does a shard need to be economically viable, which sectors produce it, and in what order should we deploy. All dollar figures are engineering estimates on stated assumptions; every fee is a lever, flagged as such, so the tables re-derive when real schedules replace ours.}

\section{The viability condition}
\label{sec:viability}

A shard is viable when monthly phlogiston revenue covers its cost. From the rent note, one ten-node shard costs $C \approx \$4{,}160$/month all-in, and dense elastic packing~\cite{fkk2025} lets it carry on the order of $\num{30000000}$ fee-bearing transactions per month before the split threshold binds. If a sector's transactions bear a blended fee $r_{\text{tx}}$ (precharge, activation, and rent combined, in dollars), then break-even monthly traffic is simply
\[
  T^\star \;=\; \frac{C_{\text{eff}}}{r_{\text{tx}}},
  \qquad
  C_{\text{eff}} \;=\; C \times (\text{compliance multiplier}).
\]
Risk pushes on \emph{both} sides of this ratio. A riskier sector can charge a higher $r_{\text{tx}}$, but it also raises $C_{\text{eff}}$: replication, audit, and dedicated infrastructure break the ``all co-resident on one cloud'' assumption. The useful dimensionless quantity is therefore not $T^\star$ but the \emph{break-even load}, $T^\star$ divided by per-shard capacity---the fraction of a shard a sector must fill to pay for it.

\begin{observation}[Capacity is never the gate; demand is.]
Every sector below is feasible on a single shard's \emph{capacity}---even the highest-volume case sits far under the split threshold per shard. Viability is gated entirely by whether a sector produces enough fee-bearing transactions, and the break-even load varies by more than two orders of magnitude across sectors. Riskier data is \emph{easier} to make pay, not harder.
\end{observation}

\paragraph{Common parameters.} Unless noted: base cost $C=\$4{,}160$/shard/month; per-payload network cost at dense packing $C/\num{30000000}\approx\$0.00014$; low-risk shards run dense (${\sim}\num{28000000}$--$\num{30000000}$ tx/mo), mission-critical shards run with headroom (${\sim}\num{20000000}$ tx/mo). Compliance multipliers: AI $1.2\times$, music $1\times$, biotech $1.5\times$, fintech $2.5\times$, medical $4\times$.

\section{AI (ASI Alliance) --- the growth engine}
\label{sec:ai}

Assumptions: $r_{\text{tx}}=\$0.01$ (inference calls, tool invocations, and agent-to-agent settlement all bear real value); $C_{\text{eff}}=\$4{,}992$; ${\sim}\num{10000}$ tx per agent per month at machine pace; dense packing.

\textbf{Break-even: ${\sim}\num{500000}$ tx/month ${=}$ ${\sim}1.8\%$ of a shard.} An AI shard pays for itself at two percent utilisation.

\begin{table}[h]
\centering
\caption*{\textbf{Table AI.} Agent population ramp at $10\times$/year (unbounded).}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
Year & Agents & Monthly tx & Shards & Cost & Revenue & Margin \\
\midrule
1 & \num{100}     & 1.0M  & 1   & \$4{,}992    & \$10{,}000  & 2.0$\times$ \\
2 & \num{1000}    & 10M   & 1   & \$4{,}992    & \$100{,}000 & 20$\times$ \\
3 & \num{10000}   & 100M  & 4   & \$19{,}968   & \$1.0M      & 50$\times$ \\
4 & \num{100000}  & 1.0B  & 36  & \$179{,}712  & \$10M       & 56$\times$ \\
5 & \num{1000000} & 10B   & 358 & \$1.79M      & \$100M      & 56$\times$ \\
\bottomrule
\end{tabular}
\end{table}

The defining trait is the absence of a TAM ceiling. Machine-generated volume is unbounded and sticky; margin saturates near $56\times$ and then revenue simply scales. AI both load-tests the dispatch/\textsc{comm} hot path under real volume \emph{and} is the only sector whose growth does not terminate---which is why it should anchor the rollout rather than a low-fee consumer sector.

\section{Fintech (Boring Financial) --- the density engine}
\label{sec:fintech}

Assumptions: $r_{\text{tx}}=\$0.25$ blended (settlement, contract-lifecycle, and loan-servicing events); $C_{\text{eff}}=\$10{,}400$ (custody, audit, regulatory load); safety headroom.

\textbf{Break-even: ${\sim}\num{42000}$ tx/month ${=}$ ${\sim}0.2\%$ of a shard}---the most revenue-dense sector; a fintech shard pays for itself at one-fifth of one percent utilisation.

\begin{table}[h]
\centering
\caption*{\textbf{Table FIN.} Transaction ramp.}
\begin{tabular}{@{}rrrrrr@{}}
\toprule
Year & Monthly tx & Shards & Cost & Revenue & Margin \\
\midrule
1 & 50k   & 1  & \$10{,}400  & \$12{,}500  & 1.2$\times$ \\
2 & 500k  & 1  & \$10{,}400  & \$125{,}000 & 12$\times$ \\
3 & 5M    & 1  & \$10{,}400  & \$1.25M     & 120$\times$ \\
4 & 50M   & 3  & \$31{,}200  & \$12.5M     & 400$\times$ \\
5 & 500M  & 25 & \$260{,}000 & \$125M      & 481$\times$ \\
\bottomrule
\end{tabular}
\end{table}

\begin{remark}[The most assumption-sensitive number in the note.]
$\$0.25$/tx is right for high-value settlement events, where it is trivial against notional, and wrong by one to two orders of magnitude for high-frequency micropayments, where fintech behaves like AI on margins. The real schedule is basis-points-on-notional for large events and sub-cent for small ones, so fintech is bimodal; model the two instrument classes separately once Boring supplies their mix.
\end{remark}

\section{Medical (VA) --- the cohort-capped annuity}
\label{sec:medical}

The Veterans Health Administration serves roughly \num{9100000} enrolled veterans~\cite{vha2026}; the $80\%$ target is ${\sim}\num{7280000}$ patients. Assumptions: $\tau=50$ provenance-logged record accesses per patient per month; $R=\$1$/patient/month in platform fees (a conservative $2$--$7\%$ of full-EHR per-patient spend); $C_{\text{eff}}=\$16{,}640$; headroom.

\textbf{Break-even: ${\sim}\num{16640}$ patients against a ${\sim}\num{400000}$-patient shard ${=}$ ${\sim}4\%$ load.} A medical shard pays for itself at four percent utilisation.

\begin{table}[h]
\centering
\caption*{\textbf{Table MED.} $10\times$/year rollout against a finite cohort (saturates at year~4).}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
Year & Patients & Monthly tx & Shards & Cost & Revenue & Margin \\
\midrule
1 & \num{7280}     & 0.36M & 1  & \$16{,}640  & \$7{,}280     & 0.4$\times$ (loss) \\
2 & \num{72800}    & 3.6M  & 1  & \$16{,}640  & \$72{,}800    & 4.4$\times$ \\
3 & \num{728000}   & 36M   & 2  & \$33{,}280  & \$728{,}000   & 21.9$\times$ \\
4 & \num{7280000}  & 364M  & 19 & \$316{,}160 & \$7{,}280{,}000 & 23.0$\times$ \\
\rowcolor{coolrow}5--10 & \num{7280000} & 364M & 19 & \$316{,}160 & \$7{,}280{,}000 & 23.0$\times$ (flat) \\
\bottomrule
\end{tabular}
\end{table}

Two facts dominate. First, the cohort is a hard wall: a literal $10\times$/year ramp saturates the finite population by year~4, so medical revenue is \emph{logistic}---explosive then flat---and past saturation it is an annuity (${\sim}\$87$M/year revenue on ${\sim}\$3.8$M/year infrastructure), not a growth story. Cumulative decade revenue is ${\sim}\$620$M, of which infrastructure is a ${\sim}4\%$ rounding error. Second, the constraint is never economic: even tripling cost for high-availability replication ($3\times\!\times4\times$ loading) leaves ${\sim}7$--$8\times$ margin at cohort. The binding constraints are regulatory and operational.

\begin{remark}[Risk sets the rent posture.]
``\textsc{comm}s delivered while funded'' (the rent note's mortality semantics) is a safe default for music but a safety failure for a patient-record access that silently never fires on fund depletion. Mission-critical residuals must be pinned to committed on-chain balance (germ line), never allowed to deplete to eviction. This is the strongest reason medical deploys last: the rent model itself is most stressed exactly where the data is least forgiving.
\end{remark}

\section{Music --- the cold-tail fill}
\label{sec:music}

A stable of ten independent artists at ${\sim}1$M non-overlapping followers, each audient streaming a conservative ten songs/month, is $100$M plays/month. Per-play on-chain, that is ${\sim}\num{4}$ shards ($\$16{,}640$/mo), and break-even is
\[
  r_{\text{tx}}^\star \;=\; \frac{\$16{,}640}{100\text{M}} \;\approx\; \$0.000166\ \text{per play}.
\]
Any network fee above ${\sim}1/60$ of a cent per play is profitable; at $\$0.0003$/play the stable runs ${\sim}1.8\times$.

\begin{table}[h]
\centering
\caption*{\textbf{Table MUS.} Margin on the ten-artist stable (100M plays/mo, 4 shards, \$16{,}640/mo).}
\begin{tabular}{@{}rrrr@{}}
\toprule
Network fee/play & Monthly revenue & Cost & Margin \\
\midrule
\$0.0001  & \$10{,}000  & \$16{,}640 & 0.6$\times$ (underwater) \\
\$0.000166 & \$16{,}640 & \$16{,}640 & break-even \\
\$0.0003  & \$30{,}000  & \$16{,}640 & 1.8$\times$ \\
\$0.001   & \$100{,}000 & \$16{,}640 & 6.0$\times$ \\
\bottomrule
\end{tabular}
\end{table}

Two design facts. Batching plays off-chain and settling (say $100{:}1$) collapses $100$M plays to $1$M transactions on one shard at ${\sim}7\times$ margin---strictly better economics \emph{unless per-play provenance is the product}. If transparent per-play royalty and capability-attenuated rights are the differentiator over incumbents, we \emph{want} the $100$M on-chain transactions, and music becomes the abundant \emph{cold-tail packing} that monetises the spare leaves of the AI and fintech shards (rent note, bid-rent stratification). Either way music is viable; the choice is provenance granularity versus margin. The caveat is concentration: $100$M/month resting on ten artists is roster-dependent and churns, which is why music is high-margin fill, never the anchor.

\section{Web 2.0 --- participatory advertising}
\label{sec:web2}

Free-at-point-of-use consumer properties have no per-transaction user fee, so the framework above appears to fail---until advertising is made \emph{participatory}. End users pay in tokenised impressions per data payload until their usage is paid off, then either switch to impression-free payloads or accumulate advertising tokens spendable in other markets. This both closes the monetisation gap and inverts the extraction relationship.

The mechanism rests on one ratio. A data payload costs $\$0.00014$ at dense packing; an ordinary display impression at a ${\sim}\$3$ CPM~\cite{gdn2026} is worth $\$0.003$:
\[
  \boxed{\ \text{one ad impression covers the network cost of } {\sim}20 \text{ data payloads.}\ }
\]
Because impressions are tokenised on-chain, each impression is itself a fee-bearing transaction that funds the payload it rides plus margin: the ad layer is self-funding volume, not overhead. When cumulative impressions exceed cumulative usage, the user crosses into the impression-free or net-earner fork.

\begin{table}[h]
\centering
\caption*{\textbf{Table WEB.} Three Google properties; payloads as emails/sessions/edits.
Volumes from Gmail~\cite{gmail2026}, Maps~\cite{maps2026}, Workspace~\cite{workspace2026}.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Property & Payloads/mo & Shards & Infra cost/mo & Break-even ads/user/day \\
\midrule
Gmail (\num{1800000000} users) & ${\sim}3.6$T & ${\sim}\num{120000}$ & ${\sim}\$500$M & ${\sim}3$ \\
Maps (\num{2000000000} users)  & ${\sim}3.0$T & ${\sim}\num{100000}$ & ${\sim}\$416$M & ${\sim}2$--$3$ \\
Docs (\num{3000000000} users)  & ${\sim}600$B & ${\sim}\num{20000}$  & ${\sim}\$83$M  & ${\sim}0.3$ \\
\bottomrule
\end{tabular}
\end{table}

Every property is funded if a user views two or three ads per day---far below real exposure---so the typical user crosses into accumulating spendable tokens. \textbf{That is the inversion}: in incumbent Web 2.0 the user is the product, their attention extracted and the surplus captured by the platform; here the user covers their own cost with attention and \emph{keeps the surplus} as portable tokens. The network takes a thin per-payload margin at astronomical volume---a $\$0.00002$/payload take across Gmail's $3.6$T payloads is still ${\sim}\$72$M/month. This is ``the community is the moat'' and the anti-enshittification thesis rendered as an income statement.

\begin{observation}[The cross-market spend saves Docs.]
Docs has the lowest break-even (${\sim}0.3$ ads/day) but the highest ad resistance---nobody wants ads in documents. The fork resolves it: users earn tokens on high-CPM, ad-tolerant surfaces (Maps' local and sponsored context) and spend them on impression-free Docs. The cross-subsidy is carried in the \emph{user's} portable tokens, structurally unlike incumbent bundling, where the cross-subsidy is the platform's lever over the user.
\end{observation}

\paragraph{Enter through a privacy wedge, not at Gmail scale.} $\num{120000}$ shards and $\$500$M/month is a hyperscaler buildout; per-payload viability is trivial but absolute capital is enormous. The wedge is a privacy-premium slice: $1\%$ of Gmail (${\sim}\num{18000000}$ users who actively want data ownership) is ${\sim}36$B payloads/month ${\approx}\num{1200}$ shards ${\approx}\$5$M/month infrastructure, trivially funded---and ``own your data, earn from your attention'' self-selects exactly those adopters. Web 2.0 is a beachhead-then-scale play entered once the participatory rails are proven on the earlier sectors.

\section{Acquisition and soft costs}
\label{sec:cac}

Folklore puts incumbent direct-to-consumer acquisition at ${\sim}\$50$/head (Facebook) and ${\sim}\$19$/head (WhatsApp). Those are \emph{retail} costs: paid per individual, one head at a time. F1R3FLY's model is \emph{wholesale}---court one supply-side anchor and acquire its entire downstream population in a single deal:
\[
  \mathrm{CAC}_{\text{eff}} \;=\; \frac{A}{n\,c},
\]
with $A$ the cost to acquire one anchor (business development, onboarding, integration, contract setup), $n$ the population that anchor brings, and $c$ the fraction that converts. The denominator is what collapses the figure. A second difference matters more than the first: \textbf{the anchor is a paying customer, not a liability to be monetised later.} Because the artist's sale embeds the transaction fee and the provider pays for the platform, acquisition is recovered from the anchor's own fees---CAC is \emph{self-liquidating}---and there is no consumer billing relationship with millions of end users, so collections, billing, and consumer-churn management are wholesale too.

\begin{table}[h]
\centering
\caption*{\textbf{Table CAC-1.} Music \u2014 court the artist, acquire the audience.}
\begin{tabular}{@{}lrrrrr@{}}
\toprule
Scenario & $A$ / artist & $c$ & Audients/artist & CAC/head & vs.\ WhatsApp \$19 \\
\midrule
Inbound / high-convert & \$2{,}000  & 10\% & \num{100000} & \$0.02 & ${\sim}950\times$ cheaper \\
Mid                    & \$5{,}000  & 5\%  & \num{50000}  & \$0.10 & ${\sim}190\times$ cheaper \\
Conservative           & \$10{,}000 & 1\%  & \num{10000}  & \$1.00 & ${\sim}19\times$ cheaper \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption*{\textbf{Table CAC-2.} Healthcare \u2014 court the provider, own the patient.}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Anchor & $A$ & Patients & CAC/head & vs.\ Facebook \$50 \\
\midrule
Mid hospital system          & \$500{,}000    & \num{1000000}  & \$0.50 & ${\sim}100\times$ cheaper \\
VA (conservative)            & \$5{,}000{,}000  & \num{7280000}  & \$0.69 & ${\sim}72\times$ cheaper \\
VA (heavy gov.\ sale)        & \$20{,}000{,}000 & \num{7280000}  & \$2.75 & ${\sim}18\times$ cheaper \\
\bottomrule
\end{tabular}
\end{table}

Patient data ownership---the private key to a data wallet---does double duty: it is a close-rate multiplier on the provider sale (a consent, portability, and compliance differentiator that lowers effective $A$) and a downstream acquisition channel (a patient who owns their data carries it to other F1R3FLY services, seeding the rest of the portfolio at zero marginal CAC). The cross-market multiplier compounds this: with token portability, a head acquired in one market is a user in $m$ markets, so blended CAC falls to $\mathrm{CAC}/m$ while LTV multiplies by $m$ against that single acquisition. Music LTV is thin on streams alone (${\sim}\$0.003$/head-month, a multi-year payback), but the same head transacting across artist direct-sales and participatory-ad surfaces pays back in weeks; healthcare needs no such rescue (${\sim}\$1$/patient-month over a multi-year relationship gives LTV/CAC ${\sim}20$--$90\times$).

\begin{remark}[The counter-forces are symmetric.]
Anchor acquisition is lumpy and front-loaded---the VA's \$5--20M is spent years before the cohort ramps, a real cash-flow-timing and sales-cycle problem. Conversion $c$ is genuinely uncertain and is the whole ballgame for music. And concentration cuts on the acquisition side too: lose an anchor and lose its entire population at once. Wholesale acquisition and wholesale churn are the same mechanism.
\end{remark}

\section{The connector shard: monetising interoperability}
\label{sec:connector}

Per-sector shards earn on work done \emph{inside} them; the \emph{connector} shard earns on value moving \emph{between} them. A provider checks a box at deploy time to join, and F1R3FLY collects a fee on every cross-shard transaction. This is the interchange position (Visa/Mastercard, SWIFT, an internet exchange point), and three properties make it the profit centre.

First, the tax base grows superlinearly. If a shard transacts with counterparties roughly in proportion to who is reachable, the cross-shard fraction $\chi$ of its traffic rises toward $(n-1)/n$ as $n$ shards connect:

\begin{table}[h]
\centering
\caption*{\textbf{Table CON-1.} Cross-shard fraction under uniform mixing.}
\begin{tabular}{@{}rrrrrr@{}}
\toprule
Connected shards $n$ & 2 & 5 & 10 & 50 & 100 \\
$\chi \to (n-1)/n$ & 50\% & 80\% & 90\% & 98\% & 99\% \\
\bottomrule
\end{tabular}
\end{table}

Early on almost everything is intra-shard and the connector earns little; at scale nearly every transaction crosses a boundary and the connector taxes essentially the entire platform volume. Second, its cost stays tiny---it routes and clears coordination, it does not store payloads---so its margin is the highest in the system. Third, the opt-in box is a flywheel: each provider that joins makes the connector more valuable to every other member, raising the value of joining, and it captures \emph{competitors} too, since any rival that wants interoperability connects and pays the toll. One need not win every sector; one needs every sector to connect.

\begin{table}[h]
\centering
\caption*{\textbf{Table CON-2.} Connector revenue (illustrative; $r_x$ and $\chi$ are levers).}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Scenario & Connectable vol./mo & Effective $\chi$ & Fee $r_x$ & Connector rev./mo \\
\midrule
Core sectors, early        & ${\sim}11$B   & 20\% & \$0.005  & ${\sim}\$11$M \\
Core sectors, dense network & ${\sim}11$B   & 60\% & \$0.005  & ${\sim}\$33$M \\
+ Web 2.0 connected        & ${\sim}7.2$T payloads & 5\% & \$0.0005 & ${\sim}\$180$M \\
\bottomrule
\end{tabular}
\end{table}

Even the conservative core case rivals the largest single-sector line; once the mass-market substrate connects, the connector dwarfs everything at a fraction of the per-transaction fee. \textbf{Travel and hospitality} is the archetypal amplifier: a single trip is inherently multi-shard---payment, loyalty/identity, location, booking, reviews---so it carries a high native $\chi$ and lights up the connector far beyond its standalone volume. The correct valuation of any new sector is therefore \emph{intra-shard fees plus its contribution to connector $\chi$}.

This closes the loop with acquisition. The cheap wholesale CAC works \emph{because} a head acquired in one market is a user in all of them, and the mechanism by which it is a user in all of them is cross-shard transactions---which are the connector's revenue. Cross-market portability is simultaneously the CAC discount and the connector's tax base: the thing that makes users cheap to acquire is the thing that makes them lucrative.

\begin{remark}[Counter-forces.]
The connector is back-loaded like medical: with few shards connected $\chi$ is small and revenue is negligible for years, so the sectors must bootstrap volume first. Providers can bypass the toll via private peering (the peering-versus-transit dynamic of internet economics), which caps $r_x$; because the box is opt-in, demand is elastic and revenue is maximised at an interior toll, not the highest one. A levy on all cross-sector flow is interchange, among the most heavily regulated prices in finance, so dominance carries a compliance tax. And concentration recurs: the connector's value rests on the set of connected anchors.
\end{remark}

\section{Bandwidth: the media exception}
\label{sec:bandwidth}

The cost model set egress to zero on the assumption that payloads fit inside OCI's free tier. That is correct for KB-scale transactional payloads but wrong for media. OCI prices internet egress at \$0.0085/GB beyond the first \SI{10}{\tera\byte}/month and inter-region transfer at \$0.005/GB~\cite{oci2026} (Oracle reportedly moved to free egress across all regions in early 2026; we model the rate as a lever rather than baking in zero). Bandwidth cost is $\text{payload size}\times\text{moves}\times\text{rate}$, and per-payload it dwarfs the \$0.00014 compute/storage cost exactly when payloads are large:

\begin{table}[h]
\centering
\caption*{\textbf{Table BW.} Per-payload egress versus the transactional cost floor.}
\begin{tabular}{@{}llrrl@{}}
\toprule
Payload & Size & Egress @ \$0.0085/GB & @ CDN \$0.04/GB & Verdict \\
\midrule
Fintech tx / AI token / EHR record & ${\sim}1$--$5$ KB & \$0.00000004 & --- & negligible \\
Audio, compressed                  & ${\sim}4$ MB   & \$0.000034   & \$0.00016 & absorbable \\
Audio, lossless / hi-res           & ${\sim}40$ MB  & \$0.00034    & \$0.0016  & exceeds fee \\
Radiology study (CT/MRI)           & ${\sim}250$ MB & \$0.002125   & \$0.010   & dominant cost \\
\bottomrule
\end{tabular}
\end{table}

The verdict by sector follows the payload size. Bandwidth is \textbf{negligible} for the transactional core (AI, fintech, EHR metadata---the zero assumption holds). It is \textbf{material but absorbed} for medical imaging: at ${\sim}1.25$ GB/patient-month the cohort moves ${\sim}9$ PB/month, costing ${\sim}\$77$k/month (${\sim}\$364$k at CDN rates)---only ${\sim}1$--$5\%$ of medical's high per-patient revenue. It is \textbf{viability-threatening} only where a large payload meets a thin fee: lossless audio streaming costs ${\sim}\$0.00034$/play in egress against a ${\sim}\$0.0003$ streaming fee, so hi-res music is bandwidth-underwater and needs either a higher fee (${\sim}\$0.0005$/play) or edge caching, whereas compressed audio (${\sim}11\%$ of the fee) is merely a margin haircut.

\begin{remark}[Move capabilities, not payloads.]
The OCap/data-wallet architecture mitigates this structurally. Cross-shard and through the connector, the system transmits an access token (KB), not the blob (GB); the large object lives once, content-addressed, in the patient's data wallet or an edge cache near where it is consumed, and bandwidth is incurred only at the final serve hop. Three consequences: the connector economics are unaffected (cross-shard traffic is tokens, so the profit centre carries no bandwidth); bandwidth becomes a per-serve, edge-localised cost that should be metered in phlogiston as a third resource---a per-byte egress pass-through alongside rent (standby) and activation (\textsc{comm})---so that whoever moves the blob pays for the bytes; and edge caching of hot blobs (radiology priors, popular tracks) amortises egress over serves, which is what rescues lossless audio. On OCI specifically, cross-data-centre movement is cheap (inter-region \$0.005/GB, and FastConnect to on-premises charges a flat port fee with no per-GB cost), so the expensive hop is not between data centres but the last mile to the consumer.
\end{remark}

\section{Synthesis and rollout}
\label{sec:synthesis}

\begin{table}[h]
\centering
\caption*{\textbf{Table SUM.} The five sectors plus Web 2.0.}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}llllll@{}}
\toprule
\rowcolor{headrow}
Sector & Revenue density & Risk & Boundedness & Break-even load & Role \\
\midrule
Web 2.0  & ad-funded (\$0.00014/pl) & low--mod & astronomical & ${\sim}3$ ads/user/day & mass-market moat (last) \\
Music    & \$0.0003/tx              & low      & unbounded, churny & ${\sim}$near full & cold-tail fill \\
AI       & \$0.01/tx                & low--mod & unbounded, sticky & ${\sim}2\%$ & \textbf{anchor / growth} \\
Fintech  & ${\sim}\$0.25$/tx        & high     & semi-bounded & ${\sim}0.2\%$ & margin engine \\
Medical  & high/patient             & highest  & hard cohort cap & ${\sim}4\%$ & cash-cow annuity \\
\bottomrule
\end{tabular}
\end{table}

The pattern that falls out: \textbf{revenue density and risk rise together, and both run opposite to volume.} The original safe-to-critical \emph{risk} ordering is correct and unchanged. Layered on top, the \emph{economic} ordering is:

\begin{enumerate}[leftmargin=1.6em,itemsep=2pt]
  \item \textbf{AI --- anchor.} Simultaneous load-testing and viability; hardens the hot path; unbounded growth.
  \item \textbf{Fintech --- margin.} Highest revenue density rewards even modest volume; builds high-assurance discipline on financially recoverable failures.
  \item \textbf{Medical --- annuity.} Highest margin per unit, capped at the cohort, deployed last so the assurance and germ-line-pinning machinery is battle-tested first.
  \item \textbf{Music --- continuous fill.} Not a phase but the cold-tail packing that monetises spare capacity across the above.
  \item \textbf{Web 2.0 --- mass-market substrate.} Entered through a privacy wedge once the participatory ad rails are proven.
\end{enumerate}

\section{Sensitivity and open levers}
\label{sec:sensitivity}

Every figure scales transparently. The fees $r_{\text{tx}}$ are the dominant levers and should be replaced with real schedules from ASI and Boring as soon as available; the fintech number in particular is bimodal and warrants two instrument classes. Acquisition adds three more levers ($A$, conversion $c$, population per anchor) and the connector adds two ($r_x$, effective $\chi$); the connector's revenue is the most sensitive of all to $\chi$, which is itself a function of how many shards connect. The per-payload cost falls linearly with packing density, so the entire Web 2.0 case strengthens with the elastic-hashing gains the rent note quantifies---for low-fee consumer volume, dense packing is the difference between profit and loss, not a refinement. The compliance multipliers are coarse stand-ins for replication, audit, and dedicated-infrastructure costs and should be replaced with per-sector infrastructure models before any commitment. Finally, the single largest non-modelled risk is customer concentration---ten artists in music, one federal buyer in medical---which the portfolio sequencing above is in part designed to diversify.

\begin{thebibliography}{9}

\bibitem{rentnote}
F1R3FLY Industries.
\newblock Rent and shard splitting: a storage-economics companion to occupancy-driven sharding.
\newblock Working note, 2026.

\bibitem{fkk2025}
M.~Farach-Colton, A.~Krapivin, and W.~Kuszmaul.
\newblock Optimal bounds for open addressing without reordering.
\newblock arXiv:2501.02305, 2025.

\bibitem{rho2005}
L.~G. Meredith and M.~Radestock.
\newblock A reflective higher-order calculus.
\newblock \emph{Electronic Notes in Theoretical Computer Science}, 141(5):49--67, 2005.

\bibitem{oci2026}
Oracle Corporation.
\newblock Oracle Cloud Infrastructure pricing (compute, block volume, and networking price lists).
\newblock \url{https://www.oracle.com/cloud/price-list/}, accessed June 2026.

\bibitem{vha2026}
U.S. Department of Veterans Affairs, Veterans Health Administration.
\newblock About VHA: enrolled-veteran population and budget.
\newblock \url{https://www.va.gov/health/aboutvha.asp}, accessed June 2026.

\bibitem{gmail2026}
Gmail usage statistics: active users and daily message volume, 2026.
\newblock Aggregated industry reporting (Statista, Litmus, Radicati), accessed June 2026.

\bibitem{maps2026}
Google Maps usage statistics: monthly active users, 2026.
\newblock Aggregated industry reporting, accessed June 2026.

\bibitem{workspace2026}
Google Workspace active-user disclosures, 2026.
\newblock Google I/O and aggregated industry reporting, accessed June 2026.

\bibitem{gdn2026}
Display advertising benchmarks 2026: Google Display Network average CPM ${\approx}\$3.12$.
\newblock Aggregated programmatic-benchmark reporting, accessed June 2026.

\end{thebibliography}

\end{document}
