%% rset.nwf.tex
%% The reflective construction as a bridge between well-founded set theory
%% (ZF, FM) and Aczel's non-well-founded set theory.
%% Drop in via \input{rset.nwf}.
%%
%% NB: this section cites Aczel, Barwise--Moss, Joyal--Moerdijk and Awodey.
%% The corresponding entries (aczel1988nwf, barwise1996vicious,
%% joyal1995algebraic, awodey2008ast) are in the updated rset.bib.
\input{rset.colors}   % red/black colour layer + FM symbols (providecommand-guarded)

\section{A bridge between well-founded and non-well-founded set theory}
\label{sec:nwf}

$\mathsf{ZF}$ and $\mathsf{FM}$ are well-founded theories: both take the axiom of
foundation, so membership admits no infinite descent and no cycles. At the
opposite pole stands Aczel's anti-foundation axiom ($\mathsf{AFA}$), which
replaces foundation outright, admits sets such as $\Omega=\{\Omega\}$, and gives
every accessible pointed graph a unique decoration as a set
\cite{aczel1988nwf,barwise1996vicious}. The reflective construction sits between
these poles, and
clarifies their relation, because here non-well-foundedness is not \emph{posited}
but \emph{produced}: it is what two well-founded theories generate when they are
intertwined.

\subsection{Each colour is locally well-founded}
\label{sec:nwf:local}

Restricted to a single colour, the construction is as well-founded as
$\mathsf{ZF}$. The red membership $\rin$ relates red sets only to their red
members, bottoming out at $\remp$ and at the red atoms; and every red set has a
rank over those atoms (finite in the core of
Section~\ref{sec:fmmodel:finitary}, an ordinal in the infinitary extension of
Section~\ref{sec:fmmodel:infinitary}). So $\rin$ is well-founded, and foundation
holds for the red theory --- as a theorem, never assumed
(Theorem~\ref{thm:rep}). The black theory is well-founded for the identical
reason. Crucially, an atom is \emph{opaque} to the membership that has it as an
atom: $\rin$ cannot descend into a black atom, and $\bin$ cannot descend into a
red one. Each colour's well-foundedness rests exactly on this opacity.

\subsection{The pair is globally non-well-founded}
\label{sec:nwf:global}

Now erase the colour barrier. Let $\cbin$ be the \emph{colour-blind} membership
that refuses to treat atoms as opaque: where a red set has a black atom
$a=\bk{b}$, the relation $\cbin$ exposes the black members of $b$, and
symmetrically. Under $\cbin$ a red set may contain a black set, which contains a
red set, which contains a black set, without end:
\[
  S_0 \;\ni_\star\; \bk{b_0} \;\ni_\star\; S_1 \;\ni_\star\; \bk{b_1}
  \;\ni_\star\; S_2 \;\ni_\star\; \cdots,
\]
an infinite descending $\cbin$-chain that alternates colours. Globally,
foundation fails. And the failure is not marginal: the two universes are
themselves \emph{defined} by the mutually recursive equations
\[
  \rX \;=\; \bSetX, \qquad \bX \;=\; \rSetX,
\]
a system whose solution is a fixed point. Read as a specification of objects by
their unfoldings, this is precisely the kind of non-well-founded equation that
Aczel's solution lemma is designed to solve \cite{aczel1988nwf}; allowing a red
name whose black unfolding refers back to it realises genuine $\Omega$-like
reflexive structure under $\cbin$.

\subsection{Non-well-foundedness as an interaction effect}
\label{sec:nwf:interaction}

This is the conceptual dividend. In Aczel's theory non-well-foundedness is an
\emph{axiom}: $\mathsf{AFA}$ is asserted in place of foundation, of a single
universe. In the reflective construction it is an \emph{emergent global
property}: well-foundedness is a feature of each colour \emph{in isolation},
non-well-foundedness a feature of the colours' \emph{interaction}. The colour
barrier is the exact seam between the two --- intra-colour $\in$ is well-founded,
inter-colour $\cbin$ is not --- so one can read off where descent terminates and
where it loops simply by tracking colour changes.

The coalgebraic gloss makes the spectrum sharp. A mono-chromatic theory is an
\emph{initial algebra} of its set-builder operator: the least fixed point, well
founded, the home of $\mathsf{ZF}$ and $\mathsf{FM}$, and the object of study of
\emph{algebraic set theory}, where the well-founded universe appears as the free
algebra for a set-builder operator on a category of classes
\cite{joyal1995algebraic,awodey2008ast}. The intertwined system
$(\rd{\mathrm{Set}}[-],\,\bk{\mathrm{Set}}[-])$, taken instead as a \emph{greatest}
fixed point, is a \emph{final coalgebra}: non-well-founded, the home of
$\mathsf{AFA}$ \cite{aczel1988nwf,barwise1996vicious}. A caveat of size belongs
here, and it falls exactly along the finitary/infinitary seam of
Sections~\ref{sec:fmmodel:finitary}--\ref{sec:fmmodel:infinitary}. The
\emph{finite} powerset functor --- the operator behind the finitary core --- is
finitary, so its final coalgebra exists already at the level of sets, and the
non-well-founded reading of the $\mathrm{HF}$ theory is literal. The
\emph{full} powerset operator has no final coalgebra at all, by Cantor's theorem
together with Lambek's lemma; the unrestricted, infinitary theory therefore lives
not in a set-sized final coalgebra but in Aczel's class-sized framework of
accessible pointed graphs and the solution lemma \cite{aczel1988nwf}. The
reflective construction places both fixed points of one system of operators within
a single framework, and the knobs of this paper --- least versus greatest fixed
point, finitary versus infinitary embrace (Section~\ref{sec:fmmodel:infinitary}),
acyclic versus cyclic colour discipline (Section~\ref{sec:future}) --- are
exactly the controls that slide a theory along the spectrum from $\mathsf{ZF}$
and $\mathsf{FM}$ at the well-founded end to Aczel's universe at the
non-well-founded end. Making that correspondence precise --- decorating every
accessible pointed graph by a colour-blind set picture, and exhibiting the
finitary intertwined universe as a genuine final coalgebra --- is a natural next
step (Sections~\ref{sec:rigor} and~\ref{sec:future}). What the present
construction already shows is the structural moral:
\emph{non-well-foundedness need not be assumed; it can be manufactured, as a
global property of locally well-founded theories.}
