\documentclass[12pt]{llncs}

\usepackage[pdftex]{hyperref}                   
\usepackage {mathpartir}
\usepackage{bigpage}
\usepackage{bcprules}
\usepackage {listings}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amstext}
\usepackage{latexsym}
\usepackage{longtable}
\usepackage{stmaryrd}
%\usepackage{fdsymbol}
\usepackage{graphicx} 
%\usepackage[margins=2.5cm,nohead,nofoot]{geometry}
%\usepackage{geometry}
\usepackage{color}
\usepackage{xcolor}


%\include{myPreamble}
\include{rset.local} 

%\ifpdf
%\usepackage[pdftex]{graphicx}
%\else
%\usepackage{graphicx}
%\fi

 % \ifpdf
%  \usepackage{pdfsync}
%  \if


%\title{Brief Article}
%\author{David F. Snyder}
%\author{L.G. Meredith}

%\address{Dept. of Math., Texas State University--San Marcos, San Marcos, TX 78666}
       
\pagestyle{plain}   % page numbers for the draft; revert to \pagestyle{empty} for LNCS camera-ready


\begin{document}

\lstset{language=[Objective]Caml,frame=shadowbox}


\input{rset.front}

In the early 2000's Pitts and Gabbay's use of Fraenkl-Mostowski set
theory ($\mathsf{FM}$ set theory, or just $\mathsf{FM}$ for short), a
theory of sets with atoms, to model nominal phenomena, sparked a
vibrant line of research. \cite{DBLP:journals/fac/GabbayP02}
\cite{DBLP:journals/jcss/Clouston14}. As Chaitin points out, the
axioms and assumptions of a theory comprise its risk
\cite{chaitin1999unknowable}. This observation raises an interesting
point: is it possible to have a set theory with atoms without taking
on \emph{infinite} risk? It turns out it is.

We describe a theory of sets that enjoys both the cleanliness of
Zermelo-Fraenkl set theory ($\mathsf{ZF}$ set theory, or just
$\mathsf{ZF}$ for short), that is arising only from the empty set, and
yet with an infinite supply of atoms.

The shape of the solution is easy to state in advance: we take a
familiar theory, make two copies of it, and tie them in a recursive
knot, each copy serving as the supply of atoms for the other. When the
author first arrived at this idea in 2006, copying a theory and binding
the copies by mutual recursion was an unusual move, justified mostly by
where it led. Two decades on it has become ordinary. In the age of AI
and the autoformalization of mathematics and logic, copies of a theory
are everywhere: autoformalization is exactly the business of producing a
formal copy of an informal theory and reasoning about the fit between
the two, proof assistants host reflective towers in which a theory
manipulates a copy of itself, and a single development now routinely
keeps several presentations of the same theory in play at once. Tying
such copies into a recursive knot --- a theory whose objects are those
of its mirror image, and conversely --- is no less natural: it is
reflection, the self-referential idiom that runs from the reflection
principles of logic through reflective calculi
\cite{DBLP:journals/entcs/MeredithR05} to present-day systems that reason
about their own output. So although the theory below was devised in
2006, it arguably reads more naturally in 2026 --- the moves from which
it is built have become part of the ordinary vocabulary of formalized
mathematics.

\input{rset.intuitions}

\input{rset.presentation}

\input{rset.stdfmset}

\input{rset.mainthm}

\input{rset.applications}

\input{rset.nwf}

\input{rset.rigor}

\input{rset.future}

\input{rset.conc}

\input{rset.ack}

\bibliographystyle{plain}   
\bibliography{rset.bib}

\end{document}
