%% rset.mainthm.tex
%% A model of FM set theory inside the reflective theory of Sections 2--3,
%% with the two copies colour-coded (red = player, black = opponent).
%% Drop in via \input{rset.mainthm} (the hook already exists in rset.tex).
\input{rset.colors}   % red/black colour layer + FM symbols (providecommand-guarded)

\section{A model of $\mathsf{FM}$ in the reflective universe}
\label{sec:fmmodel}

We can now collect the dividend of the construction \emph{Tying the second
recursive knot} above. The reflective theory was assembled from the empty set
alone, yet --- we claim --- it already contains a model of $\mathsf{FM}$, and in
fact two of them, interlocked. The atoms such a model needs are not postulated;
they are \emph{supplied by the dual copy}. Throughout we fix the two tied
theories introduced above, the red copy $\rSetX$ and the black copy $\bSetX$,
with
\[
  \rX \;=\; \bSetX, \qquad \bX \;=\; \rSetX,
\]
so that the admissible atoms of the red theory are exactly the black sets, and
\emph{vice versa}. We write the red copy in {\rd red} and the black copy in
{\bk black} throughout, so that a red comprehension $\rbr{\cdots}$ and a black
comprehension $\bbr{\cdots}$, or the red and black membership relations
$\rin$ and $\bin$, are told apart at a glance.

\subsection{Atoms out of the empty set}
\label{sec:fmmodel:atoms}

Inside the black copy, build the von Neumann numerals from the black empty set,
\[
  \bk{0} \;=\; \bemp,
  \qquad
  \bk{n+1} \;=\; \bk{n} \bcup \bbr{\bk{n}},
\]
each a black, hereditarily finite set obtained, ultimately, from $\bemp$. By
black extensionality (the equations of the formal theory above) these are
pairwise distinct: $\bk{n} \neq \bk{m}$ whenever $n\neq m$. Put
\[
  \Atoms \;:=\; \{\, \bk{n} \;:\; n\in\omega \,\}\ \subseteq\ \bSetX .
\]

Two observations make $\Atoms$ behave \emph{as a set of atoms} for the red
theory.
\begin{itemize}
\item \textbf{Admissibility.} By the second recursive knot $\rX=\bSetX$, every
  black set is an admissible red atom; in particular each $\bk{n}$ is one.
\item \textbf{Opacity.} The red membership relation $\rin$ cannot see into a
  black set (the \emph{Intuitions} above). Hence, \emph{as far as the red theory
  can tell}, an element $\bk{n}$ of $\Atoms$ has no internal structure
  whatsoever: the only facts the red theory may assert about $\bk{n}$ and
  $\bk{m}$ are $\bk{n} = \bk{m}$ or $\bk{n} \neq \bk{m}$, and those are settled,
  out of sight, by black equality.
\end{itemize}
Thus $\Atoms$ presents to the red theory precisely as $\mathsf{FM}$'s set of
atoms does: a countable family of mutually distinct, structureless objects that
may sit inside sets but admit no membership query of their own.

\begin{remark}[Where the risk went]
\label{rem:risk}
This is the sense, promised in the introduction, in which atoms cost no infinite
risk. $\mathsf{ZFA}$ and $\mathsf{FM}$ \emph{postulate} an infinite set of
primitive atoms; the postulate is the risk. Here nothing is postulated beyond
$\bemp$. The atoms are black hereditarily finite sets, and the stream $S[X]$ of
the formal theory hands them out lazily, $\mathrm{take}(\Sigma,n)$ at a time, so
that any red set ever built mentions only finitely many of them. The supply is
\emph{potentially} infinite, never an actual completed infinity: the commitment
is the black empty set together with the same successor step that underwrites
$\omega$, and not one primitive atom more.
\end{remark}

\subsection{The permutation action}
\label{sec:fmmodel:action}

The stream operation $\mathrm{swap}(S[X],\rho)$ of the formal theory applies a
finite permutation $\rho=\{\,\bk{j_1}/\bk{i_1}:\dots:\bk{j_k}/\bk{i_k}\,\}$ to
the ordering of atoms; these finite permutations generate the full group
$\Sym(\Atoms)$. A permutation $\pi\in\Sym(\Atoms)$ acts on red sets by
red-membership recursion exactly as in $\mathsf{FM}$,
\[
  \pi\cdot a = \pi(a)\ (a\in\Atoms),
  \qquad
  \pi\cdot S = \rd{\{}\, \pi\cdot e \;:\; e \rin S \,\rd{\}}.
\]
Because $\pi$ moves only atoms and never the black interior of an atom, and
because the red operations $\remp,\ \rcup,\ \rcap,\ \rsm$ and
$\mathrm{embrace}_i$ are visibly equivariant (a renaming commutes with each), the
action is by automorphisms of the red theory and respects the equations of the
formal theory. We have, verbatim, the $\Sym(\Atoms)$-action that $\mathsf{FM}$
requires.

\subsection{Finite support is automatic}
\label{sec:fmmodel:support}

Write $\Red$ for the class of well-formed red sets, taken modulo red equality,
all of whose atoms lie in $\Atoms$. The finitary rules of the formal theory ---
$\mathrm{embrace}_i$ has finite arity, $\mathrm{Atomicity}$ adjoins one atom, and
$\rcup,\rcap,\rsm$ and $\mathrm{Comprehension}$ over finite sets preserve
finiteness --- yield only \emph{hereditarily finite} red sets. For $S\in\Red$ let
$\atomsof(S)\subseteq\Atoms$ be the finite set of atoms occurring hereditarily in
$S$.

\begin{lemma}[Support]
\label{lem:support}
For every $S\in\Red$ the finite set $\atomsof(S)$ supports $S$: if
$\pi\in\Sym(\Atoms)$ fixes $\atomsof(S)$ pointwise then $\pi\cdot S = S$.
Moreover $\atomsof(S)$ is the \emph{least} support, so $\supp(S)=\atomsof(S)$.
\end{lemma}

\begin{proof}[Proof sketch]
That $\atomsof(S)$ supports $S$ is an induction on the rank of $S$: an atom
$a\in\atomsof(S)$ is fixed by hypothesis, and for a set $S=\rbr{e_1,\dots,e_m}$ a
permutation fixing $\atomsof(S)\supseteq\bigcup_i \atomsof(e_i)$ fixes each $e_i$
by induction, hence fixes $S$. Leastness uses that $\Atoms$ is infinite: if
$a\in\atomsof(S)$ then, choosing a fresh $b\notin\atomsof(S)$, the transposition
$(a\,b)$ fixes $\atomsof(S)\setminus\{a\}$ yet moves $S$, so $a$ lies in every
support. Hence no proper subset of $\atomsof(S)$ supports $S$.
\end{proof}

\begin{corollary}
\label{cor:hfs}
Every $S\in\Red$ is hereditarily finitely supported, with
$\supp(S)=\atomsof(S)$. Freshness is occurrence-failure,
$a\mathrel{\#}S \iff a\notin\atomsof(S)$, and since $\Atoms$ is infinite the
fresh quantifier $\new$ and its some/any law (Section~\ref{sec:fm}) hold in
$\Red$.
\end{corollary}

The point worth stressing is that in $\mathsf{FM}$ finite support is an
\emph{axiom}, imposed by cutting the cumulative hierarchy down to
$\mathcal{U}_{\mathsf{FM}}$. Here it is a \emph{theorem}: the finitary red rules
never build anything that is not finitely supported, so the cut is vacuous and
nothing need be assumed.

\begin{remark}[Two points of precision]
\label{rem:group}
Lemma~\ref{lem:support} rests on two facts worth making explicit. First, the
group in play is the \emph{finitary} symmetric group of $\Atoms$ --- the
permutations generated by the transpositions that $\mathrm{swap}$ realises --- for
which supports are closed under intersection and least supports therefore exist;
on an arbitrary subgroup they need not. Second, the action permutes atoms while
treating each as structureless: it is an automorphism of the \emph{red} structure
$(\Red,\rin)$ alone, and is not asked to respect the black membership $\bin$
internal to an atom. The atoms are accordingly black-distinct but red-opaque, and
the entire support theory lives at the red level.
\end{remark}

\subsection{The representation theorem}
\label{sec:fmmodel:thm}

Let $\mathcal{M}$ be the structure with universe $\Atoms\cup\Red$, with the atoms
$\Atoms$ as urelements, with $\in^{\mathcal M}$ the red membership relation and
$=^{\mathcal M}$ red equality.

\begin{theorem}[Representation]
\label{thm:rep}
$\mathcal{M}$ is a model of hereditarily-finite Fraenkel--Mostowski set theory:
the $\mathsf{FM}$ analogue of $\mathrm{HF}(\Atoms)$. Concretely, interpreting the
atoms of $\mathsf{FM}$ as $\Atoms$, its sets as $\Red$, and its membership as red
$\rin$, the structure $\mathcal{M}$ validates extensionality, the empty set,
pairing, (finite) union, (finite) separation and foundation in their
$\mathsf{ZFA}$ form, carries the $\Sym(\Atoms)$-action of
Section~\ref{sec:fmmodel:action}, and satisfies the finite-support axiom. The
fresh quantifier and the some/any theorem hold throughout.
\end{theorem}

\begin{proof}[Proof sketch]
The atoms are membersless: red $\rin$ cannot enter a black set, so no
$a\in\Atoms$ has red elements, and atoms are $\in$-minimal and disjoint from
$\Red$ in sort, as $\mathsf{ZFA}$ demands. Foundation holds because each red set
is built at a finite rank over the atoms, so $\in^{\mathcal M}$ is well-founded.
The remaining axioms come straight from the rules of the formal theory:
$\mathrm{Foundation}$ gives $\remp$; $\mathrm{embrace}$ with $\mathrm{Union}$
gives pairing $\rbr{e_1}\rcup\rbr{e_2}$ and, iterated, every finite set;
$\mathrm{Union}$ gives binary and hence finite unions; $\mathrm{Subtraction}$ and
$\mathrm{Comprehension}$ over finite domains give finite separation; and
extensionality holds for the model provided provable red equality coincides with
sameness of members --- a property the equations of the formal theory are intended
to force, and which Section~\ref{sec:rigor} records as one of the obligations a
fully rigorous development must discharge. The $\Sym(\Atoms)$-action and the
finite-support axiom are Section~\ref{sec:fmmodel:action} and
Corollary~\ref{cor:hfs}; the some/any law follows because $\Atoms$ is infinite
and every $S\in\Red$ has finite support. Powerset and infinity are absent --- as
expected of an $\mathrm{HF}$ universe --- and are the subject of
Section~\ref{sec:fmmodel:infinitary}.
\end{proof}

\begin{corollary}[Two interlocked models]
\label{cor:two}
By the red/black symmetry of the construction, the same argument run with the
colours exchanged produces a black model $\bk{\mathcal M}$ of
hereditarily-finite $\mathsf{FM}$ whose atoms are the red numerals
$\{\,\rd{n} : n\in\omega\,\}\subseteq\rSetX$. The two models interlock: the atoms
of each are sets of the other. In the game-semantic reading of the
\emph{Intuitions} above, player's atoms are opponent's strategies and conversely
--- a single reflective universe carrying both $\mathsf{FM}$ models at once, each
underwriting the other's names.
\end{corollary}

\begin{remark}[What the $\mathrm{HF}$ restriction costs]
\label{rem:hftrivial}
It should be said plainly what is and is not delivered. The phenomena that give
$\mathsf{FM}$ its character --- finite support as a genuine constraint on
\emph{infinite} objects, equivariance with non-trivial orbits, and the some/any
law quantifying over infinitely many fresh names against a fixed infinite
support --- all require infinite, finitely-supported sets, of which $\mathcal{M}$
has none. On the hereditarily finite fragment these properties hold, but
trivially: every set is finite, so finite support is automatic and the some/any
law degenerates. $\mathcal{M}$ is thus best described as a model of the
\emph{nominal hereditarily finite sets}: faithful to the $\mathsf{ZFA}$ and
$\mathsf{FM}$ axioms, but exercising the distinctively nominal content only once
the infinitary extension of Section~\ref{sec:fmmodel:infinitary} is in force. We
retain the phrase ``model of $\mathsf{FM}$'' only under this qualification.
\end{remark}

\subsection{Finitary by design: the hereditarily finite core}
\label{sec:fmmodel:finitary}

It is worth being explicit about why Theorem~\ref{thm:rep} lands on the
hereditarily finite fragment, because the reason is also the source of the
construction's advantage. The signature of the formal theory offers, as set
constructors, only $\remp$, the finite-arity embraces $\mathrm{embrace}_i$
($|\mathrm{embrace}_i| = i < \omega$), the binary $\rcup,\rcap,\rsm$, and
$\mathrm{Comprehension}$ over finite domains $S_1\times\dots\times S_n$. Each of
these maps finite sets to finite sets, and the only base case is $\remp$;
so a trivial induction shows that \emph{every} derivable red set is hereditarily
finite. The model is $\mathrm{HF}(\Atoms)$, the nominal hereditarily finite sets
over the atoms.

Two of $\mathsf{FM}$'s defining commitments are, in this regime, theorems rather
than axioms.
\begin{itemize}
\item \textbf{Foundation} holds because every red set sits at a finite rank over
  the atoms, so $\in^{\mathcal{M}}$ is well-founded (Theorem~\ref{thm:rep}); we
  never assumed regularity.
\item \textbf{Finite support} holds because a hereditarily finite set mentions
  only finitely many atoms, each of which is its own least support
  (Lemma~\ref{lem:support}); we never assumed the finite-support axiom.
\end{itemize}
Correspondingly there is no infinity axiom and no infinite risk: the stream
$S[X]$ is read only through $\mathrm{take}(\Sigma,n)$ for finite $n$, so at no
point is an actually-infinite set formed. This is the regime in which the
reflective theory strictly improves on axiomatic $\mathsf{FM}$ --- both of the
moves that define $\mathsf{FM}$, foundation and the finite-support cut, are
\emph{derived} from the construction rather than \emph{posited}.

\subsection{Going infinitary, and the return of the cut}
\label{sec:fmmodel:infinitary}

Full $\mathsf{FM}$ also has infinite, finitely-supported sets --- $\Atoms$
itself, or the equivariant powerset $\pow_{\fs}(\Atoms)$ of finitely-supported
sets of atoms --- which the finitary signature cannot build. To reach them one
adds an \emph{infinitary} constructor. The most economical is an $\omega$-ary (or
$\kappa$-ary) embrace that consumes a whole stream rather than a finite prefix,
\[
  \inferrule{\Sigma;\Gamma\vdash S \\ (\sigma,\Sigma') = \mathrm{take}(\Sigma,\omega)}
            {\Sigma';\,\sigma,\Gamma\vdash \mathrm{embrace}_\omega(\sigma)\rcup S}
  \quad\textsc{Infinitary-Embrace},
\]
where $\mathrm{take}(\Sigma,\omega)$ returns the entire stream; equivalently one
adjoins a powerset operator and an infinity axiom and climbs the red cumulative
hierarchy $\mathcal{U}_\alpha(\Atoms)$ through all ordinals. Either way two things
change, and --- this is the point the finitary core conceals --- they change
\emph{together}.

\paragraph{An actual infinity, hence one $\omega$ of risk.}
The lazy stream becomes a completed set; $\Atoms$ itself becomes a red set. By
Remark~\ref{rem:risk} the commitment incurred is exactly black infinity, a single
$\omega$-style axiom. No primitive atoms are reintroduced --- the atoms remain
black hereditarily finite sets built from $\bemp$ --- but the potential infinity
of the finitary core is now actual.

\paragraph{Finite support is no longer free; the cut returns.}
Below $\omega$, finite support was automatic. The infinitary hierarchy is not so
forgiving: it contains sets that have \emph{no} finite support. The canonical
witness is the graph of the stream enumeration,
\[
  E \;=\; \{\,(n,a_n) : n\in\omega\,\},
  \qquad a_n = \text{the $n$-th atom of } S[X],
\]
which is fixed by a permutation $\pi$ only when $\pi$ preserves the enumeration,
i.e. only when $\pi$ is the identity on the infinitely many $a_n$; so no finite
set of atoms supports $E$. This is no accident. The formal theory already tells
us that $S[X]$ ``may be thought of as an ordering on $\Atoms$'', and a total order
on an infinite set of atoms is the textbook example of non-finitely-supported
data. The finitary rules never expose that order --- $\mathrm{take}$ reads only a
finite prefix --- which is precisely why support comes for free below $\omega$.
An infinitary embrace that swallows the whole stream imports the order, and with
it the non-finitely-supported sets.

Consequently, in the infinitary regime one must do exactly what axiomatic
$\mathsf{FM}$ does: cut down to the hereditarily finitely-supported part,
\[
  \mathcal{U}_{\mathsf{FM}} \;=\; \{\, x : x \text{ is hereditarily finitely supported}\,\},
\]
discarding $E$ and its kind. On the cut, the some/any theorem and the
equivariance principle hold verbatim, and the result satisfies the Gabbay--Pitts
axioms. The instructive point is that the cut is now the \emph{only} thing
borrowed from axiomatic $\mathsf{FM}$: the atoms still come from the dual copy,
the $\Sym(\Atoms)$-action still comes from $\mathrm{swap}$, and foundation still
comes from the construction (Section~\ref{sec:nwf}); only the finite-support
axiom must be re-imposed, and only because we elected to admit actual infinities.

This locates precisely where reflection saves risk and where it cannot. Atoms:
always free, from the dual copy. Foundation: always free, from local
well-foundedness (Section~\ref{sec:nwf}). Finite support: free in the finitary
core, an imposed cut once infinities are admitted. The finitary/infinitary
boundary is therefore not a technicality but the exact ledger line between what
the construction \emph{derives} and what it must still \emph{assume} --- and a
deliberate design choice for any theory built this way. We would keep the
finitary signature as the canonical core, where the accounting is cleanest, and
treat \textsc{Infinitary-Embrace} as an optional, clearly-priced extension.
