%% rset.applications.tex
%% Applications: the Pitts--Gabbay nominal theory, founded on the reflective
%% construction via the FM model of Section~\ref{sec:fmmodel}.
%% Drop in via \input{rset.applications} (add the hook to rset.tex after
%% \input{rset.mainthm}).
\input{rset.colors}   % red/black colour layer + FM symbols (providecommand-guarded)

\section{Applications: nominal techniques founded on reflection}
\label{sec:applications}

The original motivation for $\mathsf{FM}$ in computer science was Pitts and
Gabbay's \emph{nominal} treatment of names and binding
\cite{DBLP:journals/fac/GabbayP02}. That treatment is founded squarely on
$\mathsf{FM}$: names are the atoms, $\alpha$-equivalence is an orbit relation
under the permutation action, and the freshness machinery is exactly support and
the $\new$-quantifier. Since Section~\ref{sec:fmmodel} realises $\mathsf{FM}$
inside the reflective universe --- the nominal hereditarily finite fragment
outright, and the full theory under the infinitary extension of
Section~\ref{sec:fmmodel:infinitary} --- the entire nominal edifice descends with
it. The names that nominal techniques take as primitive become, here, black sets
built from the empty set; the reflective construction \emph{founds} the nominal
theory rather than presupposing its atoms.

\subsection{Nominal sets, internally}
\label{sec:applications:nom}

A \emph{nominal set} is a set carrying a $\Sym(\Atoms)$-action in which every
element is finitely supported; nominal sets and equivariant maps form the
category $\mathbf{Nom}$, equivalent to the Schanuel topos. The model
$\mathcal{M}$ of Theorem~\ref{thm:rep} is, on the nose, a nominal set: the action
is that of Section~\ref{sec:fmmodel:action}, and finite support is
Corollary~\ref{cor:hfs}. Every construction nominal practitioners rely on ---
the equivariance principle (anything definable from equivariant data is
equivariant), least supports, and the some/any law for $\new$ --- is therefore
available inside the reflective universe, with the colour discipline of
Section~\ref{sec:fmmodel} keeping the bookkeeping honest: names live in the
{\bk black} copy, the sets that bind them in the {\rd red} copy.

\subsection{Abstract syntax with binding}
\label{sec:applications:syntax}

The headline application of \cite{DBLP:journals/fac/GabbayP02} is a uniform
account of inductively-defined syntax \emph{modulo} $\alpha$-equivalence. Its key
device is \emph{name abstraction}: for a nominal set $X$,
\[
  [\Atoms]X \;=\; (\Atoms\times X)\big/\!\sim,
  \qquad
  \langle a\rangle x \sim \langle b\rangle y
  \;\;\Longleftrightarrow\;\;
  \new c.\ (a\,c)\cdot x = (b\,c)\cdot y,
\]
where $\langle a\rangle x$ is the class of $(a,x)$ and $(a\,c)$ is a
transposition. Both the action and $\new$ are present in $\mathcal{M}$, so
$[\Atoms]X$ is definable there, and it has the property that makes it a model of
binding: $\supp(\langle a\rangle x) = \supp(x)\setminus\{a\}$, i.e. abstraction
\emph{discharges} the bound name from the support, precisely as a binder removes
a variable from the free-variable set.

Inductive syntax is then obtained as usual. Object-level $\lambda$-terms, for
instance, are the least solution of
\[
  \Lambda \;\cong\; \Atoms \;+\; \Lambda\times\Lambda \;+\; [\Atoms]\Lambda
  \qquad(\text{variable, application, abstraction}),
\]
with $\alpha$-equivalence built into the third summand. Structural recursion and
induction come with the \emph{freshness condition for binders}, justified by the
some/any law: a definition or proof that treats a bound name as fresh for its
parameters is sound because, $\Atoms$ being infinite, ``some fresh name'' and
``every fresh name'' agree (Corollary~\ref{cor:hfs}). This is exactly the
reasoning principle nominal users invoke when they ``choose a fresh variable''.

The risk accounting carries through and sharpens the point. Each individual
term --- each element of $\Lambda$ --- is a hereditarily finite nominal set, so
it already lives in the $\mathrm{HF}$ model $\mathcal{M}$ of
Theorem~\ref{thm:rep}, founded on the empty set alone: a binding tree of any
fixed shape costs no infinite risk. The completed datatype $\Lambda$, an
infinite equivariant set, is one of the finitely-supported infinities of
Section~\ref{sec:fmmodel:infinitary}, and so costs exactly one $\omega$ --- black
infinity --- and still not a single primitive name. Where Pitts and Gabbay begin
with an infinite set of atoms given by fiat, here the names are manufactured, the
$\alpha$-structure is an orbit of the $\mathrm{swap}$ action, and the whole
apparatus is refinanced down to one empty set and one successor.

\subsection{Nominal algebraic theories}
\label{sec:applications:alg}

Beyond syntax, equational reasoning in the presence of names --- nominal
equational logic and the nominal Lawvere theories of Clouston
\cite{DBLP:journals/jcss/Clouston14} --- takes its semantics in $\mathbf{Nom}$.
As $\mathcal{M}$ furnishes $\mathbf{Nom}$-structure internally, any such theory
(name-restriction, scope extrusion, binding signatures and their equational
laws) can be interpreted in the reflective universe. The reflective setting adds
one structural feature these accounts do not have: by
Corollary~\ref{cor:two} there are \emph{two} interlocking name-universes, the red
and the black, each supplying the other's names. Reading red as player and black
as opponent recovers the game-semantic picture of the \emph{Intuitions}: a name
on one side is a strategy on the other, and a nominal theory and its dual coexist
in a single reflective universe.

\subsection{The hereditarily finite fragment is the right fragment}
\label{sec:applications:hf}

Remark~\ref{rem:hftrivial} and Section~\ref{sec:fmmodel:infinitary} flagged the
hereditarily finite restriction as the limit of what the finitary theory captures
relative to full $\mathsf{FM}$. From the standpoint of \emph{computing}, that
restriction is not a limitation but an alignment with practice. The realised
nominal tools --- $\mathsf{FreshML}$ and Fresh O'Caml
\cite{shinwell2003freshml,shinwell2005freshocaml}, $\alpha$Prolog
\cite{cheney2004alphaprolog} --- manipulate only hereditarily finite data:
abstract syntax trees over a name supply, with name-swapping and nominal
unification as the engine \cite{pitts2013nominal}. The completed set of atoms, the
equivariant infinite sets, the Schanuel topos --- these inhabit the
\emph{semantics} of those languages, never a running program. A program holds a
finite term mentioning finitely many names and asks only local questions of it:
are these two names equal, is this one fresh, swap that pair.

So the fragment on which the reflective construction is sharpest --- where finite
support is a theorem rather than an imposed cut, the some/any law holds, and no
infinity and no infinite risk are incurred (Section~\ref{sec:fmmodel:finitary})
--- is exactly the fragment these implementations inhabit. The hereditarily finite
cut is where the theory and the practice coincide. This suggests a methodological
point worth making explicit: a foundation for nominal \emph{computing}, as
distinct from nominal \emph{set theory}, has no need to reach for infinite
finitely-supported sets at all. It needs the nominal hereditarily finite sets,
and those the reflective construction delivers from the empty set with the
strongest guarantees and the least risk. What looked like the construction's
ceiling is, for the applications that motivated the whole enterprise, precisely
its home.

\subsection{Reflective names: a more expressive theory of names}
\label{sec:applications:reflective}

There is a second, less obvious dividend, and it concerns the \emph{nature of
names} rather than their supply. In the Pitts--Gabbay world an atom is primitive
and structureless: the only operations on it are equality, freshness, and
swapping, and that opacity is the entire point. It is what forces every definable
function to be equivariant, and equivariance is what makes $\alpha$-equivalence
and capture-avoidance come for free \cite{DBLP:journals/fac/GabbayP02}. In the
reflective construction a name is \emph{not} primitive: it is a black set, and one
may cross the colour barrier to look inside it.

This is the set-theoretic shadow of the move the reflective higher-order calculus
makes against the $\pi$-calculus \cite{DBLP:journals/entcs/MeredithR05}: where
$\pi$ names are atomic, reflective names are quoted processes, and reflection lets
one pass between a name and the structure it names. The reflective set theory says
that ``names as quoted structure'' is not special to process calculi --- it is a
general operation on theories, two copies each serving as the other's atoms --- and
that performing it on set theory yields nominal sets whose names carry structure.
What such names buy, that opaque oracle-drawn names cannot, is fourfold, and it
tells in the distributed and mobile setting rather than in pure syntax.
\begin{itemize}
\item \textbf{Names are computed, not drawn.} ``Fresh'' becomes ``a black set not
  occurring in this finite red structure'' --- decidable and constructive, a
  deterministic fresh-name \emph{function} rather than the observable side effect
  that name generation is in $\mathsf{FreshML}$. No $\mathrm{gensym}$ oracle and no
  central allocation authority are required, so in a decentralised system any
  party may mint and verify a name locally with no coordination.
\item \textbf{Names are first-class data.} A name is a finite set, so it
  serialises, communicates, and hash-conses; content-addressed names --- names
  that \emph{are} the canonical encoding of what they name --- fall straight out.
\item \textbf{Names have a decidable structural identity}, rather than an
  abstract primitive equality.
\item \textbf{Self-representation.} The two copies form a reflection tower: the
  metalanguage's names are the object theory's sets and conversely, which is a
  principled basis for metaprogramming and staging that a flat primitive name type
  cannot express.
\end{itemize}

The honest cost fixes the shape of the contribution. Reflection breaks opacity,
and opacity was load-bearing: a program that inspects a name's black structure is
not equivariant, so it need not respect $\alpha$-equivalence, and for it the
automatic capture-avoidance and the some/any principle are forfeit. The value is
therefore not to discard opacity but to obtain a \emph{unifying} metalanguage with
a knob. Staying within one colour is ordinary nominal programming, with every
guarantee intact, recovered as the \emph{equivariant fragment}; crossing colours
is the additional reflective power. A metalanguage built on this would carry a
grading --- a modality distinguishing equivariant, colour-respecting computations
from reflective ones --- so that its type system states exactly which operations
still enjoy the nominal guarantees and which have traded them for structure. That
discipline, in the spirit of graded and modal type systems, is the genuine open
problem here: we can claim the unification and the semantic account, but the
better metalanguage --- a reflective $\mathsf{FreshML}$ in which the
Pitts--Gabbay discipline is the equivariant sublanguage and reflective names are
the whole --- is the research programme this construction opens, not an artifact
it delivers.



The chain is short and worth stating plainly. Nominal techniques rest on
$\mathsf{FM}$; $\mathsf{FM}$ embeds in the reflective universe
(Section~\ref{sec:fmmodel}); the reflective universe is built from the empty set.
Composing, the Pitts--Gabbay theory of names and binding is founded on two
recursively intertwined copies of ordinary set theory, with its primitive atoms
replaced by constructed black sets and its infinite risk replaced by the finite
risk of $\varnothing$ --- exactly the reconciliation announced in the
introduction, now delivered for the application that made $\mathsf{FM}$ matter to
computer science in the first place. Two further observations sharpen the point
for computing rather than for set theory. The hereditarily finite cut, a ceiling
for the set theory, is the natural home of the realised nominal languages
(Section~\ref{sec:applications:hf}); and reflective names, by trading opacity for
structure under a graded discipline, promise a metalanguage that contains the
nominal one as its equivariant fragment while reaching the computed,
serialisable, content-addressable names of the mobile setting
(Section~\ref{sec:applications:reflective}).
